# Introduction to Health Metrics {#sec-01-health_metrics}

> **"The concept of health also has highly subjective connotations,
> precisely because it is conditioned by the personal view of
> happiness." René Dubos[@dubos1976]**

Health metrics\index{health metrics} are key variables to understanding
the state of health of a population. In this first section of the book
we'll explore the history of these metrics, how to calculate them, and
how to account for their variation due to causes and risks. We will
examine the components of these metrics and the challenges in providing
standardised values for global comparison. Additionally, in the
following sections, we will learn how to build a model, evaluate the
effect of infectious disease spread on health metrics, and how it
impacts the overall health status of a population.

## The History of Health Metrics

The history of health metrics is a fascinating journey that spans
centuries, evolving from rudimentary observations to sophisticated
statistical methods, as illustrated by @fig-timeline.

```{r}
#| echo: false
#| eval: false
#| output: false
library(tidyverse)
timeline <- tibble(
  id = seq(1, 8, 1),
  type = c(
    "Ancient Observations",
    "John Graunt (17th Century)",
    "Life Tables (19th Century)",
    "Quality-Adjusted Life Years (QALYs)",
    "Global Burden of Disease Study (1990)",
    "Disability-Adjusted Life Years (DALYs)",
    "Advancements in Technology and Data",
    "Current Landscape"
  ),
  text = c(
    "The concept of health metrics can be traced back to ancient civilizations where observations on life expectancy, birth rates, and mortality were made. However, these were often anecdotal and lacked systematic measurement.",
    "In the 17th century, John Graunt, an English haberdasher, made significant contributions to health metrics. In 1662, he published 'Natural and Political Observations Made upon the Bills of Mortality,' which analysed weekly and annual records of deaths in London. Graunt's work laid the foundation for modern demography, introducing concepts like life expectancy and mortality rates.",
    "The 19th century saw the development of life tables, providing a systematic approach to analysing mortality and life expectancy. This period marked a transition from descriptive observations to more structured methods in health metrics.",
    "QALYs, another important health metric, emerged in the late 20th century as a measure combining both the quality and quantity of life lived. QALYs are commonly used in health economics and healthcare decision-making.",
    "A significant milestone in contemporary health metrics is the initiation of the Global Burden of Disease (GBD) Study in 1990. Led by Christopher J. L. Murray and Alan D. Lopez, this collaborative effort between the World Health organisation (WHO) and the World Bank aimed to quantify the impact of diseases and injuries on global health.",
    "The GBD Study introduced DALYs as a metric in 1990. DALYs combine years of life lost due to premature mortality and years lived with disability, providing a comprehensive measure of disease burden. This metric became instrumental in assessing the overall health impact of different conditions.",
    "With advancements in technology, including computing power and big data analytics, health metrics have become more sophisticated. Real-time tracking, machine learning applications, and global health databases contribute to a more dynamic understanding of population health.",
    "Today, health metrics encompass a wide range of measures, including traditional mortality rates, disease-specific metrics, and broader indicators like well-being and social determinants of health. These metrics play a crucial role in shaping public health policies, healthcare interventions, and global health assessments."
  )
)
```

```{r}
#| echo: false
#| eval: false

timeline %>%
  ggplot(aes(x = id, y = 0.5)) +
  geom_segment(aes(x = id, xend = id, y = 0.5, yend = 0.8), linewidth = 0.3) +
  geom_segment(aes(x = id, xend = id, y = 0.5, yend = -1), linewidth = 0.1) +
  geom_point() +
  geom_line(group = 1, linewidth = 1) +
  ggtext::geom_textbox(aes(label = type, y = 1),
    fontface = "bold",
    width = unit(0.13, "npc"),
    box.size = 0,
    fill = NA,
    hjust = 0.5,
    size = 2
  ) +
  ggtext::geom_textbox(aes(label = text, y = -1),
    width = unit(0.13, "npc"),
    box.size = 0,
    fill = "white",
    hjust = 0.5,
    size = 2
  ) +
  ylim(-3, 2) +
  xlim(0.8, 8.2) +
  coord_cartesian() +
  theme_void()
```
```{r}
#| echo: false
#| eval: false

# save as png 600 dpi, bg white
ggsave("images/01_timeline.png", 
       bg = "white",
       width = 8, height = 4, dpi = 600)
```

![Health Metrics Development
Timeline](images/01_timeline.png){#fig-timeline
fig-align="center" width="90%"
fig-alt="Health Metrics development timeline"}

One of the earliest forms of health metrics can be traced back to the
work of *John Graunt*, an Englishman, in the 17th century[@johngra2024].
In 1662, Graunt published a landmark work titled "Natural and Political
Observations Made upon the Bills of Mortality." Graunt, defined as the
**father of human demography**, was largely self-educated, and his
interest in mortality data was driven by the Bills of Mortality, which
were weekly and annual records of deaths in London, and provided
information on the number of births and deaths in the city. Graunt
meticulously analysed this **mortality data** and produced **tables**
that presented various patterns and trends. His work laid the foundation
for modern demography and statistical analysis. Among his notable
contributions were the concepts of **life expectancy** and **mortality
rates**. One of Graunt's key observations was the consistent pattern of
higher mortality among infants and young children. He noted the
difference in life expectancy between males and females and provided
insights into factors influencing mortality, such as epidemics and
seasons.

Over time, advancements in mathematics, medicine, and statistics led to
the development of more sophisticated health metrics. The 19th century
saw the emergence of **life tables**, which provided a systematic way to
analyse mortality and life expectancy. More specifically, *Mary Dempsey*
in 1940s wrote an article[@dempsey2019a] for the National Tuberculosis
Association about the concept of **Years of Life Lost (YLL)**, assessing
time-lost as a metric for health for the first time. She was then
followed by other more advanced improvements spanning all throughout the
rest of the century.

> **"The quality of life depends also on the state of "public
> health"-namely on factors that affect the biological welfare of the
> community as a whole."[@dubos1976]**

The 20th century marked significant advancements in epidemiology,
driving the expansion and refinement of health metrics. This period saw
the widespread adoption of life tables and the modification of expected
survival time to incorporate weighted metrics, such as
disability-adjusted life years (DALYs). These developments enabled more
comprehensive assessments of population health by accounting for both
mortality and the quality of life, laying the foundation for modern
health evaluation tools.

The term **Summary Measures of Population Health (SMPH)**[@summary2002]
was established in the field, originating from the need to encapsulate
both fatal and non-fatal health outcomes into a single, comprehensive
measure, representing a significant evolution in health metrics. Used
for various purposes, the SMPH compares the health of populations and
includes the contributions of different diseases, injuries and risk
factors to the total disease burden in a population. SMPH, also referred
to as a composite indicator[@hyder2012], is divided into two broad
families: health expectancies and health gaps.

**Health expectancies** are summary measures that combine information on
mortality and morbidity to represent the average number of years that an
individual can expect to live in good health. Such as **healthy life
expectancy (HALE)** and the **quality adjusted life expectancy (QALE)**
using health-related quality of life.

**Health gaps**, on the other hand, represent the difference between the
current health status of a population and an ideal health situation,
providing a measure of the potential for improvement in health outcomes.
Divided into the **disability adjusted life year (DALY)**[@murray1994a]
in the global burden of disease (GBD) study and the **quality adjusted
life year (QALY)** used as the outcome index of the cost-utility
analysis.

In the contemporary era, the integration of computing power and big data
has further transformed health metrics. Today, we have complex models,
machine learning algorithms, and global health databases that allow for
real-time tracking of diseases and health outcomes.

John Graunt's work may have been modest in its origins, but it laid the
groundwork for a scientific approach to understanding population health.
From Graunt’s early analysis of the *Bills of Mortality*[@johngra2024a]
to the development of today’s advanced health metrics, this journey
reflects the ongoing effort to measure and understand the complexities
of human health and disease.

### Quality-Adjusted Life Years (QALYs)

The concept of **Quality-Adjusted Life Years (QALYs)** emerged in the
late 1970s, originating from the idea of combining both the *quantity*
and *quality* of life by *Joseph S. Pliskin*, then it became a widely
used metric for evaluating the *cost-effectiveness* of healthcare
interventions [@quality-2023]. One *QALY* corresponds to one year in
perfect health, it ranges from 1 (perfect health) to 0 (dead). However,
this metric, used in various sectors such as health insurance, is
somewhat criticised due to its lack of addressing disparities among
individuals. It might result in discrimination against individuals with
specific medical conditions or disabilities.

A controversial result followed 25 years of research and studies due to
the concerning validity of the metric to be inclusive. Started in the
early 1989s, to step in the 2010 with the *European Commission*
[@echoutco2016] which started the largest-ever study specifically
dedicated to testing the assumptions of the *QALY* [@holmes2013], ending
in 2013 with the recommended "not using QALYs in healthcare decision
making", arguing that patients with disabilities are valued less under a
QALY-based system then individuals with no disabilities. This result has
been further criticised and remains a topic of debate.

> **"Thus, medicine cannot by itself determine the quality of life. It
> can only help people to achieve the state of health that enables them
> to cultivate the art of life-but in their own way. This implies the
> ability to enjoy the fundamental satisfactions of the biological joie
> de vivre. It also implies the ability for each person to do what he
> wants to do and become what he wants to become, according to human
> values that transcend medical judgement" R.Dubos[@dubos1976a]**

Used in health policy to measure the value of medical interventions by
balancing both the quantity and quality of life added, QALYs help
decision-makers determine which treatments or programs offer the
greatest health benefits relative to their costs. For example, a health
policy might prioritise funding for treatments that deliver high QALYs
per dollar spent, ensuring resources are allocated to maximise health
outcomes across the population. This approach guides decisions in
healthcare budgeting, insurance coverage, and prioritising interventions
for chronic illnesses, preventive care, or new therapies.

### Disability-Adjusted Life Years (DALYs)

In the 1990s, the World Bank financed a study at Harvard University to
develop a sustainable index that considered not only the health status
but also the identification of the level of disabilities concerned by
disparities. This lead to a new way of comparing the overall health and
life expectancy of different countries, released along with the Global
Burden of Disease Study in 1990 [@murray1994]; a project that involved
*Christopher J. L. Murray* [@christop2024] and *Alan Lopez*
[@alanlop2023], in collaboration with the World Health organisation
(WHO). This effort resulted in the development of the
Disability-Adjusted Life Years (DALYs) [@disabili2023a], a metric that
combines **years of life lost due to premature death** and **years lived
with disability**.

```{r}
#| echo: false
#| eval: false

library(DiagrammeR)

grViz("
  digraph {
    rankdir = LR
    node [shape = circle, width = 2]
    A [label = 'Years of Life Lost (YLLs)']
    B [label = 'Years Lived with Disability\n(YLDs)']
    C [label = 'Disability-Adjusted Life Years\n(DALYs)']
    A -> C
    B -> C
  }")
```

![YLLs and YLDs components of
DALYs](images/01_dalys-diagram.png){#fig-dalys
width="50%" fig-align="center"
fig-alt="Breakdown of DALYs into YLLs and YLDs components"}

DALYs are connected with QALYs in that both metrics seek to quantify the
impact of health on quality and length of life. However, while QALYs
focus on the benefits with and without medical intervention, DALYs
measure the overall burden of diseases.

The DALY metric is calculated by summing the **years of life lost
(YLLs)** due to premature death and the **years lived with disabilities
(YLDs)**. This metric provides a comprehensive measure of a population’s
health status by highlighting both mortality and disability. For
example, consider a population with a life expectancy of 80 years.
Individuals who die before reaching this age contribute to the YLLs,
while those living with disabilities add to the YLDs. The sum of these
values gives the total DALYs for the population, which can then be
compared globally to assess health status and identify areas needing
improvement in facilities, research, or investment.

### Healthy-Adjusted Life Years (HALY)

Healthy Adjusted Life Years (HALY) is a metric that combines aspects of
both DALYs and QALYs to provide a comprehensive measure of health
outcomes in a population. HALYs quantify the burden of disease and the
overall quality of life experienced by individuals, measuring the number
of years of healthy life lived by a population while accounting for both
mortality and morbidity. The information about the prevalence of
diseases and disabilities, as well as the impact of these conditions on
individuals' quality of life is incorporated in the calculation of
HALYs.

HALYs adjust life expectancy levels based on the prevalence of health
conditions and their associated disability weights. Disability weights
reflect the severity of different health conditions and their impact on
daily functioning. By applying these weights to the prevalence of each
condition, researchers can estimate the overall burden of disease in
terms of years of healthy life lost, similar to the calculation of
DALYs. Unlike DALYs, which focus primarily on morbidity and mortality,
HALYs incorporate measures of individuals’ quality of life, allowing for
a more comprehensive assessment of health outcomes.

### Healthy-Adjusted Life Expectancy (HALE)

Healthy Life Expectancy (HALE) is a measure of overall health and
well-being that accounts for both the quantity and quality of life. HALE
combines years of life expectancy with the prevalence and severity of
disability in a population, providing a more nuanced view of health
outcomes than traditional measures of life expectancy.

HALE provides a more comprehensive view of health outcomes than other
traditional measures of life expectancy, which only consider the
quantity of life. The calculation of HALE typically involves estimating
the number of years that an individual can expect to live in good
health, taking into account the impact of diseases and injuries on
quality of life. This information is then used to estimate the overall
health status of a population. It is a useful tool for public health
practitioners and policy makers, as it provides a more nuanced view of
the health outcomes of a population. This information can help inform
public health interventions and prioritise resources, as well as help
track changes in health outcomes over time. Additionally, HALE can be
used to compare the health outcomes of different populations and
identify disparities in health outcomes, which can help inform targeted
public health interventions.

Overall, the HALE metric provides a valuable perspective on the overall
health and well-being of a population, combining information about both
quantity and quality of life to provide a comprehensive view of health
outcomes. It adjusts overall life expectancy by the amount of time lived
in less than perfect health. This is calculated by subtracting from the
life expectancy a figure which is the number of years lived with
disability multiplied by a weighting to represent the effect of the
disability[^01-health_metrics-1][@healthy].

[^01-health_metrics-1]: [www.healthknowledge.org.uk](https://www.healthknowledge.org.uk/e-learning/health-information/population-health-specialists/lifetables-hales-dalys-pylls#:~:text=Health%20Adjusted%20Life%20Expectancy%20(HALE)&text=It%20adjusts%20overall%20life%20expectancy,the%20effect%20of%20the%20disability.)

### Healthy Life Years (HLY)

There is one more health metrics that is worth mentioning, the **Healthy
Life Years** (**HLY**) indicator, also known as **disability-free life
expectancy** (**DFLE**) or **Sullivan's Index**[@healthy2024]. It
assesses the number of years an individual can expect to live in good
health, free from significant health problems or disabilities. It is a
measure of the *quality of life-adjusted life expectancy*, focusing on
the years lived in full health rather than simply the total number of
years lived. It specifically focuses on the number of years a person can
expect to live without disability, providing valuable insights into
overall health and well-being beyond just life expectancy. It takes into
account both mortality and morbidity, providing a comprehensive picture
of overall health and well-being.

HLYs differs from HALE in that it focuses on the number of additional
years a person can expect to live in good health after reaching a
certain age, typically 65. This makes HLYs especially relevant for
assessing health in ageing populations and evaluating interventions that
aim to improve the quality of life for older adults.

For example, an organisation focusing on elder care policy might use
HLYs to measure program impact on older adults specifically, while a
global health agency might prefer HALE for its applicability to general
population health across all ages.

HLY is typically calculated by subtracting the number of years lived
with disability from life expectancy at birth. The resulting figure
represents the number of years an individual can expect to live in good
health, free from significant health-related limitations. Unlike life
expectancy, which focuses solely on the length of life, HLY incorporates
measures of the quality of life by considering the impact of health
conditions on individuals' ability to engage in daily activities and
maintain independence.

### Well-being-Adjusted Health Expectancy (WAHE)

Currently, **Well-being-Adjusted Health Expectancy (WAHE)** is still an
emerging metric in health research that combines traditional health
indicators with **well-being factors** to provide a more comprehensive
assessment of population health. WAHE is calculated by estimating the
number of life years equivalent to full health, taking into account both
physical and psychological well-being.

WAHE aims to capture the overall health and well-being of individuals by
incorporating measures of physical health, mental health, and social
well-being. This metric goes beyond traditional health metrics like life
expectancy and disability-adjusted life years to provide a more holistic
view of health outcomes by considering factors like **happiness**,
**life satisfaction**, and **social connectedness**.

Particularly relevant in the context of global health, WAHE provides a
more holistic view of population health that can inform public health
policies, healthcare interventions, and resource allocation
decisions[@ipc2021][@muszynska-spielauer2022][@well-bei].

```{r}
#| echo: false
#| label: tbl-health_metrics
#| tbl-cap: "Health Metrics Overview"

library(knitr)

# Define the data for the table
data <- data.frame(
  Metric = c(
    "QALY (Quality-Adjusted Life Year)", 
    "DALY (Disability-Adjusted Life Year)", 
    "HALY (Health-Adjusted Life Year)", 
    "HALE (Healthy Life Expectancy)",
    "HLY (Health Life Years)",
    "WAHE (Well-being Adjusted Health Expectancy)"
  ),
  Purpose = c(
    "Measures health benefits of interventions by combining quantity and quality of life gained",
    "Measures burden of disease by capturing premature death and disability impact",
    "General measure combining health quantity and quality; similar to QALY but less common",
    "Estimates life expectancy in 'healthy' years, adjusting for disability and illness",
    "Measures the additional healthy years expected, often from a specific age like 65",
    "Incorporates both physical and subjective well-being for life expectancy in 'full health'"
  ),
  Focus = c(
    "Health gains from interventions", 
    "Health loss due to disease and disability", 
    "Health gains or losses from interventions or disease impact", 
    "Overall healthy life expectancy at a population level",
    "Healthy life expectancy focused on ageing populations",
    "Overall well-being and mental health as well as physical health"
  ),
  Usage = c(
    "Used in cost-effectiveness studies, healthcare resource allocation, and insurance",
    "Used in public health to understand and compare disease burden globally",
    "Used similarly to QALY, though less frequently in decision-making",
    "Primarily used in population health and public health reports for health expectancy",
    "Used in geriatric health assessments and interventions for ageing populations",
    "Applicable for holistic health assessments, incorporating quality of life and mental wellness"
  ),
  Calculation = c(
    "Calculated by adjusting life years gained for quality of life (e.g., 1 year at half quality = 0.5 QALY)",
    "Calculated by summing years of life lost (YLL) and years lived with disability (YLD)",
    "Calculated by adjusting life years for health quality; similar to QALY",
    "Calculated by adjusting total life expectancy based on age-specific disability prevalence",
    "Typically calculated by estimating additional years of good health beyond a baseline age",
    "Calculated by weighting health states to reflect impact on well-being and quality of life"
  )
)

# Display the table
kable(data, col.names = c("Metric", "Purpose", "Focus", "Usage", "Calculation"), align = "l")
```

## How the Metrics are Used in Global Health

These metrics are used alone or in combination with other health metrics
in global health assessments conducted by organisations such as the
**World Health organisation (WHO)**, the **European Commission**, and
the **Institute for Health Metrics and Evaluation (IHME)**. They provide
a global perspective and insights into the health status of populations,
helping to track progresses towards long-term health objectives over
time. The metrics also inform public health policies, healthcare
interventions, and resource allocation decisions, helping to prioritise
health needs and target interventions effectively.

The investigation into finding updated metrics is a sustained
commitment, aiming to develop new measures tailored to the evolving
landscape of the health field. Recently, the new **Summary Measure of
Population Health (SMPH)** proposed the **Well-being-Adjusted Health
Expectancy (WAHE)** measure as an estimate of the number of life years
equivalent to full health[@muszynska-spielauer2022]. This approach
reflects a growing recognition that health metrics should account for
physical and clinical indicators and consider psychological and overall
well-being.

## Summary

Health metrics play a crucial role in understanding and improving
population health. From John Graunt’s early work in the 17th century to
today’s sophisticated metrics like DALYs, QALYs, HALYs, HALE, and HLY,
these measures have evolved to provide a comprehensive view of health
outcomes. They guide public health policies, healthcare interventions,
and global health assessments, highlighting areas for improvement and
helping to allocate resources effectively. As health metrics continue to
evolve, incorporating both traditional and well-being indicators, they
will remain essential tools in the quest to quantify and enhance the
complexities of human health and disease.

What’s next? In the next chapter we will investigate deeper into how
DALYs break down health impacts into two powerful metrics—Years of Life
Lost (YLL) and Years Lived with Disability (YLD). These calculations not
only reveal the total toll of disease but drive critical decisions on
where to focus resources and which interventions can most effectively
boost population health.
