# Introduction to Health Metrics {#sec-01-metrics_evaluation}

> **"The concept of health also has highly subjective connotations,
> precisely because it is conditioned by the personal view of
> happiness." René Dubos[@dubos1976]**

Health metrics\index{health metrics} are key variables to understanding
the state of health of a population. In this first section of the book
we'll explore about the history of these metrics, how to calculate them,
and how to account for their variation due to causes and risks. We will
examine the components of these metrics and the challenges in providing
standardized values for global comparison. Additionally, in the
following sections, we will learn how to build a model, evaluate the
effect of infectious disease spread on health metrics, and how it
impacts the overall health status of a population.

## The History of Health Metrics

The history of health metrics is a fascinating journey that spans
centuries, evolving from rudimentary observations to sophisticated
statistical methods, as illustrated by @fig-timeline.

```{r}
#| echo: false
#| eval: false
#| message: false
#| warning: false
#| output: false
library(tidyverse)
timeline <- tibble(
  id = seq(1, 8, 1),
  type = c(
    "Ancient Observations",
    "John Graunt (17th Century)",
    "Life Tables (19th Century)",
    "Quality-Adjusted Life Years (QALYs)",
    "Global Burden of Disease Study (1990)",
    "Disability-Adjusted Life Years (DALYs)",
    "Advancements in Technology and Data",
    "Current Landscape"
  ),
  text = c(
    "The concept of health metrics can be traced back to ancient civilizations where observations on life expectancy, birth rates, and mortality were made. However, these were often anecdotal and lacked systematic measurement.",
    "In the 17th century, John Graunt, an English haberdasher, made significant contributions to health metrics. In 1662, he published 'Natural and Political Observations Made upon the Bills of Mortality,' which analyzed weekly and annual records of deaths in London. Graunt's work laid the foundation for modern demography, introducing concepts like life expectancy and mortality rates.",
    "The 19th century saw the development of life tables, providing a systematic approach to analyzing mortality and life expectancy. This period marked a transition from descriptive observations to more structured methods in health metrics.",
    "QALYs, another important health metric, emerged in the late 20th century as a measure combining both the quality and quantity of life lived. QALYs are commonly used in health economics and healthcare decision-making.",
    "A significant milestone in contemporary health metrics is the initiation of the Global Burden of Disease (GBD) Study in 1990. Led by Christopher J. L. Murray and Alan D. Lopez, this collaborative effort between the World Health Organization (WHO) and the World Bank aimed to quantify the impact of diseases and injuries on global health.",
    "The GBD Study introduced DALYs as a metric in 1990. DALYs combine years of life lost due to premature mortality and years lived with disability, providing a comprehensive measure of disease burden. This metric became instrumental in assessing the overall health impact of different conditions.",
    "With advancements in technology, including computing power and big data analytics, health metrics have become more sophisticated. Real-time tracking, machine learning applications, and global health databases contribute to a more dynamic understanding of population health.",
    "Today, health metrics encompass a wide range of measures, including traditional mortality rates, disease-specific metrics, and broader indicators like well-being and social determinants of health. These metrics play a crucial role in shaping public health policies, healthcare interventions, and global health assessments."
  )
)
```

```{r}
#| echo: false
#| eval: false
#| fig-align: "center"
#| fig-cap: "Health Metrics development timeline"
#| fig-alt: "Health Metrics development timeline"
timeline %>%
  ggplot(aes(x = id, y = 0.5)) +
  geom_segment(aes(x = id, xend = id, y = 0.5, yend = 0.8), linewidth = 0.3) +
  geom_segment(aes(x = id, xend = id, y = 0.5, yend = -1), linewidth = 0.1) +
  geom_point() +
  geom_line(group = 1, linewidth = 1) +
  ggtext::geom_textbox(aes(label = type, y = 1),
    fontface = "bold",
    width = unit(0.13, "npc"),
    box.size = 0,
    fill = NA,
    hjust = 0.5,
    size = 2
  ) +
  ggtext::geom_textbox(aes(label = text, y = -1),
    width = unit(0.13, "npc"),
    box.size = 0,
    fill = "white",
    hjust = 0.5,
    size = 2
  ) +
  ylim(-3, 2) +
  xlim(0.8, 8.2) +
  coord_cartesian() +
  theme_void()
```

![Health Metrics Development
Timeline](images/01_timeline.png){#fig-timeline
fig-align="center" width="90%"}

One of the earliest forms of health metrics can be traced back to the
work of *John Graunt*, an Englishman, in the 17th century
[@johngra2024]. In 1662, Graunt published a landmark work titled
"Natural and Political Observations Made upon the Bills of Mortality."
Graunt defined as the **father of human demography** was largely
self-educated, and his interest in mortality data was driven by the
Bills of Mortality, which were weekly and annual records of deaths in
London, and provided information on the number of births and deaths in
the city. Graunt meticulously analyzed this **mortality data** and
produced **tables** that presented various patterns and trends. His work
laid the foundation for modern demography and statistical analysis.
Among his notable contributions were the concepts of **life expectancy**
and **mortality rates**. One of Graunt's key observations was the
consistent pattern of higher mortality among infants and young children.
He noted the difference in life expectancy between males and females and
provided insights into factors influencing mortality, such as epidemics
and seasons.

Over time, advancements in mathematics, medicine, and statistics led to
the development of more sophisticated health metrics. The 19th century
saw the emergence of **life tables**, which provided a systematic way to
analyze mortality and life expectancy. More specifically, *Mary Dempsey*
in 1940s wrote an article[@dempsey2019a] for the National Tuberculosis
Association about the concept of **Years of Life Lost (YLL)**, assessing
time-lost as a metric for health for the first time. She was then
followed by other more advanced improvements spanning all throughout of
the rest of the century.

> **"The quality of life depends also on the state of "public
> health"-namely on factors that affect the biological welfare of the
> community as a whole."[@dubos1976a]**

With the growth of epidemiology the 20th century was a key century for
the expansion of the health metrics, the use of life table and
modification of expected survival time to include weights metrics.

The term **Summary Measures of Population Health (SMPH)**[@summary2002]
was established in the field, originating from the need to encapsulate
both fatal and non-fatal health outcomes into a single, comprehensive
measure, representing a significant evolution in health metrics. Used
for various purposes, the SMPH compare the health of populations and
include the contributions of different diseases, injuries and risk
factors to the total disease burden in a population. SMPH, also referred
to as a composite indicator[@hyder2012], is divided into two broad
families: health expectancies and health gaps.

**Health expectancies** are summary measures that combine information on
mortality and morbidity to represent the average number of years that an
individual can expect to live in good health. Such as **healthy life
expectancy (HALE)** and the **quality adjusted life expectancy (QALE)**
using health-related quality of life.

**Health gaps**, on the other hand, represent the difference between the
current health status of a population and an ideal health situation,
providing a measure of the potential for improvement in health outcomes.
Divided into the **disability adjusted life year (DALY)**[@murray1994a]
in the global burden of disease (GBD) study and the **quality adjusted
life year (QALY)** used as the outcome index of the cost-utility
analysis.

In the contemporary era, the integration of computing power and big data
has further transformed health metrics. Today, we have complex models,
machine learning algorithms, and global health databases that allow for
real-time tracking of diseases and health outcomes.

John Graunt's work may have been a humble beginning, but it set in
motion a scientific approach to understanding population health. The
journey from *Graunt's Bills of Mortality* [@johngra2024a] to today's
sophisticated health metrics reflects the continuous quest to quantify
and comprehend the complexities of human health and disease.

### Quality-Adjusted Life Years (QALYs)

The concept of **Quality-Adjusted Life Years (QALYs)** emerged in the
late 1970s, originating from the idea of combining both the *quantity*
and *quality* of life by *Joseph S. Pliskin*, then it became a widely
used metric for evaluating the *cost-effectiveness* of healthcare
interventions [@quality-2023]. One *QALY* corresponds to one year in
perfect health, it ranges from 1 (perfect health) to 0 (dead). However,
this metric, used in various sectors such as health insurance, is
somewhat criticized due to its lack of addressing disparities among
individuals. It might result in discrimination against individuals with
specific medical conditions or disabilities.

A controversial result followed 25 years of research and studies due to
the concerning validity of the metric to be inclusive. Started in the
early 1989s, to step in the 2010 with the *European Commission*
[@echoutco2016] which started the largest-ever study specifically
dedicated to testing the assumptions of the *QALY* [@holmes2013], ending
in 2013 with the recommended "not using QALYs in healthcare decision
making", arguing that patients with disabilities are valued less under a
QALY-based system then individuals with no disabilities. This result has
been further criticized and remains a topic of debate.

> **"Thus, medicine cannot by itself determine the quality of life. It
> can only help people to achieve the state of health that enables them
> to cultivate the art of life-but in their own way. This implies the
> ability to enjoy the fundamental satisfactions of the biological joie
> de vivre. It implies also the ability for each person to do what he
> wants to do and become what he wants to become, according to human
> values that transcend medical judgment." R.Dubos[@dubos1976a]**

### Disability-Adjusted Life Years (DALYs)

In the 1990s, the World Bank financed a study at Harvard University to
develop a sustainable index that considered not only the health status
but also the identification of the level of disabilities concerned by
disparities. This lead to a new way of comparing the overall health and
life expectancy of different countries, released along with the Global
Burden of Disease Study in 1990 [@murray1994]; a project that involved
*Christopher J. L. Murray* [@christop2024] and *Alan Lopez*
[@alanlop2023], in collaboration with the World Health Organization
(WHO). This effort resulted in the development of the
Disability-Adjusted Life Years (DALYs) [@disabili2023a], a metric that
combines **years of life lost due to premature death** and **years lived
with disability**.

DALYs are connected with QALYs in that both metrics seek to quantify the
impact of health on quality and length of life. However, while QALYs
focus on the benefits with and without medical intervention, DALYs
measure the overall burden of diseases.

The DALY metric is calculated by the summing the **years of life lost
(YLLs)** due to premature death and the **years lived with disabilities
(YLDs)**. This metric provides a comprehensive measure of a population’s
health status by highlighting both mortality and disability. For
example, consider a population with a life expectancy of 80 years.
Individuals who die before reaching this age contribute to the YLLs,
while those living with disabilities add to the YLDs. The sum of these
values gives the total DALYs for the population, which can then be
compared globally to assess health status and identify areas needing
improvement in facilities, research, or investment.

### Healthy-Adjusted Life Years (HALY)

Healthy Adjusted Life Years (HALY) is a metric that combine aspects of
both DALYs and QALYs to provide a comprehensive measure of health
outcomes in a population. HALYs quantify the burden of disease and the
overall quality of life experienced by individuals, measuring the number
of years of healthy life lived by a population while accounting for both
mortality and morbidity. The information about the prevalence of
diseases and disabilities, as well as the impact of these conditions on
individuals' quality of life is incorporated in the calculation of
HALYs.

HALYs adjust life expectancy levels based on the prevalence of health
conditions and their associated disability weights. Disability weights
reflect the severity of different health conditions and their impact on
daily functioning. By applying these weights to the prevalence of each
condition, researchers can estimate the overall burden of disease in
terms of years of healthy life lost, similar to the calculation of
DALYs. Unlike DALYs, which focus primarily on morbidity and mortality,
HALYs incorporate measures of individuals’ quality of life, allowing for
a more comprehensive assessment of health outcomes.

### Healthy-Adjusted Life Expectancy (HALE)

Healthy Life Expectancy (HALE) is a measure of overall health and
well-being that accounts for both the quantity and quality of life. HALE
combines years of life expectancy with the prevalence and severity of
disability in a population, providing a more nuanced view of health
outcomes than traditional measures of life expectancy.

HALE provides a more comprehensive view of health outcomes than other
traditional measures of life expectancy, which only consider the
quantity of life. The calculation of HALE typically involves estimating
the number of years that an individual can expect to live in good
health, taking into account the impact of diseases and injuries on
quality of life. This information is then used to estimate the overall
health status of a population. It is a useful tool for public health
practitioners and policy makers, as it provides a more nuanced view of
the health outcomes of a population. This information can help inform
public health interventions and prioritize resources, as well as help
track changes in health outcomes over time. Additionally, HALE can be
used to compare the health outcomes of different populations and
identify disparities in health outcomes, which can help inform targeted
public health interventions.

Overall, the HALE metric provides a valuable perspective on the overall
health and well-being of a population, combining information about both
quantity and quality of life to provide a comprehensive view of health
outcomes. It adjusts overall life expectancy by the amount of time lived
in less than perfect health. This is calculated by subtracting from the
life expectancy a figure which is the number of years lived with
disability multiplied by a weighting to represent the effect of the
disability[^01-metrics_evaluation-1][@healthy].

[^01-metrics_evaluation-1]: [www.healthknowledge.org.uk](https://www.healthknowledge.org.uk/e-learning/health-information/population-health-specialists/lifetables-hales-dalys-pylls#:~:text=Health%20Adjusted%20Life%20Expectancy%20(HALE)&text=It%20adjusts%20overall%20life%20expectancy,the%20effect%20of%20the%20disability.)

### Healthy Life Years (HLY)

The **Healthy Life Years** (**HLY**) indicator, also known as
**disability-free life expectancy** (**DFLE**) or **Sullivan's Index**
[@healthy2024], assesses the number of years an individual can expect to
live in good health, free from significant health problems or
disabilities. It is a measure of the *quality of life-adjusted life
expectancy*, focusing on the years lived in full health rather than
simply the total number of years lived. It specifically focuses on the
number of years a person can expect to live without disability,
providing valuable insights into overall health and well-being beyond
just life expectancy. It takes into account both mortality and
morbidity, providing a comprehensive picture of overall health and
well-being.

HLY is typically calculated by subtracting the number of years lived
with disability from life expectancy at birth. The resulting figure
represents the number of years an individual can expect to live in good
health, free from significant health-related limitations. Unlike life
expectancy, which focuses solely on the length of life, HLY incorporates
measures of the quality of life by considering the impact of health
conditions on individuals' ability to engage in daily activities and
maintain independence.

In terms of policy implications, HLY is used to inform health policy
decisions by identifying areas where interventions are needed to promote
healthy aging and improve overall population health. By focusing on the
number of years lived in good health, policymakers can target
interventions to prevent and manage chronic diseases and disabilities
effectively.

## How the Metrics are Used in Global Health

These metrics are used alone or in combination with other health metrics
in global health assessments conducted by organizations such as the
World Health Organization (WHO), the European Commission, and the
Institute for Health Metrics and Evaluation (IHME). They provide a
global perspective and insights into the health status of populations,
helping to track progresses towards long-term health objectives over
time. The metrics also inform public health policies, healthcare
interventions, and resource allocation decisions, helping to prioritize
health needs and target interventions effectively.

The investigation into finding updated metrics is a sustained
commitment, aiming to develop new measures tailored to the evolving
landscape of the health field. Recently, a new **Summary Measure of
Population Health (SMPH)** was proposed, known as **Well-being-Adjusted
Health Expectancy (WAHE)**. This measure provides an estimate of the
number of life years equivalent to full
health[@muszynska-spielauer2022]. WAHE incorporates both the traditional
health status indicators and well-being factors, thus offering a more
comprehensive assessment of the population's health that aligns with
contemporary health priorities. This approach reflects a growing
recognition that health metrics should account for physical and clinical
indicators and consider psychological and overall well-being.

## Conclusion

Health metrics play a crucial role in understanding and improving
population health. From John Graunt’s early work in the 17th century to
today’s sophisticated metrics like DALYs, QALYs, HALYs, HALE, and HLY,
these measures have evolved to provide a comprehensive view of health
outcomes. They guide public health policies, healthcare interventions,
and global health assessments, highlighting areas for improvement and
helping to allocate resources effectively. As health metrics continue to
evolve, incorporating both traditional and well-being indicators, they
will remain essential tools in the quest to quantify and enhance the
complexities of human health and disease.
