# Causes and Risks {#sec-01.3-Causes_and_Risks}

::: callout
## Learning Objectives:

-   Overview of the causes and risks
-   Identify the objective of the research question
-   Causal inference
:::

## Conditions and injuries

Conditions and injuries associated with the burden of disease and injury vary according to specific causes and risks. In this book causes and risk factors include:

-   **Lifestyle choices**: Poor diet, physical inactivity, tobacco use, and excessive alcohol consumption are major risk factors for many chronic diseases and injuries, including heart disease, stroke, cancer, and liver disease.

-   **Environmental factors**: Exposure to pollutants, such as air pollution and toxic chemicals, can increase the risk of certain diseases and injuries.

-   **Infections**: Many diseases, such as tuberculosis, HIV/AIDS, and malaria, are caused by infectious agents.

-   **Poverty**: People living in poverty are often more susceptible to health problems due to limited access to healthcare, healthy food, and safe living conditions.

-   **Aging**: As people get older, they are at an increased risk of many health problems, including chronic diseases and disabilities.

-   **Genetics**: Some diseases and injuries are caused by genetic factors, such as a genetic predisposition to certain cancers.

-   **Injuries**: Injuries, such as falls, road traffic accidents, and violence, can also contribute to the burden of diseases and injuries.

\

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-cap: "Causal relationships leading to fueling DALYs value"
# install.packages("devtools")
# devtools::install_github("r-causal/ggdag")
library(ggdag)
dag <- dagify(
  "DALY"~"W"+"YLD"+"YLL", # mortality, age,sex, cause
  "Cause"~"age"+"sex"+"place"+"race"+"other", # age, sex, cause
  "W"~"Cause", # age, sex, cause
  "YLD"~"W", # age, sex, cause, disability weights
  "YLL"~"W",
  coords=time_ordered_coords()
)

ggdag(dag)+theme_dag_blank()
```

A particular health condition can have multiple causes (co-morbidities) and risk factors. For instance, a poverty status and the lack of access to healthcare facilities, is proven to be increasing the risk of infectious diseases, while poor diet and physical inactivity can increase the risk of chronic diseases. Acting in favor of addressing the *underlying causes and risk factors* for diseases and injuries is crucial for prompt public health interventions and can help reduce the overall burden of disease.

## Risk Measures

In the following section we will step into key risk measures such as:

-   Risk-specific exposures

<!-- -->

-   Relative risks (RRs)

-   Theoretical minimum-risk exposure levels (TMRELs)

-   Population attributable fractions (PAFs)

### Risk-specific exposures

The quantification of risks and causes involves the evaluation of a set of behavioral, environmental and occupational, and metabolic risks. Pairs of risk-outcome are investigated based on observations and statistical evidences. Convincing evidences consist of plausible associations between exposure and disease in terms of size, duration and effects.

Risk combinations can be **additive** (the occurrence of a least one event, A or B), **multiplicative** (the occurrence of both of two events, A and B) or just **interactive,** acting to influence other pairs, this action is generally identified as possible **confounding**, to be distinguished by factors in the causal pathway between exposure and outcome.

To have an idea of the impact of different risk factors on a cause of illness, the **Socio-demographic Index (SDI)** provide insights into the potential magnitude of social, cultural and demographic factors looking at the risk exposures and possible paths for policy interventions. The Life Expectancy level is closely correlated to the level of the SDI indicator as it is based on average income per person, educational attainment, and total fertility rate (TFR). Higher SDI values typically indicate better socio-economic conditions, including improved access to healthcare, education, and sanitation, which can mitigate various health risks. Conversely, lower SDI values are associated with higher risk exposure due to limited access to healthcare, poorer living conditions, and other socio-economic challenges.

One more element to take into consideration is the **Comparative Risk Assessment (CRA)**[@stanaway2018] divided into attributable and avoidable burden. Considering as the objective the potential reduction of future disease burden, four types of **minimum risk exposure** distributions are identified:

-   theoretical

-   plausible

-   feasible

-   cost-effective

The following provide a high level overview on quantifying attributable burden by using the theoretical minimum risk.

### Relative risks (RRs)

The Relative Risk (or Risk Ratio) for the risk-outcome pairs, mortality and morbidity calculation passes through the decomposition of the attributable burden to a risk. The sum of all-risk exposure effect is accounted for location, age, sex, and mediation-cause. And as said the exposure of all-risks for a disease is split in metabolic, behavioral and environmental risks. Each factor contributes to the overall change.

The relative risk is the ratio between the proportions of exposed and unexposed groups.

$$
RR= \frac{p_0}{p_1}
$$

where $p_0$ and $p_1$ are the proportions of unexposed and exposed group respectively. Or, in terms of population, these group would approx the values of the real population:

$$
RR = \frac{p_1}{p_0}=\frac{d_1/n_1}{d_0/n_0}
$$

### Theoretical minimum-risk exposure levels

(TMRELs)

Risk factors associated with a particular health condition are considered based on the *Theoretical minimum risk exposure levels (TMRELs)* and as a **function of the risk exposure or *relative risk (RR)* value**. Not all the variables that are thought to be risk factors increasing causes for a particular health condition are always the driving cause of the condition, for this reason a minimum level of risk exposure is established for the risk to be considered involved as effective in the outcome.

Moreover, disease attributable to a particular risk factor or combination of risk factors need to be ascertain investigating the risk-outcome relationship. Risk factors can also act indirectly on the outcome via intermediate risks, such as the association of low fruit consumption and heart disease influenced by systolic blood pressure which acts as mediator between the two.

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 0.5
#| fig-align: "center"
library(dagitty)
fhp <- dagitty('dag{ "low fruit" -> "systolic blood pressure" -> "heart disease"}')
coordinates(fhp) <- list( x=c("low fruit"=0,
                              "systolic blood pressure"=0.5,
                              "heart disease"=1),
                          y=c("low fruit"=0,
                              "systolic blood pressure"=0,
                              "heart disease"=0) )
rethinking::drawdag(fhp)
```

As an example of risk factors which met the minimum level (TMRELs) are **particulate matter air pollution**, **high systolic blood pressure**, and **smoking**, these factors are expected to be the *leading contributors* to the overall burden of diseases (cit.ihme). Although, it is not always the case when looking closer at a more specific type of disease; the outcome varies based on location, sex, and age.

In terms of DALYs, the level of the disability adjusted life years is highly influenced by behavioral, environmental, and occupational risks.

### Health Risk by Exposure

The GBD evaluated 87 risk factors from 1990 to 2019.

#### Relative Risk

### Summarizing the relationship between risk and outcome

## Causal inference

Causality concerns the relationship between two variables, where one variable (the cause) influences or determines the other variable (the effect). For example, regular exercise leads to improved cardiovascular health, or adequate sleep contributes to better cognitive function. However, it's essential to distinguish causality from correlation, which denotes the statistical association between two variables. For instance, the correlation between watching television and obesity rates does not imply a causal link. To establish causality, one must systematically consider alternative explanations or confounding factors that may influence both the cause and the effect.

As the foundation of prediction, causal inference investigates on the underlining causes of a condition or a phenomenon. These causes might not immediately visible but need to be discovered passing through various steps of data analysis of the observed data to reveal the hidden threads of cause.

Performing casual inference requires a settlement of an experiment, where there are **treatment** and **outcome** elements. The treatment is the action taken to the data, in the example of regular exercise leads to improved cardiovascular health, to state that this statement is true, an intervention is done to the data, simulating the intervening of another factor variable on the relationship. The formal name of an intervention is "treatment", and to investigate the relationship between exercise and the improvement of cardiovascular health, a treatment would be seeing how data can change if regular portions of fruits is eaten.

So, now the elements factor of the investigation are exercise and potion of fruits, and this is done to see how the combination of the two would improve or not cardiovascular health. Once the intervention is analysed looking at how the response variable, in this case cardiovascular health, changed due to treatment, then the next step would be to apply a control procedure to

It is useful, especially when the result of an intervention is uncertain to apply a **counter-factual** scenario.
