# Causes and Risks {#sec-01.3-Causes_and_Risks}

::: callout
## Learning Objectives:

-   Overview of the causes and risks
-   Identify the objective of the research question
-   Causal inference
:::

## Conditions and injuries

Conditions and injuries associated with the burden of disease and injury vary according to specific causes and risks. In this book causes and risk factors include:

-   **Lifestyle choices**: Poor diet, physical inactivity, tobacco use, and excessive alcohol consumption are major risk factors for many chronic diseases and injuries, including heart disease, stroke, cancer, and liver disease.

-   **Environmental factors**: Exposure to pollutants, such as air pollution and toxic chemicals, can increase the risk of certain diseases and injuries.

-   **Infections**: Many diseases, such as tuberculosis, HIV/AIDS, and malaria, are caused by infectious agents.

-   **Poverty**: People living in poverty are often more susceptible to health problems due to limited access to healthcare, healthy food, and safe living conditions.

-   **Aging**: As people get older, they are at an increased risk of many health problems, including chronic diseases and disabilities.

-   **Genetics**: Some diseases and injuries are caused by genetic factors, such as a genetic predisposition to certain cancers.

-   **Injuries**: Injuries, such as falls, road traffic accidents, and violence, can also contribute to the burden of diseases and injuries.

\

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-cap: "Causal relationships leading to fueling DALYs value"
# install.packages("devtools")
# devtools::install_github("r-causal/ggdag")
library(ggdag)
dag <- dagify(
  "DALY"~"W"+"YLD"+"YLL", # mortality, age,sex, cause
  "Cause"~"age"+"sex"+"place"+"race"+"other", # age, sex, cause
  "W"~"Cause", # age, sex, cause
  "YLD"~"W", # age, sex, cause, disability weights
  "YLL"~"W",
  coords=time_ordered_coords()
)

ggdag(dag)+theme_dag_blank()
```

A particular health condition can have multiple causes (co-morbidities) and risk factors. For instance, a poverty status and the lack of access to healthcare facilities, is proven to be increasing the risk of infectious diseases, while poor diet and physical inactivity can increase the risk of chronic diseases. Acting in favor of addressing the *underlying causes and risk factors* for diseases and injuries is crucial for prompt public health interventions and can help reduce the overall burden of disease.

## Risk Metrics

Risk factors associated with a particular health condition are considered based on the *Theoretical minimum risk exposure levels (TMRELs)* and as a function of the risk exposure or *relative risk (RR)* value. Not all the variables that are thought to be risk factors increasing causes for a particular health condition are always the driving cause of the condition, for this reason a minimum level of risk exposure is established for the risk to be considered involved as effective in the outcome.

Moreover, disease attributable to a particular risk factor or combination of risk factors need to be ascertain investigating the risk-outcome relationship. Risk factors can also act indirectly on the outcome via intermediate risks, such as the association of low fruit consumption and heart disease influenced by systolic blood pressure which acts as mediator between the two.

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 0.5
#| fig-align: "center"
library(dagitty)
fhp <- dagitty('dag{ "low fruit" -> "systolic blood pressure" -> "heart disease"}')
coordinates(fhp) <- list( x=c("low fruit"=0,
                              "systolic blood pressure"=0.5,
                              "heart disease"=1),
                          y=c("low fruit"=0,
                              "systolic blood pressure"=0,
                              "heart disease"=0) )
rethinking::drawdag(fhp)
```

As an example of risk factors which met the minimum level (TMRELs) are **particulate matter air pollution**, **high systolic blood pressure**, and **smoking**, these factors are expected to be the *leading contributors* to the overall burden of diseases (cit.ihme). Although, it is not always the case when looking closer at a more specific type of disease; the outcome varies based on location, sex, and age.

In terms of DALYs, the level of the disability adjusted life years is highly influenced by behavioral, environmental, and occupational risks.

### Health Risk by Exposure

The GBD evaluated 87 risk factors from 1990 to 2019.

#### Relative Risk

### Summarizing the relationship between risk and outcome

## Causal inference

Causality concerns the relationship between two variables, where one variable (the cause) influences or determines the other variable (the effect). For example, regular exercise leads to improved cardiovascular health, or adequate sleep contributes to better cognitive function. However, it's essential to distinguish causality from correlation, which denotes the statistical association between two variables. For instance, the correlation between watching television and obesity rates does not imply a causal link. To establish causality, one must systematically consider alternative explanations or confounding factors that may influence both the cause and the effect.

As the foundation of prediction, causal inference investigates on the underlining causes of a condition or a phenomenon. These causes might not immediately visible but need to be discovered passing through various steps of data analysis of the observed data to reveal the hidden threads of cause.

Performing casual inference requires a settlement of an experiment, where there are **treatment** and **outcome** elements. The treatment is the action taken to the data, in the example of regular exercise leads to improved cardiovascular health, to state that this statement is true, an intervention is done to the data, simulating the intervening of another factor variable on the relationship. The formal name of an intervention is "treatment", and to investigate the relationship between exercise and the improvement of cardiovascular health, a treatment would be seeing how data can change if regular portions of fruits is eaten.

So, now the elements factor of the investigation are exercise and potion of fruits, and this is done to see how the combination of the two would improve or not cardiovascular health. Once the intervention is analysed looking at how the response variable, in this case cardiovascular health, changed due to treatment, then the next step would be to apply a control procedure to

It is useful, expecially when the result of an intervention is uncertain to apply a **counter-factual** scenario.
