# Causes and Risks {#sec-01.3-Causes_and_Risks}

::: callout
## Learning Objectives:

-   Overview of the causes and risks
-   Identify the objective of the research question
-   Causal inference
:::

## Conditions and injuries

Conditions and injuries associated with the burden of disease and injury vary according to specific causes and risks. In this book causes and risk factors include:

-   **Lifestyle choices**: Poor diet, physical inactivity, tobacco use, and excessive alcohol consumption are major risk factors for many chronic diseases and injuries, including heart disease, stroke, cancer, and liver disease.
-   **Environmental factors**: Exposure to pollutants, such as air pollution and toxic chemicals, can increase the risk of certain diseases and injuries.
-   **Infections**: Many diseases, such as tuberculosis, HIV/AIDS, and malaria, are caused by infectious agents.
-   **Poverty**: People living in poverty are often more susceptible to health problems due to limited access to healthcare, healthy food, and safe living conditions.
-   **Aging**: As people get older, they are at an increased risk of many health problems, including chronic diseases and disabilities.
-   **Genetics**: Some diseases and injuries are caused by genetic factors, such as a genetic predisposition to certain cancers.
-   **Injuries**: Injuries, such as falls, road traffic accidents, and violence, can also contribute to the burden of diseases and injuries.

\

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-cap: "Causal relationships leading to fueling DALYs value"
# install.packages("devtools")
# devtools::install_github("r-causal/ggdag")
library(ggdag)
dag <- dagify(
  "DALY"~"W"+"YLD"+"YLL", # mortality, age,sex, cause
  "Cause"~"age"+"sex"+"place"+"race"+"other", # age, sex, cause
  "W"~"Cause", # age, sex, cause
  "YLD"~"W", # age, sex, cause, disability weights
  "YLL"~"W",
  coords=time_ordered_coords()
)

ggdag(dag)+theme_dag_blank()
```

A particular health condition can have multiple causes (co-morbidities) and risk factors. For instance, a poverty status and the lack of access to healthcare facilities, is proven to be increasing the risk of infectious diseases, while poor diet and physical inactivity can increase the risk of chronic diseases. Acting in favor of addressing the *underlying causes and risk factors* for diseases and injuries is crucial for prompt public health interventions and can help reduce the overall burden of disease.

## Risk Measures

In the following section we will step into key risk measures such as:

-   Risk-specific exposures
-   Relative risks (RRs)
-   Theoretical minimum-risk exposure levels (TMRELs)
-   Population attributable fractions (PAFs)

### Risk-specific exposures

The quantification of risks and causes involves the evaluation of a set of behavioral, environmental and occupational, and metabolic risks. Pairs of risk-outcome are investigated based on observations and statistical evidences. Convincing evidences consist of plausible associations between exposure and disease in terms of size, duration and effects.

Risk combinations can be **additive** (the occurrence of a least one event, A or B), **multiplicative** (the occurrence of both of two events, A and B) or just **interactive,** acting to influence other pairs, this action is generally identified as possible **confounding**, to be distinguished by factors in the causal pathway between exposure and outcome.

To have an idea of the impact of different risk factors on a cause of illness, the **Socio-demographic Index (SDI)** provide insights into the potential magnitude of social, cultural and demographic factors looking at the risk exposures and possible paths for policy interventions. The Life Expectancy level is closely correlated to the level of the SDI indicator as it is based on average income per person, educational attainment, and total fertility rate (TFR). Higher SDI values typically indicate better socio-economic conditions, including improved access to healthcare, education, and sanitation, which can mitigate various health risks. Conversely, lower SDI values are associated with higher risk exposure due to limited access to healthcare, poorer living conditions, and other socio-economic challenges.

One more element to take into consideration is the **Comparative Risk Assessment (CRA)**[@stanaway2018] divided into attributable and avoidable burden. Considering as the objective the potential reduction of future disease burden, four types of **minimum risk exposure** distributions are identified:

-   theoretical
-   plausible
-   feasible
-   cost-effective

The following provide a high level overview on quantifying attributable burden by using the theoretical minimum risk.

### Relative Risks (RRs)

The Relative Risk (or Risk Ratio) for the risk-outcome pairs, mortality and morbidity calculation passes through the decomposition of the attributable burden to a risk. The sum of all-risk exposure effect is accounted for location, age, sex, and mediation-cause. And as said the exposure of all-risks for a disease is split in metabolic, behavioral and environmental risks. Each factor contributes to the overall change.

The relative risk is the ratio between the proportions of exposed and unexposed groups.

$$
RR= \frac{p_0}{p_1}
$$

where $p_1$ and $p_0$ are the proportions of exposed and unexposed group respectively. Or, in terms of population, these group would approx the values of the real population:

$$
RR = \frac{p_1}{p_0}=\frac{d_1/n_1}{d_0/n_0}
$$

where $d_1/n_1$ and $d_0/n_0$ are the proportion of the population with and without the disease.

For example, let's say we are studying the association between smoking (exposure) and lung cancer (outcome). We want to calculate the relative risk of lung cancer among smokers compared to non-smokers. If the relative risk is 2, it means that smokers are twice as likely to develop lung cancer compared to non-smokers.

```{r}
# Create a 2x2 table of exposure and outcome
exposed <- c(50, 10)   # Number of exposed individuals with and without the outcome
unexposed <- c(20, 5)  # Number of unexposed individuals with and without the outcome

# Calculate the relative risk
relative_risk <- function(exposed, unexposed) {
  (exposed[1] / sum(exposed)) / (unexposed[1] / sum(unexposed))
}

# Print the relative risk
relative_risk(exposed, unexposed)

```

```{r}
# Define the number of events and person-time at risk for exposed and unexposed groups
d1 <- 50  # Number of events in the exposed group
n1 <- 10  # Person-time at risk in the exposed group
d0 <- 20  # Number of events in the unexposed group
n0 <- 5   # Person-time at risk in the unexposed group

# Calculate the relative risk 
relative_risk_d <- (d1 / n1) / (d0 / n0)

# Print the relative risk
relative_risk_d

```

-   The first formula, $RR=p_1/p_0$, calculates the relative risk directly using the proportions of events $p_1$ in the exposed group compared to the unexposed group $p_0$ . This formula provides a more simplified view of the relative risk based solely on event proportions.

-   The second formula, $RR=\frac{d_1/n_1}{d_0/n_0}$, considers both the **number of events** $(d1, d0)$ and **person-time at risk** $(n1, n0)$ for each group. This formula takes into account the **incidence rate** in addition to **event proportions**, providing a more specific understanding of the relative risk by incorporating information about the **duration of exposure**.

#### Relative Risks and Network Analysis

In some cases, relative risks can be modeled using **network analysis**, a specialized approach within statistical modeling which extend the concept of mixed effects to compare multiple treatments while accounting for various factors and dependencies.

The relationship between variables, represented by nodes and edges considering potential interactions or dependencies between them. This approach is generally favorable when exploring complex relationships among multiple variables.

To represent the network we can use a Directed Acyclic Graph (DAG), this graphs also help identifying the causal relationship between treatments or interventions and let understand the action of influence within the network. 
```{r}
# Load the library
library(ggdag)
set.seed(555)
# Define the DAG structure
dag <- dagify(outcome ~ exposure + c1 + c2 + c3, 
              exposure ~ c1 + c2 + c3,
              c1 ~ c2 + c3,
              c1 ~ c3)

# Plot the DAG
ggdag(dag)+theme_dag_blank()

```

```{r}
# Load necessary libraries
library(dagitty)
library(tidyverse)

# Create DAG
dag <- dagitty('dag {
    E -> O
    C1 -> O
    C2 -> O
    C3 -> O
}')

dat <- dag%>%tidy_dagitty()
# Generate data
set.seed(123)
n <- 1000
data <- simulateLogistic(dag)

# Fit logistic regression model
model <- glm(O ~ E + C1 + C2 + C3, data = data, 
             family = "binomial")

# Extract estimated probabilities
prob_outcome_given_exposure <- predict(model, type = "response")
prob_outcome_given_no_exposure <- predict(model, newdata = data.frame(E = '1', C1 = data$C1, C2 = data$C2, C3 = data$C3), 
                                          type = "response")

# Calculate relative risk
relative_risk <- prob_outcome_given_exposure / prob_outcome_given_no_exposure
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
data$relative_risk <- relative_risk
# Visualize relative risk distribution
ggplot(data = data, aes(x =relative_risk)) +
  geom_histogram(binwidth = 0.2, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Relative Risk",
       x = "Relative Risk",
       y = "Frequency") +
  theme_minimal()

```
```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
# Visualize relative risk distribution with density plot
ggplot(data = data, aes(x = relative_risk)) +
  geom_density(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Relative Risk",
       x = "Relative Risk",
       y = "Density") +
  theme_minimal()

```


### Theoretical minimum-risk exposure levels (TMRELs)

Risk factors associated with a particular health condition are considered based on the *Theoretical minimum risk exposure levels (TMRELs)* and as a **function of the risk exposure or *relative risk (RR)* value**. Not all the variables that are thought to be risk factors increasing causes for a particular health condition are always the driving cause of the condition, for this reason a minimum level of risk exposure is established for the risk to be considered involved as effective in the outcome.

Moreover, disease attributable to a particular risk factor or combination of risk factors need to be ascertain investigating the risk-outcome relationship. Risk factors can also act indirectly on the outcome via intermediate risks, such as the association of low fruit consumption and heart disease influenced by systolic blood pressure which acts as mediator between the two.

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 0.5
#| fig-align: "center"
library(dagitty)
fhp <- dagitty('dag{ "low fruit" -> "systolic blood pressure" -> "heart disease"}')
coordinates(fhp) <- list( x=c("low fruit"=0,
                              "systolic blood pressure"=0.5,
                              "heart disease"=1),
                          y=c("low fruit"=0,
                              "systolic blood pressure"=0,
                              "heart disease"=0) )
rethinking::drawdag(fhp)
```

As an example of risk factors which met the minimum level (TMRELs) are **particulate matter air pollution**, **high systolic blood pressure**, and **smoking**, these factors are expected to be the *leading contributors* to the overall burden of diseases (cit.ihme). Although, it is not always the case when looking closer at a more specific type of disease; the outcome varies based on location, sex, and age.

In terms of DALYs, the level of the disability adjusted life years is highly influenced by behavioral, environmental, and occupational risks.

### Population attributable fractions (PAFs)

The Population Attributable Fraction (PAF) is a measure used to quantify the proportion of disease incidence in a population that can be attributed to a specific risk factor. It represents the proportion of risk that would be reduced in a given year if the exposure to a risk factor in the past were reduced to an ideal exposure scenario.

PAF is calculated based on the prevalence of the risk factor in the population and the relative risk associated with that risk factor. The formula for calculating PAF is as follows:

$$
PAF = \frac{{P_e \times (RR - 1)}}{{1 + P_e \times (RR - 1)}}
$$

Where:

- $P_e$ is the prevalence of the risk factor in the population.
- $RR$ is the relative risk associated with the risk factor, representing the increased risk of disease among individuals exposed to the risk factor compared to those who are not exposed.

The PAF ranges from 0% to 100%. A PAF of 0% indicates that the risk factor has no impact on the incidence of the disease, while a PAF of 100% indicates that all cases of the disease in the population can be attributed to the risk factor.

PAF is useful for public health interventions as it provides insight into the potential impact of reducing or eliminating a specific risk factor on the incidence of disease in the population. By targeting interventions to reduce exposure to the risk factor, public health efforts can effectively reduce the burden of disease in the population and improve overall health outcomes.


## Causal inference

Causality concerns the relationship between two variables, where one variable (the cause) influences or determines the other variable (the effect). For example, regular exercise leads to improved cardiovascular health, or adequate sleep contributes to better cognitive function. However, it's essential to distinguish causality from correlation, which denotes the statistical association between two variables. For instance, the correlation between watching television and obesity rates does not imply a causal link. To establish causality, one must systematically consider alternative explanations or confounding factors that may influence both the cause and the effect.

As the foundation of prediction, causal inference investigates on the underlining causes of a condition or a phenomenon. These causes might not immediately visible but need to be discovered passing through various steps of data analysis of the observed data to reveal the hidden threads of cause.

Performing casual inference requires a settlement of an experiment, where there are **treatment** and **outcome** elements. The treatment is the action taken to the data, in the example of regular exercise leads to improved cardiovascular health, to state that this statement is true, an intervention is done to the data, simulating the intervening of another factor variable on the relationship. The formal name of an intervention is "treatment", and to investigate the relationship between exercise and the improvement of cardiovascular health, a treatment would be seeing how data can change if regular portions of fruits is eaten.

So, now the elements factor of the investigation are exercise and potion of fruits, and this is done to see how the combination of the two would improve or not cardiovascular health. Once the intervention is analysed looking at how the response variable, in this case cardiovascular health, changed due to treatment, then the next step would be to apply a control procedure to

It is useful, especially when the result of an intervention is uncertain to apply a **counter-factual** scenario.

## Summarizing the relationship between risk and outcome

The relationship between risk and outcome in epidemiology is fundamental for understanding the causes and prevention of disease. It involves assessing how exposure to certain risk factors influences the likelihood of developing a particular health outcome. Epidemiological studies examine the association between risk factors and health outcomes, quantifying the strength of this association through measures like relative risk (RR). Relative risk compares the risk of developing the outcome among individuals exposed to the risk factor to the risk among those who are not exposed, with an RR greater than 1 indicating an increased risk associated with exposure. Population Attributable Fraction (PAF) measures the proportion of disease incidence in a population that can be attributed to a specific risk factor, aiding in estimating the potential impact of reducing exposure to the risk factor on overall disease burden. Establishing causality requires demonstrating consistent associations, dose-response relationships, temporal precedence, and ruling out alternative explanations. Ultimately, understanding the relationship between risk factors and outcomes informs preventive strategies, public health interventions, and improves population health outcomes through evidence-based decision-making in public health.

In conclusion, the study of risk and outcome extends beyond traditional epidemiological methods to incorporate advanced techniques like transfer learning. This interdisciplinary approach allows for the application of insights gained from epidemiological research to other fields and viceversa, enhancing our understanding of complex relationships between risk factors and health outcomes. By leveraging learning techniques, we can identify patterns, extract meaningful insights, and develop predictive models that transcend traditional boundaries, offering new perspectives on population health and guiding targeted interventions. Thus, the integration of epidemiology with emerging technologies opens doors to innovative solutions and holistic approaches to address public health challenges and improve overall well-being.
