# Health Metrics {#sec-02-health_metrics}

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   A closer look at the metrics
-   Understand how to use them
-   Think about potentiality for improvements
:::
:::

## Methods and Calculations

The objective of this book is to assess the burden of diseases, focusing
on disability and premature mortality, which are captured by DALYs. The
main goal is to comprehend the impact of diseases on overall health,
with a specific emphasis on the calculation of DALYs. In the previous
chapter (@sec-01-metrics_evaluation), we defined the metrics to
establish the theoretical framework for constructing them. In this
chapter, we will proceed with the initial calculation of YLLs, YLDs, and
DALYs. This step is essential for understanding the structure of the
metric components, which will be further investigated in the following
chapter (@sec-03-components).

Used to measure the burden of disease and quantify the impact of
diseases and injuries on individuals and populations, these metrics can
help prioritize public health interventions and evaluate the
effectiveness of public health programs.

### YLLs

The Years of Life Lost (YLLs) is a metric that measures the number of
years a person would have lived if they had not died prematurely due to
a disease or injury, YLLs is calculated by subtracting the age at death
from the expected age at death in a population without the disease or
injury.

First mentioned in 1947 by Mary Dempsey[@dempsey2019] to measure the
burden of tuberculosis, it is built for identifying the area where
improvement is required for reducing the loss in health status and
clearly reducing the probability of death [@reiner2022].

This metric was first calculate in late 1940s
(@sec-01-metrics_evaluation), then defined as **Standardized Rate for
Mortality in units of Lost Years of Life** by *William Haenszel*
[@haenszel1950]. The investigation about **What is the leading cause of
death** brought to the development of a measurement that would take
consideration of the **age at death** and put it in comparison with the
**expected life expectancy**. This initially lead to understand that
early deaths had a greater impact than those at later ages. And in
particular it was ascertain that if the number of years were accounted
for a group of people, this would have led to recognize potential life
lost for a certain cause. Since this first approach to measure the cause
of deaths, some adjustments of the standard death rates were made
leading to a new calculation which was built on considering
**standardized deaths rate applied to age-specific factors** that would
have led to death for a specific cause of death.

> The impact on the society due to premature losses involves, not only
> the mere sadness for having lost a beloved one but the lost of
> working-people dying prematurely. It is cost for the entire society.

#### Calculation

$$
YLL=N*le
$$ {#eq-yll}

in @eq-yll, $N$ is the number of premature deaths, and $le$ is the
[expected live expectancy]{.underline}, or the expected numbers of years
someone could live. More specification about the components of YLL are
provided in the next @sec-03-components and a sample of the construction
of a life table with relative estimations of the life expectancy can be
found in @sec-appendixA.

In general, the Global Burden of Diseases studies (GBD) utilize standard
life tables, enabling the measurement of the impact of different
diseases at a consistent level. This standardization is crucial for
comparative studies, particularly cross-country comparisons
[@devleesschauwer2020]. Additionally, country-specific life expectancies
are valuable for investigating premature deaths resulting from
health-related causes, such as fatal diseases. Moreover, certain types
of country-specific life expectancies are frequently recommended due to
their ability to provide insight into longevity characteristics, which
can inform policymakers and aid in prevention efforts, as exemplified by
the Japanese life expectancy [@tsugane2021]. A G7 cross-country study
demonstrated that Japan has the longest average life expectancy,
primarily attributed to significantly low mortality rates from ischemic
heart disease and cancer, which are the leading causes of death in most
countries, as indicated by the GBD study.

As an initial step, we use the deaths data from the GBD 2019 study,
which is stored in the `{hmsidwR}` package. This dataset comprises 2754
observations and 7 variables, containing the estimated number of deaths
due to 9 causes across 6 regions: Global, France, Italy, Germany, the
United Kingdom, and the United States. Our specific task is to calculate
the Years of Life Lost (YLLs) attributable to stroke in the year 2019.

```{r}
#| warning: false
#| message: false
#| output: false

library(tidyverse)
library(hmsidwR)
```

```{r}
#| eval: false
hmsidwR::deaths2019
```

```{r}
#| label: tbl-deaths_stroke
#| tbl-cap: "Deaths due to Stroke"

deaths_stroke <- hmsidwR::deaths2019 %>%
  filter(location=="Global",
         str_detect(cause,"Stroke")) %>%
  select(-location,-cause)

deaths_stroke %>% head()
```

Among the indicators available in the hmsidwR::gho_lifetables dataset,
we specifically focus on the indicator denoted as $ex$, which represents
the expectation of life at age x.

```{r}
#| label: tbl-indicators
#| tbl-cap: "Life Table Indicators"
#| echo: false
hmsidwR::gho_lifetables%>%
  count(indicator)%>%
  select(-n)%>%
  mutate(description=c("person-years lived above age x",
         "expectation of life at age x",
         "number of people left alive at age x",
         "person-years lived between ages x and x+n",
         "age-specific death rate between ages x and x+n",
         "number of people dying between ages x and x+n",
         "probability of dying between ages x and x+n"))%>%
  knitr::kable()
```

The **expectation of life at age x** ($ex$) refers to the life
expectancy at age x. Typically, life expectancy represents the maximum
number of years a person can live. For example, if we consider the
Japanese life expectancy at birth, which is measured in total years,
according to the estimation released in 2021 by the United Nations
Population Division research, it is approximately 84 years.

Here, we use the estimated value of the expected life for a 5-year age
groups for both females and males, indicated by the `le`.

```{r}
#| label: tbl-ex2019
#| tbl-cap: "Expectation of life at age x - 2019"
ex2019 <- hmsidwR::gho_lifetables%>%
          filter(year==2019,
                 indicator=="ex")%>%
            select(-indicator,-year)%>%
            rename(le=value)

ex2019 %>%
  head()
```

```{r}
#| label: tbl-YLL_global_stroke
#| tbl-cap: "Global Stroke YLLs"
YLL_global_stroke <- deaths_stroke %>%
  select(-upper,-lower)%>%
  full_join(ex2019)%>%
  group_by(age,sex)%>%
  reframe(YLL=dx*le)


YLL_global_stroke%>%
  head()
```

```{r}
#| warning: false
#| message: false
#| label: fig-YLL_global_stroke
#| fig-cap: "YLLs due to stroke in the Global region"
#| echo: false
#| eval: true

dat <- YLL_global_stroke%>%
  filter(!sex=="both")

dat %>%  
  ggplot(aes(x=age))+
  geom_line(aes(y=YLL))+
  geom_point(data=dat%>%
               filter(sex=="female"),
             aes(y=YLL),color="pink")+
  geom_point(data=dat%>%
               filter(sex=="male"),
             aes(y=YLL),color="darkgreen")+
  geom_smooth(aes(y=YLL,group=1),
              se=F,color="#29306c")+
  scale_y_continuous(labels = scales::label_comma(scale = 1/1000))+
  labs(title="Global region: Years of Life Lost (YLLs) due to stroke",
       subtitle="Year 2019",
       caption = "DataSource: WHO life expectancy & IHME results Stroke data | YLLs Values in Thousands",x="Age Group",y="YLLs")+
  ggplot2::theme_minimal()+
  theme(axis.text.x = element_text(size=7,
                                   angle = 45,
                                   vjust = 0.5),
        plot.caption = element_text(size=8))
```

### YLDs

**YLDs** (Years Lived with Disability) measures the number of years a
person lives with a disability due to a disease or injury. It is
calculated by multiplying the **prevalence of a condition** by the
**disability weight**, which reflects the severity of the disability.

The key factor of the disability weights (DW) is liked to the severity
(mean of the range of health loss suffered to disease) of a non-fatal
health condition due to disease or injury. DW ranges between 0
(equivalent to full health) and 1 (equivalent to death). The estimation
of the disability weights is challenging and has been continuously
changed by modifying and adapting methodologies in various
studies[@liu2023]. The challenge is assigning disability weights to
diseases with different levels of prevalence and severity, such as cases
with high prevalence and low severity. Furthermore, then the results
need to be reported to a year-based value.

Moreover, the calculation of the YLDs has been updated over the years
passing from incidence to prevalence-based approach. More in-depth
analyses are in @sec-03-components, for now we will be focusing on using
the most updated approach based on values of prevalence estimated by the
GBD2019 which are here summarized in the `hmsidwR::g7_hmetrics` data
set.

#### Calculation

$$
YLD=p*DW
$$ {#eq-yld}

in @eq-yld, $p$ is the prevalence, $DW$ are the disability weights. The
prevalence is the incidence times the duration of the condition.

Since the release of the GBD 2010 and WHO have decided to switch to a
prevalence-based approach to calculation of YLDs. The major impact of
this is to shift the age distribution of YLDs significantly (Figure
2.2). Thus for example YLDs for congenital hearing loss will be spread
relatively evenly across all age groups in the prevalence perspective,
whereas they will all fall at age 0 in an incidence perspective.

In this case we use the `disability weights` and the `severity` levels
extracted from a dataset in the GBD 2019 study. The dataset
`disweights2019` is stored in the `{hmsidwR}` package and is made of 25
observations and 7 variables. It contains the estimated values for the
disability weights as well as the upper and lower estimates. Let's
calculate the `Years Lived with Disabilities (YLDs)` due to **Stroke**
in 2019.

```{r}
#| eval: false
hmsidwR::disweights2019
```

```{r}
#| echo: false
hmsidwR::disweights2019%>%
  select(specification,cause1,severity,dw) %>%
  drop_na()%>%
  head()
```

The level of `severity` is assigned to be:

-   mild = 20%
-   moderate = 45%
-   severe = 75%

This assigned levels are general for all ages, the values might vary for
other specification of the level of disability[@wyper2020].

```{r}
dwsev2019 <- hmsidwR::disweights2019%>%
  select(cause1,severity,dw) %>%
  drop_na() %>%
  mutate(severity_n=case_when(severity=="mild"~0.2,
                              severity=="moderate"~0.45,
                              severity=="severe"~0.75)) 

dwsev2019 %>%
  head()
```

The values for disability weights and severity are here considered for
all ages, in general the level of disability weights can differ by age
for different type of stroke, but here for simplicity consider weights
varying only based on the severity level of stroke.

```{r}
dw_stroke <- dwsev2019 %>%
  filter(cause1=="Stroke")%>%
  group_by(severity,severity_n)%>%
  reframe(avg_dw=mean(dw))
dw_stroke
```

Strokes are associated with `cardiovascular diseases (CVDs)`. In
particular, there are two main types of stroke: ischemic stroke and
hemorrhagic stroke. When looking at the age-standardized prevalence of
cardiovascular disease at the global level, we can calculate the YLD at
the global level age standardized by extracting the incidence and
prevalence values from the table `hmsidwR::g7_hmetrics`. We then
multiply these values by the severity and disability weights, as
follows. We note the difference in the YLDs if incidence or prevalence
is used in the calculation.

```{r}
inc_prev_stroke_as <- g7_hmetrics%>%
  filter(measure%in%c("Incidence","Prevalence"),
         location=="Global",
         cause=="Cardiovascular diseases",
         metric=="Percent",
         year==2019)%>%
  select(measure,sex,val)%>%
  pivot_wider(names_from = "measure",values_from = "val")
inc_prev_stroke_as
```

```{r}
#| label: tbl-YLD_global_stroke_as
#| tbl-cap: "Global YLD due to stroke age standardized"
YLD_global_stroke_as <- merge(inc_prev_stroke_as,dw_stroke) %>%
  group_by(sex)%>%
  mutate(YLD_p=round(Prevalence*severity_n*avg_dw*100,3),
         YLD_i=round(Incidence*severity_n*avg_dw*100,3))%>%
  select(sex,YLD_p,YLD_i)%>%
  summarise(YLD_p=sum(YLD_p),YLD_i=sum(YLD_i))
YLD_global_stroke_as
```

Now, what we want is to calculate the YLDs due to `stroke` for 5-year
age groups at Global level, so then with the YLL values previously
obtained in @tbl-YLL_global_stroke we can calculate the DALYs.

Here we use the incidence and prevalence values for stroke found in
`hmsidwR::g7_inc_prev_CVD2019`.

```{r}
inc_prev_stroke_5y <- hmsidwR::g7_inc_prev_CVD2019 %>% 
  filter(cause=="Stroke",
         location=="Global")%>%
  select(measure,sex,age,val)%>%
  pivot_wider(names_from = "measure",values_from = "val")


inc_prev_stroke_5y %>%
  head()
```

```{r}
#| label: tbl-YLD_global_stroke
#| tbl-cap: "Global YLD due to stroke 5Yrs"
YLD_global_stroke <- merge(inc_prev_stroke_5y,dw_stroke)%>%
  group_by(sex,age) %>%
  reframe(YLD_p=sum(Prevalence*severity_n*avg_dw*100),
         YLD_i=sum(Incidence*severity_n*avg_dw*100))

YLD_global_stroke %>%
  head()
```

```{r}
#| echo: false
#| eval: true
#| label: fig-YLD_global_stroke
#| fig-cap: "YLDs due to stroke in the Global region"

dat <- YLD_global_stroke %>%  
  filter(!sex=="both")

dat %>%
  ggplot(aes(x=age,y=YLD_p))+
  geom_line(aes(y=YLD_p))+
  geom_point(data=dat%>%filter(sex=="female"),
             color="pink")+
  geom_point(data=dat%>%filter(sex=="male"),
             color="darkgreen")+
  geom_smooth(aes(group=1), 
              se=F, color="#29306c")+
  scale_y_continuous(labels = scales::label_comma(scale = 1*1000,
                                                  accuracy = 0.1))+
  labs(title="Global region: Years Lived with Disabilities (YLDs) due to stroke",
       subtitle="Year 2019",
       caption = "DataSource: IHME results Stroke data | YLDs Values X 1000",
       x="Age Group",y="YLDs")+
  theme(axis.text.x = element_text(size=7,
                                   vjust = 0.5),
        plot.caption = element_text(size=8))
```

### DALYs

As a measure of the overall disease burden the DALYs are used to
quantify the sum of **years of potential life lost** due to premature
death **(YLL)** and **years lived with disability (YLD)**.

So, the number of DALYs indicate the number of years of life lost due
premature deaths, disease or injury. This metric takes into account both
the quantity and quality of life.

The DALY is a generalization of the well-known *Potential Years of Life
Lost measure (PYLLs)* which includes the lost of good
health[^02-health_metrics-1]. We do not consider the PYLLs in this book
but more information can be found in the references.

[^02-health_metrics-1]: [@globalh]

#### Calculation

$$
DALYs=YLLs+YLDs
$$ {#eq-daly}

One DALY is one lost year of healthy life, this measure is used to
assess how diseases and injuries impact populations, and provides a
comprehensive picture of the overall burden of disease by combining YLLs
and YLDs in different groups.

The sum of YLLs @tbl-YLL_global_stroke and YLDs @tbl-YLD_global_stroke
releases the overall value of DALYs due to stroke in the Global region.
Here we can see there is no much difference in magnitude if incidence or
prevalence is used.

```{r}
DALY_global_stroke <- merge(YLL_global_stroke,YLD_global_stroke) %>%
  mutate(DALY_p=YLL+YLD_p,DALY_i=YLL+YLD_i)

DALY_global_stroke%>%
  select(sex,age,YLL,YLD_p,YLD_i,DALY_p,DALY_i)%>%
  head()
```

```{r}
#| echo: false
#| eval: true
#| label: fig-DALY_global_stroke
#| fig-cap: "DALYs due to stroke in the Global region"
dat <- DALY_global_stroke %>% 
  select(sex,age,YLL,YLD_p,DALY_p)%>%
  filter(!sex=="both")%>%
  mutate_if(is.numeric,scale)

dat %>%
  ggplot(aes(x=age))+
  geom_line(aes(y=DALY_p))+
  geom_smooth(aes(y=DALY_p,group=1),
              se=F, color="#29306c") +
  geom_line(aes(y=YLL,group=sex,color=sex),
            #color="grey",
            linewidth=0.2)+
  geom_line(aes(y=YLD_p,group=sex,color=sex),
            #color="darkred",
            linewidth=0.2)+
  scale_color_manual(values=c("pink","darkgreen"))+
  geom_point(data=dat%>%filter(sex=="female"),
             aes(y=YLL),color="pink")+
  geom_point(data=dat%>%filter(sex=="male"),
             aes(y=YLL),color="darkgreen")+
  geom_point(data=dat%>%filter(sex=="female"),
             aes(y=YLD_p),color="pink")+
  geom_point(data=dat%>%filter(sex=="male"),
             aes(y=YLD_p),color="darkgreen")+
  scale_y_continuous(labels = scales::label_comma(scale = 1*1000,
                                                  accuracy = 0.1))+
  labs(title="Global region: Disability Adjusted Life Years (DALYs) due to stroke",
       subtitle="Year 2019",
       caption = "DataSource: IHME results Stroke data | DALYs Values ",
       x="Age Group",y="DALYs") +
  theme(axis.text.x = element_text(size=7,angle = 45,vjust = 0.5),
        plot.title = element_text(size=12),
        plot.caption = element_text(size=8),
        legend.position = "bottom")
```

#### How the DALYs are used

The health metrics of DALY, YLL, and YLD can be used in several ways to
help prioritize public health interventions, evaluate the impact of
diseases and injuries, and inform public health decision-making. Some
common uses of these metrics include:

**Prioritizing public health interventions**: By calculating the overall
burden of disease in a population, public health practitioners can
prioritize which diseases and injuries to address first. This helps
allocate resources and target interventions to the areas of greatest
need.

**Evaluating the impact of diseases and injuries**: These metrics can be
used to measure the impact of diseases and injuries on individuals and
populations and to track changes over time. This information can help
inform public health decision-making and allocate resources more
effectively.

**Comparing the burden of disease across populations**: DALY, YLL, and
YLD can be used to compare the burden of disease across populations and
between different regions. This information can help identify
disparities in health outcomes and inform targeted public health
interventions.

**Evaluating the effectiveness of public health programs**: These
metrics can be used to evaluate the impact of public health programs and
to assess the effectiveness of public health interventions. This
information can help public health practitioners identify areas for
improvement and make necessary changes to ensure that programs are
achieving their goals.

**Monitoring global health trends**: DALY, YLL, and YLD can also be used
to monitor global health trends and track changes in the burden of
disease over time. This information can be used to inform global health
policies and allocate resources to address emerging health threats.

Overall, the health metrics of DALY, YLL, and YLD provide valuable
information for public health practitioners, researchers, and policy
makers to help prioritize and allocate resources, evaluate the impact of
diseases and injuries, and inform public health decision-making.

#### General application of DALYs

As an example here is shown how the DALY metric can be used for
prevention:

Suppose we have data on the number of cases of a particular disease, as
well as the average number of years of life lost due to this disease. We
can use this information to calculate the total number of DALYs lost due
to this disease.

```{r}
# Create a data frame with the number of cases and average years of life lost
df <- data.frame(YLL = c(5, 10, 15),
                 YLD = c(1,3,4))

# Calculate the number of DALYs lost
df <- df %>% mutate(DALY = YLL + YLD)

# Sum the total number of DALYs lost
total_dalys <- sum(df$DALY)
total_dalys
```

In this example, the number of cases of the disease and the average
years of life lost for each case are used to calculate the number of
DALYs lost for each case. Finally, the total number of DALYs lost for
the entire population.

This information can be used to inform public health interventions to
prevent the spread of this disease and reduce the number of DALYs lost.
For example, the information could be used to prioritize resources for
disease control and prevention activities, such as health education
campaigns, vaccination programs, and screening and treatment programs.

### HALE

**Healthy Life Expectancy (HALE)** is a metric used to estimate the
number of years a person can expect to live in good health, taking into
account both mortality and morbidity factors. It is often used as a
**measure of overall population health and quality of life**. It takes
into account the impact of both fatal and non-fatal health outcomes on
overall life expectancy. HALE is typically calculated using data on
mortality rates and health-related quality of life measures, such as
disability-adjusted life years (DALYs) or quality-adjusted life years
(QALYs). These measures allow for the estimation of the number of years
lost due to premature death or disability.

It can be used to identify health disparities, assess the effectiveness
of healthcare interventions, and inform public health policies. By
measuring the number of years lived in good health, HALE offers a more
articulated understanding of population health rather than traditional
life expectancy measures.

#### Calculation

$$
HALE= \text{life expectancy}- YLD
$$ {#eq-hale}

The HALE is obtained by subtracting the YLDs from the life expectancy of
a population. But if we consider more specifications such as
comorbidities and other health factors that can verify in real life, the
calculation can be adjusted by assuming different levels of
**prevalence** based on disease sequelae with associated **disability
weights**, and eventually accounting for **comorbidity** level by using
a **Monte Carlo simulation approach**.

We use age-specific mortality and YLDs per capita by location, age, sex,
and year, and define the HALE as:

> the average number of years that a person at a given age can expect to
> live in good health, taking into account mortality and loss of
> functional health.

![HALE - GLOBAL HEALTH METRICS\| VOLUME 392, ISSUE 10159, P1859-1922,
NOVEMBER 10, 2018](images/hale.png){fig-align="center" width="504"}

The process of calculating the HALE for a specific population,
considering factors such as sex, country, and year, involves computing
the **average health** of individuals across different age groups within
the population and integrating information on the **prevalence** of
various health conditions and their associated **disability weights**.
**Comorbidity** is addressed using a **Monte Carlo simulation
approach**, assuming independence of comorbidities within each age
group. Simulations are conducted to model exposure to different health
conditions based on their estimated prevalence in each age group,
resulting in a simulated population reflecting the prevalence of
multi-morbidities[@stanaway2018].

**Positive health** associated with each health condition is defined as
one minus the disability weight $(1-DW)$. The **combined health** for an
individual in the simulated population is determined by multiplying
these positive health values for all relevant health conditions present.

The **average health** values are then computed as one minus the Years
Lived with Disability (YLD) $(1-YLD)$ per person in the population,
which are used to calculate **health-adjusted person years**. The
**Sullivan's method** is employed to incorporate these average health
values into the **life table**. This involves adjusting the values in
the $nLx$ column of the life table by the corresponding average health
values, recalculating the life table using these adjusted values, and
then using an iterative process to estimate health-adjusted person-years
for different age groups.

Finally, HALE is calculated by dividing the adjusted person-years for
each age group by the proportion of a hypothetical birth cohort still
alive at that age.

**Simulate life table data**

Define age intervals (e.g., every 5 years), simulate survival
probabilities for each age interval (between 0 and 1)

```{r}
set.seed(040424)
age_intervals <- seq(0, 100, by = 5)
survival_probabilities <- runif(length(age_intervals), 
                                min = 0.5, max = 1)
```

```{r}
life_table <- data.frame(Age = age_intervals, 
                         Survival_Probability = survival_probabilities,
                         nLx=100000-(age_intervals*survival_probabilities))
life_table%>%head()
```

Calculate adjusted $Tx$ for each age group as the sum of
**health-adjusted person-years** with all age intervals above the
current age interval.

```{r}
# Simulate prevalences
prevalences <- runif(length(age_intervals), 
                     min = 0, max = 0.5)
# Simulate disability weights
disability_weights <- runif(length(age_intervals), 
                            min = 0, max = 1)
# Calculate average health for each age group
average_health <- 1 - (prevalences * disability_weights)

# Adjust Tx for each age group
life_table$adjusted_px <- life_table$Survival_Probability * average_health

life_table%>%head()
```

Let's make a function to calculate the HALE.

```{r}
calculate_HALE <- function(age,px) {
  # Initialize a vector to store adjusted Tx for each age group
  adjusted_Tx <- numeric(length(age))
  # Iterate over each age interval to calculate adjusted Tx
  for (i in 1:length(age)) {
    # Calculate adjusted Tx for the current age group
    adjusted_Tx[i] <- sum(px[i:length(age)])
  }
  HALE <- numeric(length(age))
  # Calculate HALE for each age group
  for (i in 1:length(age)) {
    # Calculate HALE for the current age group
    HALE[i] <- adjusted_Tx[i] / px[i]
  }
HALE
}

```

```{r}
# Calculate HALE for the simulated life table data
HALE1 <- calculate_HALE(age_intervals,
                        life_table$Survival_Probability)
HALE2 <- calculate_HALE(age_intervals,
                        life_table$adjusted_px)
```

```{r}
#| echo: false
#| eval: true
data.frame(age_intervals, HALE1,HALE2)%>% 
  pivot_longer(cols=c(HALE1,HALE2),
               names_to="HALE",
               values_to="Years")%>% 
  ggplot(aes(x=age_intervals))+
  geom_line(aes(y=Years,group=HALE,color=HALE))+
  scale_color_viridis_d(option = "H")+
  labs(title = "Health Adjusted Life Expectancy (HALE) by Age",
       x="Age Group (5Yrs)")
```
