# Methods and Calculations {#sec-02-method_calculation}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   A closer look at the metrics
-   Understand how to use them
-   Think about potentiality for improvements
:::
::::::

\
The objective of this chapter is to provide a first level calculation of
the burden of diseases, focusing on disability and premature mortality,
which are captured by DALYs and HALE. In @sec-01-health_metrics, we
defined the metrics to establish the theoretical framework for
constructing them. In this chapter, we will proceed with a general
calculation of YLLs and YLDs to obtain the DALYs. This step is essential
for understanding the structure of the metric components, which will be
further investigated in @sec-03-components.

Used to measure the burden of disease and quantify the impact of
diseases and injuries on individuals and populations, these metrics can
help prioritise public health interventions and evaluate the
effectiveness of public health programs.

## YLLs Calculation

The **Years of Life Lost (YLLs)** is a metric that measures the number
of years a person would have lived if they had not died prematurely due
to a disease or injury. **YLLs are calculated by subtracting the age at
death from the expected age at death in a population without the disease
or injury**.

First introduced by Mary Dempsey[@dempsey2019] in 1947 to assess the
burden of tuberculosis, Years of Life Lost (YLLs) were designed to
highlight areas needing improvement to reduce health loss and lower the
risk of premature death[@reiner2022]. This metric was first calculated
in late 1940s (@sec-01-health_metrics), and defined as **Standardised
Rate for Mortality in units of Lost Years of Life** by *William
Haenszel* [@haenszel1950]. The investigation into the leading causes of
death led to the development of a measurement that considered the age at
death and compared it with the expected life expectancy. This initially
led to the understanding that early deaths had a greater impact than
those at later ages. It was ascertained that accounting for the number
of years lost for a group of people would help recognize the potential
life lost for a certain cause. Since this first approach, some
adjustments to the standard death rates were made, leading to a new
calculation that considered Standardised death rates applied to
age-specific factors for specific causes of death.

> The impact of premature losses on society extends beyond the grief of
> losing a loved one; it also includes the economic and social cost of
> losing productive members of the workforce. Such premature deaths lead
> to a significant burden on the community, affecting overall economic
> stability and societal well-being.

**Standard Formula**

$$
YLL=N*le
$$ {#eq-yll}

In @eq-yll, $N$ is the number of premature deaths, and $le$ is the
standard live expectancy at the age of death. This calculation takes
into account the number of deaths at each age and multiplies it by the
standard life expectancy remaining at that age, using global life tables
to determine life expectancy. More details about the components of YLL
are provided in the next chapter (@sec-03-components), and a sample of
the construction of a life table with relative estimations of the life
expectancy can be found in @sec-appendixA.

In general, the **Global Burden of Diseases studies (GBD)** use
standardised life tables, to consistently measure the impact of various
diseases. This approach allows for reliable comparisons across different
conditions and populations, providing a comprehensive assessment of
disease burden on a global scale[@devleesschauwer2020]. Additionally,
country-specific life expectancies are valuable for investigating
premature deaths resulting from health-related causes, such as fatal
diseases. For instance, certain types of country-specific life
expectancies are frequently recommended due to their ability to provide
insight into longevity characteristics, which can inform policymakers
and aid in prevention efforts, as exemplified by the Japanese life
expectancy [@tsugane2021]. A G7 cross-country study demonstrated that
Japan has the longest average life expectancy, primarily attributed to
significantly low mortality rates from ischemic heart disease and
cancer, which are the leading causes of death in most countries, as
indicated by the GBD study.

### Example: YLLs due to Stroke

In the following example, we calculate the YLLs due to stroke in the
year 2019 for the Global region. We use the data from the Global Burden
of Disease (GBD) study, which provides estimates of the number of deaths
due to stroke in different regions. The data can be downloaded from the
`{hmsidwR}` package, which contains the necessary datasets for this
calculation. The `deaths2019` dataset comprises
`r dim(hmsidwR::deaths2019)[1]` observations and
`r dim(hmsidwR::deaths2019)[2]` variables, containing the estimated
number of deaths due to 9 causes, including stroke, across 6 regions:
Global, France, Italy, Germany, the United Kingdom, and the United
States. We also use the **Global Health Observatory Life Tables** to
estimate the life expectancy at different ages, which is used to
calculate the YLLs due to stroke.

```{r}
#| eval: false
# install.packages("devtools")
devtools::install_github("Fgazzelloni/hmsidwR")
```

```{r}
#| output: false
library(tidyverse)
library(hmsidwR)
```

```{r}
unique(hmsidwR::deaths2019$cause)
```

Our specific task is to calculate the Years of Life Lost (YLLs)
attributable to stroke in the year 2019 for the Global region. We filter
the location to be "Global" and the cause to be "Stroke". The use of
`str_detect()` is to match the cause of death containing a specific
word, it is very useful when the cause of death is not exactly
containing just one word.

```{r}
deaths_stroke <- hmsidwR::deaths2019 %>%
  arrange(age)%>%
  filter(location == "Global",
         str_detect(cause, "Stroke")) %>%
  select(-location, -cause, -upper, -lower)

deaths_stroke %>% head()
```

Then, we visualise the number of deaths due to stroke by age group, with
a `geom_boxplot()`. The boxplot shows the distribution of the number of
deaths by age group, with the median, quartiles, and outliers. The plot
is divided by age group and shows the variation in the number of deaths
due to stroke. We can observe the difference in the number of deaths
increases for older age groups.

```{r}
#| label: fig-deaths_stroke
#| fig-cap: "Boxplots showing male and female deaths variation due to Stroke by Age Group (2019)"
#| fig-alt: "Deaths due to Stroke by Age Group (2019)"
deaths_stroke %>% 
  filter(!sex == "both") %>%
  ggplot(aes(x = age,y = dx)) +
  geom_boxplot() +
  labs(title = "Male and Female Deaths due to Stroke",
       subtitle = "Year 2019",
       caption = "IHME GBD 2019 Data",
       x = "Age Group", y = "Deaths") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Among the indicators available in the `hmsidwR::gho_lifetables` dataset,
we specifically focus on the indicator denoted as $e_x$, which
represents the expectation of life at age $x$. This dataset is part of
the Global Health Observatory (GHO) data repository, which is maintained
by the World Health Organization (WHO). The dataset also contains life
table indicators, such as the number of person-years lived above age $x$
$(l_x)$, the number of person-years lived between ages $x$ and $x+n$
$(_{n}L_x)$, the age-specific death rate between ages $x$ and $x+n$
$(_{n}M_x)$, the number of people dying between ages $x$ and $x+n$
$(_{n}d_x)$, and the probability of dying between ages $x$ and $x+n$
$(_{n}q_x)$.

```{r}
#| label: tbl-indicators
#| tbl-cap: "Life Table Indicators"
#| echo: false
hmsidwR::gho_lifetables %>%
  count(indicator) %>%
  select(-n) %>%
  rename(Indicator = indicator) %>%
  mutate(Description = c(
    "person-years lived above age x",
    "expectation of life at age x",
    "number of people left alive at age x",
    "person-years lived between ages x and x+n",
    "age-specific death rate between ages x and x+n",
    "number of people dying between ages x and x+n",
    "probability of dying between ages x and x+n")) %>%
  knitr::kable()
```

The expectation of life at age x $(e_x)$ refers to the life expectancy
at age $x$. We use the estimated value of the expected life for 5-year
age groups, such as \<1, ... , 05-09, 10-14, etc. for both females and
males.

The standard life expectancy represents the maximum number of years a
person can live at birth. For example, if we consider the **Japanese
life expectancy at birth**, which is measured in total years, according
to the estimation released in 2021 by the United Nations Population
Division research, it is approximately 84 years.

To calculate the YLLs due to stroke, we need to use the life expectancy
at different ages. We filter the `gho_lifetables` dataset to year 2019,
which is the most updated year available for the life expectancy data,
and select the indicator `ex` to get the life expectancy at different
ages, renaming it as `le`.

```{r}
ex2019 <- hmsidwR::gho_lifetables %>%
  filter( year == 2019, indicator == "ex") %>%
  select(-indicator, -year) %>%
  rename(le = value)

ex2019 %>% head()
```

Then we merge the `deaths_stroke` with the `ex2019` data to calculate
the YLLs due to stroke in the Global region with the `full_join()`
function, and group the data by age and sex before to create one more
vector named `YLL`, which is the product of the number of deaths and the
life expectancy at that age. These YLLs values are expressed in
millions, and are not necessarily the real values, but estimated values.
Their values are strongly dependent on the life expectancy and the
number of deaths due to stroke, also other adjustments can be made to
the life expectancy values to get more accurate results. In the past,
the calculation of YLLs included the use of a discount rate, which is no
longer used in the most recent calculations.

```{r}
YLL_global_stroke <- deaths_stroke %>%
  full_join(ex2019) %>%
  group_by(age, sex) %>%
  mutate(YLL = dx * le) %>%
  ungroup()

YLL_global_stroke %>% head()
```

```{r}
#| label: fig-YLL_global_stroke
#| fig-cap: "YLLs due to Stroke in the Global region"
#| fig-alt: "YLLs due to Stroke in the Global region"
#| echo: false
#| eval: true
dat <- YLL_global_stroke %>%
  filter(!sex == "both")

legend_grob <- cowplot::get_legend(
  ggplot(data = dat, aes(x = age)) +
    geom_point(aes(y = YLL, color = sex),
               size = 4) +
    scale_color_manual(values = c("pink", "darkgreen")) +
    theme(legend.direction = "horizontal",
          legend.background = element_blank(),
          legend.position = "right",
          legend.title = element_blank(),
          legend.key = element_blank(),
          legend.text = element_text(size = 8)))

dat %>%
  ggplot(aes(x = age)) +
  geom_line(aes(y = YLL)) +
  geom_point(data = dat %>% filter(sex == "female"),
             aes(y = YLL), color = "pink", size = 2) +
  geom_point(data = dat %>% filter(sex == "male"),
             aes(y = YLL), color = "darkgreen", size = 2) +
  geom_smooth(aes(y = YLL, group = 1),
              se = F, color = "#29306c") +
  scale_y_continuous(labels = function(x) 
                              scales::number(x, 
                                             scale = 1e-6, 
                                             accuracy = 0.01, 
                                             suffix = "M"),
                   name = "Value (Millions)") +
  annotation_custom(legend_grob, ymax = max(dat$YLL) * 2 + 10^6) +
  labs(title = "Global Region : Years of Life Lost (YLLs) due to stroke",
       subtitle = "Year 2019 - Millions",
       caption = "DataSource: WHO life expectancy & IHME results Stroke data",
       x = "Age Group", y = "YLLs") +
  theme(axis.text.x = element_text(size = 7,
                                   angle = 45,
                                   vjust = 0.5),
        plot.title = element_text(size = 11),
        plot.caption = element_text(size = 8))
```

In summary, stroke is a leading cause of death and disability worldwide,
and it is responsible for a significant number of premature deaths in
the Global region. The YLLs due to stroke are particularly high in older
age groups, where the number of deaths is higher, and the life
expectancy is lower. This indicates that stroke has a significant impact
on the overall burden of disease in the Global region, particularly
among older populations. The YLLs due to stroke are an important measure
of the impact of this disease on premature mortality and can help inform
public health interventions and policies to reduce the burden of stroke
in the population.

## YLDs Calculation

**YLDs (Years Lived with Disability)** measure the number of years a
person lives with a disability due to a disease or injury. It is
calculated by multiplying the **prevalence of a condition** by the
**disability weight**, which reflects the **severity** of the
disability.

The key factor of the disability weights (DW) is linked to the severity
(mean of the range of health loss suffered to disease) of a non-fatal
health condition due to disease or injury. DW ranges between 0
(equivalent to full health) and 1 (equivalent to death). The estimation
of the disability weights is challenging and has been continuously
changed by modifying and adapting methodologies in various studies
[@liu2023]. The challenge is assigning disability weights to diseases
with different levels of prevalence and severity, such as cases with
high prevalence and low severity. People’s experiences of the same
condition can vary significantly, making it difficult to establish a
Standardised weight that accurately reflects the global impact on
quality of life. Furthermore, the results need to be reported to a
year-based value.

Moreover, the calculation of YLDs has been updated over the years,
shifting from an incidence to a prevalence-based approach. More in-depth
analyses are in @sec-03-components; for now, we will show how both types
of calculations differ, but then we will focus on using the most updated
approach based on values of prevalence estimated by the GBD2021,
summarised in the `hmsidwR::incprev_stroke` dataset. This dataset
specifically contains the estimated values for the incidence and
prevalence of stroke in the Global region for the years 2019 and 2021.

#### Incidence-based Calculation

**Standard Formula**

$$
YLD_i=I*DW*L
$$ {#eq-yldi}

In @eq-yldi, $I$ is the incidence of the condition, $DW$ is the
disability weight, and $L$ is the average duration of the condition. The
incidence takes into account the number of new cases of a disease or
health condition that occur in a population over a specific period. The
disability weight reflects the severity of the health condition, and the
average duration of the condition is the average length of time a person
lives with the condition.

#### Prevalence-based Calculation

**Standard Formula**

$$
YLD_p=p*DW
$$ {#eq-yld}

In @eq-yld, $p$ is the prevalence, $DW$ are the disability weights.
While an incidence-based approach focuses on the number of new cases of
a health condition, the prevalence-based approach considers the total
number of cases in the population. The disability weight reflects the
severity of the health condition and it is applied to the prevalence to
calculate the YLDs, and the duration of the condition is not considered
in the prevalence-based calculation.

### Example: YLDs due to Stroke

Since the release of GBD 2010, the WHO has decided to switch to a
prevalence-based approach for the calculation of YLDs. The major impact
of this shift is to distribute the weights of the YLDs more evenly
across all age groups, rather than concentrating them at the age of
incidence.

In the following example we use the `disability weights` and the
`severity` levels extracted from a dataset in the GBD study. The
`disweights` dataset is stored in the `{hmsidwR}` package and is made of
`r dim(hmsidwR::disweights)[1]` observations and
`r dim(hmsidwR::disweights)[2]` variables. It contains the estimated
values for the disability weights, which are measured on a scale from 0
to 1, where 0 equals a state of full health and 1 equals death.

```{r}
hmsidwR::disweights %>%
  filter(year == 2019) %>%
  group_by(cause1, severity) %>%
  reframe(dw = mean(dw)) %>%
  filter(cause1 == "Stroke") 
```

The level of `severity` is assigned based on the **level of disability**
by the **National Institutes of Health Stroke Scale**, or
[**NIH**](https://en.wikipedia.org/wiki/National_Institutes_of_Health "National Institutes of Health")
**Stroke Scale** (**NIHSS**). The classification is used by healthcare
providers to objectively quantify the impairment caused by a
[stroke](https://en.wikipedia.org/wiki/Stroke "Stroke"). Here is assumed
a sample population affected by a stroke, categorised as **mild**,
**moderate**, or **severe** with assigned proportions.

|       |                 |                |            |
|-------|-----------------|----------------|------------|
| Score | Stroke severity | Severity Level | Severity % |
| 0-4   | Minor stroke    | Mild           | 50.3%      |
| 5–20  | Moderate stroke | Moderate       | 25.3%      |
| 21-42 | Severe stroke   | Severe         | 24.4%      |

: **Level of disability** by the **National Institutes of Health Stroke
Scale**, or
[**NIH**](https://en.wikipedia.org/wiki/National_Institutes_of_Health "National Institutes of Health")
**Stroke Scale** (**NIHSS**)

These levels are general for all ages; the values might vary for other
specifications of the level of disability [@wyper2020].

```{r}
dwsev2019 <- hmsidwR::disweights %>%
  select(cause1, severity, dw) %>%
  drop_na() %>%
  mutate(severity_n = case_when(
    severity == "mild" ~ 0.503,
    severity == "moderate" ~ 0.253,
    severity == "severe" ~ 0.244))

dwsev2019 %>% head()
```

The values for disability weights and severity are considered for all
ages; in general, they differ by age, and for different types of stroke.

```{r}
dw_stroke <- dwsev2019 %>%
  filter(cause1 == "Stroke") %>%
  group_by(severity, severity_n) %>%
  reframe(avg_dw = mean(dw))

dw_stroke 
```

Then, calculate the part of the population affected by a specific level
of severity considering the prevalence (and/or the incidence) multiplied
by the severity levels.

For instance, here we use the `incprev_stroke` and the `dw_stroke`
datasets to calculate the YLDs due to stroke, and have a look at how
incidence and prevalence differ from each other.

::: {.callout .callout-info}
Health metrics values can be expressed as numbers, percentages or rates:

-   If the prevalence of stroke is 1,541,506.96 for male with age
    between 34-39 and it is expressed in numbers, it means that there
    are approximately 1,541,507 individuals in the age group 35-39 who
    had a stroke.
-   If the prevalence of stroke is 0.58 for male with age between 34-39
    and it is expressed in percent value (%), it means 0.58% of the
    population within that age range had a stroke.
-   If the prevalence of stroke is expressed as rate, it is the number
    of cases of stroke per 100,000 individuals in the population.

Percentages and numbers are related by the size of the population in
that age group. To convert between these forms, you need to know the
total population in the age group.
:::

Let's consider the numbers of incidence and prevalence, and assign them
as two separate vectors in a new dataset named `inc_prev_stroke_5y`. We
use the `pivot_wider()` function to spread the data into a wider format,
with the `measure` column as the key column and the `val` column as the
value column.

```{r}
inc_prev_stroke_5y <- hmsidwR::incprev_stroke %>%
  filter(year == 2019) %>%
  select(measure, sex, age, val) %>%
  pivot_wider(names_from = "measure", values_from = "val")
```

Let's check the values for the age group 35-39.

```{r}
inc_prev_stroke_5y %>%
  arrange(sex) %>%
  filter(age == "35-39")
```

Then, multiply these values for the severity levels for stroke in the
Global region. In this way we obtain three values, one for each severity
level.

For calculating the prevalence-based YLDs, we use the `Prevalence`
values, the `severity` and the `average weights`, while for
incidence-based YLDs, we also need to consider the `average duration` of
the condition.

In particular, for stroke, the average duration of the condition can
vary based on:

-   duration for acute stroke: up to 28 days
-   duration for chronic stroke: beyond 28 days, often modelled for
    long-term consequences, sometimes up to the lifetime of the patient
    depending on the model used.

In this example, we consider the average duration of the condition for
28 days. It does need to be converted to be year-based:

$$ \hat{L}_{stroke}= \frac{28}{365} $$ {#eq-avg_dur}

These values strongly depend on the disability weights and the severity
values assigned to the condition.

```{r}
#| echo: false
options(scipen = 999)
```

```{r}
YLD_by_severity <- merge(inc_prev_stroke_5y, dw_stroke) %>%
  group_by(sex, age, avg_dw) %>%
  reframe(prev_sev = Prevalence*severity_n,
          inc_sev = Incidence*severity_n,
          yld_p = prev_sev* avg_dw,
          yld_i = inc_sev* avg_dw * 28/365)
```

Let's check the values for the age group 35-39.

```{r}
YLD_by_severity %>%
  filter(age == "35-39")
```

We can reverse the calculation to check the original values.

```{r}
YLD_by_severity %>%
  filter(age == "35-39") %>%
  group_by(sex, age) %>%
  reframe(prev=sum(prev_sev),
          inc=sum(inc_sev))
```

And, finally calculate the total value for YLDs as:

$$ \text{Total YLDs} = \text{YLD}_{\text{mild}} + \text{YLD}_{\text{moderate}} + \text{YLD}_{\text{severe}}$$ {#eq-tot-ylds}

```{r}
YLD_global_stroke <- YLD_by_severity %>%
  group_by(sex, age) %>%
  reframe(YLD_p = sum(yld_p),
          YLD_i = sum(yld_i))
```

For instance, for a male with age 34-39, the YLDs due to stroke
calculated on the population of that age group who have the condition
for three different severity levels, are shown below.

```{r}
YLD_global_stroke %>% 
  filter(sex == "male",
         age == "35-39")
```

This value is the sum of the severity levels of prevalence-based YLDs
due to stroke; it strongly depends on the disability weights and the
severity values assigned to the condition. Here we have considered
estimated values, which do not necessarily correspond to the real
values.

We note the difference in the YLDs if incidence or prevalence is used in
the calculation. The magnitude of the YLDs is similar, but the values
are distributed differently across the age groups.

```{r}
#| echo: false
dat <- YLD_global_stroke %>%
  filter(!sex == "both")
```

```{r}
#| layout-ncol: 2
#| label: fig-YLD_global_stroke
#| fig-cap: "YLDs due to stroke in the Global region"
#| fig-subcap:
#|  - "YLDs calculated with prevalence"
#|  - "YLDs calculated with incidence"
#| fig-alt: "YLDs due to stroke in the Global region"
#| echo: false

dat %>%
  ggplot(aes(x = age, y = YLD_p)) +
  geom_line(aes(y = YLD_p)) +
  geom_point(data = dat %>% filter(sex == "female"),
             color = "pink", size = 3) +
  geom_point(data = dat %>% filter(sex == "male"),
             color = "darkgreen", size = 3) +
  geom_smooth(aes(group = 1),
              se = F, color = "#29306c") +
  scale_y_continuous(labels = function(x) 
                              scales::number(x, 
                                             scale = 1e-6, 
                                             accuracy = 0.01, 
                                             suffix = "M"),
                   name = "Value (Millions)") +
  annotation_custom(legend_grob, ymax = max(dat$YLD_p)*2 + 10^5) +
  labs(title = "Global region: Prevalence based YLDs due to stroke",
       subtitle = "Year 2019 - Millions",
       caption = "DataSource: IHME results Stroke data | YLDs Values X 1000",
       x = "Age Group", y = "YLDs") +
  theme(axis.text.x = element_text(size = 7,
                                   angle = 45,
                                   vjust = 0.5),
    plot.title = element_text(size = 11),
    plot.caption = element_text(size = 8))


dat %>%
  ggplot(aes(x = age, y = YLD_i)) +
  geom_line(aes(y = YLD_i)) +
  geom_point(data = dat %>% filter(sex == "female"),
             color = "pink", size = 3) +
  geom_point(data = dat %>% filter(sex == "male"),
             color = "darkgreen", size = 3) +
  geom_smooth(aes(group = 1),
              se = F, color = "#29306c") +
  scale_y_continuous(labels = function(x) 
                              scales::number(x, 
                                             scale = 1e-6, 
                                             accuracy = 0.01, 
                                             suffix = "M"),
                   name = "Value (Millions)") +
  annotation_custom(legend_grob, ymax = max(dat$YLD_i)*2 + 10^3) +
  labs(
    title = "Global region: Incidence based YLDs due to stroke",
    subtitle = "Year 2019. - Millions",
    caption = "DataSource: IHME results Stroke data",
    x = "Age Group", y = "YLDs") +
  theme(axis.text.x = element_text(size = 7,
                                   angle = 45,
                                   vjust = 0.5),
    plot.title = element_text(size = 11),
    plot.caption = element_text(size = 8))
```

In summary, the YLDs due to stroke are an important measure of the
impact of this condition on the quality of life of individuals in the
Global region. In this example we saw how the level of disability and
the severity of the condition can affect the YLDs, and how the
prevalence-based and incidence-based calculations can provide different
results. The YLDs due to stroke are particularly high in older age
groups, where the prevalence of the condition is higher, and the
severity of the disability is more pronounced. This indicates that
stroke has a significant impact on the overall burden of disease in the
Global region, particularly among older populations.

## DALYs Calculation

As a measure of the overall disease burden, **Disability Adjusted Life
Years (DALYs)** are used to quantify the sum of years of potential life
lost due to premature death (YLLs) and years lived with disability
(YLDs). The number of DALYs indicates the number of years of life lost
due to premature deaths, disease, or injury. This metric takes into
account both the **quantity** and **quality** of life.

DALYs are a generalisation of the well-known **Potential Years of Life
Lost measure (PYLLs)**, which includes the loss of good health. We do
not consider PYLLs in this book, but more information can be found in
the references [@globalh].

**Standard Formula**

$$
DALYs=YLLs+YLDs
$$ {#eq-daly}

One DALY is one lost year of healthy life. This measure is used to
assess how diseases and injuries impact populations, providing a
comprehensive picture of the overall burden of disease by combining YLLs
and YLDs in different groups.

### Example: DALYs due to Stroke

The sum of YLLs and YLDs releases the overall value of DALYs due to
stroke in the Global region.

```{r}
DALY_global_stroke <- YLL_global_stroke %>%
  select(age, sex, YLL) %>%
  full_join(YLD_global_stroke %>% 
            select(age, sex, YLD=YLD_p), 
            by = c("age","sex")) %>%
  distinct() %>%
  mutate(DALY = YLL + YLD) 
  
DALY_global_stroke %>%
  head()
```

Let's have a closer look at the 35-39 age groups.

```{r}
DALY_global_stroke %>%
  filter(age == "35-39")
```

```{r}
#| echo: false
dat <- DALY_global_stroke %>%
  select(sex, age, YLL, YLD, DALY) %>%
  filter(sex == "both") %>%
  arrange(age) %>%
  mutate_if(is.numeric, scale)
```

```{r}
#| echo: false 
#| eval: true 
#| label: fig-DALY_global_stroke 
#| fig-cap: "DALYs due to stroke in the Global region. The blue line represents the DALYs, YLLs and YLDs are pink for females and green for males." 
#| fig-alt: "DALYs due to stroke in the Global region" 
 
dat %>%
  pivot_longer(cols=c(-sex,-age)) %>%
  ggplot(aes(x = age, y = value, group = name, linetype = name)) +
  geom_line() +
  labs(
  title = "Global region: Disability Adjusted Life Years (DALYs) due to stroke",
  subtitle = "Year 2019 - Standardised Values",
  caption = "DataSource: IHME results Stroke data",
  x = "Age Group", y = "DALYs", linetype = "Metric") +
  theme(axis.text.x = element_text(size = 7, angle = 45, vjust = 0.5),
        plot.title = element_text(size = 11),
        plot.caption = element_text(size = 8),
        legend.position = "bottom")  
```

In summary, DALYs are a valuable metric for assessing the overall
burden. The DALYs due to stroke provide a comprehensive measure of the
impact of this condition on the quality of life of individuals,
combining the years of life lost due to premature death and the years
lived with disability.

DALYs can vary significantly between conditions, reflecting both the
severity and impact of each disease on quality and length of life. For
instance, stroke and diabetes, both prevalent globally, differ widely in
their DALY contributions. Strokes tend to have high DALYs due to their
immediate impact on mortality (Years of Life Lost or YLL) and long-term
disability (Years Lived with Disability or YLD). On the other hand,
diabetes, a chronic condition, has a lower YLL but a higher YLD,
reflecting the long-term impact on quality of life.

## How DALYs are Used

DALYs, YLLs, and YLDs can be used in several ways to help identify
health priorities, evaluate the impact of diseases and injuries, and
inform public health decision-making. Some common uses of these metrics
include:

-   **Prioritising public health interventions**: By calculating the
    overall burden of disease in a population, public health
    practitioners can prioritise which diseases and injuries to address
    first. This helps allocate resources and target interventions to the
    areas of greatest need.
-   **Evaluating the impact of diseases and injuries**: These metrics
    can be used to measure the impact of diseases and injuries on
    individuals and populations and to track changes over time. This
    information can help inform public health decision-making and
    allocate resources more effectively.
-   **Comparing the burden of disease across populations**: DALY, YLL,
    and YLD can be used to compare the burden of disease across
    populations and between different regions. This information can help
    identify disparities in health outcomes and inform targeted public
    health interventions.
-   **Evaluating the effectiveness of public health programs**: These
    metrics can be used to evaluate the impact of public health programs
    and to assess the effectiveness of public health interventions. This
    information can help public health practitioners identify areas for
    improvement and make necessary changes to ensure that programs are
    achieving their goals.
-   **Monitoring global health trends**: DALY, YLL, and YLD can also be
    used to monitor global health trends and track changes in the burden
    of disease over time. This information can be used to inform global
    health policies and allocate resources to address emerging health
    threats.

To mention the case of Rwanda, the GBD 2021 results released by the IHME
showed that the country had made significant progress in reducing the
burden of disease over the past decade[@ferrari2024]. By using DALYs,
YLLs, and YLDs, public health leaders were able to identify key areas
for improvement and target resources to address the most pressing health
challenges. For example, the government used the DALYs metrics to
address the burden of non-communicable diseases (NCDs), public health
leaders observed that conditions like cardiovascular disease and
diabetes were contributing significantly to the country's overall
disease burden, surpassing infectious diseases in certain demographics.
This insight led to targeted policy shifts, including prioritising
funding for NCD prevention and treatment programs, investing in training
for healthcare providers on chronic disease management, and increasing
public awareness of lifestyle risk factors like smoking and poor diet.

Overall, the health metrics of DALY, YLL, and YLD provide valuable
information for public health practitioners, researchers, and policy
makers to help prioritise and allocate resources, evaluate the impact of
diseases and injuries, and inform public health decision-making.

### General application of DALYs

As an example here is shown how the DALY metric can be used for
prevention. Suppose we have data on the number of cases of a particular
disease, as well as the average number of years of life lost due to this
disease. We can use this information to calculate the total number of
DALYs lost due to this disease.

```{r}
# Create a data frame with the number of cases and average years of life lost
df <- data.frame(
  YLL = c(5, 10, 15),
  YLD = c(1, 3, 4))

# Calculate the number of DALYs lost
df <- df %>% mutate(DALY = YLL + YLD)

# Sum the total number of DALYs lost
total_dalys <- sum(df$DALY)

total_dalys
```

In this example, the number of cases of the disease and the average
years of life lost for each case are used to calculate the number of
DALYs lost for each case. Finally, the total number of DALYs lost for
the entire population.

This information can be used to inform public health interventions to
prevent the spread of this disease and reduce the number of DALYs lost.
For example, the information could be used to prioritise resources for
disease control and prevention activities, such as health education
campaigns, vaccination programs, and screening and treatment programs.

## HALE Calculation

Healthy Life Expectancy (HALE) is a metric used to estimate the number
of years a person can expect to live in good health, taking into account
both mortality and morbidity factors. It is often used as a measure of
overall population health and quality of life. It takes into account the
impact of both fatal and non-fatal health outcomes on overall life
expectancy. HALE is typically calculated using data on mortality rates
and health-related quality of life measures, such as disability-adjusted
life years (DALYs) or quality-adjusted life years (QALYs). These
measures allow for the estimation of the number of years lost due to
premature death or disability.

It can be used to identify health disparities, assess the effectiveness
of healthcare interventions, and inform public health policies. By
measuring the number of years lived in good health, HALE offers a more
nuanced understanding of population health rather than traditional life
expectancy measures.

**Standard Formula**

$$
HALE= \text{life expectancy}- YLD
$$ {#eq-hale}

HALE\index{HALE} is obtained by subtracting the YLDs from the life
expectancy of a population. However, if we consider more specifications
such as comorbidities and other health factors that can occur in real
life, the calculation can be adjusted by assuming different levels of
prevalence based on disease sequelae with associated disability weights,
and eventually accounting for comorbidity levels by using a Monte Carlo
simulation approach.

We use age-specific mortality and YLDs per capita by location, age, sex,
and year, and define the HALE as:

> the average number of years that a person at a given age can expect to live in good health, taking into account mortality and loss of functional health.

![HALE - GLOBAL HEALTH METRICS VOLUME 392, ISSUE 10159, P1859-1922,
NOVEMBER 10,
2018](images/02_hale.png){#fig-flow-hale
fig-align="center" width="504" fig-alt="HALE - GLOBAL HEALTH METRICS"}

The process of calculating HALE for a specific population, considering
factors such as sex, country, and year, involves computing the average
health of individuals across different age groups within the population
and integrating information on the prevalence of various health
conditions and their associated disability weights. Comorbidity is
addressed using a Monte Carlo simulation approach, assuming independence
of comorbidities within each age group. Simulations are conducted to
model exposure to different health conditions based on their estimated
prevalence in each age group, resulting in a simulated population
reflecting the prevalence of multi-morbidities [@stanaway2018].

**Positive health** associated with each health condition is defined as
one minus the disability weight $(1 - DW)$. The combined health for an
individual in the simulated population is determined by multiplying
these positive health values for all relevant health conditions present.

The **average health** values are then computed as one minus the Years
Lived with Disability (YLD) $(1 - YLD)$ per person in the population,
which are used to calculate health-adjusted person years. The **Sullivan
method** is employed to incorporate these average health values into the
life table. This involves adjusting the values in the $nLx$ column of
the life table by the corresponding average health values, recalculating
the life table using these adjusted values, and then using an iterative
process to estimate health-adjusted person-years for different age
groups.

Finally, HALE is calculated by dividing the adjusted person-years for
each age group by the proportion of a hypothetical birth cohort still
alive at that age.

#### Simulating Life Table Data

To understand the process of calculating HALE, we can simulate life
table data. We’ll define age intervals (e.g., every 5 years) and
simulate survival probabilities for each age interval (between 0 and 1).

```{r}
set.seed(040424)
age_intervals <- seq(0, 100, by = 5)
survival_probabilities <- runif(length(age_intervals),
                                min = 0.5, max = 1)
```

```{r}
life_table <- data.frame(
  Age = age_intervals,
  Survival_Probability = survival_probabilities,
  nLx = 100000 - (age_intervals * survival_probabilities))

life_table %>% head()
```

We then calculate adjusted $T_x$ for each age group as the sum of
health-adjusted person-years for all age intervals above the current age
interval.

```{r}
# Simulate prevalences
prevalences <- runif(length(age_intervals),
                     min = 0, max = 0.5)
# Simulate disability weights
disability_weights <- runif(length(age_intervals),
                            min = 0, max = 1)
# Calculate average health for each age group
average_health <- 1 - (prevalences * disability_weights)

# Adjust Tx for each age group
life_table$adjusted_px <- life_table$Survival_Probability * average_health

life_table %>% head()
```

Let's make a function to calculate the **HALE**:

```{r}
calculate_HALE <- function(age, px) {
  # Initialise a vector to store adjusted Tx for each age group
  adjusted_Tx <- numeric(length(age))
  # Iterate over each age interval to calculate adjusted Tx
  for (i in 1:length(age)) {
    # Calculate adjusted Tx for the current age group
    adjusted_Tx[i] <- sum(px[i:length(age)])
    }
  HALE <- numeric(length(age))
  # Calculate HALE for each age group
  for (i in 1:length(age)) {
    # Calculate HALE for the current age group
    HALE[i] <- adjusted_Tx[i] / px[i]
  }
  HALE
}
```

Now, we calculate HALE for the simulated life table data.

```{r}
# Calculate HALE for the simulated life table data
HALE1 <- calculate_HALE(age_intervals, life_table$Survival_Probability)
HALE2 <- calculate_HALE(age_intervals, life_table$adjusted_px)
```

To visualise the HALE for both the standard and adjusted life
table data we can plot the results with a line graph. 

```{r}
#| echo: false
#| eval: true
#| label: fig-HALE
#| fig-cap: "Health Adjusted Life Expectancy (HALE) by Age. The blue line represents the HALE for the simulated life table data, the dotted line represents the HALE for the adjusted life table data."
#| fig-alt: "Health Adjusted Life Expectancy (HALE) by Age"
#| fig-width: 6
#| fig-height: 4

data.frame(age_intervals, HALE1, HALE2) %>%
  pivot_longer(cols = c(HALE1, HALE2),
               names_to = "HALE",
               values_to = "Years") %>%
  ggplot(aes(x = age_intervals)) +
  geom_line(aes(y = Years, group = HALE, 
                color = HALE, linetype = HALE)) +
  scale_color_viridis_d(option = "H") +
  labs(title = "Health Adjusted Life Expectancy (HALE) by Age",
       x = "Age Group (5Yrs)")
```

The difference between the two lines represents the impact of adjusting a life table for health conditions on the calculation of HALE. The adjusted life table data considers the prevalence of health conditions and their associated disability weights, providing a more accurate estimate of the number of years a person can expect to live in good health.

In general, HALE is a valuable metric for assessing the overall health of a population, by incorporating information on the prevalence of health conditions and their associated disability weights, HALE offers a more specific understanding of population health.

## Summary

In this chapter, we have detailed the methods and calculations for
evaluating the burden of disease using key health metrics: YLLs, YLDs,
and DALYs. By understanding how to compute these metrics, we gain
insights into the impact of diseases and injuries on populations,
guiding public health decisions and interventions. We also introduced
HALE, a metric that combines mortality and morbidity data to provide a
comprehensive measure of population health. These metrics are invaluable
tools for public health practitioners, policymakers, and researchers in
their quest to improve global health outcomes.

As we confront new global health challenges like pandemics, ageing
populations, and climate-related health risks, traditional metrics like
DALYs and QALYs are evolving to meet new approaches to capture the
complexity of health outcomes. The inclusion of well-being, resilience,
and mental health is becoming increasingly essential to understanding
the full impact of health conditions on individuals and populations.

In the next chapter (@sec-03-components), we will delve deeper into the
components and variations of these metrics, further enhancing our
understanding of how to utilise them effectively in public health
assessments.
