# Methods and Calculations {#sec-02-method_calculation}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   A closer look at the metrics
-   Understand how to use them
-   Think about potentiality for improvements
:::
:::

The objective of this chapter is to provide a first level calculation of
the burden of diseases, focusing on disability and premature mortality,
which are captured by DALYs and HALE. In @sec-01-health_metrics, we
defined the metrics to establish the theoretical framework for
constructing them. In this chapter, we will proceed with a general
calculation of YLLs and YLDs to obtain the DALYs. This step is essential
for understanding the structure of the metric components, which will be
further investigated in @sec-03-components.

Used to measure the burden of disease and quantify the impact of
diseases and injuries on individuals and populations, these metrics can
help prioritize public health interventions and evaluate the
effectiveness of public health programs.

## YLLs Calculation

The Years of Life Lost (YLLs) is a metric that measures the number of
years a person would have lived if they had not died prematurely due to
a disease or injury. **YLLs are calculated by subtracting the age at
death from the expected age at death in a population without the disease
or injury**.

First mentioned in 1947 by Mary Dempsey[@dempsey2019] to measure the
burden of tuberculosis,YLLs are built to identify areas where
improvement is needed to reduce the loss in health status and reduce the
probability of death [@reiner2022]. This metric was first calculated in
late 1940s (@sec-01-health_metrics), and defined as **Standardized Rate
for Mortality in units of Lost Years of Life** by *William Haenszel*
[@haenszel1950]. The investigation into the leading causes of death led
to the development of a measurement, that considered the age at death
and compared it with the expected life expectancy. This initially led to
the understanding that early deaths had a greater impact than those at
later ages. It was ascertained that accounting for the number of years
lost for a group of people would help recognize the potential life lost
for a certain cause. Since this first approach, some adjustments to the
standard death rates were made, leading to a new calculation that
considered standardized death rates applied to age-specific factors for
specific causes of death.

> The impact on the society due to premature losses involves, not only
> the mere sadness for having lost a beloved one but the lost of
> working-people dying prematurely. It is cost for the entire society.

**Standard Formula**

$$
YLL=N*le
$$ {#eq-yll}

In @eq-yll, $N$ is the number of premature deaths, and $le$ is the
[standard live expectancy at the age of death]{.underline}. This
calculation takes into account the number of deaths at each age and
multiplies it by the standard life expectancy remaining at that age,
using global life tables to determine life expectancy. More details
about the components of YLL are provided in the next chapter
(@sec-03-components), and a sample of the construction of a life table
with relative estimations of the life expectancy can be found in
@sec-appendixA.

In general, the **Global Burden of Diseases studies (GBD)** utilize
standard life tables, enabling the measurement of the impact of
different diseases at a consistent level. This standardization is
crucial for comparative studies, particularly cross-country comparisons
[@devleesschauwer2020]. Additionally, country-specific life expectancies
are valuable for investigating premature deaths resulting from
health-related causes, such as fatal diseases. For instance, certain
types of country-specific life expectancies are frequently recommended
due to their ability to provide insight into longevity characteristics,
which can inform policymakers and aid in prevention efforts, as
exemplified by the Japanese life expectancy [@tsugane2021]. A G7
cross-country study demonstrated that Japan has the longest average life
expectancy, primarily attributed to significantly low mortality rates
from ischemic heart disease and cancer, which are the leading causes of
death in most countries, as indicated by the GBD study.

### Example: YLLs due to Stroke

In the following example, we calculate the YLLs due to stroke in the
year 2019 for the Global region. We use the data from the Global Burden
of Disease (GBD) study, which provides estimates of the number of deaths
due to stroke in different regions. The data can be downloaded from the
`{hmsidwR}` package, which contains the necessary datasets for this
calculation. The `deaths2019` dataset comprises
`r dim(hmsidwR::deaths2019)[1]` observations and
`r dim(hmsidwR::deaths2019)[2]` variables, containing the estimated
number of deaths due to 9 causes, included stroke, across 6 regions:
Global, France, Italy, Germany, the United Kingdom, and the United
States. We also use the **Global Health Observatory Life Tables** to
estimate the life expectancy at different ages, which is used to
calculate the YLLs due to stroke.

```{r}
#| eval: false
# install.packages("devtools")
devtools::install_github("Fgazzelloni/hmsidwR")
```

```{r}
#| warning: false
#| message: false
#| output: false

library(tidyverse)
library(hmsidwR)
```

```{r}
unique(hmsidwR::deaths2019$cause)
```

Our specific task is to calculate the Years of Life Lost (YLLs)
attributable to stroke in the year 2019 for the Global region.

```{r}
deaths_stroke <- hmsidwR::deaths2019 %>%
  arrange(age)%>%
  filter(location == "Global",
         str_detect(cause, "Stroke")) %>%
  select(-location, -cause, -upper, -lower)

deaths_stroke %>% head()
```

```{r}
#| label: fig-deaths_stroke
#| fig-cap: "Boxplots showing male and female deaths variation due to Stroke by Age Group (2019)"
#| fig-alt: "Deaths due to Stroke by Age Group (2019)"

deaths_stroke %>% 
  filter(!sex == "both") %>%
  ggplot(aes(x = age,y = dx)) +
  geom_boxplot() +
  labs(title = "Male and Female Deaths due to Stroke",
       subtitle = "Year 2019",
       caption = "IHME GBD 2019 Data",
       x = "Age Group", y = "Deaths") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Among the indicators available in the `hmsidwR::gho_lifetables` dataset,
we specifically focus on the indicator denoted as $ex$, which represents
the expectation of life at age $x$.

```{r}
#| label: tbl-indicators
#| tbl-cap: "Life Table Indicators"
#| echo: false

hmsidwR::gho_lifetables %>%
  count(indicator) %>%
  select(-n) %>%
  mutate(description = c(
    "person-years lived above age x",
    "expectation of life at age x",
    "number of people left alive at age x",
    "person-years lived between ages x and x+n",
    "age-specific death rate between ages x and x+n",
    "number of people dying between ages x and x+n",
    "probability of dying between ages x and x+n")) %>%
  knitr::kable()
```

The expectation of life at age x $(ex)$ refers to the life expectancy at
age $x$. We use the estimated value of the expected life for 5-year age
groups, such as \<1,..., 05-09, 10-14, etc. for both females and males
in place of the standard life expectancy $(le)$.

The standard life expectancy represents the maximum number of years a
person can live. For example, if we consider the **Japanese life
expectancy at birth**, which is measured in total years, according to
the estimation released in 2021 by the United Nations Population
Division research, it is approximately 84 years. And so in our case it
would not be suitable for representing different age groups.

```{r}
ex2019 <- hmsidwR::gho_lifetables %>%
  filter( year == 2019, indicator == "ex") %>%
  select(-indicator, -year) %>%
  rename(le = value)

ex2019 %>% head()
```

Then we merge the `deaths_stroke` with the `ex2019` data to calculate
the YLLs due to stroke in the Global region.

```{r}
YLL_global_stroke <- deaths_stroke %>%
  full_join(ex2019) %>%
  group_by(age, sex) %>%
  mutate(YLL = dx * le) %>%
  ungroup()

YLL_global_stroke %>% head()
```

```{r}
#| warning: false
#| message: false
#| label: fig-YLL_global_stroke
#| fig-cap: "YLLs due to stroke in the Global region"
#| fig-alt: "YLLs due to stroke in the Global region"
#| echo: false
#| eval: true

dat <- YLL_global_stroke %>%
  filter(!sex == "both")

legend_grob <- cowplot::get_legend(
  ggplot(data = dat, aes(x = age)) +
    geom_point(aes(y = YLL, color = sex),
               size = 4) +
    scale_color_manual(values = c("pink", "darkgreen")) +
    theme(legend.direction = "horizontal",
          legend.background = element_blank(),
          legend.position = "right",
          legend.title = element_blank(),
          legend.key = element_blank(),
          legend.text = element_text(size = 8)))

dat %>%
  ggplot(aes(x = age)) +
  geom_line(aes(y = YLL)) +
  geom_point(data = dat %>% filter(sex == "female"),
             aes(y = YLL), color = "pink", size = 2) +
  geom_point(data = dat %>% filter(sex == "male"),
             aes(y = YLL), color = "darkgreen", size = 2) +
  geom_smooth(aes(y = YLL, group = 1),
    se = F, color = "#29306c") +
  scale_y_continuous(labels = function(x) 
                              scales::number(x, 
                                             scale = 1e-6, 
                                             accuracy = 0.01, 
                                             suffix = "M"),
                   name = "Value (Millions)") +
  annotation_custom(legend_grob, ymax = max(dat$YLL) * 2 + 10^6) +
  labs(
    title = "Global Region : Years of Life Lost (YLLs) due to stroke",
    subtitle = "Year 2019 - Millions",
    caption = "DataSource: WHO life expectancy & IHME results Stroke data | YLLs Values in Thousands",
    x = "Age Group", y = "YLLs") +
  theme(axis.text.x = element_text(size = 7,
                                   angle = 45,
                                   vjust = 0.5),
        plot.title = element_text(size = 11),
        plot.caption = element_text(size = 8))
```

## YLDs Calculation

**YLDs** (Years Lived with Disability) measure the number of years a
person lives with a disability due to a disease or injury. It is
calculated by multiplying the **prevalence of a condition** by the
**disability weight**, which reflects the **severity** of the
disability.

The key factor of the disability weights (DW) is linked to the severity
(mean of the range of health loss suffered to disease) of a non-fatal
health condition due to disease or injury. DW ranges between 0
(equivalent to full health) and 1 (equivalent to death). The estimation
of the disability weights is challenging and has been continuously
changed by modifying and adapting methodologies in various studies
[@liu2023]. The challenge is assigning disability weights to diseases
with different levels of prevalence and severity, such as cases with
high prevalence and low severity. Furthermore, the results need to be
reported to a year-based value.

Moreover, the calculation of YLDs has been updated over the years,
shifting from an incidence to a prevalence-based approach. More in-depth
analyses are in @sec-03-components; for now, we will show how both types
of calculations differ, but then we will focus on using the most updated
approach based on values of prevalence estimated by the GBD2019,
summarized in the `hmsidwR::g7_hmetrics` dataset.

#### Incidence-based Calculation

**Standard Formula**

$$
YLD_i=I*DW*L
$$ {#eq-yldi}

In @eq-yldi, $I$ is the incidence of the condition, $DW$ is the
disability weight, and $L$ is the average duration of the condition. The
incidence takes into account the number of new cases of a disease or
health condition that occur in a population over a specific period. The
disability weight reflects the severity of the health condition, and the
average duration of the condition is the average length of time a person
lives with the condition.

#### Prevalence-based Calculation

**Standard Formula**

$$
YLD_p=p*DW
$$ {#eq-yld}

In @eq-yld, $p$ is the prevalence, $DW$ are the disability weights.
While an incidence-based approach focuses on the number of new cases of
a health condition, the prevalence-based approach considers the total
number of cases in the population. The disability weight reflects the
severity of the health condition and it is applied to the prevalence to
calculate the YLDs, and the duration of the condition is not considered
in the prevalence-based calculation.

### Example: YLDs due to Stroke

Since the release of GBD 2010, the WHO has decided to switch to a
prevalence-based approach for the calculation of YLDs. The major impact
of this shift is to distribute the weights of the YLDs more evenly
across all age groups, rather than concentrating them at the age of
incidence.

In the following example we use the `disability weights` and the
`severity` levels extracted from a dataset in the GBD 2019 study. The
dataset `disweights2019` is stored in the `{hmsidwR}` package and is
made of `r dim(hmsidwR::disweights2019)[1]` observations and
`r dim(hmsidwR::disweights2019)[2]` variables. It contains the estimated
values for the disability weights, which are measured on a scale from 0 to 1, where 0 equals a state of full health and 1 equals death. 

```{r}
hmsidwR::disweights2019 %>%
  select(specification, cause1, severity, dw) %>%
  drop_na() %>%
  head()
```

The level of `severity` is assigned based on the **level of
disability**, which is categorized as **mild**, **moderate**, or
**severe**. The disability weights are assigned as follows:

-   Mild = 20%
-   Moderate = 45%
-   Severe = 75%

These levels are general for all ages; the values might vary for other
specifications of the level of disability [@wyper2020].

```{r}
dwsev2019 <- hmsidwR::disweights2019 %>%
  select(cause1, severity, dw) %>%
  drop_na() %>%
  mutate(severity_n = case_when(
    severity == "mild" ~ 0.2,
    severity == "moderate" ~ 0.45,
    severity == "severe" ~ 0.75))

dwsev2019 %>% head()
```

The values for disability weights and severity are considered for all
ages; generally, the level of disability weights can differ by age for
different types of stroke, but here, for simplicity, weights are only
varied based on the severity level of stroke.

```{r}
dw_stroke <- dwsev2019 %>%
  filter(cause1 == "Stroke") %>%
  group_by(severity, severity_n) %>%
  reframe(avg_dw = mean(dw))

dw_stroke 
```

And calculate the YLDs as:

$$ \text{Total YLDs} = \text{YLD}_{\text{mild}} + \text{YLD}_{\text{moderate}} + \text{YLD}_{\text{severe}}$$

For instance, here we use the `g7_inc_prev_CVD2019` and the `dw_stroke`
datasets to calculate the YLDs due to stroke, and we want to know how
incidence and prevalence differ. They can be expressed as percent values
or numbers, so for example:

-   if the prevalence of stroke is 0.58 for male with age between 34-39
    and it is expressed in percent value (%), it means 0.58% of the
    population within that age range had a stroke.
-   if the prevalence of stroke is 1,541,506.96 for male with age
    between 34-39 and it is expressed in numbers, it means that there
    are approximately 1,541,507 individuals in the age group 35-39 had a
    stroke.

The percentage and the number are related by the size of the population
in that age group. To convert between these forms, you need to know the
total population in the age group 35-39.

```{r}
inc_prev_stroke_5y <- hmsidwR::g7_inc_prev_CVD2019 %>%
  filter(cause == "Stroke",
         location == "Global") %>%
  select(measure, sex, age, val) %>%
  pivot_wider(names_from = "measure", values_from = "val")

inc_prev_stroke_5y %>% head()
```

The calculation of YLDs due to stroke in the Global region with both
approaches incidence and prevalence is shown below.

While for calculating prevalence-based YLDs, we use the `Prevalence`
values, the `severity` and the `average weights`, for incidence-based
YLDs, we need to consider the `Incidence` values and the
`average duration` of the condition.

In particular, for stroke, the average duration of the condition can
vary based on:

-   duration for acute stroke: up to 28 days
-   duration for chronic stroke: beyond 28 days, often modeled for
    long-term consequences, sometimes up to the lifetime of the patient
    depending on the model used.

In this example, we consider the average duration of the condition for
28 days. It does need to be converted to be year-based:

$$ \hat{L}_{stroke}= \frac{28}{365} $$ These values strongly depends on
the disability weights and the severity values assigned to the
condition. Here we have considered estimated values, which do not
necessarily correspond to the real values.

```{r}
options(scipen = 999)
YLD_global_stroke <- merge(inc_prev_stroke_5y, dw_stroke) %>%
  group_by(sex, age) %>%
  reframe(YLD_p = sum(Prevalence* (1 + severity_n) * avg_dw * 100),
          YLD_i = sum(Incidence * (1 + severity_n) * avg_dw * 28/365 * 100))

YLD_global_stroke %>% 
  filter(!sex == "both",
         age == "35-39")
```

We note the difference in the YLDs if incidence or prevalence is used in
the calculation. The magnitude of the YLDs is similar, but the values
are distributed differently across the age groups.

```{r}
#| echo: false
#| message: false
#| warning: false
dat <- YLD_global_stroke %>%
  filter(!sex == "both")
```

```{r}
#| layout-ncol: 2
#| label: fig-YLD_global_stroke
#| fig-cap:
#|  - "YLDs calculated with prevalence"
#|  - "YLDs calculated with incidence"
#| fig-alt: "YLDs due to stroke in the Global region"
#| echo: false
#| message: false
dat %>%
  ggplot(aes(x = age, y = YLD_p)) +
  geom_line(aes(y = YLD_p)) +
  geom_point(data = dat %>% filter(sex == "female"),
             color = "pink", size = 3) +
  geom_point(data = dat %>% filter(sex == "male"),
             color = "darkgreen", size = 3) +
  geom_smooth(aes(group = 1),
              se = F, color = "#29306c") +
  scale_y_continuous(labels = scales::label_comma(scale = 1 * 1000)) +
  annotation_custom(legend_grob) +
  labs(
    title = "Global region: Prevalence based YLDs due to stroke",
    subtitle = "Year 2019 - Percent Values",
    caption = "DataSource: IHME results Stroke data | YLDs Values X 1000",
    x = "Age Group", y = "YLDs") +
  theme(axis.text.x = element_text(size = 7,
                                   angle = 45,
                                   vjust = 0.5),
    plot.title = element_text(size = 11),
    plot.caption = element_text(size = 8))


dat %>%
  ggplot(aes(x = age, y = YLD_i)) +
  geom_line(aes(y = YLD_i)) +
  geom_point(data = dat %>% filter(sex == "female"),
             color = "pink", size = 3) +
  geom_point(data = dat %>% filter(sex == "male"),
             color = "darkgreen", size = 3) +
  geom_smooth(aes(group = 1),
              se = F, color = "#29306c") +
  scale_y_continuous(labels = scales::label_comma(scale = 1 * 1000)) +
  annotation_custom(legend_grob) +
  labs(
    title = "Global region: Incidence based YLDs due to stroke",
    subtitle = "Year 2019. - Percent Values",
    caption = "DataSource: IHME results Stroke data | YLDs Values X 1000",
    x = "Age Group", y = "YLDs") +
  theme(axis.text.x = element_text(size = 7,
                                   angle = 45,
                                   vjust = 0.5),
    plot.title = element_text(size = 11),
    plot.caption = element_text(size = 8))
```

## DALYs Calculation

As a measure of the overall disease burden, DALYs are used to quantify
the sum of years of potential life lost due to premature death (YLL) and
years lived with disability (YLD). The number of DALYs indicates the
number of years of life lost due to premature deaths, disease, or
injury. This metric takes into account both the quantity and quality of
life.

DALYs are a generalization of the well-known **Potential Years of Life
Lost measure (PYLLs)**, which includes the loss of good health. We do
not consider PYLLs in this book, but more information can be found in
the references [@globalh].

**Standard Formula**

$$
DALYs=YLLs+YLDs
$$ {#eq-daly}

One DALY is one lost year of healthy life. This measure is used to
assess how diseases and injuries impact populations, providing a
comprehensive picture of the overall burden of disease by combining YLLs
and YLDs in different groups.

### Example: DALYs due to Stroke

The sum of YLLs and YLDs releases the overall value of DALYs due to
stroke in the Global region. Here we can see there is no much difference
in magnitude if incidence or prevalence is used.

Before to sum up the YLLs and YLDs, we need to either way transform the
YLLs numbers of years into percent values or the YLDs percent values
into numbers. This is necessary to have the same unit of measure for
both components.

Let's transform the YLDs into numbers, we need to know the total
population for each age group. We use the `data(package = "wpp2022")`
which is a collection of population datasets, which contains the
estimated population for the year 2019 for the Global region.

```{r}
library(wpp2022)
# load the data on the population  
data(popAge1dt) 
# set the age levels
age_levels <- c("<1","01-04", "05-09", "10-14", "15-19", "20-24", 
                "25-29","30-34","35-39", "40-44", "45-49", "50-54", 
                "55-59", "60-64","65-69", "70-74", "75-79", "80-84", "85+")  
  
pop5y2019 <- popAge1dt  %>%
  filter(name == "World",
         year == 2019) %>% 
  select(-country_code, - name, -year) %>%
  mutate(age = cut_number(age, 
                          n = 19, 
                          labels = age_levels, 
                          right = FALSE))%>%
  group_by(age) %>%
  reframe(pop = sum(pop), popM = sum(popM), popF = sum(popF)) %>%
  rename(male = popM, female = popF, both = pop) %>%
  pivot_longer(cols = c(male, female, both), 
               names_to = "sex", values_to = "pop") 
  
pop5y2019$age <- factor(pop5y2019$age, levels = age_levels, ordered = TRUE)  
  
pop5y2019 %>% head()
```

```{r}
YLD_global_stroke_n <- YLD_global_stroke %>%
  select(-YLD_i) %>%
  inner_join(pop5y2019, by = c("age", "sex")) %>%
  mutate(YLD = YLD_p * pop) %>%
  select(-YLD_p)

YLD_global_stroke_n %>% 
  filter(age=="35-39")
 # head()  
```



```{r}
DALY_global_stroke <- merge(YLL_global_stroke, YLD_global_stroke_n) %>%
  mutate(DALY = YLL + YLD)

DALY_global_stroke %>%
  select(sex, age, YLL, YLD, DALY) %>%
  head()
```

```{r}
#| echo: false
#| message: false
#| warning: false
dat <- DALY_global_stroke %>%
  select(sex, age, YLL, YLD, DALY) %>%
  filter(sex == "both") %>%
  arrange(age) %>%
  mutate_if(is.numeric, scale)
```

```{r}
#| warning: false 
#| message: false 
#| echo: false 
#| eval: true 
#| label: fig-DALY_global_stroke 
#| fig-cap: "DALYs due to stroke in the Global region. The blue line represents the DALYs, YLLs and YLDs are pink for females and green for males." 
#| fig-alt: "DALYs due to stroke in the Global region"  

dat %>%
  pivot_longer(cols=c(-sex,-age)) %>%
  ggplot(aes(x = age, y = value, group = name, linetype = name)) +
  geom_line() +
  labs(title = "Global region: Disability Adjusted Life Years (DALYs) due to stroke",
       subtitle = "Year 2019 - Standardized Values",
       caption = "DataSource: IHME results Stroke data | DALYs Values ",
       x = "Age Group", y = "DALYs", linetype = "Metric") +
  theme(axis.text.x = element_text(size = 7, angle = 45, vjust = 0.5),
        plot.title = element_text(size = 11),
        plot.caption = element_text(size = 8),
        legend.position = "bottom")  
 
```

## How the DALYs are Used

The health metrics of DALY, YLL, and YLD can be used in several ways to
help prioritize public health interventions, evaluate the impact of
diseases and injuries, and inform public health decision-making. Some
common uses of these metrics include:

-   **Prioritizing public health interventions**: By calculating the
    overall burden of disease in a population, public health
    practitioners can prioritize which diseases and injuries to address
    first. This helps allocate resources and target interventions to the
    areas of greatest need.
-   **Evaluating the impact of diseases and injuries**: These metrics
    can be used to measure the impact of diseases and injuries on
    individuals and populations and to track changes over time. This
    information can help inform public health decision-making and
    allocate resources more effectively.
-   **Comparing the burden of disease across populations**: DALY, YLL,
    and YLD can be used to compare the burden of disease across
    populations and between different regions. This information can help
    identify disparities in health outcomes and inform targeted public
    health interventions.
-   **Evaluating the effectiveness of public health programs**: These
    metrics can be used to evaluate the impact of public health programs
    and to assess the effectiveness of public health interventions. This
    information can help public health practitioners identify areas for
    improvement and make necessary changes to ensure that programs are
    achieving their goals.
-   **Monitoring global health trends**: DALY, YLL, and YLD can also be
    used to monitor global health trends and track changes in the burden
    of disease over time. This information can be used to inform global
    health policies and allocate resources to address emerging health
    threats.

Overall, the health metrics of DALY, YLL, and YLD provide valuable
information for public health practitioners, researchers, and policy
makers to help prioritize and allocate resources, evaluate the impact of
diseases and injuries, and inform public health decision-making.

### General application of DALYs

As an example here is shown how the DALY metric can be used for
prevention. Suppose we have data on the number of cases of a particular
disease, as well as the average number of years of life lost due to this
disease. We can use this information to calculate the total number of
DALYs lost due to this disease.

```{r}
# Create a data frame with the number of cases and average years of life lost
df <- data.frame(
  YLL = c(5, 10, 15),
  YLD = c(1, 3, 4))

# Calculate the number of DALYs lost
df <- df %>% mutate(DALY = YLL + YLD)

# Sum the total number of DALYs lost
total_dalys <- sum(df$DALY)

total_dalys
```

In this example, the number of cases of the disease and the average
years of life lost for each case are used to calculate the number of
DALYs lost for each case. Finally, the total number of DALYs lost for
the entire population.

This information can be used to inform public health interventions to
prevent the spread of this disease and reduce the number of DALYs lost.
For example, the information could be used to prioritize resources for
disease control and prevention activities, such as health education
campaigns, vaccination programs, and screening and treatment programs.

## HALE Calculation

Healthy Life Expectancy (HALE) is a metric used to estimate the number
of years a person can expect to live in good health, taking into account
both mortality and morbidity factors. It is often used as a measure of
overall population health and quality of life. It takes into account the
impact of both fatal and non-fatal health outcomes on overall life
expectancy. HALE is typically calculated using data on mortality rates
and health-related quality of life measures, such as disability-adjusted
life years (DALYs) or quality-adjusted life years (QALYs). These
measures allow for the estimation of the number of years lost due to
premature death or disability.

It can be used to identify health disparities, assess the effectiveness
of healthcare interventions, and inform public health policies. By
measuring the number of years lived in good health, HALE offers a more
nuanced understanding of population health rather than traditional life
expectancy measures.

**Standard Formula**

$$
HALE= \text{life expectancy}- YLD
$$ {#eq-hale}

HALE\index{HALE} is obtained by subtracting the YLDs from the life
expectancy of a population. However, if we consider more specifications
such as comorbidities and other health factors that can occur in real
life, the calculation can be adjusted by assuming different levels of
prevalence based on disease sequelae with associated disability weights,
and eventually accounting for comorbidity levels by using a Monte Carlo
simulation approach.

We use age-specific mortality and YLDs per capita by location, age, sex,
and year, and define the HALE as:

> the average number of years that a person at a given age can expect to
> live in good health, taking into account mortality and loss of
> functional health.

![HALE - GLOBAL HEALTH METRICS\| VOLUME 392, ISSUE 10159, P1859-1922,
NOVEMBER 10, 2018](images/hale.png){fig-align="center" width="504"}

The process of calculating HALE for a specific population, considering
factors such as sex, country, and year, involves computing the average
health of individuals across different age groups within the population
and integrating information on the prevalence of various health
conditions and their associated disability weights. Comorbidity is
addressed using a Monte Carlo simulation approach, assuming independence
of comorbidities within each age group. Simulations are conducted to
model exposure to different health conditions based on their estimated
prevalence in each age group, resulting in a simulated population
reflecting the prevalence of multi-morbidities [@stanaway2018].

**Positive health** associated with each health condition is defined as
one minus the disability weight $(1 - DW)$. The combined health for an
individual in the simulated population is determined by multiplying
these positive health values for all relevant health conditions present.

The **average health** values are then computed as one minus the Years
Lived with Disability (YLD) $(1 - YLD)$ per person in the population,
which are used to calculate health-adjusted person years. The **Sullivan
method** is employed to incorporate these average health values into the
life table. This involves adjusting the values in the $nLx$ column of
the life table by the corresponding average health values, recalculating
the life table using these adjusted values, and then using an iterative
process to estimate health-adjusted person-years for different age
groups.

Finally, HALE is calculated by dividing the adjusted person-years for
each age group by the proportion of a hypothetical birth cohort still
alive at that age.

**Simulating Life Table Data**

To understand the process of calculating HALE, we can simulate life
table data. We’ll define age intervals (e.g., every 5 years) and
simulate survival probabilities for each age interval (between 0 and 1).

```{r}
set.seed(040424)
age_intervals <- seq(0, 100, by = 5)
survival_probabilities <- runif(length(age_intervals),
                                min = 0.5, max = 1)
```

```{r}
life_table <- data.frame(
  Age = age_intervals,
  Survival_Probability = survival_probabilities,
  nLx = 100000 - (age_intervals * survival_probabilities))

life_table %>% head()
```

We then calculate adjusted $T_x$ for each age group as the sum of
health-adjusted person-years for all age intervals above the current age
interval.

```{r}
# Simulate prevalences
prevalences <- runif(length(age_intervals),
                     min = 0, max = 0.5)
# Simulate disability weights
disability_weights <- runif(length(age_intervals),
                            min = 0, max = 1)
# Calculate average health for each age group
average_health <- 1 - (prevalences * disability_weights)

# Adjust Tx for each age group
life_table$adjusted_px <- life_table$Survival_Probability * average_health

life_table %>% head()
```

Let's make a function to calculate the HALE.

```{r}
calculate_HALE <- function(age, px) {
  # Initialize a vector to store adjusted Tx for each age group
  adjusted_Tx <- numeric(length(age))
  # Iterate over each age interval to calculate adjusted Tx
  for (i in 1:length(age)) {
    # Calculate adjusted Tx for the current age group
    adjusted_Tx[i] <- sum(px[i:length(age)])
    }
  HALE <- numeric(length(age))
  # Calculate HALE for each age group
  for (i in 1:length(age)) {
    # Calculate HALE for the current age group
    HALE[i] <- adjusted_Tx[i] / px[i]
  }
  HALE
}
```

Now, we calculate HALE for the simulated life table data.

```{r}
# Calculate HALE for the simulated life table data
HALE1 <- calculate_HALE(age_intervals, life_table$Survival_Probability)
HALE2 <- calculate_HALE(age_intervals, life_table$adjusted_px)
```

Finally, we visualize the HALE for both the standard and adjusted life
table data.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Health Adjusted Life Expectancy (HALE) by Age. The blue line represents the HALE for the simulated life table data, the brown line represents the HALE for the adjusted life table data."
#| fig-alt: "Health Adjusted Life Expectancy (HALE) by Age"
#| fig-width: 6
#| fig-height: 4

data.frame(age_intervals, HALE1, HALE2) %>%
  pivot_longer(cols = c(HALE1, HALE2),
               names_to = "HALE",
               values_to = "Years") %>%
  ggplot(aes(x = age_intervals)) +
  geom_line(aes(y = Years, group = HALE, 
                color = HALE, linetype = HALE)) +
  scale_color_viridis_d(option = "H") +
  labs(
    title = "Health Adjusted Life Expectancy (HALE) by Age",
    x = "Age Group (5Yrs)")
```

## Summary

In this chapter, we have detailed the methods and calculations for
evaluating the burden of disease using key health metrics: YLLs, YLDs,
and DALYs. By understanding how to compute these metrics, we gain
insights into the impact of diseases and injuries on populations,
guiding public health decisions and interventions. We also introduced
HALE, a metric that combines mortality and morbidity data to provide a
comprehensive measure of population health. These metrics are invaluable
tools for public health practitioners, policymakers, and researchers in
their quest to improve global health outcomes.

In the next chapter (@sec-03-components), we will delve deeper into the
components and variations of these metrics, further enhancing our
understanding of how to utilize them effectively in public health
assessments.
