# Metrics Components {#sec-03-components}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
options(scipen = 999)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   Understand the construction and interpretation of life tables and
    life expectancy measures
-   Analyse mortality levels and rates to assess population health
    dynamics
-   Distinguish between incidence and prevalence and their relevance in
    disease monitoring
-   Interpret disability weights and severity levels in the context of
    burden of disease estimation
:::
::::::

\
In @sec-02-method_calculation, we discussed the importance of health
metrics\index{Health metrics !} in assessing the overall health status
of a population and guiding the allocation of health resources. We have
seen how to calculate key metrics by providing direct calculations of
each of them, but haven't yet discussed the
components\index{Health metrics ! components} that make up these
metrics. In this chapter, we will investigate: **life
expectancy**\index{Life
expectancy}, **mortality rates**\index{Mortality rates}, **incidence and
prevalence**\index{Incidence}\index{Prevalence}, and **disability
weights**\index{Disability weights}. These components are crucial for
understanding the burden of diseases\index{Burden of diseases} and
injury at both global and population levels.

## Cause-Specific or Population-Wide

Health metrics\index{Health metrics} are essential tools for evaluating
the health status of a population and guiding public health
interventions. They provide valuable insights into the burden of disease
and injury, helping policymakers and public health officials make
informed decisions about resource allocation and health policy. By
combining these metrics, policymakers and public health officials can
make informed decisions about how to improve the health outcomes of a
population and reduce the burden of disease and injury.

> **What are the components of these health metrics, and how are they
> calculated?**

The evaluation of health metrics begins with the assessment of their
components, which is crucial because results strongly depend on the
types of **life tables**\index{Life tables}, **mortality
rates**\index{Mortality rates}, and **disability weights** used in the
analysis. Furthermore, it is important to consider whether the objective
of the analyses is **cause-specific**\index{Cause-specific} or
**population-wide**\index{Population-wide}. This can significantly
impact the granularity of the health metrics components, as in alignment
with general principles in epidemiological research and health metrics
analysis, discussed in the Global Burden of Disease (GBD)
study\index{Global Burden of Disease (GBD)} methodologies[@vos2020].

Cause-specific metrics focus on disease intervention planning, such as
assessing the burden of HIV\index{HIV} or
tuberculosis\index{Tuberculosis} on specific populations.
Population-wide metrics, on the other hand, are ideal for broader health
policy decisions, like overall mortality rates or life expectancy, and
provide a comprehensive view of the health landscape.

The results of health metrics calculations can vary significantly based
on various factors, including geographic location, the type and severity
of diseases or injuries, the specific causes of death or disability, and
the age groups analysed (whether age-standardised or covering all ages).
Additional variation arises from the chosen time points of analysis and
whether the metrics are calculated across both sexes or specifically for
males or females. Moreover, the disability weights applied can differ
depending on the condition’s severity, its impact on quality of life,
and other contextual parameters, resulting in tailored calculations that
reflect local health challenges more accurately.

In this chapter, practical examples and case studies are provided to
illustrate the process of applying the components to the metrics. By
understanding the structure of the components, you will be better
equipped to analyse and interpret health data and make informed
decisions about health policy and resource allocation.

## Life Tables and Life Expectancy

The first component is the **life expectancy**\index{Life expectancy !},
used for calculating the YLLs as in @eq-yll. More specification about
**life tables**\index{Life tables} used to calculate the **life
expectancy** is in the @sec-appendixA. As a key parameter in the
calculation of health metrics, life expectancy is essential for
estimating the number of years lost due to premature death.

The **standard life expectancy**\index{Life expectancy ! standard} is
defined as the average number of years a person is expected to live
based on current mortality rates, an important indicator of the overall
health status of a population used to compare the health outcomes of
different populations over time. It is calculated based on the
**probability of survival**\index{Probability of survival} at each age,
taking into account the **mortality rates**\index{Mortality rates} for
that age group.

In life tables, the probability of survival and the mortality rates
across different age groups are used for estimating the number of
remaining expected life years. These tables provide crucial insights
into the longevity patterns within a population, enabling researchers
and policymakers to predict life expectancy trends and to assess public
health strategies effectively.

```{r}
#| echo: false
#| label: tbl-life-table
#| tbl-cap: "Extract of Global Health Observatory (GHO) Life Tables Components"

library(tidyverse)
hmsidwR::gho_lifetables %>%
  filter(year == 2019,
         sex == "both") %>%
  select(-year, -sex) %>%
  pivot_wider(names_from = indicator, values_from = value) %>%
  select(age, lx, ndx, nLx, Tx, ex) %>%
  # round all numeric columns
  mutate(across(where(is.numeric), ~round(.))) %>%
  slice(10:12) %>%
  knitr::kable()
```

Where `lx` is the number of people alive at the beginning of the age,
`ndx` is the number of people dying between age $x$ and $x+1$, `nLx` is
the number of person-years lived between age $x$ and $x+1$, `Tx` is the
total number of person-years\index{Person-years} lived by the cohort of
persons alive at age $x$, and `ex` is the expected remaining lifetime at
age $x$.

In the context of calculating Disability-Adjusted Life Years
(DALYs)\index{Disability Adjusted Life Years (DALYs)}, specifically for
the component related to Years of Life Lost
(YLLs)\index{Years of Life Lost (YLLs)},
$ex$\index{Expected life years (ex)} (the expected remaining lifetime at
age $x$) is typically used rather than overall **life expectancy from
birth**\index{Life expectancy ! at birth} ($le$)\index{Life expectancy}.
This is because YLLs are calculated based on the difference between the
age at death and the expected age at death, which is more accurately
represented by the remaining life expectancy at a specific age.

To understand why the expected remaining lifetime at a specific age is
used in the calculation of YLLs, consider the following example: if a
person dies at age 50, the YLLs are calculated based on the difference
between the expected remaining lifetime at age 50 and the actual age at
death. This approach provides a more precise measure of the years of
life lost due to premature death, as it accounts for the age-specific
mortality rates and the probability of survival at that age.

While the life expectancy at birth provides an overall measure of the
average number of years a person is expected to live, the expected
remaining lifetime at a specific age is more relevant for calculating
YLLs, as it reflects the impact of premature death on the remaining
years of life for individuals at different ages.

### Global Health Observatory Life Tables

The **Global Health Observatory (GHO) Life
Tables**\index{Global Health Observatory (GHO)} are essential tools
provided by the World Health Organization
(WHO)\index{World Health Organization (WHO)} to support global health
monitoring and assessment. They are part of the broader GHO data
repository, which serves as WHO's gateway to health-related statistics
for its 194 Member States. The GHO life tables provide data on life
tables and life expectancy for different age groups. These tables are
collected in the
`{hmsidwR}`\index{Health Metrics Spread Infectious Diseases with R (hmsidwR)}
package and used to show how to derive key health metrics.

```{r}
#| eval: false
# install.packages("devtools")
devtools::install_github("Fgazzelloni/hmsidwR")
```

```{r}
#| eval: true
library(tidyverse)
library(hmsidwR)
```

The **`hmsidwR::gho_lifetables`** dataset contains five variables:

1.  **indicator:** as shown in @tbl-indicators
2.  **age group:** from \<1 to 85+ in 5-year classes
3.  **sex:** female, male, and both
4.  **value:** indicators value
5.  **year:** from 2000 to 2019

```{r}
#| echo: true
#| label: tbl-lifetables
#| tbl-cap: "Global Health Observatory Life Tables"
gho_lifetables %>%
  filter(indicator == "ex", sex == "both") %>%
  select(-indicator, -sex) %>%
  head()
```

The life expectancy rates for each age group are calculated with
consideration of the probability of
survival\index{Probability of survival} based on key parameters such as
age, and deaths probabilities for that age. More info about how to
calculate the life expectancy can be found in the @sec-appendixA section
of this book.

@fig-expectation-life-1 shows the `ex - expectation of life at age x`
represented for each age group while in @fig-expectation-life-2 $ex$ is
represented by averaging value across all age groups from 2000 to 2019.

By averaging across all age groups rather than plotting each line for
each age group individually, we can visualise the overall trend in life
expectancy over time. This provides a more comprehensive view of how
life expectancy has changed across different time.

We can see a difference in the life expectancy values for different age
groups, with the highest values for the youngest age groups and the
lowest values for the oldest age groups. This reflects the impact of age
on life expectancy, with younger individuals having a higher life
expectancy than older individuals. The overall trend in life expectancy
shows an increase over time, indicating improvements in health outcomes
and longevity for the population. More on how to build these plots can
be found in the @sec-09-data_visualisation chapter of this book.

```{r}
#| echo: false
#| layout-ncol: 2
#| label: fig-expectation-life
#| fig-cap: "Global Health Observatory Life Tables - Expectation of Life"
#| fig-subcap:
#|   - "ex - expectation of life at age x"
#|   - "GHO Life Tables"
#| fig-alt: "ex - expectation of life at age x and Global Health Observatory Life Tables - Expectation of Life"

gho_lifetables %>%
  filter(indicator == "ex", sex == "both") %>%
  mutate(age10 = case_when(
    age %in% c("-01", "01-04", "05-09") ~ "0-09",
    age %in% c("10-14", "15-19") ~ "10-19",
    age %in% c("20-24", "25-29") ~ "20-29",
    age %in% c("30-34", "35-39") ~ "30-39",
    age %in% c("40-44", "45-49") ~ "40-49",
    age %in% c("50-54", "55-59") ~ "50-59",
    age %in% c("60-64", "65-69") ~ "60-69",
    age %in% c("85+", "80-84", "75-79", "70-74") ~ "70+",
    TRUE ~ age)) %>%
  group_by(age10, year) %>%
  summarize("value" = round(mean(value), 3)) %>%
  ggplot(aes(year, value, group = age10, 
             color = age10, linetype = age10)) +
  geom_line(linewidth = 1,show.legend = F) +
  scale_color_manual(values = rainbow(20)) +
  labs(
    title = "ex - expectation of life at age x",
    caption = "DataSource: WHO GHO data repository",
    x = "Year", y = "Value") +
  theme(axis.text.x = element_text(size = 7,vjust = 0.5),
        plot.caption = element_text(size = 8))


gho_lifetables %>%
  filter(indicator == "ex") %>%
  group_by(year) %>%
  summarize(value = mean(value)) %>%
  ggplot(aes(year, value)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Global expectation of life (all ages avg value)",
    subtitle = "for the years 2000, 2005, 2010, 2015, and 2019",
    caption = "DataSource: WHO GHO data repository")
```

The mathematical formulation for calculating the expectation of life at
age x is computed as:

$$
e(x) = \frac{T(x)}{l(x)}
$$ {#eq-ex}

where $T(x)$ is the total number of
person-years\index{Total number of person-years (Tx)} lived by the
cohort of persons alive at age x, and $l(x)$ is the number of persons
alive at age x\index{Persons alive at age x (lx)}.

It shows how $ex$ is strongly related to the value of $T(x)$ and $l(x)$.
The total number of person-years lived by the cohort of persons alive at
age x ($T(x)$) typically referred to in demographic and
actuarial\index{Actuarial} studies, as a function to measure the
cumulative amount of life that members of a cohort can expect to live
beyond a certain age. Its changes over time indicate the effects of
health policies, technological advancements, and socio-economic
conditions on the longevity and health of populations.

```{r}
#| echo: false

pyramid <- gho_lifetables %>%
  mutate(value = ifelse(sex == "female",
                        as.integer(value * -1),
                        as.integer(value))) %>%
  filter(!sex == "both", 
         indicator == "ex") %>%
  ggplot(aes(x = age, y = value, fill = sex)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#CC6666", "#9999CC", "#66CC99")) +
  coord_flip() +
  labs(title = "Global Health Observatory Life Expectancy values\nchange 2000-2019\n\n{closest_state}",
       subtitle = "",
       y = "Age",
       caption = "Data Source: WHO | Graphic: @fgazzelloni") +
  theme(axis.text.x = element_blank(),
        axis.title = element_blank())
# pyramid
```

```{r}
#| eval: false
#| echo: false
library(gganimate)
pyramid_gif <- pyramid +
  transition_states(year, transition_length = 1, state_length = 2) +
  enter_fade() +
  exit_fade() +
  ease_aes("cubic-in-out")

animate(pyramid_gif,
  fps = 40, duration = 5, width = 1200, height = 1400, res = 120,
  renderer = gifski_renderer("images/03_gho_pyramid.gif"))
```

::: {.content-visible when-format="html"}
![Global Health Observatory Life Expectancy
2000-2019](images/03_gho_pyramid.gif){#fig-gho_pyramid
fig-align="center" width="375" fig-alt="Global Health Observatory Life"}
:::

#### Exercise

For example, at age 50, what is the life expectancy value if the total
number of person-years lived by the cohort of persons alive at age 50 is
2,370,099 and the number of persons alive at age 50 is 89,867
?[^03-components-1]

[^03-components-1]: Answer: e(50) = T(50)/l(50) = 2,370,099/89,867 =
    26.4 [see @lifeexp].

## Mortality Level and Rates

Another fundamental component of health metrics is the **mortality
level**\index{Mortality} which indicates the number of deaths in a
population. In the calculation of YLLs if the number of deaths are
**cause-specific**\index{Cause-specific}, they are calculated for
different diseases or injuries. If the number of deaths are
**population-wide**\index{Population-wide}, they are calculated for the
entire population.

This component represents the count of deaths at each age or for
specific diseases within a given time period. Data for this component is
collected from vital registration\index{Vital registration} systems,
health surveys\index{Health surveys}, hospital records, and
epidemiological studies.

Here is a table containing the mortality rates for different age groups
in two regions. The mortality rates are expressed as the number of
deaths per 1,000 population in each age group. Data are used to
illustrate the variation in mortality rates across age groups and
regions, providing insights into the health risks faced by different
populations.

```{r}
#| label: tbl-mortality-rates
#| tbl-cap: "Example Mortality Rates by Age Group and Region"
#| echo: false
library(knitr)

# Create a data frame with example mortality rates by age group and region
mortality_rates <- data.frame(
  Age_Group = c("0-1", "1-5", "5-10", "10-15", "15-20", "20-30", "30-40"),
  Region_A_Mortality_Rate = c(40.5, 5.2, 1.8, 0.9, 1.2, 1.5, 2.0),
  Region_B_Mortality_Rate = c(30.0, 4.0, 1.5, 1.0, 1.3, 1.8, 2.2)
)

# Display the table
kable(mortality_rates, col.names = c("Age Group", "Region A Mortality Rate (per 1,000)", 
                                     "Region B Mortality Rate (per 1,000)"), align = "l")
```

### Understanding Death Counts and Mortality Rates

While mortality rates themselves are not directly used in the formula
for the calculation of YLLs, they are crucial for understanding the
broader context of health risks\index{Health risks} and for estimating
the death counts in populations where direct data on deaths might be
incomplete or unreliable. In regions or for populations where vital
registration data are incomplete, mortality rates derived from sample
surveys or demographic models can be used to estimate the expected
number of deaths.

```{r}
#| echo: false
#| label: tbl-mortality-rates-example
#| tbl-cap: "Example Mortality Rates by Age Group and Region"
library(knitr)

# calculate the mortality rate based on Total Population (country x =  5000000, country y = 20000000) and Total Deaths (country x =  25000, country y = 40000)
country_comparison <- data.frame(
  Metric = c("Total Population", "Total Deaths", "Mortality Rate (per 100,000)"),
  Country_X = c(5000000, 25000, (25000/5000000)*100000),
  Country_Y = c(20000000, 40000, (40000/20000000)*100000)
)

# Display the table
kable(country_comparison, 
      col.names = c("Metric", "Country X", "Country Y"), 
      align = "l")
```

An analysis of the mortality rates alongside YLLs provides a full
picture of the health challenges faced by a population. While YLLs
emphasise the impact of premature death, mortality rates help in
understanding the probability and distribution of these deaths within
the population.

The mortality rates can be calculated in different ways, depending on
the specific nature of the analysis. In general, what is considered as
the most common mortality rates is the **Crude Mortality Rate
(CMR)**\index{Crude Mortality Rate (CMR)}, which represents the number
of deaths per 100,000 population.

$$
CMR= \frac{\text{Number of deaths}}{\text{Total population}}*100,000
$$ {#eq-cmr}

::: {.callout .callout-info}
Other types of mortality rates include:

-   **Age Specific Mortality Rate
    (ASMR)**\index{Age Specific Mortality Rate (ASMR)}: the number of
    deaths per 100,000 population in a specific age group

$$
ASMR= \frac{\text{Number of deaths in a specific age group}}{\text{Total population in that age group}}*100,000
$$ {#eq-asmr}

-   **Cause-Specific Mortality Rate
    (CSMR)**\index{Cause Specific Mortality Rate (CSMR)}: the number of
    deaths per 100,000 population due to a specific cause

$$
CSMR= \frac{\text{Number of deaths due to a specific cause}}{\text{Total population}}*100,000
$$ {#eq-csmr}

-   **Infant Mortality Rate (IMR)**\index{Infant Mortality Rate (IMR)}:
    the number of deaths per 1,000 live births in the first year of life

$$
IMR= \frac{\text{Number of infant deaths}}{\text{Total number of live births}}*1,000
$$ {#eq-imr}

-   **Maternal Mortality Rate
    (MMR)**\index{Maternal Mortality Rate (MMR)}: the number of maternal
    deaths per 100,000 live births

$$
MMR= \frac{\text{Number of maternal deaths}}{\text{Total number of live births}}*100,000
$$ {#eq-mmr}

These mortality rates are used to assess the impact of diseases or
injuries on a population and to identify areas where improvements in
health care are needed.
:::

#### Example: Deaths Counts derived by Mortality Rates

Let's simulate\index{Simulation} **malaria**\index{Malaria} death counts
in an African population with a **Case Fatality Rate
(CFR)**\index{Case Fatality Rate (CFR)} ranging from 0.01% to 0.40%, and
a case-incidence\index{Case-incidence} of 59.4 per 1000 population
annually[@worldma]. Calculate the number of expected malaria cases and
the resulting deaths based on this rate for a simulated population of
100,000 individuals over a year:

-   Define the Population: Assume a total population of 100,000.
-   Set the Malaria Incidence Rate: 59.4 per 1000 population annually.
-   Calculate Total Malaria Cases.
-   Apply the Case Fatality Rate: Calculate estimated deaths using the
    CFR range.\

```{r}
#| label: synthetic-malaria-mortality
# Define the total population
total_population <- 100000

# Define the incidence rate (cases per 1,000 population)
incidence_rate <- 59.4

# Calculate the total number of malaria cases per year
total_cases <- (incidence_rate / 1000) * total_population

# Case fatality rates (min and max)
cfr_min <- 0.0001 # 0.01%
cfr_max <- 0.004 # 0.40%

# Calculate estimated deaths for the minimum and maximum case fatality rates
deaths_min_cfr <- total_cases * cfr_min
deaths_max_cfr <- total_cases * cfr_max
```

```{r}
#| echo: false
print(paste("Total population:", total_population))
print(paste("Malaria incidence rate per 1,000 population:", incidence_rate))
print(paste("Total estimated malaria cases per year:", round(total_cases)))
print(paste("Estimated deaths at minimum CFR (0.01%):", round(deaths_min_cfr)))
print(paste("Estimated deaths at maximum CFR (0.40%):", round(deaths_max_cfr)))
```

In the simulation scenario, the incidence rate\index{Incidence rate} of
59.4 cases per 1,000 annually helps to understand how pervasive malaria
is within the simulated population. By applying this rate to the entire
population, we can estimate the total number of new cases per year.
Combining this with the case fatality rate gives an estimate of the
expected number of deaths, highlighting the impact of malaria on public
health within that population.

## Incidence and Prevalence

Incidence\index{Incidence} and prevalence\index{Prevalence} are two
fundamental concepts used in epidemiology\index{Epidemiology} to measure
how often a disease occurs and how widespread it is within a population
at a specific time.

-   **Incidence** refers to the number of new cases of a disease that
    develop in a population during a specific time period. It provides
    information about the risk of contracting the disease and is usually
    expressed as a rate — for example, the number of new cases per 1,000
    people per year.

The incidence rate, even identified as the cumulative incidence
(risk)\index{Cumulative incidence
risk (cir)} - $r$ is considered as a measure of the probability the
disease occurs in a specified period of time. It quantifies the risk of
an individual in the population at risk developing the disease during
the specified time period.

$$
\text{Incidence rate}= \frac{\text{number of new cases at time t}}{\text{total population at risk}}*1000
$$ {#eq-incidence-rate}

-   **Prevalence** measures the total number of cases of a disease in a
    population at a given time, including both new and existing cases.
    It is expressed as a proportion of the population and helps to
    provide a snapshot of how widespread the disease is.

The prevalence identifies the burden of disease at a specific time,
rather than the probability of the disease to impact on the population,
it expresses the probability of suffering from a disease, it is based on
the total number of existing cases among the whole population.

$$\text{Prevalence}=\frac{\text{number of new + existing cases at time t}}{\text{total population}}$$ {#eq-prevalence}

Like incidence, it can be multiplied by a factor of 10\^n (where n is
typically 1,000, 10,000, or 100,000) to express the prevalence per a
standard number of people.

```{r}
#| label: tbl-flu-metrics
#| tbl-cap: "Example Influenza Incidence and Prevalence Data"
#| echo: false
library(knitr)

# Create a data frame with example influenza incidence and prevalence data
flu_metrics <- data.frame(
  Metric = c("Incidence (new cases per 1,000 people)", "Prevalence (existing cases per 1,000 people)"),
  January = c(5.0, 15.0),
  April = c(3.0, 10.0),
  July = c(1.0, 5.0),
  October = c(2.0, 7.0),
  December = c(6.0, 20.0)
)

# Display the table
kable(flu_metrics, col.names = c("Metric", "January", "April", "July", "October", "December"), align = "l")
```

In summary, incidence measures the rate of new cases arising within a
population over time, while prevalence measures the proportion of
existing cases within the population at a specific point in time. These
metrics are widely used in the literature and form the basis for many
decisions in public health and clinical practice. They are discussed in
foundational texts such as *"Epidemiology*\index{Epidemiology}: An
Introduction" by Kenneth J. Rothman and in guidelines by health
institutions like the Centers for Disease Control and Prevention
(CDC)\index{Centers for Disease Control and Prevention (CDC)} and the
World Health Organization (WHO)\index{World Health Organization (WHO)}.
Both metrics are crucial for understanding the dynamics and burden of
diseases within populations.

### Use of Prevalence in DALYs Calculation

In the context of the Global Burden of Diseases
(GBD)\index{Global Burden of Diseases GBD} study, the use of prevalence
in the calculation of Disability-Adjusted Life Years
(DALYs)\index{Disability Adjusted Life
Years (DALYs)} is pivotal because DALYs aim to measure both the
**current impact** of diseases and their **long-term effects** on
populations as this metric combines years of life lost due to premature
mortality (YLLs)\index{Years of Life Lost (YLLs)} with years lived with
disability (YLDs)\index{Years Lived with Disability (YLDs)}. In
particular, for YLDs calculations, prevalence rather than incidence is
used for several reasons:

-   **Direct Measurement of Burden**: Prevalence data directly reflect
    the current burden of disease in the population, including both new
    and ongoing cases. It provides a snapshot of all individuals
    affected by the disease at a given time, which is essential for
    calculating the total amount of health loss due to disabilities.
-   **Incorporation of Disease Duration**: Using prevalence allows the
    calculation of YLDs to incorporate the duration of the disease until
    recovery or death. This is because prevalence captures both existing
    and new cases during the period of measurement.
-   **Complexity of Disease Models**: Many chronic diseases or
    conditions with long duration are better captured through prevalence
    data. This approach simplifies the modelling by avoiding the need
    for complex calculations that would otherwise be required to
    estimate the duration of each new case (from incidence data) and its
    progression over time.

The choice of prevalence over incidence in the calculation of YLDs thus
reflects a pragmatic approach to capturing the current impact of
diseases within a population, particularly for chronic diseases where
ongoing management and sustained healthcare support are crucial.

#### Example: Calculating YLD for a Disease Using Prevalence

Suppose we want to calculate the **Years Lived with Disability (YLDs)**
for a **chronic respiratory disease** in a population of 100,000 people.
The prevalence rate of the disease is 2%, with a Disability Weight
(DW)\index{Disability weights} of 0.3 and an average duration of the
disease is 5 years. The YLD is calculated using the formula:

1.  $\text{Prevalent Cases} = \text{Population Size} \times \text{Prevalence Rate}$

$$\text{Prevalent Cases} = 100,000 \times 0.02 = 2,000 \text{ cases}$$

2.  $\text{YLD} = \text{Prevalent Cases} \times \text{Disability Weight} \times \text{Average Duration}$

$$\text{YLD} = 2,000 \times 0.3 \times 5 = 3,000$$

So, the YLD for chronic respiratory disease in this population over the
average duration of the condition is 3,000 years. This YLD value would
then be added to the YLL (Years of Life Lost) for the disease to
calculate the total DALY.

#### Example: COVID-19 Incidence and Prevalence

Let's simulate\index{Simulation} some COVID-19\index{COVID-19} cases for
a population of 100 000 individuals over 365 days. Consider an average
infection time of 5 days or 1/5=0.2 (20%) rate of infection, and an
average time of recovery of 20 days which corresponds to a recovery rate
of 1/20=0.05 (5%)

```{r}
#| label: tbl-covid19-synthetic
#| tbl-cap: "COVID-19 Cases and Recovered - simulate data"
library(tidyverse, quietly = T)

population_size <- 100000 # Number of individuals
days <- 365 # Number of days

set.seed(123)
dates <- seq(as.Date("2019-01-01"),
             as.Date("2019-12-31"),
             by = "day")
```

Cases and recovered are simulated using a
`Poisson distribution`\index{Poisson distribution} with $\lambda$ of 0.2
and 0.05, respectively. To model the count of rare events in a large
population happening within a fixed interval of time or space, the
Poisson distribution is particularly useful. It suits the initial spread
of COVID-19 in many regions, and assumes that the events (new cases)
occur independently of each other, which can be a reasonable
approximation in the context of infection spread over short periods.

To apply the Poisson distribution, we use the `rpois()` function, which
generates random numbers from a Poisson distribution with a specified
rate parameter $\lambda$. The probability mass function of the Poisson
distribution is given by:

$$
  P(X = k) = \frac{\lambda^k e^{-\lambda}}{k!}
$$ {#eq-rpois}

where $X$ is the random variable representing the number of events, $k$
is the number of events, $\lambda$ is the rate parameter, and $e$ is the
base of the natural logarithm.

```{r}
cases <- rpois(length(dates), lambda = 0.2)
recovered <- rpois(length(dates), lambda = 0.05)
```

The simulated data are then combined into a data frame with the date,
number of cases, and number of recovered individuals.

```{r}
covid_data <- data.frame(
  Date = dates,
  Cases = cases,
  Recovered = recovered)
```

Let's have a look at the first few rows of the dataset:

```{r}
#| echo: false
covid_data %>%
  head()
```

Calculate the cumulative incidence and prevalence over time:

```{r}
#| label: tbl-covid19-inc-prev-synthetic
#| tbl-cap: "COVID-19 Incidence and Prevalence - simulate data"
covid_data <-
  transform(covid_data,
            Cum_Cases = cumsum(Cases),
            Cum_Recovered = cumsum(Recovered))

# Calculate daily incidence and prevalence per 100,000 population
covid_data <-
  transform(
    covid_data,
    Incidence = (Cum_Cases / population_size) * 1000,
    Prevalence = ((Cum_Cases - Cum_Recovered) / population_size) * 1000)
```

```{r}
#| echo: false
covid_data %>%
  select(-Cases, -Recovered) %>%
  head()
```

```{r}
total_cases <- sum(covid_data$Cases)
total_recoveries <- sum(covid_data$Recovered)
```

#### Incidence and Prevalence Rates for the Year

```{r}
(incidence_rate <- total_cases / population_size)
(prevalence_rate <- (total_cases - total_recoveries) / population_size)
```

```{r}
#| label: fig-inc-prev
#| fig-cap: "Incidence and Prevalence COVID-19 Cases - simulate data"
#| fig-alt: "Incidence and Prevalence COVID-19 Cases - simulate data"
#| echo: false
#| eval: true

ggplot(covid_data, aes(x = Date)) +
  geom_line(aes(y = Incidence, 
                color = "Incidence"), 
            linetype = "dotted",
            size = 1) +
  geom_line(aes(y = Prevalence, 
                color = "Prevalence"), 
            size = 1) +
  scale_color_manual(values = c(
    "Incidence" = "#29306c",
    "Prevalence" = "#a60845")) +
  labs(
    title = "Incidence and Prevalence COVID-19 Cases",
    caption = "Data: Synthetic Data",
    x = "Date",
    y = "Number per 100,000 Population",
    color = "Metric") +
  theme(legend.position = "bottom",
        plot.title = element_text(face="bold"))
```

## Disability Weights and Severity Levels

The **disability weights**\index{Disability weights} measure the
**severity**\index{Severity} of the health loss associated with specific
health outcomes or diseases, ranging from 0 (no disability) to 1
(equivalent to death). These weights are used to calculate the number of
years lived with a disability
(YLDs)\index{Years Lived with Disability (YLDs)} in the calculation of
DALYs\index{Disability Adjusted Life Years (DALYs)}, and play a pivotal
role between mortality and morbidity when estimating the Health Adjusted
Life Expectancy (HALE)\index{Health Adjusted Life Expectancy (HALE)}.

The disability weights are crucial for understanding the impact of
diseases and injuries on the health status of a population, if the
disability weight of a specific disease is overestimated, the burden of
the disease may be overestimated. Conversely, if the disability weight
is underestimated, the burden of disease could result
underestimated[@ock2019].

The severity levels of a disease or injury are determined through a
combination of clinical data, epidemiological studies, and expert input.
These levels are crucial for understanding the burden of disease and for
developing strategies in healthcare management and resource allocation.
The severity levels can be assigned to be mild, moderate, or high, with
respective probabilities reflecting the likelihood of each severity
level occurring within the affected population.

```{r}
#| label: tbl-example-disability-weights
#| tbl-cap: "Example Disability Weights"
#| echo: false
library(knitr)

# Create a data frame with example disability weights
disability_weights <- data.frame(
  Condition = c("Mild Depression", "Moderate Anxiety", "Severe Stroke", 
                "Diabetes (without complications)", "Advanced Dementia"),
  Disability_Weight = c(0.15, 0.30, 0.75, 0.10, 0.85)
)

# Display the table
kable(disability_weights, col.names = c("Condition", "Disability Weight"), align = "l")
```

For determining the disability weights, the severity levels are assigned
based on the impact of the condition on an individual's quality of life.
Higher weights are assigned to more severe conditions, reflecting the
greater impact on an individual's health status, while lower weights are
assigned to less severe conditions. Hence, the disability weights are
calculated based on the severity of the condition and the impact on an
individual's quality of life.

### Methodology for Disability Weights

The disability weights can be derived or computed, depending on data
availability. Often derived through large-scale population surveys, or
expert panels through reviews of clinical descriptions and assessment
scores based on understanding and professional experience. These methods
involve trade-off exercises, such as **Time Trade Off
(TTO)**\index{Time Trade-Off (TTO)}[@lugnér2020] or **paired
comparison**\index{Paired
comparison} tasks[@rehm2013]. Through surveys the respondents are asked
to choose between different health states or to trade life years in
different health states against a standard of full health. Expert panels
would assess the severity of different health states based on clinical
descriptions and their professional experience, rating the severity on a
scale from 0 (no disability) to 1 (equivalent to death).

Computed disability weights methods can vary from applying techniques
such as **Analytical Hierarchy Process
(AHP)**[@analytic2024]\index{Analytical Hierarchy Process (AHP)} method
used where health states are systematically compared in a pairwise
fashion, and judgements about their relative severity are quantified. Or
**Combining Results**\index{Combining Results (CR)} from different
methods and different populations to produce a set of disability weights
that reflect a consensus view. Statistical models might be used to
aggregate these results and ensure that they are consistent across
different scales.

The Global Burden of Disease (GBD)\index{Global Burden of Disease (GBD)}
study led by the Institute for Health Metrics and Evaluation
(IHME)\index{Institute for Health
Metrics and Evaluation (IHME)} is one of the most comprehensive efforts
to estimate disability weights for a wide range of health conditions.
The GBD study periodically updates and refines disability weights based
on extensive global surveys and expert consultations. In the 2010
iteration, for instance, IHME used community surveys across multiple
countries, including Bangladesh, Indonesia, Peru, and Tanzania, and an
online survey to gather feedback on how people perceive the severity of
different health states. Respondents were asked to rank or rate
hypothetical health conditions, which provided data for calculating
standardised disability weights.

#### Disability Weights Data {#sec-03-disweights}

Here is an example of a table containing the disability weights for
twelve combinations of different conditions such as Infectious disease,
Non-communicable diseases, and other health conditions.

Data are from the results of a systematic analysis of Global burden of
369 diseases and injuries in 204 countries and territories, 1990–2019
[@GBD2020]. The table can be downloaded from the `{hmsidwR}` package,
and contains the first two cause-specific disability weights for the
year 2019, as well as the severity level of the condition set as to be
mild, moderate, severe, and combined for selected communicable, and
Non-communicable diseases. To access the data:

```{r}
#| eval: true
library(hmsidwR)

disability_weights <- hmsidwR::disweights
```

The table here shows the first 6 rows of the disability
weights\index{Disability weights} for cause1, severity, and disability
weights for the year 2019 for 204 countries and territories.

```{r}
#| label: tbl-disability-weights
#| tbl-cap: "Disability Weights"
disability_weights %>%
  filter(year == 2019) %>%
  select(specification, severity, dw) %>%
  head()
```

If we focus on the first cause of disability for infectious
disease\index{Infectious diseases}, we can see that the disability
weight varies along with the severity level. This indicates that the
condition associated with a moderate level of severity has a disability
weight of 0.057. This is set to reflect the impact of the severity
condition on an individual's quality of life.

```{r}
#| label: tbl-severity-disability-weights
#| tbl-cap: "Severity and Disability Weights"
library(hmsidwR)
disability_weights <- hmsidwR::disweights

disability_weights %>%
  filter(year == 2019,
         cause1 == "Infectious disease") %>%
  group_by(severity) %>%
  summarize(dw = mean(dw))
```

## Summary of the DALYs' Components

### YLLs Components

The components of YLLs\index{Years of Life Lost (YLLs)} include several
factors that contribute to the calculation of premature death due to a
disease or injury. These components are:

-   **Age at death**\index{Age at death}: The age at which a person died
    due to a disease or injury is a crucial component of YLL. The
    earlier the age at death, the greater the impact on potential years
    of life lost.

-   **Life expectancy**\index{Life expectancy}: The expected age at
    death in a population without the disease or injury is an important
    component of YLL. This value is used to compare the actual age at
    death with the expected age at death and determine the number of
    years of life lost.

-   **Standard life expectancy**\index{Standard life expectancy}: To
    make comparisons across populations and over time, YLL is often
    expressed relative to a standard life expectancy, typically set at
    an age of 70 years.

-   **Population size**: The size of the population affected by a
    disease or injury is another important component of YLL. A larger
    population will have a greater impact on overall YLL, regardless of
    the age at death.

-   **Cause of death**: The cause of death is also considered when
    calculating YLL, as different causes may have different impacts on
    potential years of life lost.

### YLDs Components

The estimation of YLDs requires certain epidemiological parameters such
as prevalence, incidence, case-fatality
rate\index{Case Fatality Rate (CFR)}, relative
risk\index{Relative Risk (RR)}, odds ratio\index{Odds Ratio (OR)},
hazard ratio\index{Hazard Ratio (HR)}, mean, incidence ratio, severity,
duration\index{Duration} and remission\\index{Remission.

In particular, we are interested in the major factors that contribute to
the calculation of the number of years lived with a disability due to a
disease or injury. These components are:

-   **Prevalence and Incidence**: The prevalence of a condition is the
    number of cases of a particular disease or injury present in a
    population at a given time. This is an important component of YLD as
    it determines the number of people who are affected by a disease or
    injury and the overall impact on the population. The incidence is
    the impact of new cases on a population in a specified time period,
    and it is calculated by considering the number of new cases affected
    by a condition divided by the size of a population. Both incidence
    and prevalence can be used in the calculation.

-   **Disability weight**: The disability weight reflects the severity
    of the disability caused by a disease or injury. It is used to
    quantify the impact of a condition on quality of life, with higher
    weights assigned to more severe conditions.

    Prior to 2010, the calculation of YLDs included elements of
    discounting and age weighting, with discount rates that typically
    ranged from 3% to 4% per year. Considering the future years of
    healthy life valued less than present years, the practice aimed to
    reflect the economic theory of time preference, where immediate
    benefits are preferred over future benefits.

    However, due to its complexity and ethical implications, this
    approach has been criticised. By 2010, the World Health Organization
    (WHO) and other health organisations began to consider simplifying
    the methodology. The focus shifted to only the disability weights
    used to quantify the severity of different health conditions,
    leaving aside the element of discounting

-   **Age**: The age of the person affected by a disease or injury is
    also considered when calculating YLD. Conditions that occur at an
    earlier age will have a greater impact on the number of years lived
    with disability.

-   **Population size**: The size of the population affected by a
    disease or injury is another important component of YLD. A larger
    population will have a greater impact on overall YLD, regardless of
    the age of the affected individuals.

-   **Duration of disability**: The duration\index{Duration} of
    disability is also considered when calculating YLD. Conditions that
    last for a longer period of time will have a greater impact on the
    number of years lived with disability.

### DALYs Components

Both YLLs and YLDs are components\index{Health metrics} of the
DALYs\index{Disability Adjusted Life Years (DALYs)} and are used to
provide a comprehensive assessment of the impact of disease and injury
on a population. By combining YLLs and YLDs, the DALYs take into account
both premature death and the impact of disease or injury on quality of
life, providing a more comprehensive view of the overall burden of
disease.

DALYs are especially useful in guiding the allocation of health
resources as they provide a common numerator, allowing for the
expression of utility in terms of dollar/DALY.

For example, in Gambia, the provision of the pneumococcal conjugate
vaccine costs \$670 per DALY saved. This number can then be compared to
other treatments for other diseases, to determine whether investing
resources in preventing or treating a different disease would be more
efficient in terms of overall health[@disabili2023].

## Case Study: Germany Lung Cancer Study {#sec-03-case-study-lungc}

For this case study, we used data from the GBD study for Germany,
related to the number of deaths due to lung-cancer in 2019. Data are
available in the `{hmsidwR}` package and, and includes the upper and
lower bounds of the estimates and the 5-yrs age group for both sexes. We
do not have information for the 0-4, 5-9 and 85+ age groups.

Here we can see the first 6 rows of the data, that we are going to use
to calculate the YLLs, the YLDs and finally the DALYs.

```{r}
#| label: tbl-germany_lungc
#| tbl-cap: "Germany lungcancer"
hmsidwR::germany_lungc %>%
  select(age, sex, prevalence, dx) %>%
  head()
```

Joining `germany_lungc` with the GHO life tables (`gho_lifetables`) we
can build a new dataset with all information needed for calculating the
YLLs for the lung cancer\index{Lung cancer} deaths in Germany in 2019.

```{r}
gelung_YLL <- germany_lungc %>%
  filter(sex == "both") %>%
  left_join(gho_lifetables %>%
    filter(year == 2019, indicator == "ex") %>%
    select(-indicator, -year) %>%
    rename(life_expectancy = value)) %>%
  group_by(age) %>%
  reframe(dx, life_expectancy, YLL = dx * life_expectancy)

gelung_YLL %>%
  head()
```

Now to calculate the YLDs due to lung-cancer we need to consider not
fatal disability status. This can appear in conjunction with other
conditions, or morbidities, that are not fatal but are acting together
with the lung-cancer[@porst2022]. The disability weights are available
from GBD study[@ihmeuw/i2024] are the same as those found in the
`disweights` dataset.

```{r}
#| label: tbl-disweights-lung
#| tbl-cap: "Lung cancer - Disability Weights"
lung_dw <- hmsidwR::disweights %>%
  filter(year == 2019,
         str_detect(sequela, "lung")) %>%
  select(sequela, dw)
```

So, 0.451 is the disability weight for the metastatic phase of lung, or
lung cancer. We can use this value together with the prevalence of the
disease to show how to calculate the YLDs for the lung cancer deaths in
Germany in 2019. The number of lung cancer cases in Germany in 2019 is
34,000, and the prevalence of the disease is 0.0004 for all ages and
sex. While considering 5-yrs age groups with a common disability weight
of 0.451. Total YLDs can be calculated as the sum of the product of the
prevalence and the disability weight for each age group.

```{r}
prev_germany_lungc <- germany_lungc %>%
  filter(sex == "both") %>%
  select(age, prevalence)

prev_germany_lungc %>% head()
```

```{r}
#| label: tbl-prev_germany_lungc_2019
#| tbl-cap: "Lung Cancer Cases in Germany 2019"
gelung_YLD <- prev_germany_lungc %>%
  mutate(dw = 0.451, YLD = prevalence * dw)

gelung_YLD %>%
  head()
```

```{r}
gelung_YLL %>%
  select(age, YLL) %>%
  left_join(gelung_YLD %>% select(age, YLD)) %>%
  mutate(DALY = YLL + YLD)
```

```{r}
(YLLs <- sum(gelung_YLL$YLL))
(YLDs <- sum(gelung_YLD$YLD))
(DALYs <- YLLs + YLDs)
```

Per 1000 population, the YLLs, YLDs, and DALYs for lung cancer deaths in
Germany in 2019 are 821, 1, and 822, respectively. These metrics provide
a comprehensive view of the impact of lung cancer on the health status
of the population, combining the years of life lost due to premature
death with the years lived with disability.

```{r}
DALYs / 1000 # DALYs per 1000 population
```

The value of the metrics changes when different disability weights are
adopted. It is noteworthy that the value for the Years Lived with
Disabilities (YLDs) is often quite small compared to the contribution
made by the Years of Life Lost (YLLs) in the calculation of
Disability-Adjusted Life Years (DALYs). This disparity highlights how
mortality frequently dominates the overall disease burden in many health
evaluations, underscoring the critical impact of premature death on
public health metrics.

For instance, consider the disability weights for lung cancer as
calculated in a study tailored to the Korean population[@ock2019]. These
weights are subdivided into four groups corresponding to the stages of
the disease, from 1 to 4. Each stage reflects increasing severity and a
corresponding increase in the disability weight. Such gradations are
crucial for accurately reflecting the progression of the disease and its
escalating impact on patients' lives. This differentiation allows health
policymakers and researchers to fine-tune interventions and allocate
resources more effectively, targeting the stages that contribute most
significantly to health deterioration.

|                                              |                    |       |       |
|-------------------|------------------|------------------|------------------|
| Cause of Disease                             | Disability Weights | lower | upper |
| Trachea, bronchus and lung cancers (stage 1) | 0.600              | 0.542 | 0.656 |
| Trachea, bronchus and lung cancers (stage 2) | 0.738              | 0.686 | 0.785 |
| Trachea, bronchus and lung cancers (stage 3) | 0.758              | 0.710 | 0.801 |
| Trachea, bronchus and lung cancers (stage 4) | 0.906              | 0.873 | 0.932 |

In conclusion, disability weights have a fundamental role in influencing
the overall calculations of
DALYs\index{Disability Adjusted Life Years (DALYs)}, as they reflect the
severity of health conditions and their impact on quality of life.
Assigning appropriate weights to different diseases and injuries might
be challenging, but it is essential for accurately assessing the burden
of disease and guiding health policy decisions. The case study on lung
cancer in Germany illustrates how the combination of
YLLs\index{Years of Life Lost (YLLs)} and
YLDs\index{Years Lived with Disability (YLDs)} provides a comprehensive
view of the health impact of a specific condition, helping to inform
public health strategies and resource allocation.
