# Introduction to Machine Learning {#sec-05-machine_learning}

> "Human beings are not just passive victims of disease. They are active
> participants in the biological confrontation between man and microbe."

<br>

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

This chapter will walk you through a basic construction of a machine
learning model, providing a simple explanation of how to model
fast-growing phenomena, such as in the case of the spread of infectious
diseases. Then in the following chapters of this second section you will
dive deep into different techniques for modeling starting from feature
engineering to making predictions.

## Deterministic and Stochastic modeling

The spread of a virus can be seen as a **random process**, since the
number of individuals who are infected at any given time can change
randomly. The exact number of individuals who will be infected in the
future cannot be determined with certainty, as it depends on various
factors such as the contagiousness of the virus, the behavior of
individuals, and the efficacy of mitigation measures.

A **deterministic model**, on the other hand, could be used to model the
spread of a virus under certain conditions, such as a fixed number of
individuals, constant contagiousness, and no mitigation measures. This
type of model can be used to make predictions about the spread of a
virus under certain assumptions, but it will not account for the
randomness and uncertainty associated with real-world scenarios.

In general, a **stochastic process**, or random process [@dobrow2016] is
the type of model which attempt to replicate uncertain outcomes, using
probabilities. At opposite, in a **deterministic system** the outcome is
obtained from a given input, and for this reason it is reproducible.

Most models used to study the spread of a virus are a **combination of
both deterministic and stochastic models**. For example, the SIR
(Susceptible-Infected-Recovered) model is a deterministic model that
describes the dynamics of the spread of a virus, but it also includes
stochastic elements such as random interactions between individuals.

## What is a machine learning model?

And how it differs from statistical models (deterministic and
stochastic) is all a matter of parameter calibration. The difference
between a deterministic and stochastic model is key to understanding
when machine learning comes into play. Machine learning models, learn
from data to automatically adjust their parameters, optimizing them to
minimize errors or maximize performance on a given task. They are more
flexible and adaptable to complex data patterns compared to traditional
statistical models.

The emergence of infections is an ongoing evolution and adaptation
[@broemeling2021]. In the late 1970s, Dr. René Dubos believed that '*the
relationship between humans and microbes is dynamic and influenced by
various factors, including environmental conditions, human behavior, and
microbial adaptation*'. The importance of understanding the complex
interactions between hosts and pathogens in the emergence and spread of
infectious diseases is described as a dynamic interplay between
organisms and their environment—**a process of continuous adaptation and
interaction**. Organisms continuously respond to changes in their
surroundings to maintain balance and survival. This is analogous to what
happens inside a machine learning model environment. It involves
continuous parameter learning and adjustment achieved through a
technique called tuning or parameter calibration, resulting in a range
of possible alternative scenarios and simulations of the observed data.

Machine learning models fall under the category of **empirically driven
models**. Unlike **mechanistic models**, which are based on explicit
equations and known relationships between variables (such as the
differential equations used in infectious disease modeling), machine
learning models derive patterns and relationships directly from data
without explicit knowledge of the underlying mechanisms. Therefore,
while both mechanistic and empirically driven models are used for
prediction, they operate on different principles:

> Mechanistic models rely on known relationships and equations, while
> machine learning models learn patterns from data[@baker2018].

Before delving deep into machine learning techniques for analyzing
health data and metrics, we need to proceed step by step in **building a
modeling framework**. This requires a thorough understanding of the
modeling procedure. Therefore, we start by building a linear regression
model, then transition to **applying a logistic regression model**,
which is the most suitable for analyzing our data. Finally, we identify
the most appropriate model adjustment with automated parameter
calibration.

## The steps of building a model

The process of constructing a mathematical or computational model that
represents the relationships between variables in a given system is
intended as model development. At the heart of this process is the
application of a model function to estimate coefficients that minimize
the difference between the model's predictions and the observed data.

Let's break this down further: Suppose we have a response variable,
denoted as $y$, which represents the number of deaths or infections in
our dataset. Additionally, we have predictor variables $x_1, x_2, ...$,
which are other factors that may influence the level of $y$. These
predictor variables could include factors such as population density,
temperature, vaccination rates, etc. The goal of model development is to
create a mathematical function that captures the relationship between
the response variable $y$ and the predictor variables $x_i$.

This function, often referred to as the model equation or model
function, is typically represented as:

$$y = f(x_1, x_2, ...) + \epsilon$$

Here, $f$ represents the relationship between the predictors and the
response, and $epsilon$ represents the error term, which captures the
difference between the observed values of $y$ and the values predicted
by the model. To estimate the coefficients of the model function, we use
statistical techniques such as linear regression (continuous response
variable), logistic regression (binary response variable), or machine
learning algorithms (with tuning parameters). These techniques analyze
the relationship between the predictors and the response in the dataset
and determine the optimal coefficients that minimize the difference
between the observed $(y)$ values and the values predicted by the model
$(\hat y)$ .

Once the model coefficients are estimated, we can use the model function
to predict the value of $y$ for new sets of predictor variables $x_i$.
This allows us to simulate the effect of different scenarios or make
predictions about future outcomes based on the relationships identified
in the data. Overall, the model development process is a crucial step in
the modeling framework, as it sets the foundation for understanding and
analyzing the relationships between variables in a given system. It
enables us to extract insights from the data, make predictions, and
inform decision-making in various fields, including epidemiology, public
health, economics, and environmental science.

### The Cholera example

The first dataset used is from the `{HistData}` package
`HistData::CholeraDeaths1849`, made of
`r dim(HistData::CholeraDeaths1849)[1]` observations and
`dim(HistData::CholeraDeaths1849)[2]` variables, with 2 causes of deaths
cholera and diarrhaea. For this example we select just the deaths due to
cholera within 12 months in 1849. Let's have a look at the first 6 rows
of the dataset:

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(HistData)
cholera <- HistData::CholeraDeaths1849 %>%
  filter(cause_of_death=="Cholera")%>%
  select(date,deaths)
  
cholera %>% head()
```

The response variable is `y = deaths` due to cholera, and the only
predictor we are considering on this first step is `x = date` in days
from 1849-01-01 to 1849-12-31.

Now let's observe in mathematical terms what happens when we start
investigating the relationship between the response and the predictor,
to explain the behavior of the variables.

The following mathematical formulation represent our observed data,
where $\beta_0$ and $\beta_1$ are the parameters, respectively the
intercept and the slope allowing the computation of the equivalence
between $y$ and $x$ [^05-machine_learning-1]

[^05-machine_learning-1]: We can compare this formulation to a general
    $Y=a+mx$ linear function. In our case $a$ would be the intercept and
    $m$ the slope of the line that will summarize the information
    between $y$ and $x$.

$$
y=\beta_0 + \beta_1 x
$$ {#eq-lm}

When a model is attempted, the values of the slope $(\beta_0)$ and the
intercept $(\beta_1)$ are estimated[^05-machine_learning-2] in order to
be able to replicate the values $(y)$ of the response by applying $x$ to
our model function.

[^05-machine_learning-2]: The hat (\^) on top of the variables indicates
    the values are estimated.

$$
\hat y = \hat \beta_0+\hat \beta_1x
$$

where $\hat \beta_0$ is the estimate value of the intercept, and
$\hat \beta_1$ the estimated value of the slope. So, the estimated data
will approximate the observed data, with some margin of error
$(\epsilon)$:

$$
y \sim \hat y + \epsilon
$$

The difference between $y$ and $\hat y$ is the error we make when
applying a model.

$$
y- \hat y = \epsilon
$$

Reducing the error $(\epsilon)$ as much as possible can lead to better
fit of the model.

While the case of being in the presence of just one predictor
influencing our response variable might be useful for educational
purposes, as in the data above, the scenario with more than one
predictor can be more realistic and it alters the model function to
become more complex. This is referred to as a multilinear model
function.

$$y = \beta_0 + \beta_1x_1 + \beta_2x_2 + ... + \beta_nx_n$$

In this case more than one predictor is used to investigate the varying
level of the response variable. A compact form of the model function
with more than one predictor can be stated as:

$$
y=\beta_0 + \sum_{i=1}^{p}{\beta_ix_i}= \beta_0+\beta X
$$

In this notation, the capital $X$ takes the form of a matrix containing
various $x_i$: $X=(x_1,x_2,x_3,...,x_p)$, the predictors, and the
$\beta=(\beta_1,\beta_2,\beta_3,...,\beta_p)$ are the coefficient to be
estimated.

> Valuable insights into the trends and drivers of cholera mortality can
> be gained in order to inform public health interventions and policies
> aimed at reducing the burden of this disease.

In the case of **CholeraDeaths1849**, the application of a model
function requires one more step to be taken. First we visualize how the
response variable relates to each of the predictors in order to
establish if it is the case of a linear or non-linear relationship. In
case of infectious diseases, the trend in time is usually starting at
some point and then grows to reach a pick which is the highest level of
infection spread to eventually slow down to were it started. This is
generally represented with a logistic model, which is not linear.

To visualize the spread of cholera in time we use the `{ggplot2}`
package with `geom_line()` and `geom_point()`
functions[^05-machine_learning-3].

[^05-machine_learning-3]: More about data visualizations techniques are
    in chapter @sec-data_visualization.

```{r}
#| eval: false
ggplot(cholera,aes(x=date,y=deaths))+
  geom_line(linetype="dashed",linewidth=0.2)+
  geom_point(shape=21,stroke=0.5)+
  geom_smooth(method = "lm",
              se=F)+
  geom_smooth(method="gam", 
              color="darkred",se=F)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-cholera1849
#| fig-cap: "Deaths due to Cholera in London (1849)"

ggplot(cholera,aes(x=date,y=deaths))+
  geom_line(linetype="dashed",linewidth=0.2)+
  geom_point(shape=21,stroke=0.5)+
  geom_smooth(method = "lm",se=F)+
  geom_smooth(method="gam", 
              color="darkred",se=F)
  labs(title="Deaths due to Cholera in London (1849)",
       x="Date",y="Cholera death")+
  annotate(geom="text",
           x=mean(cholera$date),y=-10,
           label="linear")+
  annotate("curve", 
           x = mean(cholera$date)+20, 
           xend = mean(cholera$date)+30, 
           y = 0, yend = 120,
           arrow=arrow(angle = 30, 
                       length = unit(0.05, "inches"),
                       ends = "last", type = "closed"),
           colour = "blue")+
    annotate(geom="text",
           x=mean(cholera$date)-50,y=400,
           label="logistic (non-linear)")+
  annotate("curve", 
           x = mean(cholera$date)+20, 
           xend = mean(cholera$date)-20, 
           y = 400, yend = 500,
           arrow=arrow(angle = 30, 
                       length = unit(0.05, "inches"),
                       ends = "first", type = "closed"),
           colour = "darkred")
```

#### Generalized Additive Models (GAMs)

The purpose of creating a model is to identify underlying patterns
within a series of observations. This requires the model to interpolate
given observations in order to represent the overall pattern accurately.
For instance, in the visualization above, the blue line made with the
`geom_smooth()` function, when specified with `method = "lm"`, helps us
visualize the direction of a linear pattern in the data. However, it's
evident that the data points form a bell-shaped curved as distribution
of deaths over time, indicating a non-linear relationship between date
and cholera deaths. In the second layer we used `method= "gam"`,
Generalized Additive Models (GAMs) flexible extensions of linear models,
the most suitable choice for this data.

$$
g(\mu) = \beta_0 + f(x)
$$ {#eq-gam}

where: $g()$ is a link function, $\mu$ is the expected value of the
response variable, $\beta_0$ is the intercept term, $f()$ represents the
smooth functions of the predictor variable. In case of more than one
predictor: $g(\mu) = \beta_0 + f_1(x_{i1}+f_2(x_{i2}+f_3(x_{i3}+...)$
there will be $p=1,..,j$ number of $f()$ smooth functions as many as the
number of predictors.

What is happening here is a transformation of the predictors through the
application of a function: $x ~ f(x)$. The smooth functions are
estimated using non-parametric techniques such as splines (polynomial
transformations) or kernel functions. These smooth functions allow for
flexible modeling of the relationship between the predictor variables
and the response variable, capturing non-linearities and complex
patterns in the data. The link function $g()$ is typically chosen based
on the distributional characteristics of the response variable, to
represent the data mean trend.

```{r}
# Load the mgcv package
library(mgcv)

# use the days instead of the full date
cholera$days <- row_number(cholera)

# Fit a GAM using the gam() function
gam_model <- gam(deaths ~ s(days), data = cholera)

# Print the summary of the GAM model
summary(gam_model)
```

The smooth term for days is significant, as indicated by the approximate
significance test. The effective degrees of freedom (edf) for the smooth
term are 8.969, suggesting a moderately flexible relationship between
deaths and time. The adjusted R-squared value is 0.914, indicating that
the model explains approximately 91.4% of the variance in the response
variable (deaths). The deviance explained is 91.6%, suggesting that the
model fits the data well.

We can even use the plot() function as shown below to see the shape of
the model results.

```{r}
#| label: fig-gam
#| fig-cap: "Cholera 1849 - Gam Model"
plot(gam_model, select = 1)
```

This indicates that the relationship between deaths and time is modeled
non-linearly using a smooth function.

Overall, the model suggests that there is a positive relationship
between increasing time and the number of deaths due to cholera, with
the number of deaths increasing over time to reach a pick and then
eventually slow down to zero. However, it's essential to consider the
context of the data and potential confounding variables before drawing
conclusions about causality or making predictions based solely on this
model. Other important factors influencing the spread are social
behavior and vaccination. We will talk more about infectious diseases
spread in section 4 @sec-infectious_diseases.

Now calculate the prediction, with the `predict()` function on new data,
30 days forward in time.

```{r}
# Predict future deaths
future_days <- seq(max(cholera$days) + 1, max(cholera$days) + 30)  

# Assuming you want to predict for the next 30 days
predicted_deaths <- predict(gam_model, 
                            newdata = data.frame(days = future_days),
                            type = "response")
```

We have used new data (30 days in the future) for predicting the number
of deaths due to cholera. These predictions are based on the
relationship between deaths and time captured by the GAM model.

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-gam-pred
#| fig-cap: "Deaths due to Cholera in London (1849)"
ggplot(cholera,aes(x=date,y=deaths))+
  geom_line(linetype="dashed")+
  geom_point(shape=21,stroke=0.5)+
  geom_smooth(method = "gam",se=F)+
  geom_point(data=data.frame(days = as_date("1849-01-01") +future_days,
                             predicted_deaths),
             mapping=aes(x=days,y=predicted_deaths),
             color="red")+
  labs(title="Deaths due to Cholera in London (1849)",
       x="Date",y="Cholera death")
```

::: callout
### Important considerations

There a variety of models, the model decision depends upon the type of
data, if the outcome is continuous or discrete, if the number of
predictors is higher than the number of observation and so on. We can
have non-linear models, as well as multilevel models and more complex
combination. A major subdivision is between **regression** and
**classification**, whether the outcome variable is continuous or not.

**Machine learning models** are all that types of models that can be all
of the above but allow for model parameters auto-calibration. More will
be explained below, but crucial is the understanding that when we
attempt making a model we do high level calculations, for instance when
make differential equations. The next level of a basic equation with
various parameters, included time and probabilities. In substance we
imagine to consider the proportion of variation in time when time
reduces to zero. In the case of studying infectious disease and
interested in time varying variables we would need a model that
considers time.
:::

So, let's start looking at how our simple linear equation $y=a+mx$ can
change into a differential equation. A differential equation describes
how a quantity changes with respect to another quantity, in particular
time in this case. So, we are interested in the variation of $y$ within
time.

$$
{y}'=\frac{dy}{dt}= \frac{y_1 - y_0}{t_1-t_0}
$$

And this is explained by the variation of the predictors in time, the
calculation is a derivative. Recall that the derivative of a constant is
0, in our case $a$ is the constant. $$
\frac{dy}{dt}= \frac{d(a+mx)}{dt}= m\frac{dx}{dt}
$$ This equation describes how $y$ changes with respect to time $t$,
taking into account the rate of change of $x$ with respect to time 't',
and the constant slope $m$ of the line.

$$
y'=m\frac{dx}{dt}
$$

To locate the machine learning models within the models framework, we
can think about the scope of the analysis, which in this case is
predicting the future or to make forecasts.

In particular, these types of models are subdivided in:

1.  **Mechanistic Model** For example the SIR model is a **classic
    compartmental model** used to simulate the spread of infectious
    diseases. It divides the population into three compartments:
    susceptible (S), infectious (I), and recovered (R). The equations
    describe the rate of change of each compartment over time, where
    $\beta$ represents the transmission rate and $\gamma$ represents the
    recovery rate.

Mathematical Equation:

$$ \frac{dS}{dt} = - \beta \cdot S \cdot I $$

$$\frac{dI}{dt} = \beta \cdot S \cdot I - \gamma \cdot I $$

$$ \frac{dR}{dt} = \gamma \cdot I $$

2.  **Empirically Driven Model**: Empirically driven models do not have
    explicit mathematical equations like mechanistic models. Instead,
    they learn patterns and relationships directly from data. An example
    of an empirically driven model is **Random Forest** which is a
    machine learning algorithm that builds multiple decision trees and
    combines their predictions to make accurate predictions. It works by
    splitting the data into subsets and building a decision tree for
    each subset. The final prediction is made by aggregating the
    predictions of all decision trees. Unlike mechanistic models, Random
    Forest does not rely on explicit equations or known relationships
    but instead learns patterns from the input data.

### The Covid19 example

#### SEIR and Random Forest

In this example we simulate data of an epidemic to build a SEIR
(susceptible, exposed, infected, recovered) model, a case a little more
complicated than just making a SIR (susceptible, infected, recovered)
model which we will see in more details in @sec-covid19.

For this example we use a package named `{deSolve}` which solves
differential equations.

```{r}
# Load required packages
library(deSolve)  # For solving differential equations
```

We define a function for making the SEIR model, and its parameters.
$\beta$ is the transmission rate, $\sigma$ the rate of latent
individuals becoming infectious, and $\gamma$ the rate of recovery.

```{r}
seir_model <- function(time, state, parameters) {
  # variables need to be in a list
  with(as.list(c(state, parameters)), {
    # Parameters
    beta <- parameters[1]  # Transmission rate
    sigma <- parameters[2]  # Rate of latent individuals becoming infectious
    gamma <- parameters[3]  # Rate of recovery
    
    # SEIR equations
    dS <- -beta * S * I / N
    dE <- beta * S * I / N - sigma * E
    dI <- sigma * E - gamma * I
    dR <- gamma * I
    
    # Return derivatives
    return(list(c(dS, dE, dI, dR)))
  })
}
```

Then simulate starting parameters by assigning a value to them.

```{r}
N <- 1000  # Total population size
beta <- 0.3  # Transmission rate
sigma <- 0.1  # Rate of latent individuals becoming infectious
gamma <- 0.05  # Rate of recovery
```

Set an initial state and a time vector, let's say 100 days.

```{r}
initial_state <- c(S = N - 1, E = 1, I = 0, R = 0)  # Initial state

# Time vector
times <- seq(0, 100, by = 1)
```

Here we use the `ode()` function from the `{deSolve}` package for the
solving the differential equations in the the SEIR model function that
we created above.

```{r}
output <- deSolve::ode(y = initial_state, 
                       times = times, 
                       func = seir_model,
                       parms = c(beta = beta, 
                                 sigma = sigma, 
                                 gamma = gamma))
```

The output is class "deSolve", "matrix" :

```{r}
class(output)
```

So, needs to be converted to be a dataframe. And here we have the time,
the susceptible, the exposed, the infected and finally the recovered
simulated values in 100 days of an epidemic.

```{r}
simulated_data <- as.data.frame(output)
head(simulated_data)
```

```{r}
# Plot simulated data
ggplot(simulated_data, aes(x = time)) +
  geom_line(aes(y = S, color = "Susceptible")) +
  geom_line(aes(y = E, color = "Exposed")) +
  geom_line(aes(y = I, color = "Infectious")) +
  geom_line(aes(y = R, color = "Recovered")) +
  labs(title = "Simulation of SEIR Model", x = "Time", y = "Population") +
  scale_color_manual(values = c("Susceptible" = "blue", "Exposed" = "orange",
                                 "Infectious" = "red", "Recovered" = "green")) 

```

The next step is to calibrate this model by applying a machine learning
algorithm. In this case we use the a `random forest` type of model, and
add some noise to the simulated data, to be normally distributed.

```{r}
simulated_data <- cbind(simulated_data, 
                        noise = rnorm(nrow(simulated_data), 
                                      mean = 0, sd = 5))
```

Feature engineering: Generate lagged values of infected cases

```{r}
simulated_data <- cbind(simulated_data, 
                        lagged_I = lag(simulated_data$I, 1))
```

```{r}
# Load required packages
library(randomForest)

# Define the predictors and response variable
predictors <- c("S", "E", "I", "R", "noise", "lagged_I")
response <- "deaths"  # Assuming "deaths" is the outcome variable we want to predict

# Split the data into training and testing sets
set.seed(123)  # Set seed for reproducibility
train_index <- sample(nrow(simulated_data), 0.8 * nrow(simulated_data))
train_data <- simulated_data[train_index, ]
test_data <- simulated_data[-train_index, ]

# Train the Random Forest model
rf_model <- randomForest(formula = I ~ ., 
                         data = train_data)
```

```{r}
# Make predictions on the test set
predictions <- predict(rf_model, newdata = test_data)

# Calculate Root Mean Squared Error (RMSE)
rmse <- sqrt(mean((test_data$I - predictions)^2,na.rm = T))

# Print RMSE
print(paste("RMSE:", rmse))
```

#### GAM - Generalized Additive Model (non-linear)

```{r}
library(tidymodels)
library(mgcv)  # For GAM modeling

# Prepare data for machine learning
split <- initial_split(simulated_data,prop = 0.8)
train_data <- training(split)
test_data <- testing(split)
```

```{r}
#| warning: false
#| message: false
gam_model <- gam(I ~ te(S,E) + s(time, by=I, k=20) + R + noise, 
                 data = train_data, 
                 method = "GCV.Cp", # Generalized Cross-Validation
                 family = gaussian,
                 sp = 0.1)

# Make predictions on test data
predictions <- predict(gam_model, 
                       newdata = test_data,
                       type = "link", 
                       se.fit = TRUE)

# Evaluate the model
rmse <- sqrt(mean((predictions$fit - test_data$I)^2))
cat("Root Mean Squared Error (RMSE):", rmse, "\n")
```

```{r}
#| label: fig-seir-gam
#| fig-cap: "Covid19 Infections with GAM"
#| echo: false
# Plot simulated data and predictions
simulated_data %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = I, color = "Infected")) +
  geom_line(data = test_data, 
            aes(color = "Predicted"),y = predictions$fit) +
  labs(title = "Simulation and Prediction of Covid19 Infections with GAM", 
       x = "Time", y = "Infected Population") +
  scale_color_manual(values = c("Infected" = "blue", "Predicted" = "red"))
```

### INLA: an empirical Bayes approach to GAMs

## Measures of Machine Learning Models

> The loss function is **a method for evaluating how well a machine
> learning algorithm models featured a
> dataset**[^05-machine_learning-4]. The ***cost function*** is the
> average of all loss function values.

[^05-machine_learning-4]: <https://builtin.com/machine-learning/common-loss-functions>

-   overview of modeling

### Loss functions based on modeling techniques

Modeling techniques are Regression and Classification modeling, this is
depending on whether the output or the reference variable of the
analysis is numeric or categorical, respectively.

-   Regression:

    -   Mean Square Error / Quadratic Loss / L2 Loss

        $$
        MSE=\frac{1}{n}\sum_{i=1}^{n}{(\hat{}Y_i-Y_i)^2}
        $$

    -   Mean Absolute Error / L1 Loss

        $$
        MAE=\frac{\sum_{i=1}^{n}{|y_i-x_i|}}{n}
        $$

    -   Huber Loss / Smooth Mean Absolute Error

        $$
        L_{\delta}(y,f(x))=\left\{\begin{matrix} \frac{1}{2}(y-f(x))^2 & \text{for } |y-f(x)|\leq \delta, \\ \delta|y-f(x)|-\frac{1}{2}\delta^2 & \text{otherwise.} \end{matrix}\right.
        $$

    -   Log-Cosh Loss

    -   Quantile Loss

-   Classification:

    -   Binary Cross-Entropy Loss / Log Loss

        $$
        L=-\frac{1}{m}\sum_{i=1}^{m}{(y_i*log(\hat{y_i})+(1-y)log(1-\hat{y}))}
        $$

    -   Hinge Loss

        $$
        L=max(0,1-y*f(x))
        $$

### Public health loss functions

> "The public may be less interested in environmental quality than in
> economic prosperity." Dubos

In the field of public health the measures of cost like mortality and
morbidity are used to evaluate the outcome of health and well-being of a
population.

The most common analyses methods are[@hult2021]:

-   cost-utility, using quality-adjusted life years (QALY)

-   cost-effectiveness, for specific outcomes, such as life expectancy
    or medical outcomes.

As mentioned in the history of health metrics, a cost-utility framework
is one used for the calculation of QALY, a measure that within time has
been considered as discriminating disabilities in the population by
assessing a cost which was too higher or too low in most of the
situation and was largely criticized for not being able to capture the
whole picture.

Comprehensive tools assessing the impacts through cost-effectiveness
measures to estimate the effectiveness of health interventions should
also include a measure of disability. The measure of DALY (1993)
developed for major consideration of the disability status for the
evaluation of the well-being of a population was then built adding a
further layer based on weights of disability.

So, to recap, the measures for a cost-effectiveness intervention are:

-   mortality
-   morbidity
-   disability
