# Techniques for Machine Learning Applications {#sec-06-techniques}

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   How to manipulate data through feature engineering
-   Select the most suitable model for your data
-   Learn how to apply machine learning algorithms
:::
:::

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

## Data Analytics versus Data Science

It an ongoing debate that involves the distinction of the field of data
analytics from that of data science. Essentially we need both, and for
two reasons:

-   rigorous analysis of data available
-   provide a perspective on overall results

Data analytics is mostly focused on analyzing data, which means
collecting, checking, tidying and manipulating to finally apply ad hoc
statistical techniques.

Data science on the other side provide a overall conclusion about the
data provided, reached through simulations, application of probabilities
and declaring definitive conclusions sentencing about the significance
of the data test on a particular matter of concern.

In this book we will go through both of them as we analyze data and then
provide an overall conclusion about the answer of the research topic. To
be more specific, an overview of different method of data loading and
featuring selection is provided before to get into selecting the best
model to use.

## Collecting and Analyzing Data

Collecting data to use in a research analysis involves a selection of
sources and methods to use for optimizing computational time when
downloading and reading the files. Once data is set and ready to use a
further step is required to make the data suitable for the selected
model.

The source of data is an important variable. Generally, data can be
downloaded by using an API (application programming interface) which
allow the user to get access to data directly from source, with the use
of specified back-end computations. There are alternatives at using an
API; data can be obtained by downloading it directly into the computer,
or loaded through library packages. Usually, available files are
provided under various forms such as delimited type of files, .csv,
.xls, .json, and other types.

Here is an example of how to use an API for downloading a file directly
onto your computer.

```{r}
#| eval: false
library(httr)
url <- ""
httr::GET(url = url)
```

### Download Data with APIs

#### IHME Data API

```{r}
#| warning: false
#| message: false
library(httr)
library(jsonlite)
library(tidyverse)
```

SDG query input:

-   GetGoal - GetIndicator - GetTarget - GetLocation - GetAgeGroup -
    GetScenario - GetSex

SDG query input for results:

-   GetResultsByTarget - GetResultsByIndicator - GetResultsByLocation -
    GetResultsByYear

```{r}
url="https://api.healthdata.org/sdg/v1/"
key="YOUR-KEY-HERE"
get_data <- function(url,key,sdg) {
  library(httr)
  library(jsonlite)
  url <- paste0(url,"/",sdg)
  headers <- c('Authorization'=key)
  res <- GET(url, add_headers(headers))
  data = fromJSON(rawToChar(res$content))
  data = data$results
}
```

Build the query to download results by adding specification of the data.

```         
GetResultsByLocation?location_id=86&indicator_id=1002&year=2019

? location_id & indicator_id ...
```

Let's have a look at the cases of malaria in Italy in 2019:

```{r}
#| echo: false
key="8ym9ht687svohiwtqiinfjd3gxq7vihc"
```

```{r}
#| eval: false
data <- get_data(url="https://api.healthdata.org/sdg/v1",
         key=key,
         sdg="GetResultsByLocation?location_id=86&indicator_id=1002&year=2019")
data
```

### Synthetic Data

```{r}
# Load required packages
library(deSolve)  # For solving differential equations
library(ggplot2)  # For data visualization

# Function to define the SEIR model
seir_model <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Parameters
    beta <- parameters[1]  # Transmission rate
    sigma <- parameters[2]  # Rate of latent individuals becoming infectious
    gamma <- parameters[3]  # Rate of recovery
    
    # SEIR equations
    dS <- -beta * S * I / N
    dE <- beta * S * I / N - sigma * E
    dI <- sigma * E - gamma * I
    dR <- gamma * I
    
    # Return derivatives
    return(list(c(dS, dE, dI, dR)))
  })
}

# Simulation parameters
N <- 1000  # Total population size
beta <- 0.3  # Transmission rate
sigma <- 0.1  # Rate of latent individuals becoming infectious
gamma <- 0.05  # Rate of recovery
initial_state <- c(S = N - 1, E = 1, I = 0, R = 0)  # Initial state

# Time vector
times <- seq(0, 100, by = 1)

# Solve the SEIR model
output <- ode(y = initial_state, times = times, func = seir_model,
              parms = c(beta = beta, sigma = sigma, gamma = gamma))

# Convert output to data frame
simulated_data <- as.data.frame(output)

# Plot simulated data
ggplot(simulated_data, aes(x = time)) +
  geom_line(aes(y = S, color = "Susceptible")) +
  geom_line(aes(y = E, color = "Exposed")) +
  geom_line(aes(y = I, color = "Infectious")) +
  geom_line(aes(y = R, color = "Recovered")) +
  labs(title = "Simulation of SEIR Model", x = "Time", y = "Population") +
  scale_color_manual(values = c("Susceptible" = "#29306c", "Exposed" = "orange",
                                 "Infectious" = "#a60845", "Recovered" = "#a8bd3a")) 

```

Once data is on your computer available and ready to use, the next step
is to have a look at it and decide whether to perform some adjustment to
the data to make it suitable for your model.

This step includes:

-   data manipulation/wrangling
-   featuring engineering
-   exploratory data analysis

### Feature Engineering

Identifying the most relevant information within the data representing
it in a way that enhances the model's ability to learn patterns and make
accurate predictions, involves the process of selecting, transforming,
and creating new features (variables) from raw data. This is what is
meant with **feature engineering**.

Let's look at the deaths rates due to rabies and reshape the data to
make a machine learning model that would take consideration of the
deaths rates due to all causes within a time frame of 29 years from 1990
to 2019 in Asia and the Global region. In addition we consider the
number of DALYs due to both rabies and all cause mortality. The first
thing to notice is that deaths rates and DALYs are on different units,
rates and years respectively.

```{r}
rabies <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)
rabies %>% head
```

Creating new features from existing ones provide additional predictive
power. In this case we have split the cause vector into two indicating
deaths due to both causes separately. Also, we can scale and center the
numeric variables so to have homogeneous data to use in the model, this
is useful in particular when a transformation is applied. We could use
splines, for example.

```{r}
all_causes <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(!cause=="Rabies")%>%
  rename(dx_allcauses=Deaths,dalys_allcauses=DALYs)%>%
  select(-cause)

dat <- rabies %>%
  full_join(all_causes,by=c("location","year"))

dat %>% head()
```

This is the data we have obtained, and to be able to visualize the
magnitude of difference it is necessary to scale or standardiza the
data.

```{r}
dat%>%
  select(-year,-location)%>%
  scale()%>%
  cbind(dat%>%select(year,location))%>%
  ggplot(aes(x=year,group=location,color=location))+
  geom_line(aes(y=dx_rabies),
            linewidth=1)+
  #geom_line(aes(y=dx_allcauses))+
  geom_line(aes(y=dalys_rabies))
  #geom_line(aes(y=dalys_allcauses))
```

```{r}
suppressPackageStartupMessages({library(tidymodels)})
set.seed(04042024)
split <- initial_split(dat,prop = 0.8)
training <- training(split)
test <- testing(split)
folds <- vfold_cv(training)
```

Then create a recipe with some key steps to trasnform the data into a
model digestable shape.

```{r}
rec <- recipe(dalys_rabies ~. , data=training)%>%
  step_dummy(all_nominal_predictors())%>%
  step_scale(all_predictors())%>%
  step_center(all_predictors())%>%
  step_corr(threshold = 0.8)%>%
  step_YeoJohnson(all_predictors())

rec %>%
  prep()%>%juice()%>%head
```

### Machine Learning Modeling Techniques

#### Random Forest

There are various type of modeling that can be used, let's get started
with a random forest, this choice is due to this type of algorithm which
will create multiple bootstrap samples (random samples with replacement)
from the original dataset. Each bootstrap sample is used to train a
separate decision tree.

Rabies death rates may exhibit complex relationships with predictor
variables. Random forests are capable of capturing non-linear
relationships between predictors and the target variable.

Also, it handles missing data well, provides variables importance and
are an ensemble learning method, which means they combine the
predictions of multiple individual decision trees to produce a more
accurate and stable prediction.

In our simplified case this type of model will do random samples with
replacement of data. In {tidymodels} we can select different types of
engines, in the case of random forest we could use randomForest, ranger,
and others. The difference between these engines derives from the
specific type of calculation used to make the estimation. The Ranger
engine is notably faster than randomForest, so let's use that for this
example.

```{r}
rf_mod <- rand_forest(mtry = tune(),
                      trees = tune(),
                      min_n = tune(),
                      mode = "regression",
                      engine = "ranger")
```

```{r}
wkf <- workflow(preprocessor = rec,
                spec = rf_mod)

rf_res <- tune_grid(object = wkf,
          resamples = folds,
          grid = 5,
          control = control_grid(save_pred = TRUE))
```

```{r}
#rf_res%>%unnest(.metrics)
show_best(rf_res, metric = "rmse")
```

```{r}
autoplot(rf_res, metric = "rmse") +
  scale_x_log10()
```

```{r}
rf_res_tuned <- select_best(rf_res, metric = "rmse")
```

```{r}
rf_fit <- wkf %>% 
  finalize_workflow(select_best(rf_res, 
                                metric = "rmse")) %>%
  fit(training)
```

```{r}
rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test) %>% 
   rmse(truth = dalys_rabies, estimate = .pred)
```

```{r}
rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test)  %>%
  ggplot(aes(dalys_rabies, 
              .pred, 
              color = location)) +
   geom_point() +
   geom_abline(color = "red") +
   scale_color_brewer(type = "qual") +
   coord_fixed() +
   expand_limits(x = 0, y = 0)
```

#### 

## Conclusion
