# Techniques for Machine Learning Applications {#sec-06-techniques}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   How to manipulate data through feature engineering
-   Select the most suitable model for your data
-   Learn about machine learning algorithms
:::
::::::

\
Selecting the most suitable machine learning\index{machine learning}
model involves understanding **the goals of the analysis**, the **nature
of the data**, and the **statistical and machine learning methods** that
best suit the tasks. In @sec-05-machine_learning, we learned about what
machine learning models are, provided examples for building a model
framework, and selected common metrics for model performance calibration
and evaluation.

In this chapter, we focus on the strategies for selecting appropriate
models by leveraging the strengths of different techniques, specifically
for health metrics and for infectious diseases. We will explore various
considerations involved in addressing potential biases, and discuss
actions to prevent them.

## Goals of the Analysis and Nature of Data

The identification of the primary goal of the analysis is fundamental.
Whether it involves trend analysis, investigating the relationships
between response and predictor variables, or strictly forecasting to
predict future outcomes, the strategy for model selection varies
accordingly.

**Health Metrics Data:**

-   **Composite Measures:** Health metrics like DALYs are composite
    measures that include both mortality and morbidity data, often
    requiring sophisticated regression models capable of handling
    continuous variables and multiple predictors. By examining the
    components of DALYs (e.g., Years of Life Lost (YLLs) and Years Lived
    with Disability (YLDs)), we can identify the key drivers such as
    mortality rates, disease prevalence, and risk factors.

-   **Regression Models:** Regression models, including **linear
    regression**, **Ridge regression**\index{ridge regression}, and
    **Lasso regression**\index{lasso regression}, are commonly used to
    handle these continuous variables and address challenges like
    correlation and multicollinearity\index{multicollinearity} with
    appropriate techniques such as regularisation.

**Infectious Disease Data:**

-   **Categorical and Continuous Data:** Infectious disease data can be
    categorical (e.g., disease presence or absence) or continuous (e.g.,
    incidence rates). Classification models are suitable for categorical
    outcomes, while regression models are appropriate for continuous
    data.

-   **Disease Dynamics:** Understanding the dynamics of infectious
    diseases, such as transmission rates, incubation periods, and
    immunity, informs the selection of models. Common models include
    compartmental models (e.g., SIR, SEIR) and agent-based models.

**Common considerations for health metrics and infectious diseases data
type:**

-   **Seasonality and Trends:** The data may exhibit seasonality or
    trends, necessitating the use of time series analysis models like
    **ARIMA**\index{ARIMA} or seasonal decomposition to capture these
    patterns.

-   **Simulation Models:** These models can predict the impact of
    interventions on DALYs and infectious diseases, estimating the
    effectiveness of different interventions and guiding policy
    decisions. Examples of these types of models are: **SIR** models,
    and **Agent-based models**. In addition, **confidence intervals**
    and **sensitivity analyses** help assess the uncertainty associated
    with these predictions.

-   **Bayesian Models:**\index{bayesian models} These models can
    estimate parameters and make predictions based on prior knowledge
    and observed data, incorporating uncertainty and variability.

-   **Predictive modelling:**\index{predictive modelling} Such as
    **decision trees**\index{decision trees}, **support vector machines
    (SVM)**\index{SVM}, and **LSTM (Long Short-Term Memory) neural
    networks**\index{LSTM}, can predict disease outbreaks, identify
    high-risk populations, and optimise resource allocation.

## Statistical and Machine Learning Methods

The choice of model depends on the type of data, the relationships
between variables, and the goals of the analysis. Once we have these
factors well identified, we are a step forward in restricting the range
of applicable models.

The next step involves conducting a thorough **exploratory data analysis
(EDA)**\index{EDA}. This initial exploration helps to uncover the
underlying structure of the data, the relationships between variables,
and the way the response variable—which may also be referred to as the
outcome variable—depends on predictors. This phase is critical as it
informs the necessity of subsequent data adjustments and
transformations.

The importance of data preparation and exploratory data analysis in
machine learning are the building blocks in the preparation of machine
learning digestible data. **Feature
engineering**\index{feature engineering} is a technique that involves
creating new features from existing ones based on domain knowledge or
transformation of data to improve the model's ability to discern
patterns. For example, creating features like **moving averages** or
differences between consecutive days can reveal trends and cycles that
are not immediately apparent from raw data.

Another example is the **standardisation** process, which is crucial
when dealing with variables measured in different units. It involves
rescaling the features so they have a mean of zero and a standard
deviation of one. This process is particularly important when variables
span several orders of magnitude; without standardisation, a model might
incorrectly interpret the scale of a feature as a proxy for importance.

Furthermore, the application of transformations, such as **logarithmic**
scaling or the application of **spline** functions can help in managing
skewed data or enhancing model ability to capture non-linear
relationships, which result particularly useful in complex data
modelling. In addition, tailored adjustments, and more sophisticated
manipulations have been implemented over time to allow estimation of
missing values in order to obtain customised, flexible, and more
homogeneous data. For more information on feature engineering, see
[@butcher2020] useful for effective machine learning strategy
application, covering various techniques and appropriate use cases,
focusing on practical understanding and implementation.

## Model Selection Strategies

In developing predictive models for health metrics and infectious
diseases, selecting the appropriate model is critical to ensure accurate
and reliable forecasts. Here are outlined sample strategies employed in
the model selection process, we introduce the **Rabies** dataset used
for our discussion and demonstrate the selection of a suitable model for
analysing its impact. Rabies, although nearly 100% fatal once symptoms
appear, presents a unique challenge due to the relative rarity of cases
and limited availability of comprehensive data. This scarcity
complicates efforts to model the disease accurately and develop
effective public health strategies.

To address these challenges, we explore advanced modelling techniques
that can enhance the robustness of our analyses despite data
limitations, which involves evaluating multiple models based on their
performance and selecting the best-fitting models to achieve the most
accurate predictions.

## Example: Rabies

The `rabies`\index{rabies} dataset from the `{hmsidwR}` package contains
information on death rates and disability-adjusted life years (DALYs)
per 100,000 inhabitants due to rabies and all causes of mortality in Asia
and for the Global region from 1990 to 2019. Rabies ([@cdc2024]) is a
fatal viral infection, and it is also classified as an infectious
disease that can infect all mammals causing acute encephalitis. Caused
by the rabies virus, which belongs to the Lyssavirus genus, it is
transmitted to humans through the bite of an infected animal such as
bats, raccoons, skunks, foxes, and obviously dogs, which are the main
source of human rabies deaths [@hampson2015]. Rabies defined as
**neglected tropical disease (NTD)** predominantly affects already
marginalised, poor and vulnerable populations. Although effective human
vaccines and immunoglobulins exist for rabies, these are often not
readily available or accessible to everyone [@rabies].

In this example we consider the number of DALYs per 100,000 inhabitants
due to rabies in Asia and the Global region, as our response variable,
the dataset is made available in the `{hmsidwR}` package. It is composed
of 240 observations and 7 variables: `measure`, `location`, `cause`,
`year`, `val`, `upper`, `lower`.

```{r}
library(tidyverse)
hmsidwR::rabies %>%
  select(-upper, -lower) %>%
  head()
```

Selecting only the `cause == Rabies` , the first thing to notice is that
deaths rates and DALYs are on different units, rates and years
respectively.

```{r}
library(tidyverse)
rabies <- hmsidwR::rabies %>%
  select(-upper, -lower) %>%
  pivot_wider(names_from = measure, values_from = val) %>%
  filter(cause == "Rabies") %>%
  rename(dx_rabies = Deaths, dalys_rabies = DALYs) %>%
  select(-cause)

rabies %>% head()
```

```{r}
#| echo: false
#| eval: false
breaks <- seq(min(rabies$year), max(rabies$year), by = 10)

rabies %>%
  mutate(year_10 = cut(year,
                       breaks = breaks,
                       right = FALSE,
                       labels = paste(breaks[-length(breaks)],
                                      breaks[-1] - 1,
                                      sep = "-")),
         year_10 = case_when(is.na(year_10) ~ "2019",
                             TRUE ~ as.factor(year_10))) %>%
  group_by(location, year_10) %>%
  summarize(avg_10yrs_dalys = mean(dalys_rabies))
```

It can be seen that the number of deaths due to rabies is much lower
than the number of DALYs. This difference in scale can affect the
model's ability to learn from the data. To address this issue, we can
scale and centre the numeric variables to make them more comparable.

```{r}
p1 <- rabies %>%
  ggplot(aes(x = year, group = location, linetype = location)) +
  geom_line(aes(y = dx_rabies),
            linewidth = 1) +
  geom_line(aes(y = dalys_rabies))

p2 <- rabies %>%
  # apply a scale transformation to the numeric variables
  mutate(across(where(is.double), scale)) %>%
  ggplot(aes(x = year, group = location, linetype = location)) +
  geom_line(aes(y = dx_rabies),
            linewidth = 1) +
  geom_line(aes(y = dalys_rabies))
```

```{r}
#| layout-ncol: 2
#| label: fig-rabies-deaths-dalys
#| fig-cap: "Not Scaled and Scaled and Centred"
#| fig-subcap:
#|   - "Not scaled"
#|   - "Scaled and centred"
#| fig-alt: "DALYs due to Rabies - Not scaled and Scaled and centred"
#| echo: false

legend_grob <- cowplot::get_legend(
  ggplot(data = rabies %>%
           pivot_longer(cols = c("dx_rabies", "dalys_rabies")) %>%
           mutate(name = ifelse(name == "dx_rabies", "Deaths", "DALYs")),
         aes(x = year, group = name, linewidth = name)) +
    geom_line(aes(y = value),
              color = "grey") +
    labs(linewidth = ""))

p1 +
  labs(
    title = "Dalys and Deaths due to Rabies - Not scaled",
    subtitle = "in Asia and Global Region",
    color = "Location",
    y = "Value", x = "Time(Year)") +
  scale_color_manual(values = c("orange", "navy"))

p2 +
  labs(
    title = "Dalys and Deaths due to Rabies - Scaled and centred",
    subtitle = "in Asia and Global Region",
    color = "Location",
    y = "Value", x = "Time(Year)") +
  scale_color_manual(values = c("orange", "navy")) +
  annotation_custom(legend_grob,
    xmin = 2010, xmax = 2015,
    ymin = 1, ymax = 2)
```

Creating new features from existing ones provide additional predictive
power. Then, combine the cause vector in a way to obtain two vectors for
death rates due to rabies and all causes, scale and centre the numeric
variables to obtain homogeneous data to use in the model.

```{r}
all_causes <- hmsidwR::rabies %>%
  select(-upper, -lower) %>%
  pivot_wider(names_from = measure, values_from = val) %>%
  filter(!cause == "Rabies") %>%
  rename(dx_allcauses = Deaths, dalys_allcauses = DALYs) %>%
  select(-cause)

dat <- rabies %>%
  full_join(all_causes, by = c("location", "year"))

dat %>% head()
```

To be able to visualise the magnitude of difference between death rates
and DALYs for both rabies and all causes, it is necessary to scale or
standardise the data as shown above.

```{r}
p3 <- dat %>%
  select(-year, -location) %>%
  scale() %>%
  cbind(dat %>% select(year, location)) %>%
  ggplot(aes(x = year, 
             group = location, 
             linetype = location)) +
  geom_line(aes(y = dx_rabies),
            linewidth = 1) +
  geom_line(aes(y = dx_allcauses))

p4 <- dat %>%
  select(-year, -location) %>%
  scale() %>%
  cbind(dat %>% select(year, location)) %>%
  ggplot(aes(x = year,
             group = location,
             linetype = location)) +
  geom_line(aes(y = dalys_rabies),
            linewidth = 1) +
  geom_line(aes(y = dalys_allcauses))
```

```{r}
#| layout-ncol: 2
#| label: fig-deaths-dalys-std
#| fig-cap: "Scaled and centred"
#| fig-subcap:
#|  - "Deaths due to Rabies and All Causes"
#|  - "Dalys due to Rabies and All Causes"
#| fig-alt: "Deaths and DALYs due to Rabies and All Causes - Scaled and centred"
#| echo: false

legend_grob2 <- cowplot::get_legend(
  ggplot(data = dat %>%
           select(-year, -location) %>%
           scale() %>%
           cbind(dat %>% select(year, location)) %>%
           pivot_longer(cols = c("dx_rabies", "dx_allcauses")) %>%
           mutate(name = ifelse(name == "dx_rabies", "Rabies", "All Causes")),
         aes(x = year, group = name, linewidth = name)) +
    geom_line(aes(y = value), color = "grey") +
    labs(linewidth = ""))

p3 +
  labs(
    title = "Deaths due to Rabies and All Causes",
    subtitle = "in Asia and Global Region",
    color = "Location",
    y = "Value", x = "Time(Year)") +
  scale_color_manual(values = c("orange", "navy")) +
  annotation_custom(legend_grob2,
    xmin = 2010, xmax = 2015,
    ymin = 1, ymax = 2)

p4 +
  labs(
    title = "Dalys due to Rabies and All Causes",
    subtitle = "in Asia and Global Region",
    color = "Location",
    y = "Value", x = "Time(Year)") +
  scale_color_manual(values = c("orange", "navy")) +
  annotation_custom(legend_grob2,
    xmin = 2010, xmax = 2015,
    ymin = 1, ymax = 2)
```

For this task, we will use the `{tidymodels}`\index{tidymodels}
meta-package, as it provides a consistent interface for modelling and
machine learning tasks. In particular, we define and execute modelling
workflows, to create tailored data pre-processing tasks on various
modelling specifications, and evaluate the performance using resampling
techniques, to eventually select the best model. A more detailed
explanation of the `{tidymodels}` framework can be found in the book
[@silge].

### Training Data and Resampling

Splitting data into training and test allows the model to train a
subsection of the data and then test its performance on the
remaining part of the data, the test set. In this case, we will use the
`initial_split()` function to split the data into training and test
sets. The proportion assigned to trains can vary but it is usually assigned
to be 80%, also a stratification option can be set.

```{r}
library(tidymodels)

set.seed(04042024)
split <- initial_split(dat, prop = 0.8)
training <- training(split)
test <- testing(split)
```

After that, it is important to create a set of folds, which means a set
of subgroups of the original data by grouping following specific
directions based on the type of **resampling
technique**\index{resampling}. Resampling techniques are used to
evaluate the model's performance and estimate its generalisation error.
There are various types of resampling techniques, it depends on the
specific characteristics of your dataset, and the goals of your
analysis. Some of the most common resampling techniques include:

-   **k-Fold Cross-Validation**\index{cross validation} for general
    model evaluation and hyperparameter tuning.
-   **Bootstrap Resampling**\index{bootstrap resampling} to estimate the
    variability of your model and for smaller datasets.
-   **Time Series Cross-Validation**\index{time series cross validation}
    for time-dependent data to preserve temporal structure.
-   **Spatial Resampling**\index{spatial resampling} for spatially
    correlated data to account for spatial dependencies.
-   **Stratified Resampling**\index{stratified resampling} when dealing
    with imbalanced datasets to ensure proper representation of all
    classes.

In this case, we will use **k-Fold Cross-Validation** to evaluate the
model's performance. The `vfold_cv()` function creates a set of folds
for cross-validation, which is used to train and test the model on
different subsets of the data.

```{r}
set.seed(04052024)
folds <- vfold_cv(training, v = 10)
```

### Data Preprocessing and Featuring Engineering

As already seen in the exploratory phase, preprocessing data is a
crucial step in machine learning. This process can include techniques
for handling missing values, standardisation of the data, encoding
categorical variables, and removing highly correlated variables.

In this case, we will use the `{recipes}` package to create a recipe,
with a set of preprocessing steps. The `recipe()` function allows us to
define a model formula and use various `step_<functions>()`\` for
manipulating data. We are going to set up 3 recipes, the first is a
basic one which includes all variables and does not perform any data
transformation.

```{r}
rec <- recipe(dalys_rabies ~ ., data = training)
```

The second recipe includes some key steps to transform the data into a
way specific models would be able to understand and learn from it.
Models such as k-nearest neighbours, or support vector machines, that
rely on distance metrics, can be sensitive to differences in feature
scales.

For instance, non-standardised year data can dominate the model's
decision-making process, leading to biased results. By scaling and
centring the data, we ensure that all features contribute equally to
the model's predictions.

We can create more complex recipes with more steps, but for this
example, we will use a step for encoding the location variable (Asia,
Global) into a numeric vector, and a second step to normalise (or
standardise) all predictors.

```{r}
rec1 <- recipe(dalys_rabies ~ ., data = training) %>%
  # convert nominal variables to dummy variables
  step_dummy(all_nominal_predictors()) %>%
  # scale the numeric variables
  step_normalize(all_numeric_predictors())
```

Once the recipe is created, we can apply it to the data using the
`prep()` function, which estimates the necessary parameters for the
transformations and applies them to the data. Then, to check the results
we can use the `juice()` function to extract the processed data.

```{r}
rec1 %>%
  prep() %>%
  juice() %>%
  select(1, 2, 5) %>%
  head()
```

Trained data can be also tested on new data, in this case we test them
on the `test`\`set with the `bake()` function.

```{r}
rec1 %>%
  prep() %>%
  bake(new_data = test) %>%
  select(1, 2, 5) %>%
  head()
```

DALYs often aggregate various health impacts, and can have highly skewed
distributions. This skewness arises due to several factors: the presence
of outliers, the nature of the health condition being measured, and the
distribution of the data itself. To handle the skewness of the data, we
can apply:

-   Log Transformation: $log(DALYs+1)$
-   Sqrt Transformation: $\sqrt{DALYs}$
-   Yeo-Johnson Transformation, a generalisation of the Box-Cox
    transformation that can handle both positive and negative values:
    $((DALYs+1)^p-1)/p$.

Let's apply the **Yeo-Johnson
transformations**\index{yeo-johnson transformations} to the response
variable (`dalys_rabies)` and see how the density distribution changes
with different values of $\lambda$. This is a step that can be tuned
with a machine learning algorithm.

```{r}
#| layout-ncol: 3
#| label: fig-response-transformation
#| fig-cap: "Response variable transformation"
#| fig-subcap: 
#|  - "Log10"
#|  - "Yeo-Johnson p=-2"
#| fig-alt: "Log10, Yeo-Johnson p=-2, Yeo-Johnson p=2"
#| echo: false

ggplot(dat) +
  geom_density(aes(x = dalys_rabies)) +
  scale_x_log10() +
  labs(title = "Log10 transformation")

ggplot(dat) +
  geom_density(aes(x = dalys_rabies)) +
  scale_x_continuous(transform = scales::transform_yj(p = -2)) +
  labs(title = "Yeo-Johnson transformation p=-2")

ggplot(dat) +
  geom_density(aes(x = dalys_rabies)) +
  scale_x_continuous(transform = scales::transform_yj(p = 2)) +
  labs(title = "Yeo-Johnson transformation p=2")
```

Let's now create a third recipe with the `step_YeoJohnson()` function.

```{r}
rec2 <- rec1 %>%
  # apply Yeo-Johnson transformation to the response variable
  step_YeoJohnson(dalys_rabies)

rec2 %>%
  prep() %>%
  juice() %>%
  select(1, 2, 5) %>%
  head()
```

### Correlation, Multicollinearity and Overfitting

To be noted is that we haven't applied any **correlation** selection
step on this data. Filtering out highly correlated predictors, such as
those with a correlation greater than 80% to avoid multicollinearity,
would lead to excluding crucial variables. On the other hand, including
all possible covariates in a model often yields implausible signs on
covariates or unstable coefficients, as well as overfitting
[@foreman2012].

When multiple predictors are correlated, but all are crucial for the
analysis (e.g., deaths due to rabies, total deaths, and total DALYs for
all causes), applying a correlation step that filters out correlated
variables can be problematic. One way to overcome bias arising from it
is using regularisation techniques like **Ridge Regression** or **Lasso
Regression** is often the best approach to handle multicollinearity
without removing any predictors. Alternatively, **PCA**\index{PCA} can
reduce dimensionality while retaining most of the variance. These
methods ensure all important predictors are included in the model
without the adverse effects of multicollinearity.

### Model Specification

The next step is to outline the model specification. There are various
type of models that can be used. We start with a **random forest**. This
choice is typically done due to the algorithm's features, which is able
to create **multiple bootstrap samples** (random samples with
replacement) from the original dataset. Each bootstrap sample is used to
train a separate decision tree.

### Model 1: Random Forest

Rabies death rates may exhibit complex relationships with predictor
variables. Random forests\index{random forests} are capable of capturing
non-linear relationships between predictors and the target variable.

Also, it handles multicollinearity, missing data, provides variables
importance and is an ensemble learning method, which means they combine
the predictions of multiple individual decision trees to produce a more
accurate and stable prediction.

In our simplified case this type of model will do random samples with
replacement of data. In `{tidymodels}` we can select different types of
engines, in the case of random forest we could use random forest,
ranger, and others. The difference between these engines derives from
the specific type of calculation used to make the estimation. The Ranger
engine is notably faster than random forest, so let's use that for this
example.

```{r}
rf_mod <- rand_forest(mtry = tune(),
                      trees = tune(),
                      min_n = tune(),
                      mode = "regression",
                      engine = "ranger")

wkf <- workflow(preprocessor = rec, 
                spec = rf_mod)

rf_res <- tune_grid(object = wkf,
                    resamples = folds,
                    grid = 5,
                    control = control_grid(save_pred = TRUE))

show_best(rf_res, metric = "rmse") %>%
  select(-n, -std_err)
```

```{r}
rf_res_tuned <- select_best(rf_res, metric = "rmse")

rf_res_tuned
```

```{r}
rf_fit <- wkf %>%
  finalize_workflow(select_best(rf_res,
                                metric = "rmse")) %>%
  fit(training)

rf_fit %>%
  predict(new_data = test) %>%
  bind_cols(test) %>%
  rmse(truth = dalys_rabies, estimate = .pred)
```

```{r}
#| layout-ncol: 2
#| label: fig-rf-predictions
#| fig-cap: "Predictions vs. Truth"
#| fig-subcap:
#|   - "Predictions vs. Truth"
#|   - "Predictions vs. Truth by Year"
#| fig-alt: "Predictions vs. Truth and Predictions vs. Truth by Year"   
#| echo: false

rf_fit %>%
  predict(new_data = test) %>%
  bind_cols(test) %>%
  ggplot(aes(dalys_rabies, .pred,
             color = location)) +
  geom_abline(color = "brown") +
  geom_point(size = 2) +
  scale_color_manual(values = c("orange", "navy")) +
  coord_fixed() +
  expand_limits(x = 0, y = 0) +
  labs(
    title = "Predictions vs. Truth",
    x = "Truth",
    y = "Predictions")

rf_fit %>%
  predict(new_data = test) %>%
  bind_cols(test) %>%
  ggplot(aes(x = year, 
             color = location)) +
  geom_point(aes(y = dalys_rabies)) +
  geom_line(aes(y = .pred)) +
  scale_color_manual(values = c("orange", "navy")) +
  labs(
    title = "Predictions vs. Truth by Year",
    x = "Time(Year)",
    y = "Values")
```

### Model 2: Generalised Linear Model (GLM)

Generalised Linear Models (GLMs)\index{GLM} involve statistical
estimation rather than the iterative parameter tuning, common in many
machine learning techniques. However, adding a machine learning feature
through parameter calibration can be done using techniques such as
cross-validation and grid search to find the best model settings.

To introduce a machine learning feature with parameter calibration into
our modelling of the rabies data, we can use a technique like
cross-validation combined with a **regularisation method** or an
algorithm that supports parameter tuning. Here, we can employ a model
from the `glmnet`\index{glmnet} package, which fits a generalised linear
model via **penalised maximum likelihood**. The regularisation path is
computed for the lasso or elastic-net penalty at a grid of values for
the regularisation parameter lambda.

Adding Machine Learning Features with `{glmnet}` and Cross-Validation

```{r}
if (!require(glmnet)) install.packages("glmnet")
library(glmnet)
```

For `glmnet`, we need to input matrices rather than data frames, and
create matrices for the independent variables (predictors) and the
dependent variable (response).

```{r}
predictors <- model.matrix(dalys_rabies ~ .,
                           data = dat)[, -1] # Remove intercept
response <- dat$dalys_rabies
```

Use cross-validation to find the optimal lambda value, which controls
the strength of the regularisation:

```{r}
# Set seed for reproducibility
set.seed(123)

# Fit the model with cross-validation
cv_model <- cv.glmnet(predictors,
                      response,
                      family = "gaussian")
cv_model
```

```{r}
#| label: fig-glmnet-predictions
#| fig-cap: "Cross-Validation Optimal Lambda"
#| fig-alt: "Lambda Selection"
# Plot the lambda selection
plot(cv_model)
```

Extracting the best model, we can see that $\lambda$ is 0.165.

```{r}
# Get the best lambda value
best_lambda <- cv_model$lambda.min
print(paste("Best Lambda:", best_lambda))
```

Then, fitting the final model with the selected best lambda, we can
predict and evaluate the model.

```{r}
final_model <- glmnet(predictors,
                      response,
                      family = "gaussian",
                      lambda = best_lambda)

# Predict using the final model
predictions <- predict(final_model,
                       # values of the penalty parameter lambda
                       s = best_lambda,
                       # matrix of new values for x
                       newx = predictors
                       )

# Calculate Mean Squared Error
rmse <- sqrt(mean((response - predictions)^2))
print(paste("Root Mean Squared Error:", rmse))
```

By incorporating `glmnet` and using lambda selection via
cross-validation, we introduce a **machine learning feature—parameter
calibration—into** our analysis. This approach not only helps in
minimising overfitting but also enhances model performance by selecting
the most effective regularisation parameter. The cross-validation
process used here is crucial for confirming that our model's parameters
are optimally tuned for the given data, embodying a key aspect of
machine learning methodologies.

```{r}
#| echo: false
#| layout-ncol: 2
#| label: fig-glmnet-predictions2
#| fig-cap: "Predictions vs. Truth"
#| fig-subcap: 
#|  - "Predictions vs. Truth"
#|  - "Predictions vs. Truth by Year"
#| fig-alt: "Predictions vs. Truth and Predictions vs. Truth by Year"

# plot the predictions
data.frame(dalys_rabies = response, predictors, predictions) %>%
  ggplot(aes(x = dalys_rabies, y = predictions)) +
  geom_point() +
  geom_abline(color = "brown") +
  labs(
    title = "Predictions vs Truth",
    x = "Truth",
    y = "Predictions")

data.frame(dalys_rabies = response, predictors, predictions) %>%
  ggplot(aes(x = year, 
             group = factor(locationGlobal),
             color = factor(locationGlobal))) +
  geom_point(aes(y = dalys_rabies)) +
  geom_line(aes(y = predictions), color = "brown") +
  scale_color_manual(values = c("orange", "navy"),
                     labels = c("Asia", "Global")) +
  labs(
    title = "DALYs due to Rabies Predictions vs Truth",
    subtitle = "by Year",
    x = "Truth",
    y = "Predictions")
```

### Testing Multiple Models

In the example above, we used two models to predict DALYs due to rabies,
a random forest with `{tidymodels}` and a generalised linear model with
`{glmnet}` with a Root Mean-Square Error of 0.448 and 0.257
respectively. The Random Forest model has a higher RMSE, which means it
has a higher prediction error compared to the GLM model. However, we
haven't applied any of the preprocessing steps, and there are many other
models that could be used to predict DALYs such as:

1.  **Support Vector Machines (SVM)**: SVMs are a powerful machine
    learning algorithm that can be used for both classification and
    regression tasks. They work by finding the hyperplane that best
    separates the data into different classes or groups.

2.  **Extreme Gradient Boosting (XGBoost)**\index{XGBoost}: Known for
    its high performance in various prediction tasks, XGBoost can handle
    missing values and is effective for large datasets.

3.  **K-Nearest Neighbours (KNN)**\index{KNN} models are a type of
    instance-based learning algorithm that stores all available cases
    and classifies new cases based on a similarity measure.

4.  **Long Short-Term Memory (LSTM) Networks**\index{LSTM}: For temporal
    or sequential health data, LSTM networks can capture dependencies
    over time, making them suitable for time-series prediction of
    disease progression and outcomes.

Each of these models has its own strengths and weaknesses, and the
choice of model will depend on the specific characteristics of the data
and the goals of the analysis. By testing multiple models and comparing
their performance, we can identify the best model for the given data and
task.

Let's use the `{parsnip}` package and the `workflow_set()` function to
fit a set of models to the rabies data. We will fit a **Support Vector
Machine (SVM)**, and a **K-Nearest neighbours (KNN) model** to the data
and compare their performance.

```{r}
linear_reg_spec <- 
  linear_reg(penalty = tune(),
             mixture = tune()) %>%
  set_engine("glmnet")

svm_p_spec <-
  svm_poly(cost = tune(),
           degree = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("regression")

knn_spec <-
  nearest_neighbor(neighbor = tune(),
                   dist_power = tune(),
                   weight_func = tune()) %>%
  set_engine("kknn") %>%
  set_mode("regression")
```

```{r}
#| eval: false
library(rules)
library(baguette)
# Combine workflows into a workflow set
workflow_set <- workflow_set(preproc = list(scaled = rec1,
                                            yeo_johnson = rec2),
                             models = list(linear_reg = linear_reg_spec,
                                           svm = svm_p_spec,
                                           knn = knn_spec))

grid_ctrl <-control_grid(save_pred = TRUE,
                         parallel_over = "everything",
                         save_workflow = TRUE)

# Fit and evaluate the models with hyperparameter tuning
grid_results <- workflow_set %>%
  workflow_map(seed = 1503,
               resamples = folds,
               grid = 5,
               control = grid_ctrl)
```

```{r}
#| eval: true
#| echo: false
# saveRDS(grid_results, file = "data/model-data/grid_results.rds")
grid_results <- readRDS("data/model-data/grid_results.rds")
```

```{r}
# Show the results
grid_results %>%
  collect_metrics() %>%
  arrange(mean) %>%
  select(1, 5, 7, 9) %>%
  head()
```

```{r}
#| fig-cap: "Model Performance"
#| fig-alt: "Model Performance"
#| label: fig-model-performance
autoplot(grid_results,
         rank_metric = "rmse",
         metric = "rmse",
         select_best = TRUE) +
  geom_text(aes(y = mean - 0.1, 
                label = wflow_id),
            angle = 90, 
            hjust = 1, 
            color = "black",
            size = 3.5) +
  lims(y = c(-1.5, 0.9)) +
  theme(legend.position = "none")
```

## Summary

The integration of machine learning techniques into public health data
analysis can significantly enhance the predictive power and robustness
of models. By leveraging the capabilities of machine learning
algorithms, we can extract valuable insights from complex health data,
enabling more informed decision-making and policy formulation in public
health contexts. The examples provided in this chapter illustrate the
application of machine learning techniques to health metrics data,
demonstrating the importance of feature engineering, model selection,
and parameter calibration in enhancing the predictive accuracy and
relevance of models. By following best practices in machine learning,
public health researchers and practitioners can harness the power of
data-driven insights to address critical health challenges and improve
population health outcomes.

**Best Practices for Machine Learning in Public Health:**

-   Conduct exploratory data analysis to understand the underlying
    structure of the data and relationships between variables.
-   Apply feature engineering techniques to create new variables and
    enhance the model's predictive power.
-   Select machine learning models that are contextually appropriate and
    robust for public health data analysis. Such as Random Forest,
    Generalised Linear Models, and others.
-   Use parameter calibration techniques such as cross-validation,
    regularisation, monte carlo, and grid search to optimise model
    performance.
- Evaluate model
performance using appropriate metrics and visualisation tools to assess
predictive accuracy and relevance.

The integration of machine learning methodologies into
public health data analysis represents a significant opportunity to
advance the field of public health and enhance our understanding of
health metrics and disease dynamics.
