# Techniques for Machine Learning Applications {#sec-06-techniques}

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   How to manipulate data through feature engineering
-   Select the most suitable model for your data
-   Learn about machine learning algorithms
:::
:::

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

Deciding on the best strategy for selecting and applying a specific
machine learning model technique can be challenging. Best practices
suggest a thorough understanding of the phenomenon to be modeled,
followed by a data manipulation technique known as *feature engineering*
before applying data to a specific model type. This initial step will
allow the model to run more smoothly and quickly. As a critical stage in
the preparation of machine learning data, feature engineering allows
algorithms to perform better by making the data more amenable to
processing. It involves creating new features from existing data based
on domain knowledge or transforming data to improve the model's ability
to discern patterns. For example, creating features like **moving
averages** or differences between consecutive days can reveal trends and
cycles that are not immediately apparent from raw data.

A another example is the **standardization** process, which is crucial
when dealing with variables measured in different units. It involves
rescaling the features so they have a mean of zero and a standard
deviation of one. This process is particularly important when variables
span several orders of magnitude; without standardization, a model might
incorrectly interpret the scale of a feature as a proxy for importance.

Furthermore, the application of transformations, such as **logarithmic**
scaling or the application of **spline** functions can help in managing
skewed data or enhancing model ability to capture non-linear
relationships, which result particularly useful in complex data
modelings. In addition, tailored adjustments, and more sophisticated
manipulations have been implemented over time to allow estimation of
missing values in order to obtain customized, flexible, and more
homogeneous data. For more information on feature engineering, see
[@butcher2020] useful for effective machine learning strategy
application, covering various techniques and appropriate use cases,
focusing on practical understanding and implementation.

## **Model selection strategies**

When applying machine learning techniques to health metrics and the
study of infectious diseases, it is crucial to select approaches that
are not only robust but also contextually appropriate. Available machine
learning models can greatly influence the outcomes of analyses, and
therefore, our focus in this book will be on those models that are most
pertinent to public health investigations.

The first step, once data is collected and prepared for analysis,
involves conducting a thorough exploratory data analysis (EDA). This
initial exploration helps to uncover the underlying structure of the
data, the relationships between variables, and the way the response
variable—which may also be referred to as the outcome variable—depends
on predictors. This phase is critical as it informs the necessity of
subsequent data adjustments and transformations.

Feature engineering may follow the exploratory phase. This process
involves creating new variables from existing ones to enhance the
machine learning models' ability to discern patterns and make accurate
predictions. Decisions on feature engineering should be based on
insights gained during EDA, with a focus on addressing specific
challenges posed by the complexity of health data, such as
non-linearity, high dimensionality, or the presence of interaction
effects among variables.

By strategically deploying machine learning techniques tailored to
specific data, predictive accuracy and relevancy of models can improve,
thereby enhancing their utility in public health decision-making and
policy formulation.

### Example 4: Rabies

The `rabies` dataset from the `hmsidwR` package contains information on
deaths and disability-adjusted life years (DALYs) due to rabies and
other causes in different regions of the world from 1990 to 2019.

Consider only Asia and the Global region, the number of DALYs due to
both rabies and all cause mortality. The first thing to notice is that
deaths rates and DALYs are on different units, rates and years
respectively.

```{r}
library(tidyverse)
rabies <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)
rabies %>% head
```

Creating new features from existing ones provide additional predictive
power. In this case we have split the cause vector into two indicating
deaths due to both causes separately. Also, we can scale and center the
numeric variables so to have homogeneous data to use in the model, this
is useful in particular when a transformation is applied. We could use
splines, for example.

```{r}
all_causes <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(!cause=="Rabies")%>%
  rename(dx_allcauses=Deaths,dalys_allcauses=DALYs)%>%
  select(-cause)

dat <- rabies %>%
  full_join(all_causes,by=c("location","year"))

dat %>% head()
```

This is the data we have obtained, and to be able to visualize the
magnitude of difference it is necessary to scale or standardiza the
data.

```{r}
dat%>%
  select(-year,-location)%>%
  scale()%>%
  cbind(dat%>%select(year,location))%>%
  ggplot(aes(x=year,group=location,color=location))+
  geom_line(aes(y=dx_rabies),
            linewidth=1)+
  #geom_line(aes(y=dx_allcauses))+
  geom_line(aes(y=dalys_rabies))
  #geom_line(aes(y=dalys_allcauses))
```

```{r}
suppressPackageStartupMessages({library(tidymodels)})
set.seed(04042024)
split <- initial_split(dat,prop = 0.8)
training <- training(split)
test <- testing(split)
folds <- vfold_cv(training)
```

Then create a recipe with some key steps to trasnform the data into a
model digestable shape.

```{r}
rec <- recipe(dalys_rabies ~. , data=training)%>%
  step_dummy(all_nominal_predictors())%>%
  step_scale(all_predictors())%>%
  step_center(all_predictors())%>%
  step_corr(threshold = 0.8)%>%
  step_YeoJohnson(all_predictors())

rec %>%
  prep()%>%juice()%>%head
```

## Random Forest

There are various type of modeling that can be used, let's get started
with a random forest, this choice is due to this type of algorithm which
will create multiple bootstrap samples (random samples with replacement)
from the original dataset. Each bootstrap sample is used to train a
separate decision tree.

Rabies death rates may exhibit complex relationships with predictor
variables. Random forests are capable of capturing non-linear
relationships between predictors and the target variable.

Also, it handles missing data well, provides variables importance and
are an ensemble learning method, which means they combine the
predictions of multiple individual decision trees to produce a more
accurate and stable prediction.

In our simplified case this type of model will do random samples with
replacement of data. In {tidymodels} we can select different types of
engines, in the case of random forest we could use randomForest, ranger,
and others. The difference between these engines derives from the
specific type of calculation used to make the estimation. The Ranger
engine is notably faster than randomForest, so let's use that for this
example.

```{r}
rf_mod <- rand_forest(mtry = tune(),
                      trees = tune(),
                      min_n = tune(),
                      mode = "regression",
                      engine = "ranger")
```

```{r}
wkf <- workflow(preprocessor = rec,
                spec = rf_mod)

rf_res <- tune_grid(object = wkf,
          resamples = folds,
          grid = 5,
          control = control_grid(save_pred = TRUE))
```

```{r}
#rf_res%>%unnest(.metrics)
show_best(rf_res, metric = "rmse")
```

```{r}
autoplot(rf_res, metric = "rmse") +
  scale_x_log10()
```

```{r}
rf_res_tuned <- select_best(rf_res, metric = "rmse")
```

```{r}
rf_fit <- wkf %>% 
  finalize_workflow(select_best(rf_res, 
                                metric = "rmse")) %>%
  fit(training)
```

```{r}
rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test) %>% 
   rmse(truth = dalys_rabies, estimate = .pred)
```

```{r}
rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test)  %>%
  ggplot(aes(dalys_rabies, 
              .pred, 
              color = location)) +
   geom_point() +
   geom_abline(color = "red") +
   scale_color_brewer(type = "qual") +
   coord_fixed() +
   expand_limits(x = 0, y = 0)
```
