# Techniques for Machine Learning Applications {#sec-06-techniques}

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   How to manipulate data through feature engineering
-   Select the most suitable model for your data
-   Learn about machine learning algorithms
:::
:::

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

Selecting the most suitable machine learning model involves
understanding **the goals of the analysis**, the **nature of the data**,
and the **statistical and machine learning methods** that best suit the
tasks. In @sec-05-machine_learning, we learned about what machine
learning models are, provided examples for building a model framework,
and selected common metrics for model performance calibration and
evaluation.

In this chapter, we focus on the strategies for selecting appropriate
models by leveraging the strengths of different techniques, specifically
for health metrics and for infectious diseases. We will explore various
considerations involved in addressing potential biases, and discuss
actions to prevent them.

## Goals of the analysis and nature of data

The identification of the primary goal of the analysis is fundamental.
Whether it involves trend analysis, investigating the relationships
between response and predictor variables, or strictly forecasting to
predict future outcomes, the strategy for model selection varies
accordingly.

**Health Metrics Data:**

-   **Composite Measures:** Health metrics like DALYs are composite
    measures that include both mortality and morbidity data, often
    requiring sophisticated regression models capable of handling
    continuous variables and multiple predictors. By examining the
    components of DALYs (e.g., Years of Life Lost (YLLs) and Years Lived
    with Disability (YLDs)), we can identify the key drivers such as
    mortality rates, disease prevalence, and risk factors.

-   **Regression Models:** Regression models, including **linear
    regression**, **Ridge regression**, and **Lasso regression**, are
    commonly used to handle these continuous variables and address
    challenges like correlation and multicollinearity with appropriate
    techniques such as regularization.

**Infectious Disease Data:**

-   **Categorical and Continuous Data:** Infectious disease data can be
    categorical (e.g., disease presence or absence) or continuous (e.g.,
    incidence rates). Classification models are suitable for categorical
    outcomes, while regression models are appropriate for continuous
    data.

-   **Disease Dynamics:** Understanding the dynamics of infectious
    diseases, such as transmission rates, incubation periods, and
    immunity, informs the selection of models. Common models include
    compartmental models (e.g., SIR, SEIR) and agent-based models.

**Common considerations for health metrics and infectious diseases data
type:**

-   **Seasonality and Trends:** The data may exhibit seasonality or
    trends, necessitating the use of time series analysis models like
    **ARIMA** or seasonal decomposition to capture these patterns.

-   **Simulation Models:** These models can predict the impact of
    interventions on DALYs and infectious diseases, estimating the
    effectiveness of different interventions and guiding policy
    decisions. Examples of these types of models are: **SIR** models,
    and **Agent-based models**. In addition, **confidence intervals**
    and **sensitivity analyses** help assess the uncertainty associated
    with these predictions.

-   **Bayesian Models:** These models can estimate parameters and make
    predictions based on prior knowledge and observed data,
    incorporating uncertainty and variability.

-   **Predictive Modeling:** Such as **decision trees**, **support
    vector machines (SVM)**, and **LSTM (Long Short-Term Memory) neural
    networks**, can predict disease outbreaks, identify high-risk
    populations, and optimize resource allocation.

## Statistical and machine learning methods

Once the objective of the study is clear and identificated the nature of
the data, we already a step forward in restricting the range of
applicable models. The next step involves conducting a thorough
**exploratory data analysis (EDA)**. This initial exploration helps to
uncover the underlying structure of the data, the relationships between
variables, and the way the response variable—which may also be referred
to as the outcome variable—depends on predictors. This phase is critical
as it informs the necessity of subsequent data adjustments and
transformations.

The importance of data preparation and exploratory data analysis in
machine learning are the building blocks in the preparation of machine
learning digestable data. **Feature engineering** is a technique that
involves creating new features from existing ones based on domain
knowledge or transformation of data to improve the model's ability to
discern patterns. For example, creating features like **moving
averages** or differences between consecutive days can reveal trends and
cycles that are not immediately apparent from raw data.

Another example is the **standardization** process, which is crucial
when dealing with variables measured in different units. It involves
rescaling the features so they have a mean of zero and a standard
deviation of one. This process is particularly important when variables
span several orders of magnitude; without standardization, a model might
incorrectly interpret the scale of a feature as a proxy for importance.

Furthermore, the application of transformations, such as **logarithmic**
scaling or the application of **spline** functions can help in managing
skewed data or enhancing model ability to capture non-linear
relationships, which result particularly useful in complex data
modelings. In addition, tailored adjustments, and more sophisticated
manipulations have been implemented over time to allow estimation of
missing values in order to obtain customized, flexible, and more
homogeneous data. For more information on feature engineering, see
[@butcher2020] useful for effective machine learning strategy
application, covering various techniques and appropriate use cases,
focusing on practical understanding and implementation.

## Model selection strategies

### Example 4: Rabies

The `rabies` dataset from the `{hmsidwR}` package contains information
on death rates and disability-adjusted life years (DALYs) due to rabies
and all causes mortality in Asia and for the Global region from 1990 to
2019. Rabies ([@cdc2024]) is a fatal viral infection, and it is also
classified as an infectious disease that can infect all mammals causing
acute encephalitis. Caused by the rabies virus, which belongs to the
Lyssavirus genus, it is transmitted to humans through the bite of an
infected animal such as bats, raccoons, skunks, foxes, and obviously
dogs, which are the main source of human rabies deaths [@hampson2015].
Rabies defined as neglected tropical disease (NTD) **predominantly
affects already marginalized, poor and vulnerable populations**.
Although effective human vaccines and immunoglobulins exist for rabies,
these are often not readily available or accessible to everyone
[@rabies].

In this example we consider the number of DALYs due to rabies in Asia
and the Global region, as our response variable, the dataset is made
available in the `{hmsidwR}` package. It is composed of 240 observations
and 7 variables: `measure`, `location`, `cause`, `year`, `val`, `upper`,
`lower`.

```{r}
library(tidyverse)
hmsidwR::rabies%>%
  select(-upper,-lower) %>%
  head()
```

Selecting only the `cause == Rabies` , the first thing to notice is that
deaths rates and DALYs are on different units, rates and years
respectively.

```{r}
library(tidyverse)
rabies <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)
rabies%>%head
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: false
breaks <- seq(min(rabies$year), max(rabies$year), by = 10)

rabies %>%
  mutate(year_10=cut(year, 
                     breaks= breaks,
                     right = FALSE,
                     labels = paste(breaks[-length(breaks)], 
                                    breaks[-1] - 1, 
                                    sep = "-")),
year_10=case_when(is.na(year_10)~"2019",
                  TRUE~as.factor(year_10)))%>%
  group_by(location,year_10)%>%
  summarize(avg_10yrs_dalys=mean(dalys_rabies))
```

It can be seen that the number of deaths due to rabies is much lower
than the number of DALYs. This difference in scale can affect the
model's ability to learn from the data. To address this issue, we can
scale and center the numeric variables to make them more comparable.

```{r}
#| message: false
#| warning: false

p1 <- rabies%>%
  ggplot(aes(x=year, group=location,color=location))+
  geom_line(aes(y=dx_rabies),
            linewidth=1)+
  geom_line(aes(y=dalys_rabies))
  
p2 <- rabies%>%
  # apply a scale transformation to the numeric variables
  mutate(across(where(is.double),scale))%>%
  ggplot(aes(x=year, group=location,color=location))+
  geom_line(aes(y=dx_rabies),
            linewidth=1)+
  geom_line(aes(y=dalys_rabies)) 
```

```{r}
#| layout-ncol: 2
#| label: fig-rabies-deaths-dalys
#| fig.cap: 
#|   - "Not scaled"
#|   - "Scaled and centered"
#| echo: false
#| message: false
#| warning: false

legend_grob <- cowplot::get_legend(
ggplot(data= rabies%>%
pivot_longer(cols=c("dx_rabies","dalys_rabies"))%>%
mutate(name=ifelse(name=="dx_rabies","Deaths","DALYs")),
aes(x=year, group=name,linewidth=name))+
geom_line(aes(y=value),color="grey")+labs(linewidth=""))


p1 + 
labs(title = "Dalys and Deaths due to Rabies - Not scaled",
subtitle = "in Asia and Global Region",
color = "Location",
y="Value",x="Time(Year)") +
scale_color_manual(values = c("orange", "navy"))

p2 +
labs(title = "Dalys and Deaths due to Rabies - Scaled and centered",
subtitle = "in Asia and Global Region",
color = "Location",
y="Value",x="Time(Year)") +
scale_color_manual(values = c("orange", "navy"))+
annotation_custom(legend_grob, xmin = 2010, xmax = 2015, ymin = 1, ymax = 2)
```

Creating new features from existing ones provide additional predictive
power. Then, combine the cause vector in a way to obtain two vectors for
death rates due to rabies and all causes, scale and center the numeric
variables to obtain homogeneous data to use in the model.

```{r}
all_causes <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(!cause=="Rabies")%>%
  rename(dx_allcauses=Deaths,dalys_allcauses=DALYs)%>%
  select(-cause)

dat <- rabies %>%
  full_join(all_causes,by=c("location","year"))

dat %>% head()
```

To be able to visualize the magnitude of difference between death rates
and DALYs for both rabies and all causes, it is necessary to scale or
standardize the data as shown above.

```{r}
#| message: false
#| warning: false
p3 <- dat%>%
  select(-year,-location)%>%
  scale()%>%
  cbind(dat%>%select(year,location))%>%
  ggplot(aes(x=year,group=location,color=location))+
  geom_line(aes(y=dx_rabies),
            linewidth=1)+
  geom_line(aes(y=dx_allcauses))

p4 <- dat%>%
  select(-year,-location)%>%
  scale()%>%
  cbind(dat%>%select(year,location))%>%
  ggplot(aes(x=year,
  group=location,
  color=location))+
  geom_line(aes(y=dalys_rabies),
            linewidth=1)+
  geom_line(aes(y=dalys_allcauses))
```

```{r}
#| layout-ncol: 2
#| label: fig-deaths-dalys-std
#| fig.cap: "Scaled and centered"
#| echo: false
#| message: false
#| warning: false

legend_grob2 <- cowplot::get_legend(
ggplot(data= dat%>%
  select(-year,-location)%>%
  scale()%>%
  cbind(dat%>%select(year,location))%>%
pivot_longer(cols=c("dx_rabies","dx_allcauses"))%>%
mutate(name=ifelse(name=="dx_rabies","Rabies","All Causes")),
aes(x=year, group=name,linewidth=name))+
geom_line(aes(y=value),color="grey")+labs(linewidth=""))

p3 + 
labs(title = "Deaths due to Rabies and All Causes",
subtitle = "in Asia and Global Region",
color = "Location",
y="Value",x="Time(Year)") +
scale_color_manual(values = c("orange", "navy"))+
annotation_custom(legend_grob2, xmin = 2010, xmax = 2015, ymin = 1, ymax = 2)

p4 +
labs(title = "Dalys due to Rabies and All Causes",
subtitle = "in Asia and Global Region",
color = "Location",
y="Value",x="Time(Year)") +
scale_color_manual(values = c("orange", "navy"))+
annotation_custom(legend_grob2, xmin = 2010, xmax = 2015, ymin = 1, ymax = 2)
```

For this task, we will use the `{tidymodels}` meta-package, as it
provides a consistent interface for modeling and machine learning tasks.
In particular, we define and execute modeling workflows, to create
tailored data pre-processing tasks on various modelling specification,
and evaluate the performance using resampling techniques, to eventually
select the best model. A more detailed explanation of the `{tidymodels}`
framework can be found in the book [@silge].

### Training Data and Resampling Techniques

Splitting data into training and test allows the model to train a
subsection of the data and then test the its performance on the
remaining part of the data, the test set. In this case, we will use the
`initial_split()` function to split the data into training and test
sets. The proportion assigned to train can vary but it usually assigned
to be 80%, also a stratification option can be set.

```{r}
#| message: false
#| warning: false
library(tidymodels)
set.seed(04042024)
split <- initial_split(dat, prop = 0.8)
training <- training(split)
test <- testing(split)
```

After that, it is important to create a set of folds, which means a set
of subgroups of the original data by grouping following specific
directions based on the type of **resampling technique**. Resampling
techniques are used to evaluate the model's performance and estimate its
generalization error. There are various types of resampling techniques,
it depends on the specific characteristics of your dataset, and the
goals of your analysis. Some of the most common resampling techniques
include:

-   **k-Fold Cross-Validation** for general model evaluation and
    hyperparameter tuning.
-   **Bootstrap Resampling** to estimate the variability of your model
    and for smaller datasets.
-   **Time Series Cross-Validation** for time-dependent data to preserve
    temporal structure.
-   **Spatial Resampling** for spatially correlated data to account for
    spatial dependencies.
-   **Stratified Resampling** when dealing with imbalanced datasets to
    ensure proper representation of all classes.

In this case, we will use **k-Fold Cross-Validation** to evaluate the
model's performance. The `vfold_cv()` function creates a set of folds
for cross-validation, which is used to train and test the model on
different subsets of the data.

```{r}
set.seed(04052024)
folds <- vfold_cv(training, v = 10)
```

### Data Preprocessing & Featuring Engineering

As already seen in the exploratory phase, preprocessing data is a
crucial step in machine learning, this process can include techniques
for handling missing values, standardization of the data, encoding
categorical variables, and removing highly correlated variables.

In this case, we will use the `{recipes}` package to create a recipe,
with a set of preprocessing steps. The `recipe()` function allows to
define a model formula and use various `step_<functions>()`\` for
manipulating data. We are going to set up 3 recipes, the first is a
basic one which include all variables and does not perform any data
transformation.

```{r}
rec <- recipe(dalys_rabies ~. , data=training)
```

The second recipe includes some key steps to transform the data into a
way specific models would be able to understand and learn from it.
Models such as k-nearest neighbors, or support vector machines, that
rely on distance metrics, can be sensitive to differences in feature
scales.

For instance, non-standardized year data, can dominate the model's
decision-making process, leading to biased results. By scaling and
centering the data, we ensure that all features contribute equally to
the model's predictions.

We can create more complex recipes with more steps, but for this
example, we will use a step for encoding the location variable (Asia,
Global) into a numeric vector, and a second step to normalize (or
standardize) all predictors.

```{r}
rec1 <- recipe(dalys_rabies ~. , data=training)%>%
  # convert nominal variables to dummy variables
  step_dummy(all_nominal_predictors())%>% 
  # scale the numeric variables
  step_normalize(all_numeric_predictors())
```

Once the recipe is created, we can apply it to the data using the
`prep()` function, which estimates the necessary parameters for the
transformations and apply them to the data. Then, to check the results
we can use the `juice()` function to extract the processed data.

```{r}
rec1 %>%
  prep()%>%
  juice()%>%
  select(1,2,5)%>%
  head()
```

Trained data can be also tested on new data, in this case we test them
on the `test`\`set with the `bake()` function.

```{r}
rec1 %>%
  prep()%>%
  bake(new_data = test) %>%
  select(1,2,5)%>%
  head()
```

DALYs often aggregate various health impacts, and can have highly skewed
distributions. This skewness arises due to several factors: the presence
of outliers, the nature of the health condition being measured, and the
distribution of the data itself. To handle the skewness of the data, we
can apply:

-   Log Transformation: $log(DALYs+1)$
-   Sqrt Transformation: $\sqrt{DALYs}$
-   Yeo-Johnson Transformation, a generalization of the Box-Cox
    transformation that can handle both positive and negative values:
    $((DALYs+1)^p-1)/p$.

Let's apply the **Yeo-Johnson transformations** to the response variable
(`dalys_rabies)` and see how the density distribution changes with
different values of $\lambda$. This is a step that can be tuned with a
machine learning algorithm.

```{r}
#| layout-ncol: 3
#| label: fig-response-transformation
#| fig.cap: "Response variable transformation"
#| echo: false
#| message: false
#| warning: false

ggplot(dat)+
geom_density(aes(x=dalys_rabies))+
scale_x_log10()+
labs(title = "Log10 transformation")

ggplot(dat)+
geom_density(aes(x=dalys_rabies))+
scale_x_continuous(transform = scales::transform_yj(p = -2))+
labs(title = "Yeo-Johnson transformation p=-2")

ggplot(dat)+
geom_density(aes(x=dalys_rabies))+
scale_x_continuous(transform = scales::transform_yj(p = 2))+
labs(title = "Yeo-Johnson transformation p=2")
```

Let's now create a third recipe with the `step_YeoJohnson()` function.

```{r}
rec2 <- rec1 %>%
  # apply Yeo-Johnson transformation to the response variable
  step_YeoJohnson(dalys_rabies)

rec2 %>%
  prep()%>%
  juice()%>%
  select(1,2,5)%>%
  head()
```

### Correlation, Multicollinearity and Overfitting

To be noted is that we haven't applied any **correlation** selection
step on this data. Filtering out highly correlated predictors, such as
those with a correlation greater than 80% to avoid multicollinearity,
would lead to exclude crucial variables. On the other hand, including
all possible covariates in a model often yields implausible signs on
covariates or unstable coefficients, as well as overfitting
[@foreman2012].

When multiple predictors are correlated, but all are crucial for the
analysis (e.g., deaths due to rabies, total deaths, and total DALYs for
all causes), applying a correlation step that filters out correlated
variables can be problematic. One way to overcome bias arising from it
is using regularization techniques like **Ridge Regression** or **Lasso
Regression** is often the best approach to handle multicollinearity
without removing any predictors. Alternatively, **PCA** can reduce
dimensionality while retaining most of the variance. These methods
ensure all important predictors are included in the model without the
adverse effects of multicollinearity.

### Model Specification

The next step is to define the model specification. There are various
type of modeling that can be used, let's get started with a **random
forest**, this choice is due to this type of algorithm which will create
**multiple bootstrap samples** (random samples with replacement) from
the original dataset. Each bootstrap sample is used to train a separate
decision tree.

### Model 1: Random Forest

Rabies death rates may exhibit complex relationships with predictor
variables. Random forests are capable of capturing non-linear
relationships between predictors and the target variable.

Also, it handles multicollinearity, missing data, provides variables
importance and are an ensemble learning method, which means they combine
the predictions of multiple individual decision trees to produce a more
accurate and stable prediction.

In our simplified case this type of model will do random samples with
replacement of data. In `{tidymodels}` we can select different types of
engines, in the case of random forest we could use random forest,
ranger, and others. The difference between these engines derives from
the specific type of calculation used to make the estimation. The Ranger
engine is notably faster than random forest, so let's use that for this
example.

```{r}
#| message: false
#| warning: false
rf_mod <- rand_forest(mtry = tune(),
                      trees = tune(),
                      min_n = tune(),
                      mode = "regression",
                      engine = "ranger")

wkf <- workflow(preprocessor = rec,
                spec = rf_mod)

rf_res <- tune_grid(object = wkf,
          resamples = folds,
          grid = 5,
          control = control_grid(save_pred = TRUE))

show_best(rf_res, metric = "rmse")%>%
  select(-n,-std_err)
```

```{r}
rf_res_tuned <- select_best(rf_res, metric = "rmse")
rf_res_tuned
```

```{r}
rf_fit <- wkf %>% 
  finalize_workflow(select_best(rf_res, 
                                metric = "rmse")) %>%
  fit(training)

rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test) %>% 
   rmse(truth = dalys_rabies, estimate = .pred)
```

```{r}
#| layout-ncol: 2
#| label: fig-rf-predictions
#| fig.cap:
#|   - "Predictions vs. Truth"
#|   - "Predictions vs. Truth by Year"
#| echo: false
rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test)  %>%
   ggplot(aes(dalys_rabies, 
              .pred, 
              color = location)) +
   geom_abline(color = "brown") +
   geom_point(size=2) +
   scale_color_manual(values = c("orange", "navy")) +
   coord_fixed() +
   expand_limits(x = 0, y = 0)+
  labs(title="Predictions vs. Truth",
       x="Truth",
       y="Predictions")

rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test)  %>%
  ggplot(aes(x=year,
             color = location)) +
   geom_point(aes(y=dalys_rabies)) +
   geom_line(aes(y=.pred)) +
  scale_color_manual(values = c("orange", "navy"))+
  labs(title="Predictions vs. Truth by Year",
       x="Time(Year)",
       y="Values")
```

### Model 2: Generalized Linear Model (GLM)

Generalized Linear Models (GLMs) involve statistical estimation rather
than the iterative parameter tuning, common in many machine learning
techniques. However, adding a machine learning feature through parameter
calibration can be done using techniques such as cross-validation and
grid search to find the best model settings.

To introduce a machine learning feature with parameter calibration into
our modeling of the rabies data, we can use a technique like
cross-validation combined with a **regularization method** or an
algorithm that supports parameter tuning. Here, we can employ a model
from the `glmnet` package, which fits a generalized linear model via
**penalized maximum likelihood**. The regularization path is computed
for the lasso or elastic-net penalty at a grid of values for the
regularization parameter lambda.

Adding Machine Learning Features with `{glmnet}` and Cross-Validation

```{r}
#| message: false
#| warning: false  
if (!require(glmnet)) install.packages("glmnet")
library(glmnet)
```

For `glmnet`, we need to input matrices rather than data frames, and
create matrices for the independent variables (predictors) and the
dependent variable (response).

```{r}
predictors <- model.matrix(dalys_rabies ~ ., 
  data = dat)[,-1]  # Remove intercept
response <- dat$dalys_rabies
```

Use cross-validation to find the optimal lambda value, which controls
the strength of the regularization:

```{r}
# Set seed for reproducibility
set.seed(123)

# Fit the model with cross-validation
cv_model <- cv.glmnet(predictors, 
                      response, 
                      family = "gaussian")
cv_model
```

```{r}
# Plot the lambda selection
plot(cv_model)
```

Extracting the best model, we can see that $\lambda$ is 0.165.

```{r}
# Get the best lambda value
best_lambda <- cv_model$lambda.min
print(paste("Best Lambda:", best_lambda))
```

Then, fitting the final model with the selected best lambda, we can
predict and evaluate the model.

```{r}
final_model <- glmnet(predictors, 
                      response, 
                      family = "gaussian", 
                      lambda = best_lambda)

# Predict using the final model
predictions <- predict(final_model, 
                       s = best_lambda, 
                       newx = predictors)

# Calculate Mean Squared Error
rmse <- sqrt(mean((response - predictions)^2))
print(paste("Root Mean Squared Error:", rmse))
```

By incorporating `glmnet` and using lambda selection via
cross-validation, we introduce a **machine learning feature—parameter
calibration—into** our analysis. This approach not only helps in
minimizing overfitting but also enhances model performance by selecting
the most effective regularization parameter. The cross-validation
process used here is crucial for confirming that our model's parameters
are optimally tuned for the given data, embodying a key aspect of
machine learning methodologies.

```{r}
#| message: false
#| warning: false
#| echo: false
#| layout-ncol: 2
#| label: fig-glmnet-predictions
#| fig.cap: "Predictions vs. Truth"

# plot the predictions
data.frame(dalys_rabies =response,predictors,predictions)%>%
  ggplot(aes(x=dalys_rabies,y=predictions))+
  geom_point()+
  geom_abline(color="brown")+
  labs(title="Predictions vs Truth",
       x="Truth",
       y="Predictions")

data.frame(dalys_rabies =response,predictors,predictions)%>%
  ggplot(aes(x=year,group=factor(locationGlobal),
             color=factor(locationGlobal)))+
  geom_point(aes(y=dalys_rabies))+
  geom_line(aes(y=predictions),color="brown")+
scale_color_manual(values = c("orange", "navy"),
                   labels=c("Asia","Global"))+
  labs(title="DALYs due to Rabies Predictions vs Truth",
subtitle="by Year",
       x="Truth",
       y="Predictions")
```

### Testing Multiple Models

In the example above, we used two models to predict DALYs due to rabies,
a random forest with `{tidymodels}` and a generalized linear model with
`{glmnet}` with a Root Mean-Square Error of 0.448 and 0.257
respectively. The Random Forest model has a higher RMSE, which means it
has a higher prediction error compared to the GLM model. However, we
haven't applied any of the preprocessing steps, and there are many other
models that could be used to predict DALYs such as:

1.  **Support Vector Machines (SVM)**: SVMs are a powerful machine
    learning algorithm that can be used for both classification and
    regression tasks. They work by finding the hyperplane that best
    separates the data into different classes or groups.

2.  **Extreme Gradient Boosting (XGBoost)**: Known for its high
    performance in various prediction tasks, XGBoost can handle missing
    values and is effective for large datasets.

3.  **K-Nearest Neighbors (KNN)** models are a type of instance-based
    learning algorithm that stores all available cases and classifies
    new cases based on a similarity measure.

4.  **Long Short-Term Memory (LSTM) Networks**: For temporal or
    sequential health data, LSTM networks can capture dependencies over
    time, making them suitable for time-series prediction of disease
    progression and outcomes.

Each of these models has its own strengths and weaknesses, and the
choice of model will depend on the specific characteristics of the data
and the goals of the analysis. By testing multiple models and comparing
their performance, we can identify the best model for the given data and
task.

Let's use the `{parsnip}` package and the `workflow_set()` function to
fit a set of models to the rabies data. We will fit a **Support Vector
Machine (SVM)**, and a **K-Nearest Neighbors (KNN) model** to the data
and compare their performance.

```{r}
linear_reg_spec <- 
   linear_reg(penalty = tune(), 
              mixture = tune()) %>% 
   set_engine("glmnet")

svm_p_spec <- 
   svm_poly(cost = tune(), 
            degree = tune()) %>% 
   set_engine("kernlab") %>% 
   set_mode("regression")

knn_spec <- 
   nearest_neighbor(neighbors = tune(), 
                    dist_power = tune(), 
                    weight_func = tune()) %>% 
   set_engine("kknn") %>% 
   set_mode("regression")


```

```{r}
#| eval: false
library(rules)
library(baguette)
# Combine workflows into a workflow set
workflow_set <- workflow_set(
  preproc = list(scaled=rec1, 
                 yeo_johnson=rec2),
  models = list(linear_reg = linear_reg_spec, 
                svm = svm_p_spec,
                knn = knn_spec)
)

grid_ctrl <-
   control_grid(
      save_pred = TRUE,
      parallel_over = "everything",
      save_workflow = TRUE
   )

# Fit and evaluate the models with hyperparameter tuning
grid_results <-
   workflow_set %>%
   workflow_map(
      seed = 1503,
      resamples = folds,
      grid = 5,
      control = grid_ctrl
   )
```

```{r}
#| eval: true
#| echo: false
# saveRDS(grid_results, file = "data/model-data/grid_results.rds")
grid_results <- readRDS("data/model-data/grid_results.rds")
```

```{r}
# Show the results
grid_results %>% 
  collect_metrics() %>% 
  arrange(mean)%>%
select(1,5,7,9)%>%
head()
```

```{r}
autoplot(
   grid_results,
   rank_metric = "rmse",  # <- how to order models
   metric = "rmse",       # <- which metric to visualize
   select_best = TRUE     # <- one point per workflow
) +
   geom_text(aes(y = mean - 0.1, label = wflow_id), 
             angle = 90, hjust = 1,size=3) +
   lims(y = c(-1, 0.9)) +
   theme(legend.position = "none")
```

### Summary

The integration of machine learning techniques into public health data
analysis can significantly enhance the predictive power and robustness
of models. By leveraging the capabilities of machine learning
algorithms, we can extract valuable insights from complex health data,
enabling more informed decision-making and policy formulation in public
health contexts. The examples provided in this chapter illustrate the
application of machine learning techniques to health metrics data,
demonstrating the importance of feature engineering, model selection,
and parameter calibration in enhancing the predictive accuracy and
relevance of models. By following best practices in machine learning,
public health researchers and practitioners can harness the power of
data-driven insights to address critical health challenges and improve
population health outcomes.

**Best Practices for Machine Learning in Public Health:**

-   Conduct exploratory data analysis to understand the underlying
    structure of the data and relationships between variables.
-   Apply feature engineering techniques to create new variables and
    enhance the model's predictive power.
-   Select machine learning models that are contextually appropriate and
    robust for public health data analysis. Such as Random Forest,
    Generalized Linear Models, and others.
-   Use parameter calibration techniques such as cross-validation,
    regularization, monte carlo, and grid search to optimize model
    performance.

and grid search to optimize model performance. - Evaluate model
performance using appropriate metrics and visualization tools to assess
predictive accuracy and relevance.

By following these best practices, public health researchers and
practitioners can leverage the full potential of machine learning
techniques to address complex health challenges and improve population
health outcomes. The integration of machine learning methodologies into
public health data analysis represents a significant opportunity to
advance the field of public health and enhance our understanding of
health metrics and disease dynamics.
