# Techniques for Machine Learning Applications {#sec-06-techniques}

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   How to manipulate data through feature engineering
-   Select the most suitable model for your data
-   Learn about machine learning algorithms
:::
:::

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

Deciding on the best strategy for selecting and applying a specific
machine learning model technique can be challenging. Best practices
suggest a thorough understanding of the phenomenon to be modeled,
followed by a data manipulation technique known as *feature engineering*
before applying data to a specific model type. This initial step will
allow the model to run more smoothly and quickly. As a critical stage in
the preparation of machine learning data, feature engineering allows
algorithms to perform better by making the data more amenable to
processing. It involves creating new features from existing data based
on domain knowledge or transforming data to improve the model's ability
to discern patterns. For example, creating features like **moving
averages** or differences between consecutive days can reveal trends and
cycles that are not immediately apparent from raw data.

A another example is the **standardization** process, which is crucial
when dealing with variables measured in different units. It involves
rescaling the features so they have a mean of zero and a standard
deviation of one. This process is particularly important when variables
span several orders of magnitude; without standardization, a model might
incorrectly interpret the scale of a feature as a proxy for importance.

Furthermore, the application of transformations, such as **logarithmic**
scaling or the application of **spline** functions can help in managing
skewed data or enhancing model ability to capture non-linear
relationships, which result particularly useful in complex data
modelings. In addition, tailored adjustments, and more sophisticated
manipulations have been implemented over time to allow estimation of
missing values in order to obtain customized, flexible, and more
homogeneous data. For more information on feature engineering, see
[@butcher2020] useful for effective machine learning strategy
application, covering various techniques and appropriate use cases,
focusing on practical understanding and implementation.

## Model selection strategies

When applying machine learning techniques to health metrics and the
study of infectious diseases, it is crucial to select approaches that
are not only robust but also contextually appropriate. Available machine
learning models can greatly influence the outcomes of analyses, and
therefore, our focus in this book will be on those models that are most
pertinent to public health investigations.

The first step, once data is collected and prepared for analysis,
involves conducting a thorough **exploratory data analysis (EDA)**. This
initial exploration helps to uncover the underlying structure of the
data, the relationships between variables, and the way the response
variable—which may also be referred to as the outcome variable—depends
on predictors. This phase is critical as it informs the necessity of
subsequent data adjustments and transformations.

**Feature engineering** may follow the exploratory phase. This process
involves creating new variables from existing ones to enhance the
machine learning models' ability to discern patterns and make accurate
predictions. Decisions on feature engineering should be based on
insights gained during EDA, with a focus on addressing specific
challenges posed by the complexity of health data, such as
non-linearity, high dimensionality, or the presence of interaction
effects among variables.

By strategically deploying machine learning techniques tailored to
specific data, predictive accuracy and relevancy of models can improve,
thereby enhancing their utility in public health decision-making and
policy formulation.

### Example 4: Rabies

The `rabies` dataset from the `{hmsidwR}` package contains information on
death rates and disability-adjusted life years (DALYs) due to rabies and
all causes mortality in Asia and for the Global region from 1990 to
2019.

Consider only the number of DALYs due to rabies, the first thing to
notice is that deaths rates and DALYs are on different units, rates and
years respectively.

```{r}
library(tidyverse)
rabies <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)
rabies %>% head
```

It can be seen that the number of deaths due to rabies is much lower than the number of DALYs. This difference in scale can affect the model's ability to learn from the data. To address this issue, we can scale and center the numeric variables to make them more comparable.


```{r}
#| layout-ncol: 2
#| label: fig-rabies-deaths-dalys
#| fig.cap: 
#|   - "Not scaled"
#|   - "Scaled and centered"

rabies%>%
  ggplot(aes(x=year, group=location,color=location))+
  geom_line(aes(y=dx_rabies),
            linewidth=1)+
  geom_line(aes(y=dalys_rabies))
  
rabies%>%
  # apply a scale transformation to the numeric variables
  mutate(across(where(is.double),scale))%>%
  ggplot(aes(x=year, group=location,color=location))+
  geom_line(aes(y=dx_rabies),
            linewidth=1)+
  geom_line(aes(y=dalys_rabies)) 
```

Creating new features from existing ones provide additional predictive
power. Then, combine the cause vector in a way to obtain two vectors for
death rates due to rabies and all causes, scale and center the numeric
variables to obtain homogeneous data to use in the model.

```{r}
all_causes <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(!cause=="Rabies")%>%
  rename(dx_allcauses=Deaths,dalys_allcauses=DALYs)%>%
  select(-cause)

dat <- rabies %>%
  full_join(all_causes,by=c("location","year"))

dat %>% head()
```

To be able to visualize the magnitude of difference between death rates
and DALYs for both rabies and all causes, it is necessary to scale or
standardize the data as shown above.

```{r}
dat%>%
  select(-year,-location)%>%
  scale()%>%
  cbind(dat%>%select(year,location))%>%
  ggplot(aes(x=year,group=location,color=location))+
  geom_line(aes(y=dx_rabies),
            linewidth=1)+
  geom_line(aes(y=dx_allcauses))

dat%>%
  select(-year,-location)%>%
  scale()%>%
  cbind(dat%>%select(year,location))%>%
  ggplot(aes(x=year,
  group=location,
  color=location))+
  geom_line(aes(y=dalys_rabies),
            linewidth=1)+
  geom_line(aes(y=dalys_allcauses))
```

### Modeling

Load the `{tidymodels}` meta-package and split the data into training
and test, create a resampling with cross-validation technique, and a
recipe to pre-process the data.

```{r}
#| message: false
#| warning: false
library(tidymodels)
set.seed(04042024)
split <- initial_split(dat,prop = 0.8)
training <- training(split)
test <- testing(split)
folds <- vfold_cv(training)
```

Then create a recipe with some key steps to transform the data into a
way the model would be able to understand and learn from it.

```{r}
rec <- recipe(dalys_rabies ~. , data=training)%>%
  step_dummy(all_nominal_predictors())%>%
  step_scale(all_predictors())%>%
  step_center(all_predictors())%>%
  step_corr(threshold = 0.8)%>%
  step_YeoJohnson(all_predictors())

rec %>%
  prep()%>%juice()%>%head
```

### Random Forest

There are various type of modeling that can be used, let's get started
with a random forest, this choice is due to this type of algorithm which
will create multiple bootstrap samples (random samples with replacement)
from the original dataset. Each bootstrap sample is used to train a
separate decision tree.

Rabies death rates may exhibit complex relationships with predictor
variables. Random forests are capable of capturing non-linear
relationships between predictors and the target variable.

Also, it handles missing data well, provides variables importance and
are an ensemble learning method, which means they combine the
predictions of multiple individual decision trees to produce a more
accurate and stable prediction.

In our simplified case this type of model will do random samples with
replacement of data. In `{tidymodels}` we can select different types of
engines, in the case of random forest we could use random forest,
ranger, and others. The difference between these engines derives from
the specific type of calculation used to make the estimation. The Ranger
engine is notably faster than random forest, so let's use that for this
example.

```{r}
rf_mod <- rand_forest(mtry = tune(),
                      trees = tune(),
                      min_n = tune(),
                      mode = "regression",
                      engine = "ranger")
```

```{r}
wkf <- workflow(preprocessor = rec,
                spec = rf_mod)

rf_res <- tune_grid(object = wkf,
          resamples = folds,
          grid = 5,
          control = control_grid(save_pred = TRUE))
```

```{r}
#rf_res%>%unnest(.metrics)
show_best(rf_res, metric = "rmse")
```

```{r}
autoplot(rf_res, metric = "rmse") +
  scale_x_log10()
```

```{r}
rf_res_tuned <- select_best(rf_res, metric = "rmse")
```

```{r}
rf_fit <- wkf %>% 
  finalize_workflow(select_best(rf_res, 
                                metric = "rmse")) %>%
  fit(training)
```

```{r}
rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test) %>% 
   rmse(truth = dalys_rabies, estimate = .pred)
```

```{r}
rf_fit %>% 
   predict(new_data = test) %>% 
   bind_cols(test)  %>%
  ggplot(aes(dalys_rabies, 
              .pred, 
              color = location)) +
   geom_point() +
   geom_abline(color = "red") +
   scale_color_brewer(type = "qual") +
   coord_fixed() +
   expand_limits(x = 0, y = 0)
```

### Generalized Linear Model (GLM)

In the example using a Generalized Linear Model (GLM) I provided
earlier, we didn't explicitly cover the process of model parameters
calibration that typically characterizes more complex machine learning
models. GLMs, while powerful, primarily involve statistical estimation
rather than the iterative parameter tuning that is common in many
machine learning techniques. However, adding a machine learning aspect
through parameter calibration can be done using techniques such as
cross-validation and grid search to find the best model settings.

To introduce a machine learning feature with parameter calibration into
our modeling of the rabies data, we can use a technique like
cross-validation combined with a regularization method or an algorithm
that supports parameter tuning. Here, we can employ a model from the
`glmnet` package, which fits a generalized linear model via penalized
maximum likelihood. The regularization path is computed for the lasso or
elastic-net penalty at a grid of values for the regularization parameter
lambda.

Adding Machine Learning Features with `{glmnet}` and Cross-Validation

```{r}
#| message: false
#| warning: false  
if (!require(glmnet)) install.packages("glmnet")
library(glmnet)
```

For `glmnet`, we need to input matrices rather than data frames:

```{r}
# Create matrices for the independent variables (predictors) and the dependent variable (response)
predictors <- model.matrix(dalys_rabies ~ ., 
  data = dat)[,-1]  # Remove intercept
response <- dat$dalys_rabies
```

Use cross-validation to find the optimal lambda value, which controls
the strength of the regularization:

```{r}
# Set seed for reproducibility
set.seed(123)

# Fit the model with cross-validation
cv_model <- cv.glmnet(predictors, 
response, 
family = "gaussian")

# Plot the lambda selection
plot(cv_model)
```

Extract the Best Model and Review

```{r}
# Get the best lambda value
best_lambda <- cv_model$lambda.min
print(paste("Best Lambda:", best_lambda))
```

```{r}
# Fit final model with the selected best lambda
final_model <- glmnet(predictors, 
response, 
family = "gaussian", 
lambda = best_lambda)

# Summary of the final model
print(summary(final_model))
```

Predict and Evaluate

```{r}
# Predict using the final model
predictions <- predict(final_model, 
s = best_lambda, 
newx = predictors)

# Calculate Mean Squared Error
mse <- mean((response - predictions)^2)
print(paste("Mean Squared Error:", mse))
```

### Considerations

By incorporating `glmnet` and using lambda selection via
cross-validation, we introduce a machine learning feature—parameter
calibration—into our analysis. This approach not only helps in
minimizing overfitting but also enhances model performance by selecting
the most effective regularization parameter. The cross-validation
process used here is crucial for confirming that our model's parameters
are optimally tuned for the given data, embodying a key aspect of
machine learning methodologies.

In summary, the integration of machine learning techniques into public health data analysis can significantly enhance the predictive power and robustness of models. By leveraging the capabilities of machine learning algorithms, we can extract valuable insights from complex health data, enabling more informed decision-making and policy formulation in public health contexts.

