# Essential R Packages for Machine Learning {#sec-07-packages}

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   Overview of the most suitable R-packages
-   Combination of packages and functions
-   How to find a new R-package
:::
:::

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
options(timeout = 600)
```

## Inside and outside of the library boxes

This book provides all data analysis, visualization, and modeling in R
(a hint to Python translations can be found in @sec-appendixC), and the
universe of available packages is vast. How do you untangle among all of
them? Which one to choose? Sometimes, it is rather favorable to start by
working out of the library boxes, which means using functions provided
in the built-in packages inside a fresh R installation.

On the contrary, life has become easier with the advent of wrapping
functions and the reduction of lines of code through the use of
functions found in packages. In the past, data analysis and modeling in
R required extensive coding, often involving repetitive tasks and
lengthy scripts. However, with the introduction of packages containing
pre-built functions for common tasks, such as data manipulation,
visualization, and statistical modeling, the process has been
streamlined. These packages offer a wide range of functions that can be
easily applied to perform complex analyses with just a few lines of
code. As a result, users can achieve sophisticated results more
efficiently, without the need to write lengthy and intricate scripts
from scratch. This shift has not only simplified the process of data
analysis and modeling but has also democratized access to advanced
statistical techniques, making them more accessible to a broader
audience.

## Useful packages

Some of the most used modeling packages available in R that are
particularly useful for machine learning modeling of the spread of
infectious diseases and evaluating DALYs variation due to infectious
disease include:

-   [**caret**](https://topepo.github.io/caret/)[@kuhn]: The caret
    package (short for **C**lassification **A**nd **RE**gression
    **T**raining) provides a unified interface for training and
    evaluating machine learning models. It supports a wide range of
    algorithms and offers convenient functions for hyperparameter tuning
    and model selection.
-   [**tidymodels**](https://www.tmwr.org/)[@silge]: tidymodels provides
    a collection of packages for modeling and machine learning tasks,
    including tidyr, dplyr, and ggplot2. It offers a consistent and tidy
    workflow for data preprocessing, model building, evaluation, and
    visualization.
-   [**mlr3**](https://mlr3book.mlr-org.com/)[@mlr-org]: mlr3 is a
    modern and extensible framework for machine learning in R. It
    provides a unified interface for modeling tasks, making it easy to
    train and evaluate a wide range of machine learning algorithms.
-   [**h2o**](https://h2o.ai/) is another powerful machine learning
    framework that can be used to predict DALYs. **`h2o`** offers a
    range of machine learning algorithms, including linear models,
    tree-based models, and ensemble methods, with efficient handling of
    large datasets and support for parallel computing.
-   [**INLA**](https://www.r-inla.org/what-is-inla)[@r-inlap]:
    Integrated Nested Laplace Approximations (INLA) is a package for
    Bayesian inference using the INLA method. It is particularly useful
    for spatial and spatio-temporal modeling, which can be relevant for
    studying the spread of infectious diseases.
-   [**fpp3**](https://otexts.com/fpp3/)[@forecast]: The fpp3 package is
    part of the Forecasting: Principles and Practice (3rd edition) book,
    offering tools for forecasting time series data. While not
    specifically focused on machine learning, it provides useful
    functions for time series analysis, which can be relevant for
    modeling disease spread over time.

These packages offer a diverse set of tools and techniques for modeling
infectious disease spread and assessing its impact on health metrics
like DALYs. These packages provide valuable insights into the dynamics
of infectious diseases and used correctly can be used to inform public
health interventions and policies. However, whether these packages are
the most suitable depends on various factors, including:

-   the specific requirements of the analysis
-   the expertise of the researcher
-   the nature of the data being used

Each package has its strengths and weaknesses, and the choice of package
often depends on factors such as:

-   the complexity of the modeling task
-   the type of data available
-   the preferred modeling approach (e.g., frequentist vs. Bayesian).

For example, tidymodels provides a tidy and consistent workflow for
modeling tasks, while mlr3 offers extensive support for machine learning
algorithms and hyperparameter tuning.

It's also worth considering other packages that may be relevant for
specific aspects of the analysis, such as spatial modeling (e.g., spdep)
or time series forecasting (e.g., prophet).

-   [**spdep**](https://cran.r-project.org/web/packages/spdep/index.html):
    spdep package and its piped interface
    [**sfdep**](https://cran.r-project.org/package=sfdep) are powerful
    tools for spatial econometrics and spatial statistics. With a set of
    functions for analyzing spatial dependence and autocorrelation in
    spatial data, they can be particularly relevant in infectious
    disease modeling. By accounting for spatial relationships between
    geographic units, such as regions or neighborhoods, these packages
    incorporate spatial effects into models and allow for a better
    understanding of how infectious diseases spread across space.
-   [**prophet**](https://facebook.github.io/prophet/): prophet package
    in R is specifically designed for time series forecasting with an
    emphasis on flexibility and ease of use. Developed by Facebook,
    prophet can handle various time series patterns, including
    seasonality, holiday effects, and trend changes, making it suitable
    for modeling infectious disease trends over time. The package
    automatically detects and accommodates these patterns, simplifying
    the modeling process and enabling researchers to generate accurate
    forecasts with minimal effort.

Ultimately, the features and capabilities of each package should
carefully be evaluated in relation to their specific research objectives
before to choose the one that best meets research needs and objectives.
Additionally, experimenting with different packages and techniques may
be beneficial to identify the most effective approach for a given
analysis.

### Examples

#### Tidymodels

We have already used this package in @sec-05-machine_learning and
@sec-06-techniques. This meta-package is particulrely useful for testing
out more models with the help of the workflow_set() function. Here we
see an example:

```{r}
library(tidymodels)
```

#### mlr3

This meta-package is particularly useful for benchmarking different
models. Here we see an example:

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

library(tidyverse)
data <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)%>%
  mutate(across(where(is.character),as.factor))
```

```{r}
#| message: false
#| warning: false
library(mlr3)
library(mlr3learners)
library(mlr3viz)
library(mlr3verse)
library(data.table)
#library(xgboost)

# Define the task
task <- TaskRegr$new(id = "dalys_rabies", 
                     backend = data, 
                     target = "dalys_rabies")

# Define the learners
learner_lm <- lrn("regr.lm")
learner_xgboost <- lrn("regr.xgboost", 
                       nrounds = 1000, 
                       max_depth = 6, 
                       eta = 0.01)

# Add preprocessing steps: creating dummy variables
poe <- po("encode")
task <- poe$train(list(task))[[1]]

# Define a resampling strategy
resampling <- rsmp("cv", folds = 5)

# Create a benchmark design
design <- benchmark_grid(
  tasks = task,
  learners = list(learner_lm, learner_xgboost),
  resamplings = resampling
)

design
```

```{r}
#| message: false
#| warning: false
#| output: false
# Run the benchmark
set.seed(19052024)
bmr = benchmark(design, store_models = TRUE)
```

```{r}
# Aggregate the results
measures <- msrs(c("regr.rmse", "regr.mae"))
bmr$aggregate(measures)
```

#### h2o

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```



```{r}
#| message: false
#| warning: false
#| echo: false
library(tidyverse)
data <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)%>%
  mutate(across(where(is.character),as.factor))
```

```{r}
#| message: false
#| warning: false
options(timeout = 6000)
# install.packages("h2o")
library(h2o)
```

```{r}
#| message: false
#| warning: false
#| output: false
# Initialize the H2O cluster
h2o.init()
Sys.sleep(10)
```

```{r}
#| message: false
#| warning: false
#| output: false
# Upload data to H2O
h2o_data <- as.h2o(data)

response <- "dalys_rabies"
predictors <- setdiff(names(data), response)


# Split data into training and testing sets
splits <- h2o.splitFrame(h2o_data, ratios = 0.8, seed = 1234)
train <- splits[[1]]
test <- splits[[2]]


# Train a linear regression model
model_lm <- h2o.glm(
  x = predictors,
  y = response,
  training_frame = train,
  family = "gaussian"
)

# Train a GBM model
model_gbm <- h2o.gbm(
  x = predictors,
  y = response,
  training_frame = train,
  ntrees = 1000,
  max_depth = 6,
  learn_rate = 0.01,
  seed = 1234
)

# Train a random forest model
model_rf <- h2o.randomForest(
  x = predictors,
  y = response,
  training_frame = train,
  ntrees = 100,
  max_depth = 20,
  seed = 1234
)

# Evaluate linear regression model
perf_lm <- h2o.performance(model_lm, newdata = test)
h2o.rmse(perf_lm);
h2o.mae(perf_lm)
```

```{r}
#| message: false
#| warning: false
# Evaluate GBM model
perf_gbm <- h2o.performance(model_gbm, newdata = test)
h2o.rmse(perf_gbm);
h2o.mae(perf_gbm)
```

```{r}
#| message: false
#| warning: false
# Evaluate random forest model
perf_rf <- h2o.performance(model_rf, newdata = test)
h2o.rmse(perf_rf);
h2o.mae(perf_rf)
```

```{r}
#| message: false
#| warning: false
h2o.shutdown(prompt = FALSE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```
