# Essential R Packages for Machine Learning {#sec-07-packages}

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   Overview of the most suitable R-packages
-   Apply different model frameworks
-   How to find a new R-package
:::
:::

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

This chapter provides an overview of the most suitable R-packages for
machine learning. It also discusses the combination of packages and
functions to achieve the desired results. Additionally, it explores how
to find new R-packages and evaluate their suitability for a given task.

## Inside and Outside of the Library Boxes

When it comes to R, the possibilities are endless. With over 15,000
packages available on CRAN (Comprehensive R Archive Network) and
countless others on GitHub and other repositories, universe of available
packages is vast. How to untangle among all of them? Which one to
choose? Sometimes, it is rather favorable to start by working out of the
library boxes, which means using functions provided in the built-in
packages inside a fresh R installation.

However, the vast majority of the time, the best way to start is by
using the packages that are already available in the R ecosystem. These
packages are designed to make your life easier by providing pre-built
functions for common tasks. They can help you save time and effort by
automating repetitive tasks and streamlining complex analyses.

Life has become easier with the advent of wrapping functions and the
reduction of lines of code through the use of functions found in
packages. In the past, data analysis and modelling in R required
extensive coding, often involving repetitive tasks and lengthy scripts.
As a result, users can achieve sophisticated results more efficiently,
without the need to write lengthy and intricate scripts from scratch.
This shift has not only simplified the process of data analysis and
modelling but has also democratized access to advanced statistical
techniques, making them more accessible to a broader audience.

## Essential R Packages for Machine Learning

Some of the most used modelling packages available in R that are
particularly useful for machine learning modelling of the spread of
infectious diseases and evaluating DALYs variation due to infectious
disease include:

### Meta-Packages

1.  [**`{tidymodels}`**](https://www.tmwr.org/)[@silge]:
    This\index{tidymodels} framework provides a collection of packages
    for modelling and machine learning tasks, including tidyr, dplyr,
    and ggplot2. It offers a consistent and tidy workflow for data
    preprocessing, model building, evaluation, and visualization.
2.  [**`{caret}`**](https://topepo.github.io/caret/)[@kuhn]: The caret
    framework (short for **C**lassification **A**nd **RE**gression
    **T**raining) provides a unified interface for training and
    evaluating machine learning models. It supports a wide range of
    algorithms and offers convenient functions for hyperparameter tuning
    and model selection.
3.  [**`{mlr3}`**](https://mlr3book.mlr-org.com/)[@mlr-org]: This is a
    modern and extensible framework for machine learning in R. It
    provides a unified interface for modelling tasks, making it easy to
    train and evaluate a wide range of machine learning algorithms.

### Engines

Engines are the actual implementations of machine learning algorithms.
They perform the computations necessary to train and evaluate models.
Engines can be integrated into various meta-packages/frameworks to
leverage their specific algorithms and computational efficiencies. Each
engine might have its own set of parameters and functionalities, which
can be accessed through the high-level interfaces provided by
meta-packages.

1.  [`{randomForest}`](https://www.stat.berkeley.edu/~breiman/RandomForests/):
    For random forest models.
2.  [`{xgboost}`](https://github.com/dmlc/xgboost): An optimized
    distributed gradient boosting library designed to be highly
    efficient, flexible, and portable. It implements machine learning
    algorithms under the gradient boosting framework.
3.  [`{glmnet}`](https://glmnet.stanford.edu/): For lasso and ridge
    regression. Provides efficient procedures for fitting generalized
    linear models via penalized maximum likelihood.
4.  [`{nnet}`](http://www.stats.ox.ac.uk/pub/MASS4/): For neural
    networks.
5.  [`{kernlab}`](https://cran.r-project.org/web/packages/kernlab/index.html):
    For kernel-based machine learning methods.
6.  [`{e1071}`](https://cran.r-project.org/web/packages/e1071/index.html):
    For support vector machines and other functions.
7.  [`{lme4}`](https://github.com/lme4/lme4/): For linear and
    generalized linear mixed-effects models.
8.  [`{mgcv}`](https://cran.r-project.org/web/packages/mgcv/index.html):
    For generalized additive models.
9.  [`{rpart}`](https://github.com/bethatkinson/rpart): For recursive
    partitioning and regression trees.
10. [`{h2o}`](https://github.com/h2oai/h2o-3): A powerful machine
    learning framework that can be efficiently integrated with
    `{tidymodels}`. `{h2o}` offers a range of machine learning
    algorithms, including linear models, tree-based models, and ensemble
    methods, with efficient handling of large datasets and support for
    parallel computing.
11. [`{keras3}`](https://keras.posit.co/), the R interface to
    [Keras](https://keras.io/), is a deep learning technique which
    provides high-level **neural networks** API written in Python.
    Designed to enable fast experimentation with deep learning models on
    top of [`{tensorflow}`](https://github.com/rstudio/tensorflow) (an
    open-source machine learning framework developed by the Google).

### Time Series Analysis

1.  [`{forecast}`](https://pkg.robjhyndman.com/forecast/): For
    forecasting functions.
2.  [`{prophet}`](https://github.com/facebook/prophet): Specifically
    designed for time series forecasting with an emphasis on flexibility
    and ease of use. Developed by Facebook, prophet can handle various
    time series patterns, including seasonality, holiday effects, and
    trend changes, making it suitable for modelling infectious disease
    trends over time.
3.  [`{fpp3}`](https://github.com/robjhyndman/fpp3package)[@forecast]:
    This package is part of the Forecasting: Principles and Practice
    (3rd edition) book, offering data and tools for forecasting time
    series data, such as the import of the
    [`{fable}`](https://fable.tidyverts.org/) package with the `ARIMA()`
    model which is relevant for modelling disease spread over time.

### Bayesian Analysis

1.  [`{brms}`](https://github.com/paul-buerkner/brms): For Bayesian
    generalized multivariate non-linear multilevel models using Stan.
2.  [`{rstan}`](https://mc-stan.org/rstan/): For Bayesian inference
    using Stan.

### Specialized Tools

1.  [`{spdep}`](https://github.com/r-spatial/spdep/) is a powerful tools
    for spatial statistics, it provides a set of functions for analyzing
    spatial dependencies and autocorrelation in spatial data,
    particularly relevant in infectious disease modelling. It also
    incorporates spatial effects into models and allows for a better
    understanding of how infectious diseases spread across space.
2.  [`{INLA}`](https://www.r-inla.org/)[@r-inlap]: Integrated Nested
    Laplace Approximations (INLA) is a package for Bayesian inference
    using the INLA method. It is particularly useful for spatial and
    spatio-temporal modelling, which can be relevant for studying the
    spread of infectious diseases.

These packages offer a diverse set of tools and techniques for modelling
infectious disease spread and assessing its impact on health metrics
like DALYs. These packages provide valuable insights into the dynamics
of infectious diseases and can be used to inform public health
interventions and policies. However, whether these packages are the most
suitable depends on various factors, including:

-   the specific requirements of the analysis
-   the expertise of the researcher
-   the nature of the data being used

Each package has its strengths and weaknesses, and the choice of package
often depends on factors such as:

-   the complexity of the modelling task
-   the type of data available
-   the preferred modelling approach (e.g., frequentist vs. Bayesian).

For example, `tidymodels` provides a tidy and consistent workflow for
modelling tasks, while `mlr3` offers extensive support for machine
learning algorithms and hyperparameter tuning.

It's also worth considering other packages that may be relevant for
specific aspects of the analysis, such as spatial modelling (e.g.,
`spdep`) or time series forecasting (e.g., `prophet`).

The features and capabilities of each package should carefully be
evaluated in relation to their specific research objectives before to
choose the one that best meets research needs and objectives. And it is
worth considering that experimenting with different packages and
techniques may be beneficial to identify the most effective approach for
a given analysis.

## Model Framework Application Examples

In this section we provide some examples of the application of different
model frameworks. The data is used, as introduced in @sec-06-techniques,
to demonstrate the application of model frameworks such as `mlr3`,
`h2o`, and `keras` for machine learning tasks.

We have already shown an application of the `{tidymodels}` meta-package
in @sec-05-machine_learning and @sec-06-techniques, particularly useful
for testing out different types of models with the help of key functions
such as the `workflow()`, `workflow_set()`, and `workflow_map()`.

### Example: DALYs due to Dengue with mlr3

The dataset for this example is from the `{hmsidwR}` package, which
contains data on Deaths, DALYs, YLDs, YLLs, Prevalence, and Incidence
due to selected infectious diseases from 1980 to 2021.

```{r}
#| message: false
#| warning: false
library(tidyverse)
hmsidwR::infectious_diseases %>% names
```

Let's have a look at the locations where Dengue has been reported:

```{r}
hmsidwR::infectious_diseases %>%
  filter(cause_name == "Dengue") %>%
  distinct(location_name)
```

And look at how the health metrics shape up over the years for Dengue:

```{r}
#| message: false
#| warning: false
#| echo: false
#| layout-ncol: 3
#| label: fig-dengue-metrics
#| fig-cap:
#|  - "YLLs"
#|  - "YLDs"
#|  - "DALYs"
#| fig-alt: "Health Metrics due to Dengue"

hmsidwR::infectious_diseases %>%
  filter(cause_name == "Dengue") %>%
  group_by(location_name) %>%
  ggplot(aes(x = year, y = YLLs)) +
  geom_line(aes(linetype = location_name)) +
  labs(title = "YLLs due to Dengue from 1980 to 2021")

hmsidwR::infectious_diseases %>%
  filter(cause_name == "Dengue") %>%
  drop_na() %>%
  group_by(location_name) %>%
  ggplot(aes(x = year, y = YLDs)) +
  geom_line(aes(linetype = location_name)) +
  labs(title = "YLDs due to Dengue from 1990 to 2021")


hmsidwR::infectious_diseases %>%
  filter(cause_name == "Dengue") %>%
  drop_na() %>%
  group_by(location_name) %>%
  ggplot(aes(x = year, y = DALYs)) +
  geom_line(aes(linetype = location_name)) +
  labs(title = "DALYs due to Dengue from 1990 to 2021")
```

The trend of the metrics in some locations is quite flat ranging around
zero, while in others it is more evident. YLLs are the leading cause of
DALYs, but YLDs are also contributing to the burden of disease.

We will focus on DALYs due to `Dengue` in three locations, Malawi,
Zambia, and Central African Republic from 1990 to 2016 (26 years). The
data from 2017 to 2021 is not included in the analysis and held out to
be used as new data for testing the model's performance, as shown in
@sec-08-predictions.

Data is preprocessed to remove missing values, the values of DALYs only
appeared in 1990, and then grouped by location id; the location name is
removed.

```{r}
dalys_dengue <- hmsidwR::infectious_diseases %>%
  arrange(year) %>%
  filter(cause_name == "Dengue",
         year<=2016,
         !location_name %in% c("Eswatini", "Lesotho")) %>%
  drop_na() %>%
  group_by(location_id) %>%
  select(-location_name, -cause_name)

dalys_dengue %>%
  head()
```

The `{mlr3}` is a meta-package particularly useful for its syntax, as it
guides the user through the machine learning steps by providing a
consistent naming convention. It is also useful for benchmarking
different models. In addition to the `{mlr3}` package, the
`{mlr3learners}`, `{mlr3viz}`, and `{mlr3verse}`, and `{data.table}`
packages are also loaded. Each of these packages provides additional
functionality for machine learning tasks. More information on these
packages can be found in the respective documentation, and further
details can be found in the [mlr3 book](https://mlr3book.mlr-org.com/).
In this example we use two types of models:

-   `regr.cv_glmnet`: A cross-validated `glmnet` model with an alpha of
    0.5 and a lambda of 0.1.
-   `regr.xgboost`: An `xgboost` model with 1000 rounds, a maximum depth
    of 6, and an eta of 0.01. For this model the `{xgboost}` package is
    required.

```{r}
#| message: false
#| warning: false
library(mlr3)
library(mlr3learners)
library(mlr3viz)
library(mlr3verse)
library(data.table)
# library(xgboost)
```

Define the task, where the `id`, `backend`, and `target` are specified.
The `backend` is the data that will be used for training the models. The
`task` is the interface between the data and the learners; in this
use-case the preprocessing steps are done before hand, by removing
missing and categorical values.

```{r}
task <- TaskRegr$new(id = "DALYs",
                     backend = dalys_dengue,
                     target = "DALYs"
                     )

task$data() %>% head()
```

The learners are the models specifications. A full list of learners can
be obtained with the function `as.data.table(mlr_learners)`.

```{r}
learner_cv_glmnet <- lrn("regr.cv_glmnet", 
                         alpha = 0.5, 
                         s = 0.1)
learner_xgboost <- lrn("regr.xgboost",
                       nrounds = 1000,
                       max_depth = 6,
                       eta = 0.01)
```

Then define a **resampling strategy** to estimate the generalization
performance, then create a **benchmark design** to compare different
learners:

```{r}
resampling <- rsmp("cv", folds = 5)

design <- benchmark_grid(tasks = task,
                         learners = list(learner_cv_glmnet,
                                         learner_xgboost),
                         resamplings = resampling
                         )
design
```

Run the benchmark:

```{r}
#| message: false
#| warning: false
#| output: false
set.seed(19052024)
bmr <- benchmark(design,
                 store_models = TRUE,
                 store_backends = TRUE
                 )
```

To access the content of an object such as `bmr`, use the `$` operator,
or for more options use the `View()` function.

The `score()` function is used to extract the performance metrics of the
models, in this case the mean squared error (MSE) is used.

```{r}
bmr$score()[, .(learner_id,
                task_id,
                iteration,
                regr.mse)]
```

Here we see the results of the benchmark, the `regr.mse` is the mean
squared error of the models. The `regr.rmse` and `regr.mae` are also
available.

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "BMR Regression RMSE"
#|   - "BMR Regression MAE"
#| fig-alt: "BMR Regression"
#| label: fig-bmr-egression
set.seed(349)
autoplot(bmr, measure = msr("regr.rmse"))
autoplot(bmr, measure = msr("regr.mae"))
```

Aggregate the results:

```{r}
#| message: false
#| warning: false
measures <- msrs(c("regr.rmse", "regr.mae"))
aggr <- bmr$aggregate(measures)
```

```{r}
rr1 <- aggr$resample_result[[1]]
rr2 <- aggr$resample_result[[2]]
```

```{r}
#| echo: false
#| eval: false
saveRDS(rr1, "data/model-data/rr1.rds")
saveRDS(rr2, "data/model-data/rr2.rds")
```

Create a data frame with the results, for plotting with ggplot2,
consider the data are shuffled in the resampling process, and need to be
reordered:

```{r}
regr.cv_glmnet <- as.data.table(rr1$prediction()) %>%
  mutate(learner = "regr.cv_glmnet") %>%
  # order data to match the original order
  cbind(rr1$task$data()[rr1$prediction()$row_ids])

regr.xgboost <- as.data.table(rr2$prediction()) %>%
  mutate(learner = "regr.xgboost") %>%
  # order data to match the original order
  cbind(rr2$task$data()[rr2$prediction()$row_ids])


dat <- rbind(regr.cv_glmnet, regr.xgboost)
```

Plot the results:

```{r}
#| fig-cap:
#|   - "Truth vs Response for cv.glmnet and xgboost"
#| fig-alt: "Truth vs Response"
#| label: fig-glmnet_xgboost
#| fig-width: 5
#| fig-height: 3

dat %>%
  ggplot(aes(x = truth, y = response,
             group = learner)) +
  geom_abline() +
  geom_point(aes(color = learner)) +
  scale_color_manual(values = c("regr.cv_glmnet" = "navy",
                                "regr.xgboost" = "orange")) +
  facet_wrap(~learner) +
  labs(x = "Truth", y = "Response") +
  theme(legend.position = "none")
```

```{r}
#| label: fig-dengue-daly
#| fig-cap: "Models results on observed DALYs due to Dengue 1990-2016; Test Resampled Data with cv.glmnet and xgboost."
#| fig-alt: "DALYs due to Dengue"
#| fig-width: 8
#| fig-height: 4
dat %>%
  group_by(location_id, learner) %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = truth),
             size = 0.2) +
  geom_line(aes(y = response,
                linetype = learner),
            linewidth = 0.3) +
  # create facets for each location and
  # rename the location_id with the location_name
  # with the labeller function
  facet_wrap(vars(location_id),
    labeller = as_labeller(c(`182` = "Malawi",
                             `191` = "Zambia",
                             `169` = "Central African Republic")),
    scales = "free") +
  labs(title = "Dalys due to Dengue 1990-2016: Truth vs Responses",
       caption = "Scale Free - Dots represent the Truth Values",
       x = "Year", y = "DALYs", linetype = "Model") +
  theme(plot.title = element_text(hjust = 0.5))
```

In the next @sec-08-predictions we will see an application of these two
models on remaing years from 2017 to 2021, to test the model performance
as it was on new data.

### Example: DALYs due to Rabies with H2O

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

The following example shows how to use the `{h20}` r package for running
[**H2O**](https://docs.h2o.ai/h2o/latest-stable/h2o-docs/index.html) via
its REST API. It is particularly useful for handling large datasets.
However, this is not the case seen the size of the rabies dataset, we
can use it to compare the results with the previous model frameworks.

```{r}
#| message: false
#| warning: false
#| echo: false
library(tidyverse)
data <- hmsidwR::rabies %>%
  select(-upper, -lower) %>%
  pivot_wider(names_from = measure, values_from = val) %>%
  filter(cause == "Rabies") %>%
  rename(dx_rabies = Deaths, dalys_rabies = DALYs) %>%
  select(-cause) %>%
  mutate(across(where(is.character), as.factor))
```

```{r}
#| message: false
#| warning: false
#| eval: false
options(timeout = 6000)
install.packages("h2o")
```

```{r}
#| message: false
#| warning: false
#| eval: false
library(h2o)
```

```{r}
#| message: false
#| warning: false
#| output: false
#| eval: false
# Initialize the H2O cluster
h2o.init()
```

```{r}
#| message: false
#| warning: false
#| output: false
#| eval: false
# Upload data to H2O
h2o_data <- as.h2o(data)
```

```{r}
response <- "dalys_rabies"
```

```{r}
#| eval: false
predictors <- setdiff(names(data), response)
# Split data into training and testing sets
splits <- h2o.splitFrame(h2o_data,
                         ratios = 0.8,
                         seed = 1234)
train <- splits[[1]]
test <- splits[[2]]
```

Train three models: a linear regression model, a GBM model, and a random
forest model.

```{r}
#| message: false
#| warning: false
#| output: false
#| eval: false
model_lm <- h2o.glm(x = predictors,
                    y = response,
                    training_frame = train,
                    family = "gaussian")
model_gbm <- h2o.gbm(x = predictors,
                     y = response,
                     training_frame = train,
                     validation_frame = test,
                     ntrees = 1000,
                     max_depth = 6,
                     learn_rate = 0.01,
                     seed = 1234)
model_rf <- h2o.randomForest(x = predictors,
                             y = response,
                             training_frame = train,
                             ntrees = 100,
                             max_depth = 20,
                             seed = 1234)
```

Evaluate the models:

```{r}
#| eval: false
perf_lm <- h2o.performance(model_lm, newdata = test)
perf_gbm <- h2o.performance(model_gbm, newdata = test)
perf_rf <- h2o.performance(model_rf, newdata = test)
```

Then extract the RMSE and MAE metrics:

```         
h2o.rmse(<model>)
h2o.mae(<model>)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: false

h20.rabies.metrics <- rbind(c(h2o.rmse(perf_lm),
                              h2o.mae(perf_lm)),
                            c(h2o.rmse(perf_gbm),
                              h2o.mae(perf_gbm)),
                            c(h2o.rmse(perf_rf),
                              h2o.mae(perf_rf))) %>%
  as.data.frame() %>%
  mutate(model = c("lm", "gbm", "rf")) %>%
  relocate(model) %>%
  rename(rmse = V1, mae = V2)

saveRDS(h20.rabies.metrics, "data/model-data/h20.rabies.metrics.rds")
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: true
h20.rabies.metrics <- readRDS("data/model-data/h20.rabies.metrics.rds")
```

```{r}
#| echo: false
h20.rabies.metrics
```

```{r}
#| eval: false
model_gbm@model
```

```{r}
#| output: false
#| eval: false
# Predict on test data
predictions <- h2o.predict(model_rf,
                           newdata = test)
```

```{r}
#| echo: false
#| eval: false
saveRDS(predictions, "data/model-data/h20.rabies.predictions.rds")
```

```{r}
#| eval: true
#| echo: false
predictions <- readRDS("data/model-data/h20.rabies.predictions.rds")
```

Convert H2O frames to data frames for ggplot2, it requires the h2o
library to be loaded and initiated:

```{r}
#| eval: false
pred_df <- as.data.frame(predictions)
actual_df <- as.data.frame(test)
```

```{r}
#| echo: false
#| eval: false
saveRDS(pred_df, "data/model-data/h20.rabies.pred_df.rds")
saveRDS(actual_df, "data/model-data/h20.rabies.actual_df.rds")
```

```{r}
#| eval: true
#| echo: false
pred_df <- readRDS("data/model-data/h20.rabies.pred_df.rds")
actual_df <- readRDS("data/model-data/h20.rabies.actual_df.rds")
```

```{r}
# Combine actual and predicted values
results_df <- data.frame(Actual = actual_df[, response],
                         Predicted = pred_df[, 1])
# Plot Actual vs Predicted
p1 <- ggplot(results_df,
             aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0,
              color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted Values",
       x = "Actual", y = "Predicted")
```

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Residuals vs Predicted Values"
#|   - "Normality Test of the Residulas"
#| fig-alt: "Residuals Distribution"
#| label: fig-h20-residuals
# Calculate residuals
results_df$Residuals <- results_df$Actual - results_df$Predicted
# Plot Residuals vs Predicted
ggplot(results_df,
       aes(x = Predicted, y = Residuals)) +
  geom_point(color = "navy", alpha = 0.5) +
  geom_hline(yintercept = 0,
             color = "orange", linetype = "dashed") +
  labs(title = "Residuals vs Predicted Values",
       x = "Predicted", y = "Residuals")
# Normality Test of the Residuals
qqnorm(results_df$Residuals)
qqline(results_df$Residuals, col = 2)
```

```{r}
# Assuming there is a time column in your test data
results_df$Time <- as.Date(actual_df$year)
# Plot Actual vs Predicted over time
p2 <- ggplot(results_df, aes(x = Time)) +
  geom_line(aes(y = Actual,
                color = "Actual"),
            linetype = 1) +
  geom_line(aes(y = Predicted,
                color = "Predicted"),
            linewidth = 1,
            linetype = "dashed") +
  labs(title = "Actual vs Predicted Values Over Time",
       x = "Time", y = "Value") +
  scale_color_manual(name = "Legend",
                     values = c("Actual" = "navy",
                                "Predicted" = "orange"))
```

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Actual vs Predicted Values"
#|   - "Actual vs Predicted Over Time"
#| fig-alt: "Actual vs Predicted Values"
#| label: fig-act-pred
#| echo: false

p1 + theme_minimal()
p2 + theme_minimal()
```

```{r}
#| message: false
#| warning: false
#| output: false
#| eval: false
h2o.shutdown(prompt = FALSE)
```

### Example: General Infection with Keras

The following example shows how to use the `{keras}`, a package that
facilitates the creation and training of **neural networks** and **deep
learning** models in R. It provides a user-friendly interface to build
models, whether they are simple neural networks or complex deep learning
architectures, particularly useful for handling network connections.
More detailed explanation of how a neural network works can be found in

In this case the **SEIR**\index{SEIR} model (susceptible, exposed,
infected and recovered) is used, as seen in @sec-05-machine_learning
chapter, to simulate the spread of a general infection in a population
of 1000 individuals over 160 days. For some type of infections, the
latency period is significant as it is the time during which infected
individuals are not yet infectious themselves. During this period the
individual is in compartment E (for exposed).

Once the simulation of the infected population is done, the **deep
learning** model acts by training social media data to predict the
probability of infection based on social media activity. The model
results are then used to adjust the SEIR parameters based on the
predicted infection status and re-run again on new data.

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

```{r}
#| eval: false
install.packages("keras3")
keras3::install_keras(backend = "tensorflow")
```

```{r}
library(keras3)
library(deSolve)
library(ggplot2)
```

To setup the SEIR compartments we start by defining the differential
equations built by considering the rates of change of each compartment
$(S,E,I,R)$, the parameters $(\beta,\sigma,\gamma)$, the initial state
value of the population $(N = 1000)$, and the length of time in days
$(t=160)$. Finally the `ode()` function is used to solve the
differential equations and provide the output dataset.

```{r}
SEIR <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Differential equations
    dS <- -beta * S * I / N
    dE <- beta * S * I / N - sigma * E
    dI <- sigma * E - gamma * I
    dR <- gamma * I
    # Return the rates of change
    list(c(dS, dE, dI, dR))
  })
}
# Parameters
parameters <- c(beta = 0.3, # Infection rate
                sigma = 0.2, # Incubation rate
                gamma = 0.1 # Recovery rate
                )
# Initial state values
N <- 1000
initial_state <- c(S = 999, E = 1, I = 0, R = 0)
# Time points
times <- seq(0, 160, by = 1)
# Solve the model
output <- ode(y = initial_state,
              times = times,
              func = SEIR,
              parms = parameters
              )
# Convert output to a data frame
output <- as.data.frame(output)
```

The social media data, composed of features and labels, is obtained by
simulating a matrix of 1000 individuals $(n)$ over 10 $(p)$ days used
for training the model:

```{r}
set.seed(123)
n <- 1000 # number of samples
p <- 10 # number of features
# Generate random features and labels
social_features <- matrix(rnorm(n * p),
                          nrow = n, ncol = p)
infection_labels <- sample(0:1, n, replace = TRUE)
# Combine into a data frame
social_data <- data.frame(social_features)
social_data$infection <- infection_labels

head(social_data)
```

Define the model by using the `keras_model_sequential()` function, and
add layers with the `layer_dense()` and `layer_activation()` functions.
On each layer the model will apply a transformation to the input data,
and the activation function will introduce non-linearity to the model.

Here we use a simple model with just two layers and a **ReLU** and a
**Sigmoid** as activation functions. These functions have specific
shapes that allows the model to learn complex patterns in the data. Any
time one layer is done the output is passed to the next layer.

Although the `layer_dropout()` is not actually used in this application,
but only commented out for testing model adjustments, it could be added
to the model as a layer that drops out output results under specified
requirements. The function is used to prevent overfitting by randomly
setting a fraction of input units to zero during training.

The model concludes with one more `layer_dense()` with `activation` used
to finally activate the output with a tailored function. In this case a
**sigmoid**, but also a **softmax** function could be used.

The **sigmoid** function is used in binary classification problems,
where the output is a probability between 0 and 1. The **softmax**
function is used in multi-class classification problems, where the
output is a probability distribution over multiple classes.

Given an input vector $x=[x_1,x_2,...,x_p]$

1.  **First Dense Layer Transformation**:

$$
z^{(1)}=\sum_{i=1}^p{W_i^{(1)} x_i+b^{(1)}}
$$ {#eq-first_layer}

where $W \epsilon R^{1\times p}$ are the weights, $b$ is the bias, and
$z$ is the output.

2.  **First Activation (ReLU)**:

$$
a^{(1)}=ReLu(z^{(1)})=max(0,z^{(1)})
$$ {#eq-first_activation}

3.  **Second Dense Layer Transformation**:

$$
z^{(2)}=\sum_{i=1}^p{W_i^{(2)} x_i+b^{(2)}}
$$ {#eq-second_layer}

4.  **Output Activation (Sigmoid)**:

$$
a^{(2)}=Sigmoid(z^{(2)})= \frac{1} {1+e^{-z^{(2)}}}
$$ {#eq-second-activation}

```{r}
model <- keras_model_sequential(input_shape = c(p))
# simple model
model %>%
  layer_dense(units = 1) %>%
  layer_activation("relu") %>%
  # layer_dropout(rate = 0.3) %>%
  # layer_dense(units = 64) %>%
  # layer_activation('relu') %>%
  # layer_dropout(rate = 0.4) %>%
  layer_dense(units = 1, activation = "sigmoid")
```

The model is then compiled using the `compile()` function which with the
use of a **loss function** optimizes the results matching them against a
minimum required value, and specified metrics to reduce the error.

$$
y= \hat{y} + \epsilon
$$

The loss function is used to measure how well the model is performing,
in this case a **binary crossentropy** loss function is used, to match
the difference between original data and the model output, and apply
model adjustments in case the difference is too high. The formula for
binary crossentropy is:

$$
L(y,\hat{y})=-\frac{1}{n}\sum_{i=1}^n{y_i \log(\hat{y}_i)+(1-y_i) \log(1-\hat{y}_i)}
$$ The `optimizer_adam()` function, used to update the model's weights
during training, in **Keras** specifies the use of the **Adaptive Moment
Estimation (Adam)** optimization algorithm for training a neural
network. It updates the parameters of the neural network.

Then finally, the model is evaluated on performance with the
**accuracy** metric.

```{r}
# Compile the model
model %>% compile(loss = "binary_crossentropy",
                  optimizer = optimizer_adam(),
                  metrics = c("accuracy"))
```

Once the model is defined and compiled, the model is then trained using
the `fit()` function with the training data, number of **epochs**,
**batch size**, and **validation** split specified. The number of epochs
is the number of times the model will go through the training data, the
batch size is the number of samples used in each training step, and the
validation split is the fraction of the training data that will be used
for validation.

A particular object that is created in this type of models is the
`history` object, which is used to store the training history of the
model. This object contains information about the loss and accuracy of
the model on the training and validation data during training.

```{r}
#| output: false
#| message: false
#| warning: false
history <- model %>% fit(x = as.matrix(social_data[, 1:p]),
                         y = social_data$infection,
                         epochs = 30,
                         batch_size = 128,
                         validation_split = 0.2
                         )
```

The subsequent times the `history` object is created, the model is
trained with half the number of epochs and batch size. And this
specifications can further be adjusted as needed.

```{r}
#| output: false
history2 <- model %>% fit(x = as.matrix(social_data[, 1:p]),
                          y = social_data$infection,
                          epochs = 30 / 2,
                          batch_size = 128 / 2,
                          validation_split = 0.2
                          )
```

Here are two attempts of the trained model with different parameters,
the first with 30 epochs and a batch size of 128, and the second with 15
epochs and a batch size of 64. The training history of both models is
then plotted to compare their performance.

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Model 1 History"
#|   - "Model 2 History"
#| fig-alt: "Model History"
#| label: fig-model-history
#| echo: false
plot(history)
plot(history2)
```

Now that the model is trained and have provided an idea of how it works
on our data, a prediction of infection status is done new social media
data.

The model is then used to predict the probability of infection based on
the social media data. The predictions are then converted to binary
values using a threshold of 0.5, where values greater than 0.5 are
classified as 1 (infected) and values less than or equal to 0.5 are
classified as 0 (not infected).

```{r}
new_social_data <- matrix(rnorm(p * 160),
                          nrow = 160, ncol = p)
predicted_infections <- model %>%
  predict(new_social_data)
predicted_infections <- ifelse(predicted_infections > 0.5, 1, 0)
```

The adjustment is then done on the SEIR parameters based on predictions
results. The beta parameter is adjusted by multiplying it by the mean of
the predicted infections.

```{r}
adjusted_parameters <- parameters
adjusted_parameters["beta"] <-
  adjusted_parameters["beta"] * (1 + mean(predicted_infections))
```

The SEIR model is then re-run with the adjusted parameters to simulate
the spread of the infection in the population.

```{r}
adjusted_output <- ode(y = initial_state,
                       times = times,
                       func = SEIR,
                       parms = adjusted_parameters
                       )
# Convert output to a data frame
adjusted_output <- as.data.frame(adjusted_output)
```

The output shows the impact of the social media adjustments on the
spread of the infection.

```{r}
#| fig-cap: "SEIR Model with and without Social Media Adjustments"
#| fig-alt: "SEIR Model"
#| label: fig-seir
#| fig-width: 5
#| fig-height: 3

ggplot() +
  geom_line(data = output,
            aes(x = time, y = I,
                color = "Original Infections")) +
  geom_line(data = adjusted_output,
            aes(x = time, y = I,
                color = "Adjusted Infections")) +
  labs(title = "SEIR Model - Social Media Adjustments",
       y = "Infectious Population",
       color = "Scenario") +
  scale_color_manual(values = c(
    "Original Infections" = "navy",
    "Adjusted Infections" = "orange"))
```

This is just a simple example of how the `{keras}` package can be used
to train a deep learning model on social media data to predict the
probability of infection. The model can tested on using different
layers, more specification are found in the [keras
documentation](https://keras.io/).

## How to Find a New R-Package

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

Finding a new R-package can be a daunting task, given the vast number of
packages available on CRAN and other repositories. A strategy to
identify relevant packages for your specific needs would be to use some
prebuilt packages which provide search functionalities.

For example, the `package_search()` function from the `{pkgsearch}`
package takes a text string as input and uses basic text mining
techniques to search all of CRAN.

```{r}
#| message: false
#| warning: false
library(tidyverse) # for data manipulation
library(dlstats) # for package download stats
library(pkgsearch) # for searching packages
```

We search for packages related to *"excess of mortality"* and
"infectious diseases":

```{r}
excessPkg <- pkg_search(query = "excess of mortality", size = 200)
```

```{r}
head(excessPkg$maintainer_name)
```

```{r}
excessPkgShort <- excessPkg %>%
  filter(maintainer_name != "ORPHANED", score > 100) %>%
  select(score, package, downloads_last_month) %>%
  arrange(desc(downloads_last_month))

head(excessPkgShort)
```

```{r}
excess_shortList <- excessPkgShort$package
```

```{r}
excess_downloads <- cran_stats(excess_shortList)
```

```{r}
ggplot(excess_downloads, aes(end, downloads,
                             group = package, 
                             color = package)) +
  geom_line() +
  geom_point(aes(shape = package)) +
  scale_y_continuous(trans = "log2")
```

Another example is the `{cranly}` package, which provides a simple
interface to search for packages on CRAN. The package provides
comprehensive methods for cleaning up and organizing the information in
the CRAN package database, making it easier to find relevant packages
for your specific needs.

```{r}
#| message: false
#| warning: false
#| eval: false

library(cranly)
p_db <- tools::CRAN_package_db()
package_db <- clean_CRAN_db(p_db)
```

```{r}
#| echo: false
#| eval: false
saveRDS(p_db, "data/model-data/p_db.rds")
saveRDS(package_db, "data/model-data/package_db.rds")
```

```{r}
#| eval: true
#| echo: false
p_db <- readRDS("data/model-data/p_db.rds")
package_db <- readRDS("data/model-data/package_db.rds")
```

```{r}
package_db %>%
  filter(grepl("infectious diseases",
               description,
               ignore.case = TRUE)) %>%
  select(package)
```
