# Essential R Packages for Machine Learning {#sec-07-packages}

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   Overview of the most suitable R-packages
-   Combination of packages and functions
-   How to find a new R-package
:::
:::

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
options(timeout = 600)
```

## Inside and outside of the library boxes

This book provides all data analysis, visualization, and modeling in R
(a hint to Python translations can be found in @sec-appendixC), and the
universe of available packages is vast. How do you untangle among all of
them? Which one to choose? Sometimes, it is rather favorable to start by
working out of the library boxes, which means using functions provided
in the built-in packages inside a fresh R installation.

On the contrary, life has become easier with the advent of wrapping
functions and the reduction of lines of code through the use of
functions found in packages. In the past, data analysis and modeling in
R required extensive coding, often involving repetitive tasks and
lengthy scripts. However, with the introduction of packages containing
pre-built functions for common tasks, such as data manipulation,
visualization, and statistical modeling, the process has been
streamlined. These packages offer a wide range of functions that can be
easily applied to perform complex analyses with just a few lines of
code. As a result, users can achieve sophisticated results more
efficiently, without the need to write lengthy and intricate scripts
from scratch. This shift has not only simplified the process of data
analysis and modeling but has also democratized access to advanced
statistical techniques, making them more accessible to a broader
audience.

## Essential R Packages for Machine Learning

Some of the most used modeling packages available in R that are
particularly useful for machine learning modeling of the spread of
infectious diseases and evaluating DALYs variation due to infectious
disease include:

-   [**caret**](https://topepo.github.io/caret/)[@kuhn]: The caret
    package (short for **C**lassification **A**nd **RE**gression
    **T**raining) provides a unified interface for training and
    evaluating machine learning models. It supports a wide range of
    algorithms and offers convenient functions for hyperparameter tuning
    and model selection.
-   [**tidymodels**](https://www.tmwr.org/)[@silge]: tidymodels provides
    a collection of packages for modeling and machine learning tasks,
    including tidyr, dplyr, and ggplot2. It offers a consistent and tidy
    workflow for data preprocessing, model building, evaluation, and
    visualization.
-   [**mlr3**](https://mlr3book.mlr-org.com/)[@mlr-org]: mlr3 is a
    modern and extensible framework for machine learning in R. It
    provides a unified interface for modeling tasks, making it easy to
    train and evaluate a wide range of machine learning algorithms.
-   [**h2o**](https://h2o.ai/) is another powerful machine learning
    framework that can be used to predict DALYs. **`h2o`** offers a
    range of machine learning algorithms, including linear models,
    tree-based models, and ensemble methods, with efficient handling of
    large datasets and support for parallel computing.
-   [**INLA**](https://www.r-inla.org/what-is-inla)[@r-inlap]:
    Integrated Nested Laplace Approximations (INLA) is a package for
    Bayesian inference using the INLA method. It is particularly useful
    for spatial and spatio-temporal modeling, which can be relevant for
    studying the spread of infectious diseases.
-   [**fpp3**](https://otexts.com/fpp3/)[@forecast]: The fpp3 package is
    part of the Forecasting: Principles and Practice (3rd edition) book,
    offering tools for forecasting time series data. While not
    specifically focused on machine learning, it provides useful
    functions for time series analysis, which can be relevant for
    modeling disease spread over time.

These packages offer a diverse set of tools and techniques for modeling
infectious disease spread and assessing its impact on health metrics
like DALYs. These packages provide valuable insights into the dynamics
of infectious diseases and used correctly can be used to inform public
health interventions and policies. However, whether these packages are
the most suitable depends on various factors, including:

-   the specific requirements of the analysis
-   the expertise of the researcher
-   the nature of the data being used

Each package has its strengths and weaknesses, and the choice of package
often depends on factors such as:

-   the complexity of the modeling task
-   the type of data available
-   the preferred modeling approach (e.g., frequentist vs. Bayesian).

For example, `tidymodels` provides a tidy and consistent workflow for
modeling tasks, while `mlr3` offers extensive support for machine
learning algorithms and hyperparameter tuning.

It's also worth considering other packages that may be relevant for
specific aspects of the analysis, such as spatial modeling (e.g.,
`spdep`) or time series forecasting (e.g., `prophet`).

-   [**spdep**](https://cran.r-project.org/web/packages/spdep/index.html)
    is a powerful tools for spatial statistics, it provides a set of
    functions for analyzing spatial dependencies and autocorrelation in
    spatial data, particularly relevant in infectious disease modeling.
    It also incorporates spatial effects into models and allows for a
    better understanding of how infectious diseases spread across space.
-   [**prophet**](https://facebook.github.io/prophet/) is specifically
    designed for time series forecasting with an emphasis on flexibility
    and ease of use. Developed by Facebook, prophet can handle various
    time series patterns, including seasonality, holiday effects, and
    trend changes, making it suitable for modeling infectious disease
    trends over time.
-   [keras3](https://keras.posit.co/), the R interface to
    [Keras](https://keras.io/), is a high-level **neural networks** API
    written in Python. Designed to enable fast experimentation with deep
    learning models on top of
    [TensorFlow](https://tensorflow.rstudio.com/) (an open-source
    machine learning framework developed by the Google).

The features and capabilities of each package should carefully be
evaluated in relation to their specific research objectives before to
choose the one that best meets research needs and objectives. And it is
worth considering that experimenting with different packages and
techniques may be beneficial to identify the most effective approach for
a given analysis.

## Examples

### Tidymodels

We have already used this package in @sec-05-machine_learning and
@sec-06-techniques. This meta-package is particularly useful for testing
out more models with the help of the `workflow_set()` function.

### mlr3

This meta-package is particularly useful for benchmarking different
models. Here we see an example:

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

```{r}
#| message: false
#| warning: false
library(tidyverse)
data <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)%>%
  mutate(across(where(is.character),as.factor))
data%>%head
```

Load necessary libraries:

```{r}
#| message: false
#| warning: false
library(mlr3)
library(mlr3learners)
library(mlr3viz)
library(mlr3verse)
library(data.table)
#library(xgboost)
```

Define the task, the learners, and add the preprocessing steps creating
dummy variables:

```{r}
task <- TaskRegr$new(id = "dalys_rabies", 
                     backend = data, 
                     target = "dalys_rabies")


learner_lm <- lrn("regr.lm")
learner_xgboost <- lrn("regr.xgboost", 
                       nrounds = 1000, 
                       max_depth = 6, 
                       eta = 0.01)


poe <- po("encode")
task <- poe$train(list(task))[[1]]
task$data()%>%head()
```

Define a resampling strategy, and create a benchmark design:

```{r}
resampling <- rsmp("cv", folds = 5)


design <- benchmark_grid(
  tasks = task,
  learners = list(learner_lm, learner_xgboost),
  resamplings = resampling
)

design
```

Run the benchmark:

```{r}
#| message: false
#| warning: false
#| output: false
set.seed(19052024)
bmr = benchmark(design, store_models = TRUE,store_backends = TRUE)
```

```{r}
bmr$score()[,.(learner_id,task_id, iteration, regr.mse)]
```

```{r}
#| layout-ncol: 2
autoplot(bmr,measure = msr("regr.rmse"))
autoplot(bmr,measure = msr("regr.mae"))
```

Aggregate the results:

```{r}
#| message: false
#| warning: false
measures <- msrs(c("regr.rmse", "regr.mae"))
aggr <- bmr$aggregate(measures)
```

```{r}
rr1 <- aggr$resample_result[[1]]
rr2 <- aggr$resample_result[[2]]

dat <- as.data.table(rr1$prediction())%>%
mutate(learner="regr.lm")%>%
rbind(as.data.table(rr2$prediction())%>%mutate(learner="regr.xgboost"))%>%
cbind(task$data())%>%
select(truth,response,learner,year,location.Asia)
```

Plot the results:

```{r}
#| layout-ncol: 2
dat %>%
  ggplot(aes(x = truth, y = response,group=learner,color=learner)) +
  geom_line() +
  #geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Truth", y = "Response") +
  theme_minimal()

dat %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = response,group=learner,color=learner))+
  geom_point(aes(y=truth)) +
  facet_wrap(~location.Asia,
  # rename the facets based on the location column
  labeller = as_labeller(c(`1`="Asia",`0`="Global")))+
  labs(x = "Time(Year)", y = "Response") +
  theme_minimal()
```

### h2o

The h20 package is particularly useful for handling large datasets and
this is not the case here. However, we can use it to compare the results
with the previous model frameworks.

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
library(tidyverse)
data <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)%>%
  mutate(across(where(is.character),as.factor))
```

```{r}
#| message: false
#| warning: false
options(timeout = 6000)
# install.packages("h2o")
library(h2o)
```

```{r}
#| message: false
#| warning: false
#| output: false
# Initialize the H2O cluster
h2o.init()
Sys.sleep(10)
```

```{r}
#| message: false
#| warning: false
#| output: false
# Upload data to H2O
h2o_data <- as.h2o(data)

response <- "dalys_rabies"
predictors <- setdiff(names(data), response)


# Split data into training and testing sets
splits <- h2o.splitFrame(h2o_data, ratios = 0.8, seed = 1234)
train <- splits[[1]]
test <- splits[[2]]
```

Train three models: a linear regression model, a GBM model, and a random
forest model.

```{r}
#| message: false
#| warning: false
#| output: false
model_lm <- h2o.glm(
  x = predictors,
  y = response,
  training_frame = train,
  family = "gaussian"
)


model_gbm <- h2o.gbm(
  x = predictors,
  y = response,
  training_frame = train,
  validation_frame = test,
  ntrees = 1000,
  max_depth = 6,
  learn_rate = 0.01,
  seed = 1234
)

model_rf <- h2o.randomForest(
  x = predictors,
  y = response,
  training_frame = train,
  ntrees = 100,
  max_depth = 20,
  seed = 1234
)

# Evaluate linear regression model
perf_lm <- h2o.performance(model_lm, newdata = test)
perf_gbm <- h2o.performance(model_gbm, newdata = test)
perf_rf <- h2o.performance(model_rf, newdata = test)
```

```{r}
#| message: false
#| warning: false
#| echo: false

rbind(c(h2o.rmse(perf_lm),h2o.mae(perf_lm)),
c(h2o.rmse(perf_gbm),h2o.mae(perf_gbm)),
c(h2o.rmse(perf_rf),h2o.mae(perf_rf)))%>%
as.data.frame()%>%
mutate(model=c("lm","gbm","rf"))%>%
relocate(model)%>%
rename(rmse=V1,mae=V2)
```

```{r}
#| eval: false
model_gbm@model
```

```{r}
#| output: false
# Predict on test data
predictions <- h2o.predict(model_rf, newdata = test)

# Convert H2O frames to data frames for ggplot2
pred_df <- as.data.frame(predictions)
actual_df <- as.data.frame(test)

# Combine actual and predicted values
results_df <- data.frame(Actual = actual_df[, response], Predicted = pred_df[, 1])


# Plot Actual vs Predicted
p1 <- ggplot(results_df, aes(x = Actual, y = Predicted)) +
  geom_point(color = 'blue', alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = 'red', linetype = 'dashed') +
  labs(title = "Actual vs Predicted Values", x = "Actual", y = "Predicted") +
  theme_minimal()
```

```{r}
# Calculate residuals
results_df$Residuals <- results_df$Actual - results_df$Predicted

# Plot Residuals vs Predicted
p2 <- ggplot(results_df, aes(x = Predicted, y = Residuals)) +
  geom_point(color = 'blue', alpha = 0.5) +
  geom_hline(yintercept = 0, color = 'red', linetype = 'dashed') +
  labs(title = "Residuals vs Predicted Values", x = "Predicted", y = "Residuals") +
  theme_minimal()
```

```{r}
# Assuming there is a time column in your test data
results_df$Time <- as.Date(actual_df$year)

# Plot Actual vs Predicted over time
p3 <- ggplot(results_df, aes(x = Time)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +
  geom_line(aes(y = Predicted, color = "Predicted"), size = 1, linetype = 'dashed') +
  labs(title = "Actual vs Predicted Values Over Time", x = "Time", y = "Value") +
  scale_color_manual(name = "Legend", values = c("Actual" = "blue", "Predicted" = "green")) +
  theme_minimal()
```

```{r}
#| layout-ncol: 2
#| echo: false
p1
p3
```

```{r}
#| message: false
#| warning: false
h2o.shutdown(prompt = FALSE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

### Keras

```{r}
#| eval: false
install.packages("keras3")
keras3::install_keras(backend = "tensorflow")
```

```{r}
library(keras3)
library(deSolve)
library(ggplot2)
```

Define the SEIR model:

```{r}
SEIR <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Differential equations
    dS <- -beta * S * I / N
    dE <- beta * S * I / N - sigma * E
    dI <- sigma * E - gamma * I
    dR <- gamma * I
    
    # Return the rates of change
    list(c(dS, dE, dI, dR))
  })
}

# Parameters
parameters <- c(
  beta = 0.3,  # Infection rate
  sigma = 0.2, # Incubation rate
  gamma = 0.1 # Recovery rate
)

# Initial state values
N <- 1000
initial_state <- c(S = 999, E = 1, I = 0, R = 0)

# Time points
times <- seq(0, 160, by = 1)

# Solve the model
output <- ode(y = initial_state, times = times, func = SEIR, parms = parameters)

# Convert output to a data frame
output <- as.data.frame(output)
```

Example social media data (features and labels):

```{r}
set.seed(123)
n <- 1000  # number of samples
p <- 10    # number of features

# Generate random features and labels
social_features <- matrix(rnorm(n * p), nrow = n, ncol = p)
infection_labels <- sample(0:1, n, replace = TRUE)

# Combine into a data frame
social_data <- data.frame(social_features)
social_data$infection <- infection_labels
```

Define the model

```{r}
model <- keras_model_sequential(input_shape = c(p)) 

# simple model
model %>% 
  layer_dense(units = 1) %>% 
  layer_activation('relu') %>%
  #layer_dropout(rate = 0.3) %>%
  #layer_dense(units = 64) %>% 
  #layer_activation('relu') %>%
  #layer_dropout(rate = 0.4) %>%
  layer_dense( units = 1, activation = 'sigmoid')


# Compile the model
model %>% compile(
  loss = 'binary_crossentropy',
  optimizer = optimizer_adam(),
  metrics = c('accuracy')
)
```

```{r}
#| output: false
#| message: false
#| warning: false
history <- model %>% fit(
  x = as.matrix(social_data[, 1:p]),
  y = social_data$infection,
  epochs = 30, 
  batch_size = 128, 
  validation_split = 0.2
)
```

```{r}
#| output: false  
history2 <- model %>% fit(
  x = as.matrix(social_data[, 1:p]),
  y = social_data$infection,
  epochs = 30/2, 
  batch_size = 128/2, 
  validation_split = 0.2
)
```

```{r}
#| layout-ncol: 2
#| echo: false
plot(history)
plot(history2)
```

Predict infection status for new social media data

```{r}
new_social_data <- matrix(rnorm(p * 160), nrow = 160, ncol = p)
predicted_infections <- model %>% predict(new_social_data)
predicted_infections <- ifelse(predicted_infections > 0.5, 1, 0)
```

Adjust the SEIR parameters based on predictions, and re-run the model.

```{r}
adjusted_parameters <- parameters
adjusted_parameters["beta"] <- 
  adjusted_parameters["beta"] * (1 + mean(predicted_infections))


adjusted_output <- ode(y = initial_state, 
                       times = times, 
                       func = SEIR, 
                       parms = adjusted_parameters)

# Convert output to a data frame
adjusted_output <- as.data.frame(adjusted_output)
```

Plot the results:

```{r}
ggplot() +
  geom_line(data = output, aes(x = time, y = I, color = "Original Infections")) +
  geom_line(data = adjusted_output, aes(x = time, y = I, color = "Adjusted Infections")) +
  labs(y = "Infectious Population", color = "Scenario") +
  theme_minimal() +
  ggtitle("SEIR Model with and without Social Media Adjustments")
```
