# Essential R Packages for Machine Learning {#sec-07-packages}

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   Overview of the most suitable R-packages
-   Combination of packages and functions
-   How to find a new R-package
:::
:::

This chapter provides an overview of the most suitable R-packages for
machine learning. It also discusses the combination of packages and
functions to achieve the desired results. Additionally, it explores how
to find new R-packages and evaluate their suitability for a given task.

## Inside and Outside of the Library Boxes

When it comes to R, the possibilities are endless. With over 15,000
packages available on CRAN (Comprehensive R Archive Network) and
countless others on GitHub and other repositories, universe of available
packages is vast. How to untangle among all of them? Which one to
choose? Sometimes, it is rather favorable to start by working out of the
library boxes, which means using functions provided in the built-in
packages inside a fresh R installation.

However, the vast majority of the time, the best way to start is by
using the packages that are already available in the R ecosystem. These
packages are designed to make your life easier by providing pre-built
functions for common tasks. They can help you save time and effort by
automating repetitive tasks and streamlining complex analyses.

Life has become easier with the advent of wrapping functions and the
reduction of lines of code through the use of functions found in
packages. In the past, data analysis and modelling in R required
extensive coding, often involving repetitive tasks and lengthy scripts.
As a result, users can achieve sophisticated results more efficiently,
without the need to write lengthy and intricate scripts from scratch.
This shift has not only simplified the process of data analysis and
modelling but has also democratized access to advanced statistical
techniques, making them more accessible to a broader audience.

## Essential R Packages for Machine Learning

Some of the most used modelling packages available in R that are
particularly useful for machine learning modelling of the spread of
infectious diseases and evaluating DALYs variation due to infectious
disease include:

### Meta-Packages

1.  [**`{tidymodels}`**](https://www.tmwr.org/)[@silge]:
    This\index{tidymodels} framework provides a collection of packages
    for modelling and machine learning tasks, including tidyr, dplyr,
    and ggplot2. It offers a consistent and tidy workflow for data
    preprocessing, model building, evaluation, and visualization.
2.  [**`{caret}`**](https://topepo.github.io/caret/)[@kuhn]: The caret
    framework (short for **C**lassification **A**nd **RE**gression
    **T**raining) provides a unified interface for training and
    evaluating machine learning models. It supports a wide range of
    algorithms and offers convenient functions for hyperparameter tuning
    and model selection.
3.  [**`{mlr3}`**](https://mlr3book.mlr-org.com/)[@mlr-org]: This is a
    modern and extensible framework for machine learning in R. It
    provides a unified interface for modelling tasks, making it easy to
    train and evaluate a wide range of machine learning algorithms.

### Engines

Engines are the actual implementations of machine learning algorithms.
They perform the computations necessary to train and evaluate models.
Engines can be integrated into various meta-packages/frameworks to
leverage their specific algorithms and computational efficiencies. Each
engine might have its own set of parameters and functionalities, which
can be accessed through the high-level interfaces provided by
meta-packages.

1.  [`{randomForest}`](https://www.stat.berkeley.edu/~breiman/RandomForests/):
    For random forest models.
2.  [`{xgboost}`](https://github.com/dmlc/xgboost): An optimized
    distributed gradient boosting library designed to be highly
    efficient, flexible, and portable. It implements machine learning
    algorithms under the gradient boosting framework.
3.  [`{glmnet}`](https://glmnet.stanford.edu/): For lasso and ridge
    regression. Provides efficient procedures for fitting generalized
    linear models via penalized maximum likelihood.
4.  [`{nnet}`](http://www.stats.ox.ac.uk/pub/MASS4/): For neural
    networks.
5.  [`{kernlab}`](https://cran.r-project.org/web/packages/kernlab/index.html):
    For kernel-based machine learning methods.
6.  [`{e1071}`](https://cran.r-project.org/web/packages/e1071/index.html):
    For support vector machines and other functions.
7.  [`{lme4}`](https://github.com/lme4/lme4/): For linear and
    generalized linear mixed-effects models.
8.  [`{mgcv}`](https://cran.r-project.org/web/packages/mgcv/index.html):
    For generalized additive models.
9.  [`{rpart}`](https://github.com/bethatkinson/rpart): For recursive
    partitioning and regression trees.
10. [`{h2o}`](https://github.com/h2oai/h2o-3): A powerful machine
    learning framework that can be efficiently integrated with
    `{tidymodels}`. `{h2o}` offers a range of machine learning
    algorithms, including linear models, tree-based models, and ensemble
    methods, with efficient handling of large datasets and support for
    parallel computing.
11. [`{keras3}`](https://keras.posit.co/), the R interface to
    [Keras](https://keras.io/), is a deep learning technique which
    provides high-level **neural networks** API written in Python.
    Designed to enable fast experimentation with deep learning models on
    top of [`{tensorflow}`](https://github.com/rstudio/tensorflow) (an
    open-source machine learning framework developed by the Google).

### Time Series Analysis

1.  [`{forecast}`](https://pkg.robjhyndman.com/forecast/): For
    forecasting functions.
2.  [`{prophet}`](https://github.com/facebook/prophet): Specifically
    designed for time series forecasting with an emphasis on flexibility
    and ease of use. Developed by Facebook, prophet can handle various
    time series patterns, including seasonality, holiday effects, and
    trend changes, making it suitable for modelling infectious disease
    trends over time.
3.  [`{fpp3}`](https://github.com/robjhyndman/fpp3package)[@forecast]:
    This package is part of the Forecasting: Principles and Practice
    (3rd edition) book, offering data and tools for forecasting time
    series data, such as the import of the
    [`{fable}`](https://fable.tidyverts.org/) package with the `ARIMA()`
    model which is relevant for modelling disease spread over time.

### Bayesian Analysis

1.  [`{brms}`](https://github.com/paul-buerkner/brms): For Bayesian
    generalized multivariate non-linear multilevel models using Stan.
2.  [`{rstan}`](https://mc-stan.org/rstan/): For Bayesian inference
    using Stan.

### Specialized Tools

1.  [`{spdep}`](https://github.com/r-spatial/spdep/) is a powerful tools
    for spatial statistics, it provides a set of functions for analyzing
    spatial dependencies and autocorrelation in spatial data,
    particularly relevant in infectious disease modelling. It also
    incorporates spatial effects into models and allows for a better
    understanding of how infectious diseases spread across space.
2.  [`{INLA}`](https://www.r-inla.org/)[@r-inlap]: Integrated Nested
    Laplace Approximations (INLA) is a package for Bayesian inference
    using the INLA method. It is particularly useful for spatial and
    spatio-temporal modelling, which can be relevant for studying the
    spread of infectious diseases.

These packages offer a diverse set of tools and techniques for modelling
infectious disease spread and assessing its impact on health metrics
like DALYs. These packages provide valuable insights into the dynamics
of infectious diseases and can be used to inform public health
interventions and policies. However, whether these packages are the most
suitable depends on various factors, including:

-   the specific requirements of the analysis
-   the expertise of the researcher
-   the nature of the data being used

Each package has its strengths and weaknesses, and the choice of package
often depends on factors such as:

-   the complexity of the modelling task
-   the type of data available
-   the preferred modelling approach (e.g., frequentist vs. Bayesian).

For example, `tidymodels` provides a tidy and consistent workflow for
modelling tasks, while `mlr3` offers extensive support for machine
learning algorithms and hyperparameter tuning.

It's also worth considering other packages that may be relevant for
specific aspects of the analysis, such as spatial modelling (e.g.,
`spdep`) or time series forecasting (e.g., `prophet`).

The features and capabilities of each package should carefully be
evaluated in relation to their specific research objectives before to
choose the one that best meets research needs and objectives. And it is
worth considering that experimenting with different packages and
techniques may be beneficial to identify the most effective approach for
a given analysis.

## Examples

### Tidymodels

We have already used this package in @sec-05-machine_learning and
@sec-06-techniques. This meta-package is particularly useful for testing
out more models with the help of the `workflow_map()` function.

### Applied Machine Learning Using mlr3

This meta-package is particularly useful, the syntax guides the user
through the machine learning steps for benchmarking different models.
Here we see an example:

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

```{r}
#| message: false
#| warning: false
library(tidyverse)
data <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)%>%
  mutate(across(where(is.character),as.factor))
data%>%head
```

Load necessary libraries:

```{r}
#| message: false
#| warning: false
library(mlr3)
library(mlr3learners)
library(mlr3viz)
library(mlr3verse)
library(data.table)
#library(xgboost)
```

Define the task, the learners, and add the preprocessing steps creating
dummy variables:

```{r}
task <- TaskRegr$new(id = "dalys_rabies", 
                     backend = data, 
                     target = "dalys_rabies")
learner_lm <- lrn("regr.lm")
learner_xgboost <- lrn("regr.xgboost", 
                       nrounds = 1000, 
                       max_depth = 6, 
                       eta = 0.01)
poe <- po("encode")
task <- poe$train(list(task))[[1]]
task$data()%>%head()
```

Define a resampling strategy, and create a benchmark design:

```{r}
resampling <- rsmp("cv", folds = 5)
design <- benchmark_grid(
  tasks = task,
  learners = list(learner_lm, learner_xgboost),
  resamplings = resampling
  )
design
```

Run the benchmark:

```{r}
#| message: false
#| warning: false
#| output: false
set.seed(19052024)
bmr = benchmark(design, 
                store_models = TRUE,
                store_backends = TRUE)
```

```{r}
bmr$score()[,.(learner_id,
               task_id, 
               iteration, 
               regr.mse)]
```

```{r}
#| layout-ncol: 2
#| fig-cap: 
#|   - "BMR Regression RMSE"
#|   - "BMR Regression MAE"
#| fig-alt: "BMR Regression"  
#| label: fig-bmr-egression 

autoplot(bmr,measure = msr("regr.rmse"))
autoplot(bmr,measure = msr("regr.mae"))
```

Aggregate the results:

```{r}
#| message: false
#| warning: false
measures <- msrs(c("regr.rmse", "regr.mae"))
aggr <- bmr$aggregate(measures)
```

```{r}
rr1 <- aggr$resample_result[[1]]
rr2 <- aggr$resample_result[[2]]
dat <- as.data.table(rr1$prediction())%>%
  mutate(learner="regr.lm")%>%
  rbind(as.data.table(rr2$prediction())%>%
          mutate(learner="regr.xgboost"))%>%
  cbind(task$data())%>%
  select(truth,response,learner,year,location.Asia)
```

Plot the results:

```{r}
#| fig-cap: 
#|   - "Truth vs Response for lm and xgboost"
#| fig-alt: "Truth vs Response" 
#| label: fig-lm_xgboost
#| fig-width: 5
#| fig-height: 3

dat %>%
  ggplot(aes(x = truth, y = response,
             group=learner)) +
  geom_abline()+
  geom_line(aes(color=learner)) +
  geom_point(aes(color=learner)) +
  scale_color_manual(values = c("regr.lm" = "navy", 
                                "regr.xgboost" = "orange")) +
  facet_wrap(~learner)+
  labs(x = "Truth", y = "Response") +
theme(legend.position = "none")
```

### H2O.ai

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

The following example shows how to use the `{h20}` r package for running
[**H2O**](https://docs.h2o.ai/h2o/latest-stable/h2o-docs/index.html) via
its REST API. It is particularly useful for handling large datasets.
However, this is not the case seen the size of the rabies dataset, we
can use it to compare the results with the previous model frameworks.

```{r}
#| message: false
#| warning: false
#| echo: false
library(tidyverse)
data <- hmsidwR::rabies%>%
  select(-upper,-lower)%>%
  pivot_wider(names_from = measure,values_from = val)%>%
  filter(cause=="Rabies")%>%
  rename(dx_rabies=Deaths,dalys_rabies=DALYs)%>%
  select(-cause)%>%
  mutate(across(where(is.character),as.factor))
```

```{r}
#| message: false
#| warning: false
#| eval: false
options(timeout = 6000)
install.packages("h2o")
```

```{r}
#| message: false
#| warning: false
#| eval: false
library(h2o)
```

```{r}
#| message: false
#| warning: false
#| output: false
#| eval: false
# Initialize the H2O cluster
h2o.init()
```

```{r}
#| message: false
#| warning: false
#| output: false
#| eval: false
# Upload data to H2O
h2o_data <- as.h2o(data)
```

```{r}
response <- "dalys_rabies"
```

```{r}
#| eval: false
predictors <- setdiff(names(data), response)
# Split data into training and testing sets
splits <- h2o.splitFrame(h2o_data, 
                         ratios = 0.8, 
                         seed = 1234)
train <- splits[[1]]
test <- splits[[2]]
```

Train three models: a linear regression model, a GBM model, and a random
forest model.

```{r}
#| message: false
#| warning: false
#| output: false
#| eval: false
model_lm <- h2o.glm(
  x = predictors,
  y = response,
  training_frame = train,
  family = "gaussian"
  )
model_gbm <- h2o.gbm(
  x = predictors,
  y = response,
  training_frame = train,
  validation_frame = test,
  ntrees = 1000,
  max_depth = 6,
  learn_rate = 0.01,
  seed = 1234
  )
model_rf <- h2o.randomForest(
  x = predictors,
  y = response,
  training_frame = train,
  ntrees = 100,
  max_depth = 20,
  seed = 1234
  )
```

Evaluate the models:

```{r}
#| eval: false  
perf_lm <- h2o.performance(model_lm, newdata = test)
perf_gbm <- h2o.performance(model_gbm, newdata = test)
perf_rf <- h2o.performance(model_rf, newdata = test)
```

Then extract the RMSE and MAE metrics:

```         
h2o.rmse(<model>)
h2o.mae(<model>)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: false

h20.rabies.metrics <- rbind(c(h2o.rmse(perf_lm),
        h2o.mae(perf_lm)),
      c(h2o.rmse(perf_gbm),
        h2o.mae(perf_gbm)),
      c(h2o.rmse(perf_rf),
        h2o.mae(perf_rf)))%>%
  as.data.frame()%>%
  mutate(model=c("lm","gbm","rf"))%>%
  relocate(model)%>%
  rename(rmse=V1,mae=V2)

saveRDS(h20.rabies.metrics,"data/model-data/h20.rabies.metrics.rds")
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: true
h20.rabies.metrics <- readRDS("data/model-data/h20.rabies.metrics.rds")
```

```{r}
#| echo: false
h20.rabies.metrics
```

```{r}
#| eval: false
model_gbm@model
```

```{r}
#| output: false
#| eval: false
# Predict on test data
predictions <- h2o.predict(model_rf, 
                           newdata = test)
```

```{r}
#| echo: false
#| eval: false
saveRDS(predictions,"data/model-data/h20.rabies.predictions.rds")
```

```{r}
#| eval: true
#| echo: false
predictions <- readRDS("data/model-data/h20.rabies.predictions.rds")
```

Convert H2O frames to data frames for ggplot2, it requires the h2o
library to be loaded and initiated:

```{r}
#| eval: false
pred_df <- as.data.frame(predictions)
actual_df <- as.data.frame(test)
```

```{r}
#| echo: false
#| eval: false
saveRDS(pred_df,"data/model-data/h20.rabies.pred_df.rds")
saveRDS(actual_df,"data/model-data/h20.rabies.actual_df.rds")
```

```{r}
#| eval: true
#| echo: false
pred_df <- readRDS("data/model-data/h20.rabies.pred_df.rds")
actual_df <- readRDS("data/model-data/h20.rabies.actual_df.rds")
```

```{r}
# Combine actual and predicted values
results_df <- data.frame(Actual = actual_df[, response], 
                         Predicted = pred_df[, 1])
# Plot Actual vs Predicted
p1 <- ggplot(results_df, 
             aes(x = Actual, y = Predicted)) +
  geom_point(color = 'blue', alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, 
              color = 'red', linetype = 'dashed') +
  labs(title = "Actual vs Predicted Values", 
       x = "Actual", y = "Predicted") 
```

```{r}
#| layout-ncol: 2
#| fig-cap: 
#|   - "Residuals vs Predicted Values"
#|   - "Normality Test of the Residulas"
#| fig-alt: "Residuals Distribution"
#| label: fig-h20-residuals
# Calculate residuals
results_df$Residuals <- results_df$Actual - results_df$Predicted
# Plot Residuals vs Predicted
ggplot(results_df, 
             aes(x = Predicted, y = Residuals)) +
  geom_point(color = 'navy', alpha = 0.5) +
  geom_hline(yintercept = 0, 
             color = 'orange', linetype = 'dashed') +
  labs(title = "Residuals vs Predicted Values", 
       x = "Predicted", y = "Residuals")
# Normality Test of the Residuals
qqnorm(results_df$Residuals); 
qqline(results_df$Residuals, col = 2)
```

```{r}
# Assuming there is a time column in your test data
results_df$Time <- as.Date(actual_df$year)
# Plot Actual vs Predicted over time
p2 <- ggplot(results_df, aes(x = Time)) +
  geom_line(aes(y = Actual, 
                color = "Actual"), 
                linetype = 1) +
  geom_line(aes(y = Predicted, 
                color = "Predicted"), 
                linewidth = 1, 
                linetype = 'dashed') +
  labs(title = "Actual vs Predicted Values Over Time", 
       x = "Time", y = "Value") +
  scale_color_manual(name = "Legend", 
                     values = c("Actual" = "navy", 
                                "Predicted" = "orange")) 
```

```{r}
#| layout-ncol: 2
#| fig-cap: 
#|   - "Actual vs Predicted Values"
#|   - "Actual vs Predicted Over Time"
#| fig-alt: "Actual vs Predicted Values"
#| label: fig-act-pred  
#| echo: false

p1+theme_minimal()
p2+theme_minimal()
```

```{r}
#| message: false
#| warning: false
#| output: false
#| eval: false
h2o.shutdown(prompt = FALSE)
```

### Keras

The following example shows how to use the `{keras}` r package for deep
learning. It is particularly useful for handling large datasets.
However, this is not the case seen the size of the rabies dataset, we
can use it to compare the results with the previous model frameworks.

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

```{r}
#| eval: false
install.packages("keras3")
keras3::install_keras(backend = "tensorflow")
```

```{r}
library(keras3)
library(deSolve)
library(ggplot2)
```

Define the SEIR model:

```{r}
SEIR <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Differential equations
    dS <- -beta * S * I / N
    dE <- beta * S * I / N - sigma * E
    dI <- sigma * E - gamma * I
    dR <- gamma * I
    # Return the rates of change
    list(c(dS, dE, dI, dR))
  })
}
# Parameters
parameters <- c(
  beta = 0.3,  # Infection rate
  sigma = 0.2, # Incubation rate
  gamma = 0.1 # Recovery rate
)
# Initial state values
N <- 1000
initial_state <- c(S = 999, E = 1, I = 0, R = 0)
# Time points
times <- seq(0, 160, by = 1)
# Solve the model
output <- ode(y = initial_state, 
              times = times, 
              func = SEIR, 
              parms = parameters)
# Convert output to a data frame
output <- as.data.frame(output)
```

Example social media data (features and labels):

```{r}
set.seed(123)
n <- 1000  # number of samples
p <- 10    # number of features
# Generate random features and labels
social_features <- matrix(rnorm(n * p), 
                          nrow = n, ncol = p)
infection_labels <- sample(0:1, n, replace = TRUE)
# Combine into a data frame
social_data <- data.frame(social_features)
social_data$infection <- infection_labels
```

Define the model

```{r}
model <- keras_model_sequential(input_shape = c(p)) 
# simple model
model %>% 
  layer_dense(units = 1) %>% 
  layer_activation('relu') %>%
  #layer_dropout(rate = 0.3) %>%
  #layer_dense(units = 64) %>% 
  #layer_activation('relu') %>%
  #layer_dropout(rate = 0.4) %>%
  layer_dense( units = 1, activation = 'sigmoid')
# Compile the model
model %>% compile(
  loss = 'binary_crossentropy',
  optimizer = optimizer_adam(),
  metrics = c('accuracy')
)
```

```{r}
#| output: false
#| message: false
#| warning: false
history <- model %>% fit(
  x = as.matrix(social_data[, 1:p]),
  y = social_data$infection,
  epochs = 30, 
  batch_size = 128, 
  validation_split = 0.2
  )
```

```{r}
#| output: false  
history2 <- model %>% fit(
  x = as.matrix(social_data[, 1:p]),
  y = social_data$infection,
  epochs = 30/2, 
  batch_size = 128/2, 
  validation_split = 0.2
  )
```

```{r}
#| layout-ncol: 2
#| fig-cap: 
#|   - "Model 1 History"
#|   - "Model 2 History"
#| fig-alt: "Model History"
#| label: fig-model-history   
#| echo: false
plot(history)
plot(history2)
```

Predict infection status for new social media data and adjust the SEIR
parameters based on predictions.

```{r}
new_social_data <- matrix(rnorm(p * 160), 
                          nrow = 160, ncol = p)
predicted_infections <- model %>% 
  predict(new_social_data)
predicted_infections <- ifelse(predicted_infections > 0.5, 1, 0)
```

Adjust the SEIR parameters based on predictions, and re-run the model.

```{r}
adjusted_parameters <- parameters
adjusted_parameters["beta"] <- 
  adjusted_parameters["beta"] * (1 + mean(predicted_infections))
adjusted_output <- ode(y = initial_state, 
                       times = times, 
                       func = SEIR, 
                       parms = adjusted_parameters)
# Convert output to a data frame
adjusted_output <- as.data.frame(adjusted_output)
```

Plot the results of the SEIR model with and without social media
adjustments.

```{r}
#| fig-cap: "SEIR Model with and without Social Media Adjustments"
#| fig-alt: "SEIR Model"
#| label: fig-seir
#| fig-width: 5
#| fig-height: 3

ggplot() +
  geom_line(data = output, 
            aes(x = time, y = I, 
                color = "Original Infections")) +
  geom_line(data = adjusted_output, 
            aes(x = time, y = I, 
                color = "Adjusted Infections")) +
  labs(title= "SEIR Model with and without Social Media Adjustments",
       y = "Infectious Population", 
       color = "Scenario") +
scale_color_manual(values = c("Original Infections" = "navy", 
                              "Adjusted Infections" = "orange")) 
```

## How to Find a New R-Package

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

Finding a new R-package can be a daunting task, given the vast number of
packages available on CRAN and other repositories. A strategy to
identify relevant packages for your specific needs would be to use some
prebuilt packages which provide search functionalities.

For example, the `package_search()` function from the `{pkgsearch}`
package takes a text string as input and uses basic text mining
techniques to search all of CRAN.

```{r}
#| message: false
#| warning: false
library(tidyverse)  # for data manipulation
library(dlstats)    # for package download stats
library(pkgsearch)  # for searching packages
```

We search for packages related to *"excess of mortality"* and
"infectious diseases":

```{r}
excessPkg<-pkg_search(query="excess of mortality",size=200)
```

```{r}
head(excessPkg$maintainer_name)
```

```{r}
excessPkgShort <- excessPkg %>% 
  filter(maintainer_name != "ORPHANED", score > 100) %>%
  select(score, package, downloads_last_month) %>%
  arrange(desc(downloads_last_month))
head(excessPkgShort)
```

```{r}
excess_shortList <- excessPkgShort$package
```

```{r}
excess_downloads <- cran_stats(excess_shortList)
```

```{r}
ggplot(excess_downloads, 
       aes(end, downloads, 
           group=package, color=package)) +
  geom_line() + 
  geom_point(aes(shape=package)) +
  scale_y_continuous(trans = 'log2')
```

Another example is the `{cranly}` package, which provides a simple
interface to search for packages on CRAN. The package provides
comprehensive methods for cleaning up and organizing the information in
the CRAN package database, making it easier to find relevant packages
for your specific needs.

```{r}
#| message: false
#| warning: false
#| eval: false

library(cranly)
p_db <- tools::CRAN_package_db()
package_db <- clean_CRAN_db(p_db)
```

```{r}
#| echo: false
#| eval: false
saveRDS(p_db,"data/model-data/p_db.rds")
saveRDS(package_db,"data/model-data/package_db.rds")
```

```{r}
#| eval: true
#| echo: false
p_db <- readRDS("data/model-data/p_db.rds")
package_db <- readRDS("data/model-data/package_db.rds")
```

```{r}
package_db %>%
  filter(grepl("infectious diseases",
               description,
               ignore.case = TRUE))%>%
  select(package)
```
