# Predictive Modelling and Beyond {#sec-08-predictions}

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   How to use the Predict function
-   Evaluate prediction results
-   Improve predictions
:::
:::

## Prediction about the Future

In the dynamic landscape of public health, the ability to forecast and
predict future trends is of paramount importance. The "Predictions"
chapter provides an overview of predictive modelling, leveraging time
series analysis and mixed models to anticipate the trajectory of
infectious diseases and health metrics. Time series analysis serves as a
powerful tool for understanding and forecasting temporal patterns in
health data. By examining historical trends and patterns, researchers
can identify underlying dynamics and make informed predictions about
future outcomes. Techniques such as ARIMA (AutoRegressive Integrated
Moving Average) models offer a systematic approach to modelling
time-dependent data, capturing seasonal variations, trends, and
irregularities to generate accurate forecasts.

Furthermore, mixed models provide a versatile framework for
incorporating various sources of variability and correlation within the
data. By combining fixed effects with random effects, mixed models
accommodate complex data structures, such as hierarchical or
longitudinal data, while accounting for individual-level and group-level
factors that may influence the outcome of interest. Throughout this
chapter, we explore the application of these predictive modelling
techniques in the context of infectious disease spread and health metric
estimation. We demonstrate how time series analysis and mixed models can
be tailored to address specific research questions, such as predicting
disease outbreaks, estimating disease burden, and assessing the impact
of interventions on population health.

From forecasting the trajectory of emerging pathogens to evaluating the
long-term implications of public health policies, predictive modelling
offers valuable insights into the future of global health.

The **Social Demographic Index (SDI)** is an index that considers the
social and economic background conditions influencing health outcomes in
specific locations. In terms of calculation it is composed of the
geometric mean of the total fertility rate (TFR) for those younger than
25 years old (TFU25), the mean education for those 15 years old and
older (EDU15+), and lag-distributed income (LDI) per capita. It ranges
from 0 to 1 but it is usually multiplied by 100 for a scale of 0 to 100.

We use the **Social Demographic Index (SDI)** as predictor in our model:

$$logit(y) = \alpha + \beta SDI$$

### Time Series Analysis

Time series analysis is a statistical technique used to analyze temporal
data, where observations are collected at regular intervals over time.
It involves studying the patterns, trends, and dependencies present in
the data to make forecasts or infer relationships. Time series data
often exhibit inherent characteristics such as trend, seasonality,
cyclic patterns, and irregular fluctuations, which can be explored using
various methods such as decomposition, smoothing, and modelling. This
analysis is widely used in fields like economics, finance, epidemiology,
and environmental science to understand and predict future trends,
identify anomalies, and make informed decisions based on historical
patterns.

**Decomposition methods** play a crucial role in time series analysis by
unraveling the underlying structure of temporal data. By decomposing a
time series into its constituent components, such as trend, seasonality,
and random fluctuations, these methods enable researchers to gain deeper
insights into the underlying patterns and dynamics driving the data.
Understanding the trend allows for the identification of long-term
changes or shifts in the data, while detecting seasonal patterns helps
uncover recurring fluctuations that may follow seasonal or cyclical
patterns. By separating these components, decomposition methods provide
a clearer understanding of the temporal dynamics within the data,
facilitating more accurate forecasting and predictive modelling.
Moreover, decomposing time series data can aid in anomaly detection,
trend analysis, and seasonal adjustment, making it an indispensable tool
for studying and interpreting complex temporal phenomena.

#### Modelling

Modelling time series data often requires a combination of techniques to
capture its complex nature. **Mixed models**, **splines**, and **ARIMA**
(AutoRegressive Integrated Moving Average) are powerful tools commonly
used for this purpose. Mixed models, also known as hierarchical models,
allow for the incorporation of both fixed and random effects, making
them flexible in capturing various sources of variability in the data.
Splines provide a flexible way to capture nonlinear relationships and
smooth out the data, which is particularly useful for capturing trends
or seasonal patterns. ARIMA models, on the other hand, are adept at
capturing the autocorrelation structure present in the data, including
both short-term and long-term dependencies. By integrating these
techniques, analysts can build comprehensive models that account for
different aspects of the time series data, leading to more accurate
forecasts and insights into underlying trends and patterns.

-   time series analysis with mixed effect model, spline interpolation
    and ARIMA(1,0,0)

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(hmsidwR)
library(fpp3)
```

```{r}
hmsidwR::sdi90_19 |> head()
```

```{r}
sdi90_19|>
  group_by(location)|>
  reframe(avg=round(mean(value),3))|>
  filter(location%in%c("Global","Italy","France","Germany"))
```

```{r}
sdi_fr <- sdi90_19 |>
  filter(location=="France")

sdi_fr |> head() |> str()
```

```{r}
library(tsibble)
sdi_fr_ts <- tsibble::as_tsibble(sdi_fr,index = year)

sdi_fr_ts|>
  autoplot()
```

```{r}
dcmp <- sdi_fr_ts |>
  model(stl = STL(value))

components(dcmp)|> head()
```

> "three types of time series patterns: trend, seasonality and cycles.
> When we decompose a time series into components, we usually combine
> the trend and cycle into a single trend-cycle component (often just
> called the trend for simplicity). Thus we can think of a time series
> as comprising three components: a trend-cycle component, a seasonal
> component, and a remainder component (containing anything else in the
> time series). (https://otexts.com/fpp3/decomposition.html)"

$$y_t=S_t+T_t+R_t$$

$$y_t=S_t*T_t*R_t--->log(y_t)=log(S_t)+log(T_t)+log(R_t)$$

The **remainder** component shown in the bottom panel is what is left
over when the seasonal and trend-cycle components have been subtracted
from the data.(https://otexts.com/fpp3/components.html)

```{r}
components(dcmp) |> autoplot()
```

```{r}
components(dcmp) |>
  as_tsibble() |>
  autoplot(value, colour="gray") +
  geom_line(aes(y=trend), colour = "#D55E00") 
```

##### Autocorrelation function

Time series that show *no autocorrelation* are called white noise. In
this case we have autocorrelation. So, no white noise.

```{r}
sdi_fr_ts |>
  ACF(value) |> 
  autoplot()
```

```{r}
sdi_fr_ts|>
  features(value, 
           features = unitroot_kpss)
```

Seasonal difference is not required

```{r}
sdi_fr_ts|>
  features(value, 
           features = unitroot_nsdiffs)
```

### Auto-ARIMA

to establish the best fit

```{r}
fit <- sdi_fr_ts|>
   model(ARIMA(value))

fit|>
  report()
```

```{r}
fit |>
  forecast(h=10) |>
  ggplot(aes(x=year,y=.mean))+
  geom_line(data=sdi_fr_ts,
            aes(y=value))+
  geom_line(color="purple")
```

### Partial Autocorrelations

```{r}
sdi_fr_ts|>
  PACF(value) |>
  autoplot()
```

```{r}
sdi_fr_ts|>
  gg_tsdisplay(y = value,plot_type = "partial")
```

### Model Ensembles

To improve the predictive performance, predictions of multiple single
learners can be aggregated to a **model ensemble** by combining the
forecasts of individual models. Ensemble methods can reduce the variance
and bias of the predictions, leading to more accurate and reliable
results. Ensemble techniques such as bagging, boosting, and stacking
leverage the diversity of individual models to create a stronger
collective model that outperforms its components. These methods are
widely used in machine learning and predictive modelling to enhance the
predictive power of the model and achieve better generalization
performance on unseen data.

```{r}
caf_fit <- sdi_fr_ts|>
  model(arima210 = ARIMA(value ~ pdq(2,1,0)),
        arima013 = ARIMA(value ~ pdq(0,1,3)),
        stepwise = ARIMA(value),
        search = ARIMA(value, stepwise=FALSE))
```

```{r}
caf_fit |> 
  pivot_longer(cols=everything(),
               names_to = "Model name",
               values_to = "Orders")
```

```{r}
glance(caf_fit) |> arrange(AICc) |> select(.model:BIC)
```

### ARIMA chosen is ARIMA(1,1,0)

```{r}
#| eval: false
caf_fit |>
  select(search) |>
  gg_tsresiduals()
```

```{r}
caf_fit |>
  forecast(h=5) |>
  filter(.model=='search') |>
  ggplot(aes(x=year,y=.mean))+
  geom_line(data=sdi_fr_ts,
            aes(y=value))+
  geom_line(color="purple")
```

### Ensemble Learning

A machine learning technique that combines multiple models could be used
to improve predictive performance is **ensemble learning**. It leverages
the diversity of individual models to make more accurate predictions and
reduce overfitting. Common ensemble methods include bagging, boosting,
and stacking.

-   Bagging (Bootstrap Aggregating) creates multiple models by training
    each model on a random subset of the data and then aggregating their
    predictions.
-   Boosting builds a sequence of models, where each model focuses on
    the errors of the previous model to improve prediction accuracy.
-   Stacking combines the predictions of multiple models using a
    meta-learner to generate the final prediction.

By combining the strengths of different models, ensemble learning can
enhance predictive performance and provide more robust results.
