# Predictive Modelling and Beyond {#sec-08-predictions}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   How to use models' predictions on new data
-   Time series analysis
-   Mixed models
:::
:::

In the dynamic landscape of public health, the ability to forecast and
predict future trends is essential for effective decision-making and the
implementation of timely interventions. This chapter provides an
overview of **predictive modelling**\index{predictive modelling},
focusing on the challenges of applying predictions to new data, the use
of **time series analysis**\index{time series}, and **mixed
models**\index{mixed models} to anticipate the trajectory of infectious
diseases and health metrics. By exploring the underlying patterns and
analyzing historical data, we estimate the disease burden and evaluate
the impact of interventions on population health. We demonstrate how
time series analysis and mixed models can be tailored to address
specific research questions, such as predicting disease outbreaks,
estimating disease burden, and assessing the impact of interventions on
population health.

## Predictions About the Future

The previous chapters provided an attempt at estimating future outcomes
based on the application of various type of statistical and machine
learning techniques. Here we look at how to use these predictions on new
data.

When making predictions on new data, it is important to consider the
**generalizability** of the model. A model that performs well on the
training data may not necessarily generalize well to new, unseen data.
To evaluate the performance of the model we used the
**cross-validation** technique, but there are more such as **hold-out
validation**, **k-fold cross-validation**, **leave-one-out
cross-validation**, and **bootstrapping**. These methods help assess the
model's ability to make accurate predictions on data not used during
training, providing a more reliable estimate of the model's performance
in real-world scenarios.

Once the model has been trained and evaluated, it is used to make
predictions on new data. This process involves applying the model to the
new data to generate forecasts or estimates of the outcome of interest.
For example, in the case of infectious diseases, predictive models can
be used to forecast the trajectory of an outbreak, estimate the number
of cases, or evaluate the impact of interventions on disease
transmission. By leveraging historical data and the insights gained from
the model, we can effectively evaluate the model performance and make
informed decisions based on the predictions.

## Example: Dengue Test Predictions for 2017-2021

**Predictive modelling** is a powerful tool for forecasting future
trends. Here we test the Dengue's model made with `{mlr3}` meta-package
in @sec-07-packages. The model was trained on data from 1990 to 2016 and
now tested for years 2017 to 2021.

First we load the original data and name it `old_data`.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(data.table)
```

```{r}
old_data <- hmsidwR::infectious_diseases %>%
  arrange(year)%>%
  filter(cause_name == "Dengue",
         year<=2017,
         !location_name %in% c("Eswatini", "Lesotho")) %>%
  drop_na() %>%
  group_by(location_id) %>%
  select(-location_name, -cause_name)
```

Then, we load Dengue data from 2017 to 2021 and name it `new_data`.

```{r}
new_data <- hmsidwR::infectious_diseases %>%
  arrange(year)%>%
  filter(cause_name == "Dengue",
         year>=2017,
         !location_name %in% c("Eswatini", "Lesotho")) %>%
  drop_na() %>%
  group_by(location_id) %>%
  select(-location_name, -cause_name)
```

```{r}
#| echo: false
rr1 <- readRDS("data/model-data/rr1.rds")
rr2 <- readRDS("data/model-data/rr2.rds")
```

In the next step, we use the trained models from @sec-07-packages to
make predictions on the new data. The predictions are stored in
`new_pred_regr.cv_glmnet` and `new_pred_regr.xgboost`. The resampling
results: `rr1$learners` and `rr2$learners`, contain the trained models,
in particular the first of 5 folds cross validation sample is used for
the predictions.

```{r}
new_pred_regr.cv_glmnet <- 
  rr1$learners[[1]]$predict_newdata(new_data, task = rr1$task)
```

We notice that the models are doing well, when old data is compared to
the new data. The predictions are close to the actual values. Much more
can be done to improve the models, such as **hyperparameter tuning**,
**feature engineering**, and **model ensembles**, but this is a good
point to start. We could also look at the confidence intervals of the
predictions, to see how certain we are about the predictions, and so on.

This is the result of the `cv_glmnet` application on new data:

```{r}
#| fig-cap: "Dengue Predictions for 2017-2021: cv_glmnet"
#| fig-alt: "Dengue Predictions for 2017-2021: cv_glmnet"
#| label: fig-dengue-predictions-cv-glmnet
#| fig-width: 8
#| fig-height: 4
regr.cv_glmnet_new <- as.data.table(new_pred_regr.cv_glmnet) %>%
  mutate(learner = "regr.cv_glmnet") %>%
  cbind(new_data) 

regr.cv_glmnet_new %>%
  ggplot(aes(x = year)) +
  geom_line(data=old_data,aes(y = DALYs))+
  geom_line(aes(y = DALYs),color="brown") +
  geom_line(aes(y = response), linetype="dashed") +
  facet_wrap(vars(location_id),
             labeller = as_labeller(c(`182` = "Malawi",
                                      `191` = "Zambia",
                                      `169` = "Central African Republic")),
             scales = "free_y") +
  labs(title = "Dengue Predictions for 2017-2021: cv_glmnet")
```

The same steps are taken for the `xgboost`, and the result is shown
below.

```{r}
new_pred_regr.xgboost <- 
  rr2$learners[[1]]$predict_newdata(new_data, task = rr2$task)
```

```{r}
#| fig-cap: "Dengue Predictions for 2017-2021: xgboost"
#| fig-alt: "Dengue Predictions for 2017-2021: xgboost"
#| label: fig-dengue-predictions-xgboost  
#| fig-width: 8
#| fig-height: 4
regr.xgboost_new <- as.data.table(new_pred_regr.xgboost) %>%
  mutate(learner = "regr.xgboost") %>%
  cbind(new_data) 

regr.xgboost_new %>%
  ggplot(aes(x = year)) +
  geom_line(data=old_data,aes(y = DALYs))+
  geom_line(aes(y = DALYs),color="brown") +
  geom_line(aes(y = response), linetype="dashed") +
  facet_wrap(vars(location_id),
             labeller = as_labeller(c(`182` = "Malawi",
                                      `191` = "Zambia",
                                      `169` = "Central African Republic")),
             scales = "free_y") +
  labs(title = "Dengue Predictions for 2017-2021: xgboost")
```

In conclusion, predictive modelling is a valuable tool for forecasting
future trends and making informed decisions based on historical data. By
leveraging the insights gained from the models, we can anticipate the
trajectory of infectious diseases, estimate disease burden, and evaluate
the impact of interventions on population health. The models can be
further refined and optimized to improve their predictive performance
and provide more accurate forecasts. By combining predictive modelling
with time series analysis and mixed models, we can gain a comprehensive
understanding of the complex dynamics of public health and make
data-driven decisions to improve population health outcomes.

## Time Series Analysis

One important side of the analysis is that to consider the evolution of
the phenomenon in time. We can do this by using time as a factor in the
model. **Time series**\index{time series} analysis serves as a tool for
understanding and forecasting temporal patterns. Techniques such as
**ARIMA (AutoRegressive Integrated Moving Average)** models offer a
systematic approach to modelling time-dependent data, capturing seasonal
variations, trends, and irregularities to generate accurate forecasts.

**Mixed models**, on the other side, provide a versatile framework for
incorporating various sources of variability and correlation within the
data. Also known as **hierarchical models**, **multilevel models**, or
**random effects models**, are a type of statistical model that include
both fixed effects and random effects. By combining fixed effects with
random effects, mixed models accommodate complex data structures, such
as hierarchical or longitudinal data, while accounting for
individual-level and group-level factors that may influence the outcome
of interest.

To analyze temporal data, where observations are collected at regular
intervals over time, time series analysis\index{time series} allows for
studying the patterns, trends, and dependencies present in the data to
make forecasts or infer relationships. Time series data often exhibit
inherent characteristics such as trend, seasonality, cyclic patterns,
and irregular fluctuations, which can be explored using various methods
such as **decomposition**, **smoothing**, and **modelling**. This
analysis is widely used in fields like economics, finance, epidemiology,
and environmental science to understand and predict future trends,
identify anomalies, and make informed decisions based on historical
patterns.

**Decomposition methods**\index{decomposition} play a crucial role by
pulling out the underlying structure of temporal data. The components of
a time series are `trend`, `seasonality`, and `random fluctuations`.
Understanding the trend allows for the identification of long-term
changes or shifts in the data, while detecting seasonal patterns helps
uncover recurring fluctuations that may follow seasonal or cyclical
patterns. By separating these components, through decomposition methods,
it is possible to discover the temporal dynamics within the data,
facilitating more accurate forecasting and predictive modelling.
Moreover, decomposing time series data can aid in anomaly detection,
trend analysis, and seasonal adjustment, making it fundamental tool for
interpreting complex temporal phenomena.

Modelling time series data often requires a combination of techniques to
capture its complex nature. **Mixed models**\index{mixed models},
**splines**\index{splines}, and **ARIMA**\index{ARIMA} are powerful
tools commonly used for this purpose. Splines provide a flexible way to
capture nonlinear relationships and smooth out the data, which is
particularly useful for capturing trends or seasonal patterns. ARIMA
models, on the other hand, are adept at capturing the autocorrelation
structure present in the data, including both short-term and long-term
dependencies.

Time series analysis can also be performed on estimated values produced
by a machine learning model. This involves using vectors of estimates,
actual values (ground truth), and time (such as years) to apply a time
series model. By incorporating these elements, the time series analysis
can help identify trends, seasonal patterns, and other temporal
structures within the data. This approach enhances the robustness and
accuracy of the predictions by combining the strengths of machine
learning models with traditional time series techniques.

## Example: SDI Time Series Analysis with Mixed Effect Models

From forecasting the trajectory of emerging pathogens to evaluating the
long-term implications of public health policies, predictive modelling
offers valuable insights into the future of global health. As seen in
@sec-04-causes_risks, the consideration of other factors such as the
**Social Demographic Index (SDI)**\index{SDI} can provide a
comprehensive understanding of the impact social and economic
determinants can have on health outcomes.

As an index, SDI is composed of the geometric mean of the total
fertility rate (TFR) for those younger than 25 years old (TFU25), the
mean education for those 15 years old and older (EDU15+), and
lag-distributed income (LDI) per capita. It ranges from 0 to 1 but it is
usually multiplied by 100 for a scale of 0 to 100.

**Standard Calculation of TFR**

$$\text{TFR} = \sum (\text{ASFR}_i \times 5)$$ {#eq-tfr}

where $\text{ASFR}_i$ represents the age-specific fertility rates for
age group $i$ , and the sum is typically calculated over all
childbearing age groups (often 5-year intervals from ages 15-49).

**Standard Calculation of SDI**

The SDI is calculated as the geometric mean of these three components:

$$SDI = \sqrt[3]{TFU25 \times EDU15+ \times LDI}$$ {#eq-sdi}

Each component (TFU25, EDU15+, and LDI) is normalized before taking the
geometric mean to ensure they are on a comparable scale.

In summary, the TFR formula is specific to calculating fertility rates,
while the SDI formula incorporates a specific subset of the TFR (TFU25)
along with education and income metrics to create a broader
socio-demographic index.

In general, the level of this index helps identify key drivers in the
development of health outcomes. SDI can be used as a predictor in
predictive models to estimate the burden of diseases, mortality rates,
and other health metrics. For example, the relationship between SDI and
the incidence of a disease can be modeled using a logistic regression
model:

$$ \log \left( \frac{p}{1 - p} \right) = \beta_0 + \beta_1 \times \text{SDI} + \beta_2 \times X_2 + \cdots + \beta_k \times X_k$$ {#eq-logit-sdi}

Where, $p$ is the probability of the incidence of the disease,
$\log \left( \frac{p}{1 - p} \right)$ is the log-odds (logit) of the
disease incidence, $\beta_0$ is the intercept term,
$\beta_1, \beta_2, \ldots, \beta_k$ are the coefficients for the
predictor variables, and $X_2, \ldots, X_k$ are other potential
predictor variables that might influence the disease incidence.

To convert the `log-odds` back to the probability of disease incidence
given the value of SDI:

$$p = \frac{1}{1 + e^{-(\beta_0 + \beta_1 \times \text{SDI})}}$$ {#eq-prob-sdi}

In this example we use the SDI values from 1990 to 2019, for 3 location
plus the Global region and evaluate the difference in SDI average across
the time.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(hmsidwR)
library(fpp3)
```

```{r}
hmsidwR::sdi90_19 |> head()
```

```{r}
sdi90_19 |>
  filter(location %in% c("Global", "Italy", 
                         "France", "Germany")) |>
  ggplot(aes(x = year, y = value, 
             linetype = location)) +
  geomtextpath::geom_textline(aes(label = location), 
                              hjust = 0, vjust = 0) +
  labs(title = "Social Demographic Index (SDI) from 1990 to 2019")+
  theme(legend.position = "none")
  
```

```{r}
sdi90_19 |>
  group_by(location) |>
  reframe(avg = round(mean(value), 3)) |>
  filter(location %in% c("Global", "Italy", "France", "Germany"))
```

Then selecting only France as location to analyze the time series using
spline interpolation and ARIMA(1,0,0), we can decompose the time series
into its components and analyze the autocorrelation function.

```{r}
sdi_fr <- sdi90_19 |>
  filter(location == "France")

sdi_fr |>
  head() |>
  str()
```

```{r}
#| fig-cap: "Social Demographic Index (SDI) in France"
#| fig-subcap: "Time Series Plot"
#| fig-alt: "Social Demographic Index (SDI) in France"
#| label: fig-sdi-france
#| fig-width: 5
#| fig-height: 3
#| message: false
#| warning: false

library(tsibble)
sdi_fr_ts <- tsibble::as_tsibble(sdi_fr, index = year)

sdi_fr_ts |> autoplot()
```

```{r}
dcmp <- sdi_fr_ts |>
  model(stl = STL(value))

components(dcmp) |> head()
```

> "...three types of time series patterns: trend, seasonality and
> cycles. When we decompose a time series into components, we usually
> combine the trend and cycle into a single trend-cycle component (often
> just called the trend for simplicity). Thus we can think of a time
> series as comprising three components: a trend-cycle component, a
> seasonal component, and a remainder component (containing anything
> else in the time series). *Rob J Hyndman*"

$$y_t=S_t+T_t+R_t$$ {#eq-additive}
$$y_t=S_t*T_t*R_t$$ {#eq-multiplicative}
$$log(y_t)=log(S_t)+log(T_t)+log(R_t)$$ {#eq-log-decomposition}

The **remainder** component shown in the bottom panel is what is left
over when the seasonal and trend-cycle components have been subtracted
from the data.

```{r}
#| layout-ncol: 2
#| label: fig-decomposition
#| fig-cap:
#|   - "Components of SDI in France"
#|   - "Trend line of SDI in France"
#| fig-alt: "Components of SDI in France"

components(dcmp) |> autoplot()

components(dcmp) |>
  as_tsibble() |>
  autoplot(value) +
  geom_line(aes(y = trend), 
            linetype="dashed")+
labs(title = "Trend line of SDI in France")
```

### Autocorrelation function

Time series that show *no autocorrelation* are called white noise. In
this case we have autocorrelation. So, no white noise.

```{r}
#| fig-cap: "Autocorrelation Function of SDI in France"
#| fig-alt: "Autocorrelation Function of SDI in France"
#| label: fig-acf
#| fig-width: 5
#| fig-height: 3

sdi_fr_ts |>
  ACF(value) |>
  autoplot()
```

```{r}
sdi_fr_ts |>
  features(value,
           features = unitroot_kpss)
```

Seasonal difference is not required

```{r}
sdi_fr_ts |>
  features(value,
           features = unitroot_nsdiffs)
```

**Auto-ARIMA** model is used to establish the best fit:

if needed `install.packages("urca")`

```{r}
fit <- sdi_fr_ts |>
  model(ARIMA(value))

fit |> report()
```

```{r}
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "Forecast of SDI in France"
#| fig-alt: "Forecast of SDI in France"
#| label: fig-forecast

fit |>
  forecast(h = 10) |>
  ggplot(aes(x = year, y = .mean)) +
  geom_line(data = sdi_fr_ts,
            aes(y = value)) +
  geom_line(color = "purple", linetype = "dashed")+
  labs(title = "Forecast of SDI in France")
```

### Partial Autocorrelations

```{r}
#| layout-ncol: 2
#| fig-cap: "Partial Autocorrelations of SDI in France"
#| fig-alt: "Partial Autocorrelations of SDI in France"
#| label: fig-pacf

sdi_fr_ts |>
  PACF(value) |>
  autoplot()

sdi_fr_ts |>
  gg_tsdisplay(y = value, plot_type = "partial")
```

### Model Ensembles

To improve the predictive performance, predictions of multiple single
learners can be aggregated to a **model ensemble**\index{model ensemble}
by combining the forecasts of individual models.

**Ensemble learning**\index{ensemble learning} is a machine learning
technique, that leverages the diversity of individual models to make
more accurate predictions and reduce overfitting. Ensemble methods can
reduce the variance and bias of the predictions, leading to more
accurate and reliable results. Ensemble techniques such as
bagging\index{bagging}, boosting\index{boosting}, and
stacking\index{stacking} leverage the diversity of individual models to
create a stronger collective model that outperforms its components.
These methods are widely used in machine learning and predictive
modelling to enhance the predictive power of the model and achieve
better generalization performance on unseen data.

In case is required install the `urca` package:
```{r}
#| eval: false
install.packages("urca")
```


```{r}
caf_fit <- sdi_fr_ts |>
  model(
    arima210 = ARIMA(value ~ pdq(2, 1, 0)),
    arima013 = ARIMA(value ~ pdq(0, 1, 3)),
    stepwise = ARIMA(value),
    search = ARIMA(value, stepwise = FALSE)
    )
```

```{r}
caf_fit |>
  pivot_longer(cols = everything(),
               names_to = "Model name",
               values_to = "Orders")
```

```{r}
glance(caf_fit) |>
  arrange(AICc) |>
  select(.model:BIC)
```

Best ARIMA model selected is ARIMA(1,1,0):

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Residuals of ARIMA Models"
#|   - "Forecast of SDI in France"
#| fig-alt: "Residuals of ARIMA Models"
#| label: fig-residuals

caf_fit |>
  select(search) |>
  gg_tsresiduals()


caf_fit |>
  forecast(h = 5) |>
  filter(.model == "search") |>
  ggplot(aes(x = year, y = .mean)) +
  geom_line(data = sdi_fr_ts,
            aes(y = value)) +
  geom_line(color = "purple", linetype = "dashed")
```

## Mixed Models

Mixed models are a powerful statistical tool for analyzing data with a
hierarchical or nested structure, where observations are grouped within
higher-level units. By incorporating both fixed and random effects,
mixed models can account for the variability within and between groups,
providing a flexible framework for modelling complex data structures.
Mixed models are widely used in various fields, and particularly useful
in fields like healthcare and infectious diseases where data might be
collected across different subjects or time points.

### Mixed-Effects Models in Estimating YLDs

Mixed-effects models are particularly useful in estimating YLDs because
they can handle data that is hierarchically structured (e.g., patients
within clinics, repeated measures over time) and account for both fixed
effects (e.g., age, gender) and random effects (e.g., variability
between patients or clinics).

**Fixed Effects**\index{fixed effects}: These are the primary effects of
interest and include variables such as age, gender, year, and specific
interventions or treatments.

**Random Effects**\index{random effects}: These capture the variability
that is not explained by the fixed effects, such as differences between
patients, clinics, or regions.

By using mixed-effects models, we can better account for the
within-subject and between-subject variability, providing more accurate
estimates of YLDs.

### Different Ratios Used in Forecasting Non-Fatal Disease Burden

When forecasting the non-fatal disease burden, different mortality
ratios are used to assess the dynamics and progression of diseases. Here
are three important ratios:

**MI Ratio (Mortality to Incidence Ratio)**: $MI=M/I$

This ratio compares the number of deaths due to a disease to the number
of new cases of the disease in a specific time period. It is considered
a **severity indicator** as it indicates the **lethality of the
disease** after diagnosis. A high MI suggests that a higher proportion
of diagnosed cases result in death, indicating a severe condition.

**MP Ratio (Mortality to Prevalence Ratio)**: $MP=M/P$

This ratio compares the number of deaths due to a disease to the total
number of existing cases (both new and old) at a given time. It is used
in the **Current Risk Assessment (CRA)** as it reflects the **current
fatality risk** among those living with the disease. A high MP suggests
a significant risk of death for those currently affected. It helps in
understanding the long-term burden and survivability of a disease.

**PI Ratio (Prevalence to Incidence Ratio)**: $PI=P/I$

This ratio compares the total number of existing cases to the number of
new cases in a specific time period. As **chronicity indicator**, it
indicates how long individuals live with the disease. A high PI suggests
that individuals live longer with the disease, indicating it is more
chronic. It provides insight into the chronic nature of a disease and
the duration that individuals typically live with the disease.

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

In R, mixed-effects models can integrated with Eigen, a C++ template
library for linear algebra which significantly speed up the computations
involved in fitting mixed-effects models.

## Example: Tuberculosis YLDs Estimation Using Mixed-Effects Models

To estimate Years Lived with Disability (YLDs) using mixed-effects
models in R, we can use the `lme4` package which provides functions to
fit linear and generalized linear mixed-effects models. We will create a
hypothetical example of how to estimate YLDs with mixed-effects models.
We use the **Respiratory infections and tuberculosis** dataset found in
`{hmsidwR}` package containing YLDs, deaths, incidence and prevalence
for 2010 and 2019 to demonstrate the estimation of YLDs using
mixed-effects models.

```{r}
library(lme4)
library(tidyverse)

data <- hmsidwR::g7_hmetrics %>%
  filter(measure %in% c("Incidence",
                        "Prevalence",
                        "YLDs", 
                        "Deaths"),
         str_detect(cause, "tuberculosis"),
         metric == "Rate",
         sex == "both") %>%
  select(-metric, -sex, -cause, -upper, -lower) %>%
  pivot_wider(names_from = measure,
              values_from = val)

data %>% head()
```

```{r}
#| fig-cap: "YLDs for Tuberculosis by Location and Year"
#| fig-alt: "YLDs for Tuberculosis by Location and Year"
#| label: fig-hist-tuberculosis
#| fig-width: 5
#| fig-height: 3

library(ggpattern)
ggplot(data,
       aes(x = fct_reorder(location, YLDs),
           y = YLDs, group = year)) +
  ggpattern::geom_col_pattern(aes(pattern = factor(year)),
                              position = "dodge",
                              fill= 'white',
                              colour= 'black') +
  labs(title="YLDs for Tuberculosis by Location and Year",
       x = "Location", pattern = "Year")
    
```

```{r}
hmsidwR::disweights %>%
  filter(str_detect(sequela, "tuberculosis")) %>%
  select(cause1, severity, dw)
```

The disability weights is 0.333 for tuberculosis, and the duration
derived from the GBD 2021 was estimated based on a systematic review and
adjusted it for treated and untreated cases. The duration of
tuberculosis was estimated to be 3 years[@tiemersma2011] for untreated
cases and 6 months for treated cases.

```{r}
# Define the formula
formula <- YLDs ~ Prevalence + year + (1 | location)

# Fit the mixed-effects model
model <- lmer(formula, data = data)

model
```

```{r}
# Extract fixed effects
fixed_effects <- fixef(model)

fixed_effects
```

```{r}
# Extract random effects
random_effects <- ranef(model)

random_effects
```

```{r}
# Create new data for prediction
new_data <- data.frame(Prevalence = c(25000, 12000, 50000),
                       year = c(2020, 2021, 2022),
                       location = c("Japan", "UK", "US"))

# Predict YLDs for new data
predictions <- predict(model, 
                       new_data, 
                       allow.new.levels = TRUE
                       )
```

```{r}
#| fig-cap: "Predicted YLDs for Tuberculosis"
#| fig-alt: "Predicted YLDs for Tuberculosis"
#| label: fig-moderes-tuberculosis
#| fig-width: 5
#| fig-height: 3

# Predict YLDs for new data
predictions <- predict(model, 
                       data,
                       allow.new.levels = TRUE
                       )

data$predicted_YLD <- predictions

ggplot(data %>% filter(year == 2019,
                       !location=="Global"),
       aes(x = YLDs, y = predicted_YLD,
       group = location)) +
  geom_point(aes(shape = factor(location))) +
  geom_abline() +
  labs(title = "YLDs Predicted vs Estimated",
       x = "YLDs",
       y = "Predicted YLD",
       shape = "Location")
```

## Summary

Predictive modelling is a valuable tool for forecasting future trends
and making informed decisions based on historical data. By leveraging
the insights gained from the models, we can anticipate the trajectory of
infectious diseases, estimate disease burden, and evaluate the impact of
interventions on population health. The models can be further refined
and optimized to improve their predictive performance and provide more
accurate forecasts. By combining predictive modelling with time series
analysis and mixed models, we can gain a comprehensive understanding of
the complex dynamics of public health and make data-driven decisions to
improve population health outcomes.
