# Predictive Modelling and Beyond {#sec-08-predictions}

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   How to use the Predict function
-   Evaluate prediction results
-   Improve predictions
:::
:::

## Prediction about the Future

In the dynamic landscape of public health, the ability to forecast and
predict future trends is essential for effective decision-making and the
implementation of timely interventions. In this chapter there is an
overview of predictive modelling, in particular the use of **time series
analysis** and **mixed models**\index{mixed models} to anticipate the
trajectory of infectious diseases and health metrics. By exploring the
underlying patterns analyzing historical data to provide an estimation
of the disease burden, and evaluate the impact of interventions on
population health. We demonstrate how time series analysis and mixed
models can be tailored to address specific research questions, such as
predicting disease outbreaks, estimating disease burden, and assessing
the impact of interventions on population health.

Time series\index{time series} analysis serves as a tool for
understanding and forecasting temporal patterns in health data.
Techniques such as **ARIMA (AutoRegressive Integrated Moving Average)**
models offer a systematic approach to modelling time-dependent data,
capturing seasonal variations, trends, and irregularities to generate
accurate forecasts.

Mixed models provide a versatile framework for incorporating various
sources of variability and correlation within the data. By combining
fixed effects with random effects, mixed models accommodate complex data
structures, such as hierarchical or longitudinal data, while accounting
for individual-level and group-level factors that may influence the
outcome of interest.

From forecasting the trajectory of emerging pathogens to evaluating the
long-term implications of public health policies, predictive modelling
offers valuable insights into the future of global health.

## Social Demographic Index (SDI)

The **Social Demographic Index (SDI)**\index{SDI} is an index that
considers the social and economic background conditions influencing
health outcomes in specific locations. It is composed of the geometric
mean of the total fertility rate (TFR) for those younger than 25 years
old (TFU25), the mean education for those 15 years old and older
(EDU15+), and lag-distributed income (LDI) per capita. It ranges from 0
to 1 but it is usually multiplied by 100 for a scale of 0 to 100.

We use the **Social Demographic Index (SDI)** as predictor in our model:

$$logit(y) = \alpha + \beta SDI$$ {#eq-logit-sdi}

where $y$ is the outcome variable, $\alpha$ is the intercept, $\beta$ is
the coefficient for SDI, and $SDI$ is the Social Demographic Index.

## Time Series Analysis

To analyze temporal data, where observations are collected at regular
intervals over time, time series analysis\index{time series} involves
studying the patterns, trends, and dependencies present in the data to
make forecasts or infer relationships. Time series data often exhibit
inherent characteristics such as trend, seasonality, cyclic patterns,
and irregular fluctuations, which can be explored using various methods
such as decomposition, smoothing, and modelling. This analysis is widely
used in fields like economics, finance, epidemiology, and environmental
science to understand and predict future trends, identify anomalies, and
make informed decisions based on historical patterns.

**Decomposition methods**\index{decomposition} play a crucial role in
time series analysis by unraveling the underlying structure of temporal
data. By decomposing a time series into its constituent components, such
as trend, seasonality, and random fluctuations, these methods enable
researchers to gain deeper insights into the underlying patterns and
dynamics driving the data. Understanding the trend allows for the
identification of long-term changes or shifts in the data, while
detecting seasonal patterns helps uncover recurring fluctuations that
may follow seasonal or cyclical patterns. By separating these
components, decomposition methods provide a clearer understanding of the
temporal dynamics within the data, facilitating more accurate
forecasting and predictive modelling. Moreover, decomposing time series
data can aid in anomaly detection, trend analysis, and seasonal
adjustment, making it an indispensable tool for studying and
interpreting complex temporal phenomena.

Modelling time series data often requires a combination of techniques to
capture its complex nature. **Mixed models**\index{mixed models},
**splines**\index{splines}, and **ARIMA**\index{ARIMA} (AutoRegressive
Integrated Moving Average) are powerful tools commonly used for this
purpose. Mixed models, also known as hierarchical models, allow for the
incorporation of both fixed and random effects, making them flexible in
capturing various sources of variability in the data. Splines provide a
flexible way to capture nonlinear relationships and smooth out the data,
which is particularly useful for capturing trends or seasonal patterns.
ARIMA models, on the other hand, are adept at capturing the
autocorrelation structure present in the data, including both short-term
and long-term dependencies. By integrating these techniques, analysts
can build comprehensive models that account for different aspects of the
time series data, leading to more accurate forecasts and insights into
underlying trends and patterns.

-   time series analysis with mixed effect model, spline interpolation
    and ARIMA(1,0,0)

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(hmsidwR)
library(fpp3)
```

```{r}
hmsidwR::sdi90_19 |> head()
```

```{r}
sdi90_19 |>
  group_by(location) |>
  reframe(avg = round(mean(value), 3)) |>
  filter(location %in% c("Global", "Italy", "France", "Germany"))
```

```{r}
sdi_fr <- sdi90_19 |>
  filter(location == "France")

sdi_fr |>
  head() |>
  str()
```

```{r}
#| fig-cap: "Social Demographic Index (SDI) in France"
#| fig-subcap: "Time Series Plot"
#| fig-alt: "Social Demographic Index (SDI) in France"
#| label: fig-sdi-france
#| fig-width: 5
#| fig-height: 3

library(tsibble)
sdi_fr_ts <- tsibble::as_tsibble(sdi_fr, index = year)

sdi_fr_ts |>
  autoplot()
```

```{r}
dcmp <- sdi_fr_ts |>
  model(stl = STL(value))

components(dcmp) |> head()
```

> "...three types of time series patterns: trend, seasonality and
> cycles. When we decompose a time series into components, we usually
> combine the trend and cycle into a single trend-cycle component (often
> just called the trend for simplicity). Thus we can think of a time
> series as comprising three components: a trend-cycle component, a
> seasonal component, and a remainder component (containing anything
> else in the time series). *Rob J Hyndman*"

$$y_t=S_t+T_t+R_t$$ {#eq-additive}
$$y_t=S_t*T_t*R_t$$ {#eq-multiplicative}
$$log(y_t)=log(S_t)+log(T_t)+log(R_t)$$ {#eq-log-decomposition}

The **remainder** component shown in the bottom panel is what is left
over when the seasonal and trend-cycle components have been subtracted
from the data.

```{r}
#| layout-ncol: 2
#| label: fig-decomposition
#| fig-cap:
#|   - "Components of SDI in France"
#|   - "Trend line of SDI in France"
#| fig-alt: "Components of SDI in France"

components(dcmp) |> autoplot()

components(dcmp) |>
  as_tsibble() |>
  autoplot(value, colour = "gray") +
  geom_line(aes(y = trend), colour = "#D55E00")
```

### Autocorrelation function

Time series that show *no autocorrelation* are called white noise. In
this case we have autocorrelation. So, no white noise.

```{r}
#| fig-cap: "Autocorrelation Function of SDI in France"
#| fig-alt: "Autocorrelation Function of SDI in France"
#| label: fig-acf
#| fig-width: 5
#| fig-height: 3

sdi_fr_ts |>
  ACF(value) |>
  autoplot()
```

```{r}
sdi_fr_ts |>
  features(value,
    features = unitroot_kpss
  )
```

Seasonal difference is not required

```{r}
sdi_fr_ts |>
  features(value,
    features = unitroot_nsdiffs
  )
```

**Auto-ARIMA** model is used to establish the best fit:

if needed `install.packages("urca")`

```{r}
fit <- sdi_fr_ts |>
  model(ARIMA(value))

fit |>
  report()
```

```{r}
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "Forecast of SDI in France"
#| fig-alt: "Forecast of SDI in France"
#| label: fig-forecast

fit |>
  forecast(h = 10) |>
  ggplot(aes(x = year, y = .mean)) +
  geom_line(
    data = sdi_fr_ts,
    aes(y = value)
  ) +
  geom_line(color = "purple")
```

### Partial Autocorrelations

```{r}
#| layout-ncol: 2
#| fig-cap: "Partial Autocorrelations of SDI in France"
#| fig-alt: "Partial Autocorrelations of SDI in France"
#| label: fig-pacf

sdi_fr_ts |>
  PACF(value) |>
  autoplot()

sdi_fr_ts |>
  gg_tsdisplay(y = value, plot_type = "partial")
```

### Model Ensembles

To improve the predictive performance, predictions of multiple single
learners can be aggregated to a **model ensemble**\index{model ensemble}
by combining the forecasts of individual models. Ensemble methods can
reduce the variance and bias of the predictions, leading to more
accurate and reliable results. Ensemble techniques such as
bagging\index{bagging}, boosting\index{boosting}, and
stacking\index{stacking} leverage the diversity of individual models to
create a stronger collective model that outperforms its components.
These methods are widely used in machine learning and predictive
modelling to enhance the predictive power of the model and achieve
better generalization performance on unseen data.

```{r}
caf_fit <- sdi_fr_ts |>
  model(
    arima210 = ARIMA(value ~ pdq(2, 1, 0)),
    arima013 = ARIMA(value ~ pdq(0, 1, 3)),
    stepwise = ARIMA(value),
    search = ARIMA(value, stepwise = FALSE)
  )
```

```{r}
caf_fit |>
  pivot_longer(
    cols = everything(),
    names_to = "Model name",
    values_to = "Orders"
  )
```

```{r}
glance(caf_fit) |>
  arrange(AICc) |>
  select(.model:BIC)
```

Best ARIMA model selected is ARIMA(1,1,0):

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Residuals of ARIMA Models"
#|   - "Forecast of SDI in France"
#| fig-alt: "Residuals of ARIMA Models"
#| label: fig-residuals

caf_fit |>
  select(search) |>
  gg_tsresiduals()


caf_fit |>
  forecast(h = 5) |>
  filter(.model == "search") |>
  ggplot(aes(x = year, y = .mean)) +
  geom_line(
    data = sdi_fr_ts,
    aes(y = value)
  ) +
  geom_line(color = "purple")
```

## Mixed Models

Mixed models are a powerful statistical tool for analyzing data with a
hierarchical or nested structure, where observations are grouped within
higher-level units. By incorporating both fixed and random effects,
mixed models can account for the variability within and between groups,
providing a flexible framework for modelling complex data structures.
Mixed models are widely used in various fields, and particularly useful
in fields like healthcare and infectious diseases where data might be
collected across different subjects or time points.

### Mixed-Effects Models in Estimating YLDs for Tuberculosis

Mixed-effects models are particularly useful in estimating YLDs because
they can handle data that is hierarchically structured (e.g., patients
within clinics, repeated measures over time) and account for both fixed
effects (e.g., age, gender) and random effects (e.g., variability
between patients or clinics).

**Fixed Effects**\\index{fixed effects}: These are the primary effects
of interest and include variables such as age, gender, year, and
specific interventions or treatments.

**Random Effects**\index{random effects}: These capture the variability
that is not explained by the fixed effects, such as differences between
patients, clinics, or regions.

By using mixed-effects models, we can better account for the
within-subject and between-subject variability, providing more accurate
estimates of YLDs.

### Different Ratios Used in Forecasting Non-Fatal Disease Burden

When forecasting the non-fatal disease burden, different mortality
ratios are used to assess the dynamics and progression of diseases. Here
are three important ratios:

**MI Ratio (Mortality to Incidence Ratio)**: $MI=M/I$

This ratio compares the number of deaths due to a disease to the number
of new cases of the disease in a specific time period. It is considered
a **severity indicator** as it indicates the **lethality of the
disease** after diagnosis. A high MI suggests that a higher proportion
of diagnosed cases result in death, indicating a severe condition.

**MP Ratio (Mortality to Prevalence Ratio)**: $MP=M/P$

This ratio compares the number of deaths due to a disease to the total
number of existing cases (both new and old) at a given time. It is used
in the **Current Risk Assessment (CRA)** as it reflects the **current
fatality risk** among those living with the disease. A high MP suggests
a significant risk of death for those currently affected. It helps in
understanding the long-term burden and survivability of a disease.

**PI Ratio (Prevalence to Incidence Ratio)**: $PI=P/I$

This ratio compares the total number of existing cases to the number of
new cases in a specific time period. As **chronicity indicator**, it
indicates how long individuals live with the disease. A high PI suggests
that individuals live longer with the disease, indicating it is more
chronic. It provides insight into the chronic nature of a disease and
the duration that individuals typically live with the disease.

```{r}
#| message: false
#| warning: false
#| echo: false
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

In R, mixed-effects models can integrated with Eigen, a C++ template
library for linear algebra which significantly speed up the computations
involved in fitting mixed-effects models. To estimate Years Lived with
Disability (YLDs) using mixed-effects models in R, we can use the `lme4`
package which provides functions to fit linear and generalized linear
mixed-effects models. We will create a hypothetical example of how to
estimate YLDs with mixed-effects models. We use the **Respiratory
infections and tuberculosis** dataset found in `{hmsidwR}` package
containing YLDs, deaths, incidence and prevalence for 2010 and 2019 to
demonstrate the estimation of YLDs using mixed-effects models.

```{r}
library(lme4)
library(tidyverse)

data <- hmsidwR::g7_hmetrics %>%
  filter(
    measure %in% c(
      "Incidence",
      "Prevalence",
      "YLDs", "Deaths"
    ),
    str_detect(cause, "tuberculosis"),
    metric == "Rate",
    sex == "both"
  ) %>%
  select(
    -metric, -sex,
    -cause, -upper, -lower
  ) %>%
  pivot_wider(
    names_from = measure,
    values_from = val
  )
data %>% head()
```

```{r}
#| fig-cap: "YLDs for Tuberculosis by Location and Year"
#| fig-alt: "YLDs for Tuberculosis by Location and Year"
#| label: fig-hist-tuberculosis
#| fig-width: 5
#| fig-height: 3

ggplot(
  data,
  aes(
    x = fct_reorder(location, YLDs),
    y = YLDs, group = year,
    fill = factor(year)
  )
) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d() +
  labs(x = "Location", fill = "Year")
```

```{r}
hmsidwR::disweights2019 %>%
  filter(str_detect(sequela, "tuberculosis")) %>%
  select(cause1, severity, dw)
```

The disability weights is 0.333 for tuberculosis, and the duration
derived from the GBD 2021 was estimated based on a systematic review and
adjusted it for treated and untreated cases. The duration of
tuberculosis was estimated to be 3 years[@tiemersma2011] for untreated
cases and 6 months for treated cases.

```{r}
# Define the formula
formula <- YLDs ~ Prevalence + year + (1 | location)

# Fit the mixed-effects model
model <- lmer(formula, data = data)
model
```

```{r}
# Extract fixed effects
fixed_effects <- fixef(model)
fixed_effects
```

```{r}
# Extract random effects
random_effects <- ranef(model)
random_effects
```

```{r}
# Create new data for prediction
new_data <- data.frame(
  Prevalence = c(25000, 12000, 50000),
  year = c(2020, 2021, 2022),
  location = c("Japan", "UK", "US")
)

# Predict YLDs for new data
predictions <- predict(model, new_data,
  allow.new.levels = TRUE
)
```

```{r}
#| fig-cap: "Predicted YLDs for Tuberculosis"
#| fig-alt: "Predicted YLDs for Tuberculosis"
#| label: fig-moderes-tuberculosis
#| fig-width: 5
#| fig-height: 3

# Predict YLDs for new data
predictions <- predict(model, data,
  allow.new.levels = TRUE
)
data$predicted_YLD <- predictions

ggplot(
  data %>% filter(year == 2019),
  aes(
    x = YLDs, y = predicted_YLD,
    group = location
  )
) +
  geom_point(aes(color = factor(location))) +
  geom_abline() +
  labs(
    title = "YLDs Predicted vs Estimated",
    x = "YLDs",
    y = "Predicted YLD",
    color = "Location"
  )
```

## Ensemble Learning

A machine learning technique that combines multiple models could be used
to improve predictive performance is **ensemble
learning**\index{ensemble learning}. It leverages the diversity of
individual models to make more accurate predictions and reduce
overfitting. Common ensemble methods include bagging, boosting, and
stacking.

-   Bagging (Bootstrap Aggregating) creates multiple models by training
    each model on a random subset of the data and then aggregating their
    predictions.
-   Boosting builds a sequence of models, where each model focuses on
    the errors of the previous model to improve prediction accuracy.
-   Stacking combines the predictions of multiple models using a
    meta-learner to generate the final prediction.

By combining the strengths of different models, ensemble learning can
enhance predictive performance and provide more robust results.
