# Introduction to Data Visualization {#sec-09-data_visualization}

```{r}
#| echo: false
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_minimal())
```

One of my favorite topics to practice with for understanding more about
data analysis is visualization. Through visualizing data, we can
summarize the information contained in it, tell a story about the
interaction of key variables, represent statistical values, and
ultimately design the background of the story. This chapter will focus
on the basics of data visualization, how to customize a plot, and how to
tell a story with data.

## Practicing Data Visualization

To practice making data visualizations, there are many free channels
available that allow to gain information on how to do it and how to
improve abilities by practicing and sharing final results. Received
feedback might be harsh, as the world can be tough, but you will improve
if you don't give up.

By participating in competitions such as `#TidyTuesday` and
`#30DayChartChallenge`, or the `#30DayMapChallenge` practicing making
maps (which we will explore further in the @sec-11-spatials chapter of
this book), will help to get better. At some point, you'll feel
proficient!

## History of Data Visualization

The history of data visualization spans centuries. The roots of data
visualization can be traced back to ancient times with rudimentary
visual representations, but its modern form began to take shape with the
growth of statistics and graphical methods in the 17th and 18th
centuries. One notable milestone was the publication of **William
Playfair**'s "Commercial and Political Atlas" in 1786, which introduced
innovative graphical techniques like the line graph, bar chart, and pie
chart.

Throughout the 19th and 20th centuries, pioneers such as **Florence
Nightingale** (1820-1910), **John Snow** (1813-1858), and **Jacques
Bertin** further advanced the field, using visualizations to communicate
complex data and uncover insights.

Data visualization was largely done manually or with the help of basic
plotting tools. One example are the hand-drown visualizations made by
**W.E.B. Du Bois** (1868-1963) in the early 20th century to illustrate
the social and economic conditions of African Americans in the United
States. These visualizations are now considered iconic examples of data
visualization and have been widely studied and digitally reproduced.

## The Digital Revolution

The digital revolution of the late 20th century, with the advent of
powerful computing technologies, enabled the creation of interactive and
dynamic visualizations. Early programming languages like Fortran and
BASIC were used to create simple plots, then the development of
specialized software like SAS and SPSS provided more robust tools for
data analysis and visualization.

The emergence of the internet and web technologies in the 1990s led to
the introduction of more sophisticated statistical software and
programming environments. Tools like R, Python, and JavaScript, along
with libraries like ggplot2, matplotlib, and D3.js, revolutionized the
field of data visualization, making it more accessible and powerful.

Today, data visualization plays a crucial role in fields like data
science, business analytics, scientific research, journalism, and public
policy. It serves as a potent tool for summarizing, communicating, and
interpreting data. In particular, in public health, tracking disease
outbreaks, understanding health trends, and communicating findings to
policymakers and the public is vital. The COVID-19 pandemic highlighted
the importance of real-time data visualization. Tools like dashboards
and interactive maps are used extensively to monitor the spread of the
virus and the effectiveness of interventions.

## The Grammar of Graphics

One of the fundamental concepts in modern data visualization is the
**Grammar of Graphics**, which provides a structured approach to
creating visualizations layer by layer. This concept is clearly
interpreted in tools like the `{ggplot2}` R package, part of the
`{tidyverse}` ecosystem, developed by **Hadley Wickham** in 2005.

The Grammar of Graphics allows for the creation of complex
visualizations by combining simple building blocks such as **data**,
**aesthetics**, and **geoms** (geometric objects) in a structured
manner. It provides a flexible framework for customizing visualizations
and supports the creation of a wide range of plots, from basic scatter
plots to intricate multi-layered visualizations.

In the previous chapters, we've introduced various types of plots using
`{ggplot2}`. Now, we'll dig deeper into these visualizations and expand
on their construction: from customizing the appearance of plots to
adding annotations, labels, and themes. We'll also explore how to create
interactive visualizations using the `{plotly}` package and how to
create maps using the `{sf}` and `{leaflet}` packages.

Example: Visualizing Lung Cancer Deaths by Age in Germany\*\*

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
options(timeout = 600)
```

```{r}
library(ggplot2)
p <- hmsidwR::germany_lungc |>
  ggplot(aes(x = age, y = dx, group=sex, color=sex)) +
  geom_line()+
  geom_point() +
  labs(title = "Lung Cancer Deaths by Age in Germany",
       x = "Age",
       y = "Deaths")
p
```

### Colors and Palettes

Choosing the right colors for your visualizations is crucial. Colors can
highlight important aspects of your data and improve readability. Many
tools provide built-in color palettes, but you can also customize your
own to match specific themes or accessibility needs.

```{r}
# Example: Customizing colors in a bar chart
p1 <- p +
  scale_color_manual(values = c("brown", "navy", "orange")) 
p1
```

### Legends and Guides

Legends and guides are essential for interpreting visualizations. They
help the audience understand what different colors, shapes, or sizes
represent in the plot.

```{r}
# Customising a legend to a plot
p2 <- p1 +
  labs(color = "Sex")

p2
```

### Plot Layouts

The layout of your plots can significantly impact their effectiveness.
Arranging multiple plots in a grid can help compare different aspects of
your data side-by-side.

```{r}
#| eval: false
# Example: Arranging multiple plots
library(gridExtra)

grid.arrange(p, p2, ncol = 2)
```

```{r}
#| echo: false
#| layout-ncol: 2
p
p2
```

### Saving as an Image

Once you've created your visualization, you might want to save it as an
image file for sharing or publication.

```{r}
#| eval: false
ggsave("line_plot.png", plot = p2, width = 6, height = 4)
```
