# Introduction to Data Visualization {#sec-09-data_visualization}

```{r}
#| echo: false
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_minimal())
```

One of my favorite topics to practice for gaining a deeper understanding
of data analysis is visualization. Data visualization is a powerful tool
that allows us to transform complex data sets into comprehensible and
insightful graphical representations. Through visualizing data, we can
summarize the information contained within it, highlight important
patterns and trends, and uncover hidden insights that might not be
apparent from raw data alone.

Visualization helps us to tell a compelling story about the interaction
of key variables, represent statistical values clearly, and ultimately
design the narrative behind the data. This chapter will focus on the
fundamentals of data visualization, including the basic principles and
techniques that are essential for creating effective and informative
visualizations. We will explore various types of plots and charts,
understanding when and how to use each type to best represent different
kinds of data.

We will also delve into customization techniques that can enhance the
clarity and aesthetic appeal of your plots. This includes modifying
colors, labels, scales, and other elements to ensure your visualizations
are not only accurate but also engaging and easy to interpret.

## Practicing Data Visualization

To practice making data visualizations, there are numerous free
resources available that provide valuable information and opportunities
to enhance your skills. By engaging with these platforms, you can
practice your abilities, share your final results, and receive feedback
from the community. While the feedback may sometimes be critical, it is
an essential part of the learning process and will help you improve if
you stay persistent.

Participating in competitions and challenges such as **#TidyTuesday**,
**#30DayChartChallenge**, and **#30DayMapChallenge** offers a great way
to refine your skills. These competitions encourage you to create and
share visualizations on various themes, pushing you to experiment with
different techniques and styles. By consistently participating in these
challenges and actively seeking feedback, you will steadily enhance your
proficiency. Over time, you will find yourself becoming more adept and
confident in your data visualization skills.

## History of Data Visualization

The history of data visualization spans centuries. The roots of data
visualization can be traced back to ancient times with rudimentary
visual representations, but its modern form began to take shape with the
growth of statistics and graphical methods in the 17th and 18th
centuries. One notable milestone was the publication of **William
Playfair**'s "Commercial and Political Atlas" in 1786, which introduced
innovative graphical techniques like the line graph, bar chart, and pie
chart.

Throughout the 19th and 20th centuries, pioneers such as **Florence
Nightingale** (1820-1910), **John Snow** (1813-1858), and **Jacques
Bertin** further advanced the field, using visualizations to communicate
complex data and uncover insights.

Data visualization was largely done manually or with the help of basic
plotting tools. One example are the hand-drown visualizations made by
**W.E.B. Du Bois** (1868-1963) in the early 20th century to illustrate
the social and economic conditions of African Americans in the United
States. These visualizations are now considered iconic examples of data
visualization and have been widely studied and digitally reproduced.

## The Digital Revolution

The digital revolution of the late 20th century, with the advent of
powerful computing technologies, enabled the creation of interactive and
dynamic visualizations. Early programming languages like Fortran and
BASIC were used to create simple plots, then the development of
specialized software like SAS and SPSS provided more robust tools for
data analysis and visualization.

The emergence of the internet and web technologies in the 1990s led to
the introduction of more sophisticated statistical software and
programming environments. Tools like R, Python, and JavaScript, along
with libraries like ggplot2, matplotlib, and D3.js, revolutionized the
field of data visualization, making it more accessible and powerful.

Today, data visualization plays a crucial role in fields like data
science, business analytics, scientific research, journalism, and public
policy. It serves as a potent tool for summarizing, communicating, and
interpreting data. In particular, in public health, tracking disease
outbreaks, understanding health trends, and communicating findings to
policymakers and the public is vital. The COVID-19 pandemic highlighted
the importance of real-time data visualization. Tools like dashboards
and interactive maps are used extensively to monitor the spread of the
virus and the effectiveness of interventions.

## The Grammar of Graphics

One of the fundamental concepts in modern data visualization is the
**Grammar of Graphics**, which provides a structured approach to
creating visualizations layer by layer. This concept is clearly
interpreted in tools like the `{ggplot2}` R package, part of the
`{tidyverse}` ecosystem, developed by **Hadley Wickham** in 2005.

The Grammar of Graphics allows for the creation of complex
visualizations by combining simple building blocks such as **data**,
**aesthetics**, and **geoms** (geometric objects) in a structured
manner. It provides a flexible framework for customizing visualizations
and supports the creation of a wide range of plots, from basic scatter
plots to intricate multi-layered visualizations.

In the previous chapters, we've introduced various types of plots using
`{ggplot2}`. Now, we'll dig deeper into these visualizations and expand
on their construction: from customizing the appearance of plots to
adding annotations, labels, and themes. We'll also explore how to create
interactive visualizations using the `{plotly}` package and how to
create maps using the `{sf}` and `{leaflet}` packages.

## General Guidelines for Data Visualization

There are many types of plots that can be used to visualize different
types of data, such as cross tabulations for categorical variables,
scatter plots for continuous variables, side-by-side box plots, and
other summaries.

Here are some specifications about the usage of common types of plots:

-   **Scatter Plots**: Used to show the relationship between two
    variables.

-   **Line Plots**: Used to show trends over time or across categories.

-   **Bar Plots**: Used to compare values across categories.

-   **Histograms**: Used to show the distribution of a single variable.

-   **Box Plots**: Used to show the distribution of a variable across
    categories.

-   **Heatmaps**: Used to show the relationship between two categorical
    variables.

-   **Network Diagrams**: Used to show the relationships between nodes
    in a network.

-   **Choropleth Maps**: Used to show the distribution of a variable
    across geographic regions.

-   **Sankey Diagrams**: Used to show the flow of data between different
    categories.

-   **Treemaps**: Used to show hierarchical data as a series of nested
    rectangles.

## Example: Visualizing Lung Cancer Deaths by Age in Germany

This is an example of visualizing lung cancer deaths by age in Germany.
It is a line plot showing the number of deaths by age group. We have
already seen how to create a line plot using `{ggplot2}` in previous
chapters. Here, we will focus on customizing the plot to make it more
visually appealing and informative.

This is a basic output, and we will enhance it by customizing the
colors, adding a legend, and adjusting the plot layout. We will also
explore how to save the plot as an image file for sharing or
publication.

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

```{r}
library(ggplot2)
p <- hmsidwR::germany_lungc |>
  ggplot(aes(x = age, y = dx, group = sex, color = sex)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Lung Cancer Deaths by Age in Germany",
    x = "Age",
    y = "Deaths"
  )
p
```

### Colors and Palettes

Choosing the right colors for your visualizations is crucial. Colors can
highlight important aspects of your data and improve readability. Many
tools provide built-in color palettes, but you can also customize your
own. Here we just add the `scale_color_manual()` function to customize
the original plot, with a new color palette.

By typing `?scale_color_manual()` in the R console, you can access the
documentation and explore other options for customizing colors.

```{r}
# Example: Customizing colors in a bar chart
p1 <- p +
  scale_color_manual(values = c("brown", "navy", "orange"))
p1
```

### Theme, Legends and Guides

Legends and guides are essential for interpreting visualizations. They
help the audience understand what different colors, shapes, or sizes
represent in the plot. Here we have just changed the title of the
legend, but much more can be done, such as changing the position, and
labels. To change the position of the legend we can use the `theme()`
function as shown below.

In the R console, typing `?theme()` will provide more information on
customizing the appearance of your plots. In this case is useful to
adjust the angle of the text in the x-axis, and even this can be done
specifying the `angle` parameter in the `theme()` function.

The `guides()` function can be used to customize the legend, such as
reversing the order of the legend items.

```{r}
# Customising a legend to a plot
p2 <- p1 +
  labs(
    color = "Sex",
    subtitle = "Line Plot",
    caption = "DataSource: hmsidwR::germany_lungc"
  ) +
  guides(color = guide_legend(reverse = TRUE)) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

p2
```

### Plot Layouts

The layout of your plots can significantly impact their effectiveness.
Arranging multiple plots in a grid can help compare different aspects of
your data side-by-side. We can use the `grid.arrange()` function from
the `gridExtra` package to arrange multiple plots in a grid layout. Or,
we can use the `layout()` function to specify the layout of the plots.
There are other packages that can be used such as `patchwork` and
`cowplot`, which provide additional functionalities for arranging plots.

```{r}
#| eval: false
# Example: Arranging multiple plots
library(gridExtra)

grid.arrange(p, p1, p2, ncol = 3)
```

```{r}
#| echo: false
#| layout-ncol: 3
p
p1
p2
```

### Saving as an Image

Once you've created your visualization, you might want to save it as an
image file for sharing or publication.

```{r}
#| eval: false
ggsave("line_plot.png", plot = p2, width = 6, height = 4)
```
