# Introduction to Data Visualisation {#sec-09-data_visualisation}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

Data visualisation\index{Data visualisation} is a powerful tool for
transforming complex datasets into clear and insightful graphical
representations. It plays a central role in **Exploratory Data Analysis
(EDA)**\index{Exploratory Data Analysis (EDA)}, helping to reveal
important patterns, trends, and hidden insights that are often obscured
in raw data. The information contained within data highlights important
patterns and trends, and uncover hidden insights that might not be
apparent from raw data alone. In the context of health
metrics\index{Health metrics} and infectious
diseases\index{Infectious diseases}, data visualisation plays a critical
role in tracking disease outbreaks, understanding health trends, and
communicating findings to policymakers and the public.

We have already seen how to create various types of plots using the
`{ggplot2}` package. In this chapter, we will discuss the history and
evolution of data visualisation, delve deeper into the topic,
introducing the concept of the **Grammar of
Graphics**\index{Grammar of Graphics}, exploring techniques for
customising plots, adding annotations, labels, and themes, and creating
interactive visualisations.

## History of Data Visualisation

The history of data visualisation spans centuries. The roots of data
visualisation can be traced back to ancient times with rudimentary
visual representations, but its modern form began to take shape with the
growth of statistics and graphical methods in the 17th and 18th
centuries. One notable milestone was the publication of **William
Playfair**'s\index{William
Playfair} **"Commercial and Political Atlas"**[@william2025] in 1786,
which introduced innovative graphical techniques like the line graph,
bar chart, and pie chart.

:::::: {.content-visible when-format="html"}
::::: columns
::: {.column width="50%"}
![William Playfair's Commercial and Political Atlas -
1786](images/09_playfair.png){#fig-playfair
fig-align="center"
fig-alt="William Playfair's Commercial and Political Atlas - 1786"}
:::

::: {.column width="50%"}
![Florence Nightingale's Coxcomb Chart -
1858](images/09_nightingale.png){#fig-nightingale
fig-align="center"
fig-alt="Florence Nightingale's Coxcomb Chart - 1858"}
:::
:::::
::::::

Throughout the 19th and 20th centuries, pioneers such as **Florence
Nightingale**\index{Florence
Nightingale} (1820-1910), **John Snow**\index{John Snow} (1813-1858),
and **Jacques Bertin** further advanced the field, using visualisations
to communicate complex data and uncover insights with his **Semiology of
Graphics** - 1967.

Data visualisation was largely done manually or with the help of basic
plotting tools. One example are the hand-drown visualisations made by
**W.E.B. Du Bois**\index{W.E.B. Du Bois} (1868-1963) in the early 20th
century to illustrate the social and economic conditions of African
Americans in the United States. These visualisations are now considered
iconic examples of data visualisation and have been widely studied and
digitally reproduced.

:::::: {.content-visible when-format="html"}
::::: columns
::: {.column width="50%"}
![John Snow's Cholera Map -
1854](images/09_johnsnow.png){#fig-johnsnow
fig-align="center" fig-alt="John Snow's Cholera Map - 1854"}
:::

::: {.column width="50%"}
![Jacques Bertin's Semiology of Graphics -
1967](images/09_bertin.png){#fig-bertin
fig-align="center"
fig-alt="Jacques Bertin's Semiology of Graphics - 1967"}
:::
:::::
::::::

**The Digital Revolution**

The digital revolution of the late 20th century, with the advent of
powerful computing technologies, enabled the creation of interactive and
dynamic visualisations. Early programming
languages\index{Programming languages !} like
Fortran\index{Programming languages ! Fortran} and
BASIC\index{Programming languages ! BASIC} were used to create simple
plots, then the development of specialised software like
SAS\index{Programming languages ! SAS} and
SPSS\index{Programming languages ! SPSS} provided more robust tools for
data analysis and visualisation.

The emergence of the internet and web technologies in the 1990s led to
the introduction of more sophisticated statistical software and
programming environments. Tools like R\index{Programming languages ! R},
Python\index{Programming languages ! Python}, and
JavaScript\index{Programming languages ! JavaScript}, along with
libraries like `ggplot2`, `matplotlib`, and `D3.js`, revolutionised the
field of data visualisation, making it more accessible and powerful.

In summary, data visualisation plays a crucial role in fields like data
science, business analytics, scientific research, journalism, and public
policy. It serves as a potent tool for communicating data while
visualising patterns, trends, and relationships that might not be
apparent from raw data alone. For instance, in public health, tracking
disease outbreaks in real time, such as with COVID19, tools like
dashboards and interactive maps are used extensively to monitor the
spread of the virus and communicate the effectiveness of interventions.

## The Grammar of Graphics

One of the fundamental concepts in modern data visualisation is the
**Grammar of Graphics**, which provides a structured approach to
creating visualisations layer by layer. This concept is clearly
interpreted in tools like the `{ggplot2}` R package, part of the
`{tidyverse}` ecosystem, developed by **Hadley
Wickham**\index{Hadley Wickham} in 2005.

The Grammar of Graphics allows for the creation of complex
visualisations by combining simple building blocks such as **data**,
**aesthetics**\index{Aesthetics}, and
**geoms**\index{geoms (geometric objects)} (geometric objects) in a
structured manner. It provides a flexible framework for customising
visualisations and supports the creation of a wide range of plots, from
basic scatter plots to intricate multi-layered visualisations.

It starts with the `ggplot()` function, which initialises the plot. The
function takes a data frame as its first argument and then additional
arguments to specify the aesthetics of the plot, such as the x and y
variables, colour, shape, and size.

```{r}
#| eval: false
ggplot()
```

The aesthetics are defined using the `aes()` function, which maps
variables in the data frame to visual properties of the plot, such as x
and y coordinates, colour, shape, and size. The `aes()` function is the
mapping part of the plot, and it can be called with the `mapping`
argument in the layers of the plot.

```{r}
#| eval: false
ggplot(data = df, 
       mapping = aes(x = x, y = y, color = z))
```

Data can also be called outside of the `ggplot()` function.

```{r}
#| eval: false
data %>%
   ggplot(aes(x = x, y = y, color = z))
```

And then additional layers are added using functions like `geom_point()`
for a scatterplot, or `geom_line()` for a line plot, and so on. Then,
with the `labs()` function, we can add titles, labels, and captions to
the plot.

```{r}
#| eval: false
ggplot(data = df, aes(x = x, y = y, color = z))+
  geom_point()+
  labs(title = "Scatter Plot")
```

The `{ggplot2}` package provides a wide range of geoms, scales, and
themes to customise the appearance of the plot. By combining these
elements, you can create visually appealing and informative
visualisations that effectively communicate your data.

It allows the user to create complex visualisations by combining simple
building blocks, with each layer of the plot allowing for an aesthetic
mapping of data to visual properties, such as colour, fill, shape, and
size, making it easy to create a wide range of visualisations, or even
new data. In this example there is also the use of an alternative to the
`labs()` function with the `ggtitle()` function, which differs in the
way the title is placed in the plot.

```{r}
#| eval: false
ggplot(data = df, aes(x = x, y = y, color = z))+
  geom_point()+
  geom_point(data = df2, aes(x = x2, y = y2, color = z2))+
  ggtitle("Scatter Plot")
```

In addition to functions provided by `{ggplot2}`, there are ggplot
**extensions**, other functions and packages that can be used to create
visualisations, such as `{plotly}`, `{ggplotly}`, `{leaflet}`, `{tmap}`,
and `{shiny}`. These tools provide additional functionality for creating
interactive plots, maps, and dashboards, allowing for more engaging and
dynamic data visualisations.

## General Guidelines for Data Visualisation

There are many types of plots that can be used to visualise different
types of data, such as cross tabulations for categorical variables,
scatter plots for continuous variables, side-by-side box plots, and
other summaries.

Here are some specifications about the usage of common types of plots:

```{r}
#| echo: false
#| eval: false
library(tidyverse)
library(gridExtra)
library(gt)

# Sample data for plots
df <- hmsidwR::infectious_diseases %>%
  filter(year>=1990)%>%
  group_by(year,location_name)%>%
  reframe(x=mean(Deaths,na.rm=T), y= mean(DALYs,na.rm=T))

ggplot2::theme_set(ggplot2::theme_void())

# Scatter Plot
scatter_plot <- ggplot(df, aes(x, y)) + 
  geom_point() + 
  ggtitle("Scatter Plot")

# Line Plot
line_plot <- ggplot(df, aes(year, y,group=location_name)) + 
  geom_line() + 
  ggtitle("Line Plot")

# Bar Plot
bar_plot <- ggplot(df, 
                   aes(x = factor(location_name), y = y)) + 
  geom_col() + 
  ggtitle("Bar Plot")

# Histogram
histogram_plot <- df %>%
  ggplot(aes(x = y)) + 
  geom_histogram() + 
  ggtitle("Histogram")

# Box Plot
box_plot <- ggplot(df, 
                   aes(x = factor(location_name), y = y)) + 
  geom_boxplot() + 
  ggtitle("Box Plot")

# Heatmap (using geom_tile for illustration purposes)
heatmap_plot <- ggplot(df, 
                       aes(x = factor(year), 
                           y = factor(location_name), 
                           fill = y)) + 
  geom_tile() +
  ggtitle("Heatmap")

# Network graph
# devtools::install_github("briatte/ggnet")
library(ggnet) 
library(network)
library(sna)
net = rgraph(10, mode = "graph", tprob = 0.5)
net = network(net, directed = FALSE)
network_plot <-ggnet2(net)+ 
  ggtitle("Network Diagram")

# Choropleth Map 
map <- map_data("world")
choropleth_plot <- ggplot(map, 
                          aes(x = long, y = lat, group = group,
                              fill = region)) + 
  geom_polygon(show.legend = F) + 
  coord_sf()+
  ggtitle("Choropleth Map")

# Sankey Diagram 
# devtools::install_github("davidsjoberg/ggsankey")
library(ggsankey) 
sankey_plot <- df %>% 
  group_by(location_name) %>%
  reframe(x=mean(x),y=mean(y)) %>%
  make_long(x,y) %>%
  ggplot(aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = factor(node))) +
  geom_sankey(show.legend = F) +
  ggtitle("Sankey Diagram")

# Treemap
library(treemapify)
treemap_plot <- df %>% 
  group_by(location_name) %>%
  reframe(x=mean(x),y=mean(y)) %>%
ggplot(aes(area = y, fill = location_name)) +
  geom_treemap()+
  ggtitle("Treemap")

# Save plots as images
ggsave("images/09-dataviz_gt_table/09-scatter_plot.png", scatter_plot)
ggsave("images/09-dataviz_gt_table/09-line_plot.png", line_plot)
ggsave("images/09-dataviz_gt_table/09-bar_plot.png", bar_plot)
ggsave("images/09-dataviz_gt_table/09-histogram_plot.png", histogram_plot)
ggsave("images/09-dataviz_gt_table/09-box_plot.png", box_plot)
ggsave("images/09-dataviz_gt_table/09-heatmap_plot.png", heatmap_plot)
ggsave("images/09-dataviz_gt_table/09-network_plot.png", network_plot)
ggsave("images/09-dataviz_gt_table/09-choropleth_plot.png", choropleth_plot)
ggsave("images/09-dataviz_gt_table/09-sankey_plot.png", sankey_plot)
ggsave("images/09-dataviz_gt_table/09-treemap_plot.png", treemap_plot)
```

```{r}
#| echo: false
#| eval: false
library(gt)

# Data for the gt table
plot_data <- data.frame(
  plot = c("images/dataviz_gt_table/09-scatter_plot.png", 
           "images/dataviz_gt_table/09-line_plot.png", 
           "images/dataviz_gt_table/09-bar_plot.png", 
           "images/dataviz_gt_table/09-histogram_plot.png", 
           "images/dataviz_gt_table/09-box_plot.png", 
           "images/dataviz_gt_table/09-heatmap_plot.png", 
           "images/dataviz_gt_table/09-network_plot.png", 
           "images/dataviz_gt_table/09-choropleth_plot.png", 
           "images/dataviz_gt_table/09-sankey_plot.png", 
           "images/dataviz_gt_table/09-treemap_plot.png"),
  type_of_plot = c("Scatter Plots", "Line Plots", 
                   "Bar Plots", "Histograms", 
                   "Box Plots", "Heatmaps", 
                   "Network Diagrams", "Choropleth Maps", 
                   "Sankey Diagrams", "Treemaps"),
  description = c("Used to show the relationship between two variables.",
                  "Used to show trends over time or across categories.",
                  "Used to compare values across categories.",
                  "Used to show the distribution of a single variable.",
                  "Used to show the distribution of a variable across categories.",
                  "Used to show the relationship between two categorical variables.",
                  "Used to show the relationships between nodes in a network.",
                  "Used to show the distribution of a variable across geographic regions.",
                  "Used to show the flow of data between different categories.",
                  "Used to show hierarchical data as a series of nested rectangles.")
)

# Create the gt table
gt_table <- gt(plot_data) %>%
  text_transform(
    locations = cells_body(c(plot)),
    fn = function(x) {
      lapply(x, function(path) {
        local_image(filename = path, height = px(100))
      })
    }) %>%
  cols_label(plot = "Plot",
             type_of_plot = "Type of Plot",
             description = "Description") %>%
  tab_options(table.width = pct(100))

# Save the gt table as an image
gt::gtsave(gt_table, "images/09_gt_table.png")
```

![Basic Data Visualisation
Rules](images/09_gt_table.png){#fig-dv-rules
fig-align="center" fig-alt="Basic Data Viz Rules"}

## Example: Visualising Lung Cancer Deaths by Age in Germany

This is an example of visualising lung cancer\index{Lung cancer} deaths
by age in Germany. It is a line plot showing the number of deaths by age
group. We have already seen how to create a line plot using `{ggplot2}`
in previous chapters, the focus here is on customising the plot to make
it more visually appealing and informative.

This is a basic output, and we will enhance it by customising patterns,
colours, legend, and adjusting the layout. We will also explore how to
save the plot as an image file for sharing or publication.

```{r}
library(ggplot2)
library(ggpattern)

scatterplot <- hmsidwR::germany_lungc |>
  ggplot(aes(x = age, y = dx)) +
  geom_point(aes(shape =sex)) + 
  labs(title = "Lung Cancer Deaths by Age in Germany",
       x = "Age",
       y = "Deaths")

lineplot <- hmsidwR::germany_lungc |>
  ggplot(aes(x = age, y = dx, group = sex)) +
  geom_line(aes(linetype=sex)) +
  geom_point() +
  labs(title = "Lung Cancer Deaths by Age in Germany",
       x = "Age",
       y = "Deaths")

barplot <- hmsidwR::germany_lungc |>
  ggplot(aes(x = age, y = dx, group = sex))  +
  ggpattern::geom_col_pattern(aes(pattern=sex),
                              position="stack",
                              fill= 'white',
                              colour= 'black') +
  labs(title = "Lung Cancer Deaths by Age in Germany",
       x = "Age",
       y = "Deaths")
```

```{r}
#| layout-ncol: 3
#| label: fig-data-viz-germany
#| fig-cap: "Scatterplot and Barplot of Lung Cancer Deaths by Age in Germany"
#| fig-subcap: 
#|  - "Scatterplot: Lung Cancer Deaths by Age in Germany"
#|  - "Lineplot: Lung Cancer Deaths by Age in Germany"
#|  - "Barplot: Lung Cancer Deaths by Age in Germany"
#| fig-alt: "Scatterplot and Barplot of Lung Cancer Deaths by Age in Germany"
#| echo: false
scatterplot
lineplot
barplot
```

All of these three plots have a grammar of graphic structure, which
means that they are built using layers of data, aesthetics, and
geometric objects. The `ggplot()` function initializes the plot, and
then we add layers using `geom_point()`, `geom_line()`, and
`geom_col()`, (or `geom_col_pattern()` in this particular example), to
create the visualisations. We can further customise these plots by
adding titles, labels, and adjusting the appearance of the plot
elements.

Finally, another way to visualise Lung Cancer Deaths by Age in Germany
is to use the `geom_col()` function and create a bar plot for each sex
category. This is a basic bar plot, and we will enhance it by adjusting
the layout. The `facet_<>` functions are used to create a grid of plots,
where each plot represents a different subset of the data. In this case,
we are using `facet_grid()` to create a grid of plots based on the `sex`
variable.

```{r}
#| label: fig-data-viz-germany-facer
#| fig-cap: "Barplots Layout - Lung Cancer Deaths by Age in Germany"
#| fig-alt: "Barplots Layout - Lung Cancer Deaths by Age in Germany"
#| fig-width: 6
#| fig-height: 4
hmsidwR::germany_lungc |>
  ggplot(aes(x = age, y = dx, group = sex))  +
  geom_col(position="identity",
           fill= 'white',
           colour= 'black') +
  # adjust the layout with three barplot one for each category
  facet_grid(sex ~ .) +
  labs(title = "Lung Cancer Deaths by Age in Germany",
       x = "Age",
       y = "Deaths") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Colors and Patterns

Choosing the right colours for your visualisations is crucial. Colors
can highlight important aspects of your data and improve readability.
Many tools provide built-in color palettes, but you can also customise
your own. Here we just add the `scale_color_manual()` function to
customise the original plot, with a new color palette.

By typing `?scale_color_manual()` in the R console, you can access the
documentation and explore other options for customising colours using
the `scale_<..>_<..>()` functions.

```{r}
#| eval: false
# Example: customising colours in a bar chart
lineplot +
  scale_color_manual(values = c("brown", "navy", "orange"))
```

In addition to colours, patterns can be used to differentiate between
the groups in the plot. The `{ggpattern}` package provides a variety of
patterns that can be used to fill the bars in a bar plot. Here we have
used the `geom_col_pattern()` function to add patterns to the bars in
the plot. An example is in show in the plot with the barplot. Another
way is to use the `linetype` or `shape` aesthetics to differentiate
between groups in a line plot or scatter plot.

### Theme, Legends and Guides

Legends and guides are essential for interpreting visualisations. They
help the audience understand what different colours, shapes, or sizes
represent in the plot. Here we have just changed the title of the
legend, but much more can be done, such as changing the position, and
labels. To change the position of the legend we can use the `theme()`
function as shown below.

In the R console, typing `?theme()` will provide more information on
customising the appearance of your plots. In this case it is useful to
adjust the angle of the text in the x-axis, and even this can be done
specifying the `angle` parameter in the `theme()` function.

The `guides()` function can be used to customise the legend, such as
reversing the order of the legend items.

```{r}
# Customising a legend to a plot
lineplot1 <- lineplot +
  labs(linetype = "Sex",
    subtitle = "Year 2019",
    caption = "DataSource: hmsidwR::germany_lungc") +
  guides(linetype = guide_legend(reverse = TRUE)) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))

lineplot2 <- lineplot1 + 
  scale_y_log10() + 
  coord_cartesian(clip = "off") +
  annotate("text", x = Inf, y=Inf, 
           hjust = 1, vjust = 0,
           label = "Log Scale")
```

### Plot Layouts

The layout of your plots can significantly impact their effectiveness.
Arranging multiple plots in a grid can help compare different aspects of
your data side-by-side. We can use the `grid.arrange()` function from
the `{gridExtra}` package to arrange multiple plots in a grid layout.
Or, we can use the `layout()` function to specify the layout of the
plots. There are other packages that can be used such as `{patchwork}`
and `{cowplot}`, which provide additional functionalities for arranging
plots.

```{r}
#| eval: false
#| labs: fig-plot-layout
# Example: Arranging multiple plots
library(gridExtra)

grid.arrange(lineplot, lineplot1, lineplot2, ncol = 3)
```

```{r}
#| echo: false
#| layout-ncol: 3
#| label: fig-data-viz-lineplot
#| fig-cap: "Customised Lineplots and Logarithmic Scale Transformation"
#| fig-subcap: 
#|  - "Basic Lineplot"
#|  - "Customised Lineplot"
#|  - "Customised Lineplot on a Log Scale"
#| fig-alt: "Lineplot 1, Lineplot 2, and Lineplot 3"
lineplot
lineplot1
lineplot2
```

#### Exercise

Replicate the plots above, customizing the legend and axis text to
improve readability. Apply a logarithmic scale to the y-axis for better
data visualization, and consider adding a caption to indicate the data
source.

### Saving as an Image

Once you've created your visualisation, you might want to save it as an
image file for sharing or publication.

```{r}
#| eval: false
ggsave("lineplot.png", plot = lineplot2, width = 6, height = 4)
```

## Practising Data Visualisation

To practise making data visualisations, there are numerous free
resources available that provide valuable information and opportunities
to enhance your skills. By engaging with these platforms, you can
practise your abilities, share your final results, and receive feedback
from the community. While the feedback may sometimes be critical, it is
an essential part of the learning process and will help you improve if
you stay persistent.

Participating in competitions and challenges such as
**#TidyTuesday**\index{TidyTuesday},
**#30DayChartChallenge**\index{30DayChartChallenge}, and
**#30DayMapChallenge**\index{30DayMapChallenge} offers a great way to
refine your skills. These competitions encourage you to create and share
visualisations on various themes, pushing you to experiment with
different techniques and styles. By consistently participating in these
challenges and actively seeking feedback, you will steadily enhance your
proficiency. Over time, you will find yourself becoming more adept and
confident in your data visualisation skills.
