# Interpreting Model Results Through Visualisation {#sec-10-applications}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   Visualize predicted vs. observed values and assess residuals.
-   Interpret model metrics with VIP, accuracy, and confusion matrix
    plots.
-   Create and customize ROC curves and compute AUC for classification
    models.
:::
::::::

\
In the introduction to data visualisation (@sec-09-data_visualisation),
we explored foundational techniques for effectively communicating model
results and data insights through plots. We reviewed essential plots
used to visualise predictions and outcomes, such as scatter plots, line
graphs, and bar charts. The chapter also covered how to customise these
plots using colours, palettes, legends, and guides to improve clarity
and visual appeal. Additionally, we discussed how to craft compelling
data stories by organising multiple plots in a layout and saving them as
image files for reports or presentations. These skills are critical for
interpreting model outputs and sharing findings with a wider audience.

In this chapter, an essential part of the data
visualisation\index{Data visualisation} section, we focus on how to
effectively interpret and use the results from machine
learning\index{Machine learning} models in the context of health
metrics\index{Health metrics} and infectious
diseases\index{Infectious diseases}. The aim is to provide actionable
insights, enhance decision-making, and communicate findings to various
stakeholders:

-   **Visualising Predictions and Outcomes**:
    -   Techniques for visualising predicted vs. actual values.
    -   Using scatter plots, line plots, and bar charts to compare
        predictions and observed data.
    -   Highlighting discrepancies and trends through residual plots and
        error distribution charts.
-   **Case Studies and Applications**:
    -   Real-world examples of model results application in public
        health.
    -   Case studies demonstrating the impact of machine learning and
        spatial modelling on health metrics.
    -   Lessons learned and best practices from successful
        implementations.

## Practical Insights and Examples

In this section, we will explore practical examples of how to visualise
models and their results. The objective is to learn how to interpret the
results of a model and how to communicate them effectively. Several
considerations are important when visualising the results of a model,
such as:

-   **Understanding the data**: Before visualising the results, it is
    important to understand the data and the model used. This includes
    understanding the variables, their relationships, and the
    assumptions of the model.
-   **Choosing the right visualisation**: Different types of data and
    models require different types of visualisations. It is important to
    choose the right visualisation to effectively communicate the
    results.

### Example: COVID-19 Analytics

Data are from the **ECDC (European Centre for Disease Prevention and
Control)**\index{European Centre for Disease Prevention and
Control (ECDC)} - updated daily. The data include information on the
**number of deaths for each country due to COVID-19**. The dataset can
be accessed through the `{covid19.analytics}` package\index{COVID-19}.

```{r}
#| eval: false
install.packages("covid19.analytics")
```

```{r}
#| output: false
library(tidyverse)
library(covid19.analytics)

covid19.ALL.agg.cases <- covid19.data("aggregated")
```

We select the variables of interest, such as `Country_Region`,
`Last_Update`, `Confirmed`, `Deaths`, `Recovered`.

```{r}
data <- covid19.ALL.agg.cases %>% 
  select(Country_Region, Last_Update, 
         Confirmed, Deaths, Recovered) %>%
  mutate(Last_Update = as.Date(Last_Update)) 

data %>% 
  head()
```

We group the data by `Last_Update` and calculate the mean of the
Confirmed, Deaths, Recovered variables. This is a rough aggregation and
not population-weighted.

```{r}
data %>%
  group_by(Last_Update) %>%
  reframe(across(Confirmed:Recovered, 
                   ~ mean(.x, na.rm = TRUE))) 
```

We can visualise the data with a scatter plot using `geom_point()` and
overlay a line with `geom_line()` functions and then add a
`geom_smooth()`. This last function adds the most suitable model line,
in this case a linear regression line, to the plot based on the data.

The plot shows the relationship between the number of confirmed cases
and the number of deaths.

```{r}
#| label: fig-covid19_data
#| fig-cap: "COVID-19 Deaths for all Countries. The plot shows the relationship between the number of confirmed cases and the number of deaths. The dashed line shows the linear regression line, while the solid line shows the smoothed line."
#| fig-alt: "COVID-19 data"
#| fig-width: 6
#| fig-height: 4

data %>%
  ggplot(aes(x = Confirmed, y = Deaths)) +
  geom_line(linetype = "dashed") +
  geom_point(shape = 21) +
# add a smoothed line to approximate the trend
  geom_smooth(method = "lm", linetype = "dashed") +
  geom_smooth(method = "gam", 
              formula = y ~ s(x, bs = "cs"), 
              linetype = "solid") +
# add a scale to the x-axis  
  scale_x_continuous(trans = "sqrt", 
                     expand = c(0, 0), # remove space
                     labels = scales::label_comma()) +
  labs(title = "COVID-19 Deaths for all Countries",
       subtile = "Comparison of Linear Model vs GAM Fit",
       x = "Confirmed", y = "Number of Deaths") 
```

The plot shows a positive correlation between the number of confirmed
cases and the number of deaths. As the `geom_smooth()` fits a smooth
line using method = 'gam' and formula = 'y \~ s(x, bs = "cs"), using a
spline method; it provides a useful first approximation of the
relationship between Confirmed and Deaths.

To estimate the number of Deaths due to Confirmed and Recovered cases,
we start by fitting three models:

-   the first one without predictors, to obtain the expected value of
    the number of `Deaths` in case of no epidemic in action
-   the second model is fit with `Confirmed` cases as the only predictor
-   the third model is fit with `Confirmed` and `Recovered` as
    predictors.

In all cases we use the `lm()` function, which fits a linear model to
the data.

```{r}
mod1 <- lm(Deaths ~ 1, data = data) 
mod2 <- lm(Deaths ~ Confirmed, data = data)
mod3 <- lm(Deaths ~ Confirmed + Recovered, data = data)
summary(mod1)$coefficients;
summary(mod2)$coefficients;
summary(mod3)$coefficients
```

The `summary()` function provides the model coefficients, which indicate
that, in the absence of an epidemic\index{Epidemic}, the average number
of Deaths is estimated at 562.48. For each additional Confirmed case,
the number of deaths increases by approximately 0.025, while for each
additional Recovered case, the number of deaths decreases by about
0.0038. The intercept is negative, suggesting that when both confirmed
and recovered cases are zero, the model predicts a negative number of
deaths—an implausible result. This highlights the importance of
interpreting the coefficients within the context of the data.

> "Mortality trends following an epidemic typically show a decline once
> transmission ceases, assuming no further cases occur." CDC’s
> Principles of Epidemiology.

```{r}
#| label: fig-covid19_model-comparison
#| layout-ncol: 2
#| fig-cap: "COVID-19 Model Comparison. Source ECDC: covid19.analytics"
#| fig-subcap:
#| - "Model Fit Comparison"
#| - "Observed vs. Fitted Values"
#| fig-alt: "COVID-19 Model Comparison"
#| fig-width: 6
#| fig-height: 4

# Model Fit Comparison
model_metrics <- data.frame(
  model = c("mod1", "mod2", "mod3"),
  R2 = c(summary(mod1)$r.squared, 
                summary(mod2)$r.squared,
                summary(mod3)$r.squared),
  AIC = c(AIC(mod1), AIC(mod2), AIC(mod3)))

model_metrics_long <- pivot_longer(model_metrics, -model)

ggplot(model_metrics_long, aes(x = model, y = value, fill = name)) +
  geom_col(position = "dodge") +
  facet_wrap(~name, scales = "free_y") +
  labs(title = "Model Fit Comparison", y = "Value", x = "Model") +
  theme(legend.position = "none")

# Observed vs. Fitted Values
data$fit2 <- fitted(mod2)
data$fit3 <- fitted(mod3)

data_long <- data |>
  select(Deaths, fit2, fit3) |>
  pivot_longer(cols = starts_with("fit"), 
               names_to = "model", 
               values_to = "fitted")

ggplot(data_long, aes(x = Deaths, y = fitted,
                      color = model)) +
  geom_point(alpha = 0.6, aes(shape = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Predicted vs Observed Deaths due to COVID-19", 
       x = "Observed Deaths", y = "Predicted") +
  theme_minimal()
```

The first plot shows the model fit comparison, with the
R-squared\index{R-squared} and
AIC\index{Akaike Information Criterion (AIC)} values for each model. The
second plot shows the observed vs. fitted values for the models. The
dashed line represents a perfect fit, while the points represent the
actual data.

The model with the highest R-squared value is the one that best explains
the variance in the data. The AIC value is a measure of the relative
quality of a statistical model for a given dataset, with lower values
indicating a better fit. The models explains \~85% of the variance in
the number of deaths, suggesting a strong linear relationship. The
`r.squared` value is close to 1, which means that the model fits the
data well.

To estimate the number of deaths based on the number of confirmed and
recovered cases. We can use the `predict()` function.

```{r}
Pred <- predict(mod3, newdata = data)

data_pred <- cbind(data %>% select(Confirmed, Deaths), Pred) %>%
  # calculate the residuals to see how well the model fits the data
  mutate(Residuals = Deaths - Pred) 

data_pred %>%
  head()
```

The `Residuals` column shows the difference between the actual number of
deaths and the predicted number of deaths. A positive residual means
that the model underestimated the number of deaths, while a negative
residual means that the model overestimated the number of deaths.

Let's see the results in a plot:

```{r}
#| layout-ncol: 2
#| label: fig-covid19_model
#| fig-cap: "COVID-19 data"
#| fig-subcap: 
#|  - "Predicted vs. Actual Deaths due to COVID-19"
#|  - "Model Residuals vs. Observed Deaths due to COVID-19"
#| fig-alt: "COVID-19 data"
data_pred %>%
  ggplot(aes(Deaths, Pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  geom_segment(aes(xend = Deaths, yend = Pred), alpha = 0.5) +
  tune::coord_obs_pred() +
  labs(title = "Predicted vs. Actual Deaths due to COVID-19",
       x = "Actual Deaths", y = "Predicted Deaths") 

data_pred %>%
  ggplot(aes(Deaths, Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs. Deaths due to COVID-19",
       x = "Deaths", y = "Residuals") 
```

Finally, we can visualise the results of the model with a scatter plot
and a line plot. The first plot shows the observed vs. predicted values,
while the second plot shows the residuals. The segments represent the
difference between the observed and predicted values.

```{r}
#| label: fig-lm_results
#| layout-ncol: 2
#| fig-cap: "COVID-19 Model Results"
#| fig-subcap: 
#|  - "Observed vs. Predicted" 
#|  - "Segments represent the residuals"
#| fig-alt: "COVID-19 Model Results"
data_pred %>%
  ggplot() +
  geom_point(aes(Confirmed, Deaths)) +
  geom_line(aes(Confirmed, Pred), linetype = "dashed") +
  labs(title = "Observed vs. Predicted",
       x = "Confirmed Cases", y = "Deaths",
       caption = "Source: ECDC") 
  
data_pred %>%
  ggplot(aes(x = Confirmed, y = Deaths)) +
  geom_point() +
  geom_line(aes(Confirmed, Pred), linetype = "dashed") +
  geom_segment(aes(xend = Confirmed, yend = Pred)) +
  labs(title = "Observed vs. Predicted",
       subtitle = "Segments represent the residuals",
       x = "Confirmed Cases", y = "Deaths",
       caption = "Source: ECDC")   
```

#### Exercise: COVID-19 Case Fatality Ratio

Use the `covid19.ALL.agg.cases` data to visualise the **Case Fatality
Rate**\index{Case Fatality Rate (CFR)} estimated for each country, by
last update and include the latitude and longitude as further
information. Build a suitable model and explain the results with a plot.

```{r}
cfr <- covid19.ALL.agg.cases %>% 
  select(Country_Region, Last_Update, 
         Lat, Long_, Case_Fatality_Ratio)

cfr %>%
  head()
```

### Example: Ischemic Stroke Decision Tree

In this example we have a look at how to visualise the results of a
decision tree\index{Decision tree} model for predicting Ischemic
Stroke\index{Stroke}.

Load necessary libraries and the data for the Ischemic
Stroke\index{Stroke}[@fes/data].

```{r}
# Ischemic Stroke decision tree
library(tidymodels)
library(rpart)
library(rpart.plot)
```

```{r}
#| echo: false
load(url("https://github.com/topepo/FES/blob/master/Data_Sets/Ischemic_Stroke/stroke_data.RData?raw=true"))
```

Data are already split into training and test sets, we will combine them
for the analysis. The `stroke_train` and `stroke_test` datasets contain
`r nrow(stroke_train)`, `r nrow(stroke_test)` observations respectively
and `r ncol(stroke_test)` variables. Within the variables we have
information on volume, proportion, area, thickness of the arterial wall,
among others. The target variable is `Stroke`, which indicates whether
the patient has had a stroke or not[@johnson].

We select the variables of interest for the analysis with `any_of()`, a
function to select the variables based on their names.

`?any_of()` for more infromation about the function.

```{r}
selected_train <- stroke_train %>%
  dplyr::select(any_of(VC_preds), Stroke)
```

To check which columns in `stroke_train` dataset were not selected:

```{r}
setdiff(names(stroke_train), names(selected_train))
```

Set up the recipe for the data with the `recipe()` function from the
`tidymodels` package. We will use the `step_corr()` function to remove
highly correlated predictors, `step_center()` and `step_scale()` to
standardise the predictors, `step_YeoJohnson()` to transform the
predictors, and `step_zv()` to remove zero variance predictors.

```{r}
is_recipe <- recipe(Stroke ~ ., data = selected_train) %>%
  #step_interact(int_form)  %>%
  step_corr(all_predictors(), threshold = 0.75) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>%
  step_YeoJohnson(all_predictors()) %>%
  step_zv(all_predictors())

is_recipe %>%
  prep() %>% 
  bake(new_data=NULL) %>%
  select(1:5) %>%
  head(5)
```

Set up the decision tree model with the `decision_tree()` function from
the `tidymodels` package. We will use the `rpart` engine for the
decision tree model and set the mode to classification.

```{r}
class_tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")
```

Finally, we will fit the model with the `fit()` function from the
`tidymodels` package and visualise the results with the `rpart.plot`
package.

```{r}
#| label: fig-dv-tree
#| fig-cap: "Decision Tree for Ischemic Stroke"
#| fig-alt: "Decision Tree for Ischemic Stroke"
is_wfl <- workflow() %>%
  add_model(class_tree_spec) %>%
  add_recipe(is_recipe)

is_dt_fit_wfl <- is_wfl %>%
  fit(data = selected_train,
      control = control_workflow())

is_dt_fit_wfl%>%
  extract_fit_engine() %>%
  rpart.plot::rpart.plot(roundint = FALSE)
```

In this plot, the decision tree is visualised with the `rpart.plot` and
the information released evidences the importance of some specific
predictors, such as max wall thickness, max dilatation by area, volume
proportion, and max remodelling ratio. The decision tree is a useful
tool for visualising the results of the model and understanding the
relationships between the predictors and the target variable.

The interpretation of the decision tree is straightforward: each node
represents a decision based on the value of a predictor, and the leaves
represent the final classification. The tree can be pruned to reduce its
complexity and improve its interpretability. The decision tree can be
used to make predictions on new data by following the path from the root
to a leaf node based on the values of the predictors.

### Example: Ischemic Stroke Classification

In this second example we will demonstrate
**classification**\index{Classification} (stroke vs. no stroke) using
patient and plaque imaging features, and visualise model insights for
stakeholder communication.

The objective is to visualise how well the model predict whether a
patient experienced a stroke (Stroke) based on imaging features (e.g.,
MaxStenosisByArea, CALCVolProp) and risk factors (e.g., age,
DiabetesHistory).

Visualise:

-   Variable importance
-   Prediction performance (e.g., ROC curve)
-   Partial dependence of key features

Load necessary libraries, and fit a **random
forest**\index{Random Forest} model to the data with the `rand_forest()`
function from the `{tidymodels}` package. We will use the `ranger`
engine for the random forest model and set the mode to classification.

```{r}
library(tidyverse)
library(tidymodels)
library(vip)
library(DALEXtra)
library(pROC)
```

The model specify the number of trees to grow, the minimum number of
observations in a node before a split is attempted, and the maximum
depth of the tree. The `set_engine()` function specifies the engine to
use for the model, in this case `ranger`, and the `set_mode()` function
specifies the mode of the model, in this case classification.

```{r}
rf_spec <- rand_forest(trees = 500, min_n = 5) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

rf_wf <- workflow() %>%
  add_formula(Stroke ~ .) %>%
  add_model(rf_spec)

rf_fit <- rf_wf %>% fit(data = stroke_train)
```

#### Variable Importance

The object `rf_fit` contains the fitted model. We can extract the the
model specification and use the `vip()` function to visualise what are
the predictors that most influence the model. The `vip()` function
creates a variable importance plot, which shows the importance of each
predictor in the model. The importance is calculated based on the **Mean
Decrease in Impurity (MDI)**\index{Mean Decrease in Impurity (MDI)} for
each predictor.

```{r}
#| label: fig-rf-importance
#| fig-cap: "Variable Importance for Ischemic Stroke"
#| fig-alt: "Variable Importance for Ischemic Stroke"
#| fig-width: 6
#| fig-height: 4
rf_fit %>%
  extract_fit_parsnip() %>%
  vip::vip(num_features = 10)
```

We can conclude that for this data set, the most important predictors
are `MaxStenosisByArea`, `CALCVolProp`, `MaxWallThickness`, and
`MaxRemodellingRatio`.

Next step is to evaluate the model performance. We will use the
`predict()` function to make predictions on the test data set
(`stroke_test`) and calculate the **accuracy**\index{Accuracy} of the
model.

```{r}
set.seed(05122025)
rf_preds <- predict(rf_fit, stroke_test, 
                    type = "prob") %>%
  bind_cols(predict(rf_fit, stroke_test)) %>%
  bind_cols(stroke_test %>% select(Stroke))

rf_preds %>% head()
```

#### Accuracy

The accuracy of the model is calculated as the proportion of correct
predictions. The `accuracy()` function from the `yardstick` package
calculates the accuracy of the model based on the predicted and observed
values.

```{r}
rf_preds %>%
  accuracy(truth = Stroke, .pred_class)
```

In this case, the accuracy of the model is 0.70, which means that the
model correctly predicts whether a patient experienced a stroke or not
70% of the time. The accuracy is calculated as the number of correct
predictions divided by the total number of predictions.

#### ROC Curve

The **Receiver Operating Characteristic
(ROC)**\index{Receiver Operating Characteristic (ROC)} curve is a
graphical representation of the performance of a binary classifier
system as its discrimination threshold is varied. The ROC curve plots
the true positive rate (TPR) against the false positive rate (FPR) at
various threshold settings.

We can use the `roc_curve()` function from the `yardstick` package to
calculate the ROC curve and the area under the curve (AUC). The
`roc_curve` object contains the specificity and sensitivity values for
each threshold, which can be used to calculate the
**AUC**\index{Area Under the Curve (AUC)}.

```{r}
roc_curve <- yardstick::roc_curve(data = rf_preds, 
                       truth = Stroke, .pred_N)
roc_curve %>% head()
```

The relationship between the **True Positive Rate
(TPR)**\index{True Positive Rate (TPR)} and **False Positive Rate
(FPR)**\index{False Positive Rate (FPR)} is visualised in the ROC curve.
The area under the curve (AUC) is a measure of the model's performance,
with values closer to 1 indicating better performance. The
**specificity**\index{Specificity} and
**sensitivity**\index{Sensitivity} of the model indicate the proportion
of true positives and true negatives, respectively, and we can use the
`autoplot()` function to visualise the ROC curve.

```{r}
#| label: fig-roc-curve
#| fig-cap: "ROC Curve for Ischemic Stroke"
#| fig-alt: "ROC Curve for Ischemic Stroke"
#| fig-width: 6
#| fig-height: 4
autoplot(roc_curve)
```

Another way to extract the ROC curve is to use the `pROC` package and
the `roc()` function. The `roc()` function takes the observed values and
the predicted probabilities as arguments and returns an object
containing the ROC curve.

```{r}
roc_obj <- pROC::roc(rf_preds$Stroke, 
                     rf_preds$.pred_N)
roc_obj %>%
  pROC::auc() %>%
  round(2)
```

The `roc_obj` object provides information about the area under the ROC
curve. In this case we have an AUC is `r round(roc_obj$auc,2)`. The AUC
is a measure of the model's performance, values closer to 1 indicate
better performance.

#### Partial Dependence

> "How does the prediction change when a specific feature changes,
> holding all others constant?"

Finally, we attempt an explanation of the model using `DALEX` and
`DALEXtra` packages, and some specific functions, such as:
`explain_tidymodels()` and `model_profile()`.

```{r}
#| output: false
explainer_rf <- 
  DALEXtra::explain_tidymodels(rf_fit, 
                               data = select(stroke_train, -Stroke), 
                               y = stroke_train$Stroke)
```

The `explainer_rf` object contains the model, the data, and the target
variable and it will be used to create the partial dependence plot with
the `model_profile()` function.

A **Partial Dependence Plot (PDP)**\index{Partial Dependence Plot (PDP)}
is a tool used in machine learning to help interpret black-box models
like random forests. It shows the marginal effect of one (or two)
features on the predicted outcome, while averaging out the influence of
all other features.

In this case we specify the `MaxStenosisByArea` variable, to see how
changes in `MaxStenosisByArea` influence the model’s predicted outcome,
on average. A higher value of MaxStenosisByArea indicates more severe
arterial narrowing, which is a risk factor for stroke or heart attack.

```{r}
#| label: fig-partial-dependence
#| fig-cap: "Partial Dependence Plot for Ischemic Stroke: MaxStenosisByArea (arterial narrowing)"
#| fig-alt: "Partial Dependence Plot for Ischemic Stroke"
#| fig-width: 6
#| fig-height: 4
DALEX::model_profile(explainer_rf, 
                     variables = "MaxStenosisByArea") %>%
  plot() +
  labs(
    title = "Partial Dependence Plot for Ischemic Stroke",
    subtitle = "",
    x = "Value",
    y = "AVG Prediction") +
  theme_minimal() 
```

The plot shows the partial dependence of the `MaxStenosisByArea`
variable on the predicted probability of having a stroke. The results
indicate that `MaxStenosisByArea` has a positive influence on stroke
risk, though this effect plateaus beyond a certain threshold.

## Summary

In this chapter, we have learned how to visualise the results of a
model. There are several important considerations when visualising the
results of a model, and differences might arise due to the type of model
used. We have seen how to use several packages and functions, such as
`ggplot2`, `vip`, `pROC`, and `DALEX` to visualise the results of a
model. We have also seen how to interpret the results of a model and how
to communicate them effectively. The examples provided in this chapter
demonstrate how to use visualisation techniques to enhance
decision-making and communicate findings to various stakeholders.
