# Interpreting Model Results Through Visualization {#sec-10-applications}

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   Overview of the basic plots for visualizing predictions and outcomes
-   How to customize a plot using colors, palettes, legends, and guides
-   Tell the story with data through plot layouts and saving as an image
:::
:::

This chapter is an essential part of the data visualization section, it
focus on how to effectively interpret and utilize the results from
machine learning models in the context of health metrics and infectious
diseases. The aim is to provide actionable insights, enhance
decision-making, and communicate findings to various stakeholders.

1.  **Visualizing Predictions and Outcomes**:
    -   Techniques for visualizing predicted vs. actual values.
    -   Using scatter plots, line plots, and bar charts to compare
        predictions and observed data.
    -   Highlighting discrepancies and trends through residual plots and
        error distribution charts.
2.  **Case Studies and Applications**:
    -   Real-world examples of model results application in public
        health.
    -   Case studies demonstrating the impact of machine learning and
        spatial modelling on health metrics.
    -   Lessons learned and best practices from successful
        implementations.

## Practical Insights and Examples

### Example 1: Visualizing COVID-19 Data

Data is from the ECDC (European Centre for Disease Prevention and
Control) and is updated daily. The data includes information on the
number of deaths for each country. The dataset is available on the ECDC
website and can be accessed through the `{covid19.analytics}` package.

```{r}
#| eval: false
install.packages("covid19.analytics")
```

```{r}
#| message: false
#| warning: false
library(tidyverse)
cfr <- covid19.analytics::covid19.data() %>%
  select(Country_Region, Last_Update, Lat, Long_, Case_Fatality_Ratio)

cfr %>%
  head()
```

### Example 2: Visualizing Predicted vs. Actual DALYs:

-   Creating scatter plots to compare predicted Disability-Adjusted Life
    Years (DALYs) with actual values.
-   Analyzing residuals to identify patterns and areas for model
    improvement.
