# Interpreting Model Results Through visualisation {#sec-10-applications}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   Overview of the basic plots for visualising predictions and outcomes
-   How to customise a plot using colours, palettes, legends, and guides
-   Tell the story with data through plot layouts and saving as an image
:::
::::::

\
This chapter is an essential part of the data visualisation section, it
focuses on how to effectively interpret and use the results from
machine learning models in the context of health metrics and infectious
diseases. The aim is to provide actionable insights, enhance
decision-making, and communicate findings to various stakeholders.

1.  **Visualising Predictions and Outcomes**:
    -   Techniques for visualising predicted vs. actual values.
    -   Using scatter plots, line plots, and bar charts to compare
        predictions and observed data.
    -   Highlighting discrepancies and trends through residual plots and
        error distribution charts.
2.  **Case Studies and Applications**:
    -   Real-world examples of model results application in public
        health.
    -   Case studies demonstrating the impact of machine learning and
        spatial modelling on health metrics.
    -   Lessons learned and best practices from successful
        implementations.

## Practical Insights and Examples

### Example: visualising COVID-19 Data

Data is from the **ECDC (European Centre for Disease Prevention and
Control)**, data is updated daily. The data includes information on the
number of deaths for each country. The dataset can be accessed through
the `{covid19.analytics}` package.

```{r}
#| eval: false
install.packages("covid19.analytics")
```

```{r}
library(tidyverse)
library(covid19.analytics)

covid19.ALL.agg.cases <- covid19.data("aggregated")
```

```{r}
data <- covid19.ALL.agg.cases %>% 
  select(Country_Region, Last_Update, 
         Confirmed, Deaths, Recovered) %>%
  mutate(Last_Update = as.Date(Last_Update)) 

data %>% 
  head
```

```{r}
data %>%
  group_by(Last_Update) %>%
  summarise(across(Confirmed:Recovered, ~ mean(.x, na.rm = TRUE)))
```

```{r}
#| label: fig-covid19_data
#| fig-cap: "COVID-19 Deaths by Country"
#| fig-alt: "COVID-19 data"
data %>%
  ggplot(aes(x = Confirmed, y = Deaths)) +
  geom_point() +
  geom_smooth() +
  labs(title = "COVID-19 Deaths by Country",
       x = "Confirmed", y = "Number of Deaths") 
```

The relationship is linear between the number of confirmed cases and the
number of deaths. The model can be fitted using linear regression to
predict the number of deaths based on the number of confirmed cases.

```{r}
mod <- lm(Deaths ~ Confirmed + Recovered, data = data) 

summary(mod)$coefficients
```

```{r}
Pred <- predict(mod, newdata = data)

data_pred <- cbind(data %>% select(Confirmed, Deaths),Pred) %>%
  mutate(Residuals = Deaths - Pred) 

data_pred %>%
  head()
```

Let's see the results in a plot:

```{r}
#| layout-ncol: 2
#| label: fig-covid19_model
#| fig-cap: "COVID-19 data"
#| fig-subcap: 
#|  - "COVID-19 Deaths by Country"
#|  - "Predicted vs. Actual Deaths"
#| fig-alt: "COVID-19 data"
data_pred %>%
  ggplot(aes(Deaths, Pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  geom_segment(aes(xend = Deaths, yend = Pred), alpha = 0.5) +
  labs(title = "Predicted vs. Actual Deaths",
       x = "Actual Deaths", y = "Predicted Deaths") 

data_pred %>%
  ggplot(aes(Confirmed, Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs. Confirmed Cases",
       x = "Confirmed Cases", y = "Residuals") 
```

```{r}
#| label: fig-lm_results
#| layout-ncol: 2
#| fig-cap: "COVID-19 Model Results"
#| fig-subcap: 
#|  - "Observed vs. Predicted" 
#|  - "Segments represent the residuals"
#| fig-alt: "COVID-19 Model Results"
data_pred %>%
  ggplot() +
  geom_point(aes(Confirmed, Deaths)) +
  geom_line(aes(Confirmed, Pred), linetype = "dashed") +
  labs(title = "Observed vs. Predicted",
       x = "Confirmed Cases", y = "Deaths",
       caption = "Source: ECDC") 
  
data_pred %>%
  ggplot(aes(x = Confirmed, y = Deaths)) +
  geom_point() +
  geom_line(aes(Confirmed, Pred), linetype = "dashed") +
  geom_segment(aes(xend = Confirmed, yend = Pred)) +
  labs(title = "Observed vs. Predicted",
       subtitle = "Segments represent the residuals",
       x = "Confirmed Cases", y = "Deaths",
       caption = "Source: ECDC")   
```

```{r}
cfr <- covid19.ALL.agg.cases %>% 
  select(Country_Region, Last_Update, Lat, Long_, Case_Fatality_Ratio)

cfr %>%
  head()
```

```{r}
covid19.TS.deaths <- covid19.data("ts-deaths")
covid19.TS.deaths %>% 
    pivot_longer(cols = c(-Province.State, 
                          -Country.Region, 
                          -Lat, -Long),
               names_to = "Date", values_to = "Deaths") %>%
  head()
```

### Example: Ischemic Stroke Decision Tree

In this example we have a look at how to visualise the results of a
decision tree model for predicting Ischemic Stroke.

Load necessary libraries:

```{r}
# Ischemic Stroke decision tree
library(tidymodels)
library(rpart)
library(rpart.plot)
```

Load the data for the Ischemic Stroke from the GitHub repository of the
book: "Feature Engineering and Selection: A Practical Approach for
Predictive Models" by Max Kuhn and Kjell Johnson.

```{r}
load(url("https://github.com/topepo/FES/blob/master/Data_Sets/Ischemic_Stroke/stroke_data.RData?raw=true"))
```

Data is already split into training and test sets, we will combine them
for the analysis. We just need to select the variables of interest for
the analysis with `any_of()`, a function to select the variables of
interest for the analysis.

`?any_of()`

```{r}
selected_train <- 
  stroke_train %>%
  dplyr::select(any_of(VC_preds), Stroke)
```

Set up the recipe for the data with the `recipe()` function from the
`tidymodels` package. We will use the `step_corr()` function to remove
highly correlated predictors, `step_center()` and `step_scale()` to
standardise the predictors, `step_YeoJohnson()` to transform the
predictors, and `step_zv()` to remove zero variance predictors.

```{r}
is_recipe <- recipe(Stroke ~ ., data = selected_train) %>%
  #step_interact(int_form)  %>%
  step_corr(all_predictors(), threshold = 0.75) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>%
  step_YeoJohnson(all_predictors()) %>%
  step_zv(all_predictors())

is_recipe%>%prep()%>%bake(new_data=NULL)%>%select(1:5)%>%head(5)
```

Set up the decision tree model with the `decision_tree()` function from
the `tidymodels` package. We will use the `rpart` engine for the
decision tree model and set the mode to classification.

```{r}
class_tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")
```

Finally, we will fit the model with the `fit()` function from the
`tidymodels` package and visualise the results with the `rpart.plot`
package.

```{r}
#| label: fig-dv-tree
#| fig-cap: "Decision Tree for Ischemic Stroke"
#| fig-alt: "Decision Tree for Ischemic Stroke"
is_wfl <- workflow() %>%
  add_model(class_tree_spec) %>%
  add_recipe(is_recipe)

is_dt_fit_wfl <- is_wfl %>%
  fit(data = selected_train,
      control = control_workflow())

is_dt_fit_wfl%>%
  extract_fit_engine() %>%
  rpart.plot::rpart.plot(roundint = FALSE)
```
