# Interpreting Model Results Through Visualisation {#sec-10-applications}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   Visualize predicted vs. observed values and assess residuals
-   Interpret model metrics with VIP, accuracy, and partial dependency
    plots
-   Create and customize ROC curves and compute AUC for classification
    models
:::
::::::

\
In the introduction to data visualisation (@sec-09-data_visualisation),
we explored foundational techniques for effectively communicating model
results and data insights through plots. We reviewed essential plots
used to visualise predictions and outcomes, such as scatter plots, line
graphs, and bar charts. The chapter also covered how to customise these
plots using colours, palettes, legends, and guides to improve clarity
and visual appeal. Additionally, we discussed how to craft compelling
data stories by organising multiple plots in a layout and saving them as
image files for reports or presentations.

In this chapter, which is an essential part of the data
visualisation\index{Data visualisation} section, we focus on how to
effectively interpret and use the results from machine
learning\index{Machine learning} models in the context of health
metrics\index{Health metrics} and infectious
diseases\index{Infectious diseases}.

We will explore how to visualise the results of:

-   **Regression models**\index{Regression models} with a linear
    regression, and a generalized additive model (GAM)
-   **Classification models**\index{Classification models} with decision
    trees, and random forest models.

## Practical Insights and Examples

The practice of visualising model results is essential for communicating
insights from a model to stakeholders.

Considerations for visualising model results include:

-   **Understanding the data**: Before visualising the results, it is
    important to understand the data and the model used. This includes
    understanding the variables, their relationships, and the
    assumptions of the model.
-   **Choosing the right visualisation**: Different types of data and
    models require different types of visualisations. It is important to
    choose the right visualisation to effectively communicate the
    results.

### Example: Deaths due to Meningitis

Data are from the **Institute for Health Metrics and Evaluation
(IHME)**\index{Institute for Health Metrics and Evaluation (IHME)} and
include **death rates due to meningitis**\index{Meningitis}, as well as
two exposure levels of risk factors—particulate matter (PM2.5) and
smoking—in the Sub-Saharan Africa (Central African Republic, Zambia,
Eswatini, Lesotho, and Malawi) from 1990 to 2021. The data are in the
`{hmsidwR}` package and can be loaded as `hmsidwR::meningitis`.

```{r}
#| eval: false
install.packages("hmsidwR") 
# or
# install the development version of book package
devtools::install_github("Fgazzelloni/hmsidwR")
```

Let's have a look at the first few rows of the data:

```{r}
# Load libraries and data
library(tidyverse)
library(hmsidwR)
meningitis %>% head
```

**Meningitis**[@meningit] is a serious global health concern
characterised by inflammation of the membranes surrounding the brain and
spinal cord. It can arise from both infectious and non-infectious causes
and is often associated with a high risk of mortality and long-term
complications.

In this example, we explore how two environmental risk factors —
**particulate matter (PM2.5)** and **smoking** — may influence
meningitis mortality. These variables represent **aggregate exposure
levels** that could potentially contribute to meningitis-related deaths.

Meningitis death rates vary considerably across countries and over time.
To begin, we visualise how these rates have changed from 1990 to 2021 in
five Sub-Saharan African countries. A scatter plot will show annual
death rates by country, while a smooth line will reveal the overall
trend, helping to contextualize patterns before moving to modelling.

```{r}
#| label: fig-aggregate-meningitis
#| fig-cap: "Deaths due to Menigitis from 1990 to 2021 in Sub-Saharan Africa. The solid ticker line represents the smooth line revealing the overall trend, while the other lines represent the death rates for each country."
#| fig-alt: "Deaths due to Menigitis from 1990 to 2021"
#| fig-width: 6
#| fig-height: 4
meningitis %>%
  ggplot(aes(x = year, y = deaths)) +
  geom_line(aes(group = location, 
                linetype = location)) +
  geom_smooth() +
  labs(title = "Deaths due to Menigitis",
       subtitle = "from 1990 to 2021 in Sub-Saharan Africa",
       y = "Death Rates", x = "Time (Year)") +
  theme(legend.position = "bottom",
        legend.title = element_blank()) 
```

To gain an idea of the estimated average of death rates in the
population, we can fit a simple linear regression model with `lm()`
function, with the formula `deaths ~ 1`, which means that we are fitting
a model with only an intercept (i.e., no predictors). Then we draw a
simple **Q-Q plot**[@q-qplot2025] (a scatterplot created by plotting two
sets of quantiles against one another) to check the normality of the
residuals.

```{r}
mod0 <- lm(deaths ~ 1, data = meningitis)
summary(mod0)
```

The summary of this model shows the intercept value to be the estimate average of the death rates. We can look at the distribution of the death rates with an histogram:
```{r}
hist1 <- meningitis %>%
  ggplot(aes(x = deaths)) +
  geom_histogram(bins = 10, 
                 fill = "grey70", color = "grey40") +
  geom_vline(aes(xintercept = mean(deaths)),
             color = "black", 
             size = 2,
             linetype = "dashed") +
  labs(title = "Histogram of Death Rates due to Meningitis",
       x = "Death Rates", y = "Count") +
  theme(legend.position = "bottom",
        legend.title = element_blank())

# Add log scale to x axis
hist2 <- hist1 +
   scale_x_log10() 
```

```{r}
#| label: fig-meningitis-histogram
#| layout-ncol: 2
#| fig-cap: "The histogram of Death Rates due to Meningitis clearly show the right-skewness of the data. The dashed line represents the mean value of the death rates."
#| fig-subcap: 
#| - "Histogram of Death Rates due to Meningitis"
#| - "Histogram of Death Rates due to Meningitis (log scale)"
#| fig-alt: "Histogram of Death Rates due to Meningitis"
#| fig-width: 6
#| fig-height: 4
#| echo: false

hist1
hist2
```
To investigate the non linear relationship between the number of deaths
and the two risk factors, we can use a **Generalized Additive Model
(GAM)** with `s()` function from the `{mgcv}` package. The `s()`
function is used to fit smooth terms in the model, which allows us to
capture non-linear relationships between the predictors and the response
variable.

```{r}
library(mgcv)
mod1 <- gam(deaths ~ s(smoking), data = meningitis)
mod2 <- gam(deaths ~ s(smoking) + s(pm25),
            data = meningitis)
```

To summarise, the models results in a table format, with the estimated coefficients for each model, we can see that the values for the intercept and the coefficients for smoking and PM2.5 are all positive and statistically significant (p < 0.05), indicating that **higher levels of these risk factors are associated with higher death rates due to meningitis**. 
```{r}
#| echo: false
#| label: tab-meningitis-coefficients
tibble(
  Model = c("mod0 - linear model", 
            "mod1 - gam 1 predictor", "mod2 gam 2 predictors"),
  Beta0 = c(summary(mod0)$coefficients[1],
            mod1$coefficients[1],
            mod2$coefficients[1]),
  Beta1 = c(NA, 
            mod1$coefficients[2],
            mod2$coefficients[2]),
  Beta2 = c(NA, NA, mod2$coefficients[3])) %>%
  mutate(across(-Model, round, 2)) %>%
  knitr::kable(caption = "Estimated coefficients for the models")
```

```{r}
#| layout-ncol: 3
#| label: fig-meningitis-predictors
#| fig-cap: "Summary Exposure Values: Smocking and PM2.5"
#| fig-subcap:
#| - "Smocking - mod1"
#| - "Smocking - mod2"
#| - "PM2.5 - mod2"
#| fig-alt: "Summary Exposure Values: Smocking and PM2.5"
#| fig-width: 6
#| fig-height: 4
plot(mod1)
plot(mod2)
```

A further comparison is done with the `AIC()` function:

```{r}
AIC(mod1, mod2)
```

The second GAM model (mod2) is clearly better than the first one (mod1):

-   The AIC drops by \~42 points, which is a substantial improvement (a
    drop of \>10 is considered strong evidence).
-   The additional smooth term s(pm25) significantly improves model fit,
    even after accounting for increased complexity (degrees of freedom
    increase from 10.6 to 16.1).

Let's now include the `year` variable in the model, to account for the
temporal trend in the data. We can also include an interaction term
between `year` and `location` to account for the different trends in
different countries. The `by` argument in the `s()` function allows us
to fit separate smooth terms for each level of the `location` variable.

We set the location to be a factor variable:
```{r}
meningitis$location <- as.factor(meningitis$location)
```

Then we can fit the third model:
```{r}
mod3 <- gam(deaths ~ s(smoking) + s(pm25) + s(year, by = location),
            data = meningitis)
```

The `s(year, by = location)` term allows for different smooth terms:
```{r}
#| label: fig-meningitis-location
#| fig-cap: "Meningitis Death Rates by Country. The solid line represents the smooth line revealing the overall trend, while the other lines represent the death rates for each country."
#| fig-alt: "Meningitis Death Rates by Country"
#| fig-width: 6
#| fig-height: 4
par(mfrow = c(2, 3))
for (i in 1:5) {
  plot(mod3, select = i + 2, shade = TRUE,
       main = levels(meningitis$location)[i])
}
```
```{r}
#| echo: false
#| label: tab-meningitis-model-metrics
# summarise the three models r squared
tibble(
  Model = c("mod0 - linear model", 
            "mod1 - gam 1 predictor", "mod2 gam 2 predictors"),
  R2 = c(summary(mod1)$r.sq,
         summary(mod2)$r.sq,
         summary(mod3)$r.sq),
  DevExp = c(
    summary(mod1)$dev.expl,
    summary(mod2)$dev.expl,
    summary(mod3)$dev.expl)) %>%
  knitr::kable(caption = "Model Metrics")
```

In conclusion, we began with a generalised additive model with just smoking as influencing factor, then we fit a second model where both
smoking and PM2.5 were modelled as smooth functions. This second model
explained nearly 80% of the variance in meningitis death rates. However,
by incorporating country-specific nonlinear time trends, we dramatically
improved the model fit to over 99% of variance explained. This
highlights the importance of accounting for temporal and spatial
structure in health data, particularly when modelling long time spans
across multiple countries.

There is still a consideration to be made related to the exposure effect of smocking and PM2.5 on meningitis death rates. 
```{r}
#| echo: false
#| label: tab-meningitis-p-values
library(broom)
cbind(tidy(mod2) %>% select(term),
      tidy(mod1) %>% select(p.value_mod1 = p.value),
      tidy(mod2) %>% select(p.value_mod2 = p.value),
      tidy(mod3)[1:2,] %>% select(p.value_mod3 = p.value)) %>%
  knitr::kable(caption = "Model p-values")
```
In the second model (mod2), both the smoking and PM2.5 terms have p-values below 0.05, indicating that they are statistically significant predictors of meningitis mortality. However, when incorporating temporal and spatial effects in mod3, the significance of these exposure variables changes — most notably, PM2.5 is no longer statistically significant. This shift suggests that the apparent association in the simpler model may be confounded by time trends or country-level differences, highlighting the importance of accounting for structured variation in space and time.

Let's use both models to predict the number of deaths due to
meningitis in the dataset. We can use the `predict()` function to obtain
the predicted values for the model. The predicted values are added to
the original dataset as a new column called `predicted`.

```{r}
meningitis$predicted_mod2 <- predict(mod2)
meningitis$predicted_mod3 <- predict(mod3)
```

```{r}
#| label: fig-meningitis
#| layout-ncol: 2
#| fig-cap: "Meningitis Death Rates by Country. The dashed line represents the predicted values from the model, while the solid lines represent the observed values."
#| fig-subcap: 
#| - "Observed vs Predicted - Mod2"
#| - "Observed vs Predicted - Mod3"
#| fig-alt: "Meningitis Death Rates by Country"
#| fig-width: 6
#| fig-height: 4
ggplot(meningitis, 
       aes(x = year, y = deaths, color = location)) +
  geom_line() +
  geom_line(aes(y = predicted_mod2), 
            linetype = "dashed") +
  facet_grid(location ~ ., scale = "free") +
  labs(title = "Meningitis Death Rates by Country",
       subtitle = "Observed vs Predicted",
       y = "Death Rate", x = "Year") +
  theme(legend.position = "none")

ggplot(meningitis, 
       aes(x = year, y = deaths, color = location)) +
  geom_line() +
  geom_line(aes(y = predicted_mod3), 
            linetype = "dashed") +
  facet_grid(location ~ ., scale = "free") +
  labs(title = "Meningitis Death Rates by Country",
       subtitle = "Observed vs Predicted",
       y = "Death Rate", x = "Year") +
  theme(legend.position = "none")
```

We can notice that the model fits the data well, with the predicted
values closely following the observed values, with mod3 clearly overfitting the data. 

Let's check the residuals of mod2:
```{r}
meningitis %>%
  # calculate the residuals to see how well the model fits the data
  mutate(residuals = deaths - predicted_mod2) %>%
  select(deaths, predicted_mod2, residuals) %>%
  head()
```

The residuals column represents the difference between the observed and predicted values. A positive residual indicates that the model underestimates the number of deaths, while a negative residual indicates an overestimate.

To evaluate model performance, we can visualize the residuals against the predicted values. In this plot, the dashed line marks the zero-residual baseline. Points above the line correspond to underestimation, and those below represent overestimation. The fact that most points cluster closely around the zero line suggests that the model provides a reasonably good fit to the data.

```{r}
#| layout-ncol: 2
#| label: fig-meningitis-residuals
#| fig-cap: "Residuals vs Predicted"
#| fig-subcap: 
#| - "Residuals vs Predicted"
#| - "QQ-plot of the residuals"
#| fig-width: 6
#| fig-height: 4

meningitis %>%
  mutate(residuals = deaths - predicted_mod2) %>%
  ggplot(aes(x = predicted_mod2, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_grid(location ~ ., scale = "free") +
  labs(title = "Residuals vs Predicted",
       x = "Predicted Values", y = "Residuals") +
  theme(legend.position = "none")

meningitis %>%
  mutate(residuals = deaths - predicted_mod2) %>%
  ggplot(aes(sample = residuals)) +
  geom_qq() +
  geom_qq_line() +
  facet_grid(location ~ ., scale = "free") +
  labs(title = "QQ-plot of the residuals",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme(legend.position = "none")
```

**Heteroskedasticity**\index{Heteroskedasticity} is a common problem in regression analysis, and it
occurs when the variance of the residuals is not constant across all levels of the predictor variables. This can lead to biased estimates of the coefficients and incorrect conclusions about the significance of the predictors. In this case, we can see that the residuals are not evenly distributed around zero, indicating that there may be some
**heteroskedasticity**\index{Heteroskedasticity} in the data.

#### Exercise: One Country Focus

Improve the model specifically for Lesotho by refining the smooth terms and including year as a covariate to capture temporal patterns:

1.	Subset the data to include only observations from Lesotho
2.	Adjust the smooth functions as needed for a better fit
3.	Refit the model incorporating the year variable
4.	Perform cross-validation by splitting the data into training and test sets, and simulate model performance across multiple samples

Evaluate the model fit and compare it to previous results. Consider visualizing the residuals and predicted values for further insight.

```{r}
Lesotho_data <- meningitis %>%
  filter(location == "Lesotho") 

Lesotho_data$predicted_mod2 <- predict(mod2, newdata = Lesotho_data)
```


```{r}
#| layout-ncol: 2
#| label: fig-lesotho
#| fig-cap: "Lesotho Death Rates by Country. The dashed line represents the predicted values from the model, while the solid lines represent the observed values."
#| fig-subcap:
#| - "Observed vs Predicted - Mod2"
#| - "Observed vs Predicted - Mod3"
#| fig-alt: "Lesotho Death Rates by Country"
Lesotho_data %>%
  ggplot() +
  geom_point(aes(year, deaths)) +
  geom_line(aes(year, predicted_mod2), 
            linetype = "dashed") +
  labs(title = "Lesotho_data: Observed vs. Predicted",
       x = "Time(Year)", y = "Death Rates") 
  
Lesotho_data %>%
  ggplot(aes(x = year, y = deaths)) +
  geom_point() +
  geom_line(aes(year, predicted_mod2), 
            linetype = "dashed") +
  geom_segment(aes(xend = year, 
                   yend = predicted_mod2),
               linewidth = 0.1) +
  labs(title = "Lesotho: Observed vs. Predicted",
       subtitle = "Segments represent the residuals",
       x = "Time(Year)", y = "Death Rates")  
```

### Example: Ischemic Stroke Decision Tree

In this example we have a look at how to visualise the results of a
decision tree\index{Decision tree} model for predicting Ischemic
Stroke\index{Stroke}.

Load necessary libraries and the data for the Ischemic
Stroke\index{Stroke}[@fes/data].

```{r}
# Ischemic Stroke decision tree
library(tidymodels)
library(rpart)
library(rpart.plot)
```

```{r}
#| echo: false
load(url("https://github.com/topepo/FES/blob/master/Data_Sets/Ischemic_Stroke/stroke_data.RData?raw=true"))
```

Data are already split into training and test sets, we will combine them
for the analysis. The `stroke_train` and `stroke_test` datasets contain
`r nrow(stroke_train)`, `r nrow(stroke_test)` observations respectively
and `r ncol(stroke_test)` variables. Within the variables we have
information on volume, proportion, area, thickness of the arterial wall,
among others. The target variable is `Stroke`, which indicates whether
the patient has had a stroke or not[@johnson].

We select the variables of interest for the analysis with `any_of()`, a
function to select the variables based on their names.

`?any_of()` for more infromation about the function.

```{r}
selected_train <- stroke_train %>%
  dplyr::select(any_of(VC_preds), Stroke)
```

To check which columns in `stroke_train` dataset were not selected:

```{r}
setdiff(names(stroke_train), names(selected_train))
```

Set up the recipe for the data with the `recipe()` function from the
`tidymodels` package. We will use the `step_corr()` function to remove
highly correlated predictors, `step_center()` and `step_scale()` to
standardise the predictors, `step_YeoJohnson()` to transform the
predictors, and `step_zv()` to remove zero variance predictors.

```{r}
is_recipe <- recipe(Stroke ~ ., data = selected_train) %>%
  #step_interact(int_form)  %>%
  step_corr(all_predictors(), threshold = 0.75) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>%
  step_YeoJohnson(all_predictors()) %>%
  step_zv(all_predictors())

is_recipe %>%
  prep() %>% 
  bake(new_data=NULL) %>%
  select(1:5) %>%
  head(5)
```

Set up the decision tree model with the `decision_tree()` function from
the `tidymodels` package. We will use the `rpart` engine for the
decision tree model and set the mode to classification.

```{r}
class_tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")
```

Finally, we will fit the model with the `fit()` function from the
`tidymodels` package and visualise the results with the `rpart.plot`
package.

```{r}
#| label: fig-dv-tree
#| fig-cap: "Decision Tree for Ischemic Stroke"
#| fig-alt: "Decision Tree for Ischemic Stroke"
is_wfl <- workflow() %>%
  add_model(class_tree_spec) %>%
  add_recipe(is_recipe)

is_dt_fit_wfl <- is_wfl %>%
  fit(data = selected_train,
      control = control_workflow())

is_dt_fit_wfl%>%
  extract_fit_engine() %>%
  rpart.plot::rpart.plot(roundint = FALSE)
```

In this plot, the decision tree is visualised with the `rpart.plot` and
the information released evidences the importance of some specific
predictors, such as max wall thickness, max dilatation by area, volume
proportion, and max remodelling ratio. The decision tree is a useful
tool for visualising the results of the model and understanding the
relationships between the predictors and the target variable.

The interpretation of the decision tree is straightforward: each node
represents a decision based on the value of a predictor, and the leaves
represent the final classification. The tree can be pruned to reduce its
complexity and improve its interpretability. The decision tree can be
used to make predictions on new data by following the path from the root
to a leaf node based on the values of the predictors.


### Example: Ischemic Stroke Classification

In this second example we will demonstrate
**classification**\index{Classification} (stroke vs. no stroke) using
patient and plaque imaging features, and visualise model insights for
stakeholder communication.

The objective is to visualise how well the model predict whether a
patient experienced a stroke (Stroke) based on imaging features (e.g.,
MaxStenosisByArea, CALCVolProp) and risk factors (e.g., age,
DiabetesHistory).

Visualise:

-   Variable importance
-   Prediction performance (e.g., ROC curve)
-   Partial dependence of key features

Load necessary libraries, and fit a **random
forest**\index{Random Forest} model to the data with the `rand_forest()`
function from the `{tidymodels}` package. We will use the `ranger`
engine for the random forest model and set the mode to classification.

```{r}
library(tidyverse)
library(tidymodels)
library(vip)
library(DALEXtra)
library(pROC)
```

The model specify the number of trees to grow, the minimum number of
observations in a node before a split is attempted, and the maximum
depth of the tree. The `set_engine()` function specifies the engine to
use for the model, in this case `ranger`, and the `set_mode()` function
specifies the mode of the model, in this case classification.

```{r}
rf_spec <- rand_forest(trees = 500, min_n = 5) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

rf_wf <- workflow() %>%
  add_formula(Stroke ~ .) %>%
  add_model(rf_spec)

rf_fit <- rf_wf %>% fit(data = stroke_train)
```

#### Variable Importance

The object `rf_fit` contains the fitted model. We can extract the the
model specification and use the `vip()` function to visualise what are
the predictors that most influence the model. The `vip()` function
creates a variable importance plot, which shows the importance of each
predictor in the model. The importance is calculated based on the **Mean
Decrease in Impurity (MDI)**\index{Mean Decrease in Impurity (MDI)} for
each predictor.

```{r}
#| label: fig-rf-importance
#| fig-cap: "Variable Importance for Ischemic Stroke"
#| fig-alt: "Variable Importance for Ischemic Stroke"
#| fig-width: 6
#| fig-height: 4
rf_fit %>%
  extract_fit_parsnip() %>%
  vip::vip(num_features = 10)
```

We can conclude that for this data set, the most important predictors
are `MaxStenosisByArea`, `CALCVolProp`, `MaxWallThickness`, and
`MaxRemodellingRatio`.

Next step is to evaluate the model performance. We will use the
`predict()` function to make predictions on the test data set
(`stroke_test`) and calculate the **accuracy**\index{Accuracy} of the
model.

```{r}
set.seed(05122025)
rf_preds <- predict(rf_fit, stroke_test, 
                    type = "prob") %>%
  bind_cols(predict(rf_fit, stroke_test)) %>%
  bind_cols(stroke_test %>% select(Stroke))

rf_preds %>% head()
```

#### Accuracy

The accuracy of the model is calculated as the proportion of correct
predictions. The `accuracy()` function from the `yardstick` package
calculates the accuracy of the model based on the predicted and observed
values.

```{r}
rf_preds %>%
  accuracy(truth = Stroke, .pred_class)
```

In this case, the accuracy of the model is 0.70, which means that the
model correctly predicts whether a patient experienced a stroke or not
70% of the time. The accuracy is calculated as the number of correct
predictions divided by the total number of predictions.

#### ROC Curve

The **Receiver Operating Characteristic
(ROC)**\index{Receiver Operating Characteristic (ROC)} curve is a
graphical representation of the performance of a binary classifier
system as its discrimination threshold is varied. The ROC curve plots
the true positive rate (TPR) against the false positive rate (FPR) at
various threshold settings.

We can use the `roc_curve()` function from the `yardstick` package to
calculate the ROC curve and the area under the curve (AUC). The
`roc_curve` object contains the specificity and sensitivity values for
each threshold, which can be used to calculate the
**AUC**\index{Area Under the Curve (AUC)}.

```{r}
roc_curve <- yardstick::roc_curve(data = rf_preds, 
                       truth = Stroke, .pred_N)
roc_curve %>% head()
```

The relationship between the **True Positive Rate
(TPR)**\index{True Positive Rate (TPR)} and **False Positive Rate
(FPR)**\index{False Positive Rate (FPR)} is visualised in the ROC curve.
The area under the curve (AUC) is a measure of the model's performance,
with values closer to 1 indicating better performance. The
**specificity**\index{Specificity} and
**sensitivity**\index{Sensitivity} of the model indicate the proportion
of true positives and true negatives, respectively, and we can use the
`autoplot()` function to visualise the ROC curve.

```{r}
#| label: fig-roc-curve
#| fig-cap: "ROC Curve for Ischemic Stroke"
#| fig-alt: "ROC Curve for Ischemic Stroke"
#| fig-width: 6
#| fig-height: 4
autoplot(roc_curve)
```

Another way to extract the ROC curve is to use the `pROC` package and
the `roc()` function. The `roc()` function takes the observed values and
the predicted probabilities as arguments and returns an object
containing the ROC curve.

```{r}
roc_obj <- pROC::roc(rf_preds$Stroke, 
                     rf_preds$.pred_N)
roc_obj %>%
  pROC::auc() %>%
  round(2)
```

The `roc_obj` object provides information about the area under the ROC
curve. In this case we have an AUC is `r round(roc_obj$auc,2)`. The AUC
is a measure of the model's performance, values closer to 1 indicate
better performance.

#### Partial Dependence

> "How does the prediction change when a specific feature changes,
> holding all others constant?"

Finally, we attempt an explanation of the model using `DALEX` and
`DALEXtra` packages, and some specific functions, such as:
`explain_tidymodels()` and `model_profile()`.

```{r}
#| output: false
explainer_rf <- 
  DALEXtra::explain_tidymodels(rf_fit, 
                               data = select(stroke_train, -Stroke), 
                               y = stroke_train$Stroke)
```

The `explainer_rf` object contains the model, the data, and the target
variable and it will be used to create the partial dependence plot with
the `model_profile()` function.

A **Partial Dependence Plot (PDP)**\index{Partial Dependence Plot (PDP)}
is a tool used in machine learning to help interpret black-box models
like random forests. It shows the marginal effect of one (or two)
features on the predicted outcome, while averaging out the influence of
all other features.

In this case we specify the `MaxStenosisByArea` variable, to see how
changes in `MaxStenosisByArea` influence the model’s predicted outcome,
on average. A higher value of MaxStenosisByArea indicates more severe
arterial narrowing, which is a risk factor for stroke or heart attack.

```{r}
#| label: fig-partial-dependence
#| fig-cap: "Partial Dependence Plot for Ischemic Stroke: MaxStenosisByArea (arterial narrowing)"
#| fig-alt: "Partial Dependence Plot for Ischemic Stroke"
#| fig-width: 6
#| fig-height: 4
DALEX::model_profile(explainer_rf, 
                     variables = "MaxStenosisByArea") %>%
  plot() +
  labs(
    title = "Partial Dependence Plot for Ischemic Stroke",
    subtitle = "",
    x = "Value",
    y = "AVG Prediction") +
  theme_minimal() 
```

The plot shows the partial dependence of the `MaxStenosisByArea`
variable on the predicted probability of having a stroke. The results
indicate that `MaxStenosisByArea` has a positive influence on stroke
risk, though this effect plateaus beyond a certain threshold.

## Summary

In this chapter, we have learned how to visualise the results of a
model. There are several important considerations when visualising the
results of a model, and differences might arise due to the type of model
used. We have seen how to use several packages and functions, such as
`ggplot2`, `vip`, `pROC`, and `DALEX` to visualise the results of a
model. We have also seen how to interpret the results of a model and how
to communicate them effectively. The examples provided in this chapter
demonstrate how to use visualisation techniques to enhance
decision-making and communicate findings to various stakeholders.
