# Interpreting Model Results Through Visualisation {#sec-10-applications}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   Overview of the basic plots for visualising predictions and outcomes
-   How to customise a plot using colours, palettes, legends, and guides
-   Tell the story with data through plot layouts and saving as an image
:::
::::::

\
This chapter is an essential part of the data
visualisation\index{Data visualisation} section, it focuses on how to
effectively interpret and use the results from machine
learning\index{Machine learning} models in the context of health
metrics and infectious
diseases.\index{Infectious diseases} The aim is to provide actionable
insights, enhance decision-making, and communicate findings to various
stakeholders.

1.  **Visualising Predictions and Outcomes**:
    -   Techniques for visualising predicted vs. actual values.
    -   Using scatter plots, line plots, and bar charts to compare
        predictions and observed data.
    -   Highlighting discrepancies and trends through residual plots and
        error distribution charts.
2.  **Case Studies and Applications**:
    -   Real-world examples of model results application in public
        health.
    -   Case studies demonstrating the impact of machine learning and
        spatial modelling on health metrics.
    -   Lessons learned and best practices from successful
        implementations.

## Practical Insights and Examples

### Example: Visualising COVID-19 Data

Data are from the **ECDC (European Centre for Disease Prevention and
Control)**\index{European Centre for Disease Prevention and
Control (ECDC)} - updated daily. Data include information on the
**number of deaths for each country**. The dataset can be accessed
through the `{covid19.analytics}` package.\index{COVID-19}

```{r}
#| eval: false
install.packages("covid19.analytics")
```

```{r}
#| output: false
library(tidyverse)
library(covid19.analytics)

covid19.ALL.agg.cases <- covid19.data("aggregated")
```

We select the variables of interest, such as `Country_Region`,
`Last_Update`, `Confirmed`, `Deaths`, `Recovered`.

```{r}
data <- covid19.ALL.agg.cases %>% 
  select(Country_Region, Last_Update, 
         Confirmed, Deaths, Recovered) %>%
  mutate(Last_Update = as.Date(Last_Update)) 

data %>% 
  head
```

We group the data by `Last_Update` and calculate the mean of the
Confirmed, Deaths, Recovered variables.

```{r}
data %>%
  group_by(Last_Update) %>%
  summarise(across(Confirmed:Recovered, ~ mean(.x, na.rm = TRUE)))
```

We can visualise the data with a scatter plot with `geom_line`() and
`geom_point()` functions and then add a `geom_smooth()`. This last
function adds the most suitable model line, in this case a linear
regression line, to the plot based on the data.

The plot shows the relationship between the number of confirmed cases
and the number of deaths.

```{r}
#| label: fig-covid19_data
#| fig-cap: "COVID-19 Deaths by Country"
#| fig-alt: "COVID-19 data"
data %>%
  ggplot(aes(x = Confirmed, y = Deaths)) +
  geom_point() +
  geom_line(linetype = "dashed") +
  # add a linear regression line
  geom_smooth() +
  labs(title = "COVID-19 Deaths by Country",
       x = "Confirmed", y = "Number of Deaths") 
```

We don't know if this is a good model, but we can attempt fitting a
linear regression to predict the number of deaths based on the number of
confirmed cases. We fit two models the first one without predictors,
just with the intercept and we obtain the expected values of the number
of `Deaths` in case of no epidemic in action. The second model is fit
with `Confirmed` and `Recovered` as predictors. The`lm()` function is
used to fit the model.

```{r}
mod1 <- lm(Deaths ~ 1, data = data) 
mod2 <- lm(Deaths ~ Confirmed + Recovered, data = data)
summary(mod1)$coefficients;
summary(mod2)$coefficients
```

The `summary()` function provides the coefficients of the model, which
tell us that in absence of epidemics the average number of **Deaths** is
562.4838, while for each additional **Confirmed** case, the number of
deaths increases by 0.025050716, and for each additional **Recovered**
case, the number of deaths decreases by 0.003785725. The intercept is
negative, which means that when the number of confirmed and recovered
cases is zero, the number of deaths is negative. This is not possible,
so we need to interpret the coefficients in the context of the data.

The model is not perfect, but it can be used to estimate the number of
deaths based on the number of confirmed and recovered cases. We can use
the `predict()` function.

```{r}
Pred <- predict(mod2, newdata = data)

data_pred <- cbind(data %>% select(Confirmed, Deaths), Pred) %>%
  # calculate the residuals to see how well the model fits the data
  mutate(Residuals = Deaths - Pred) 

data_pred %>%
  head()
```

The `Residuals` column shows the difference between the actual number of
deaths and the predicted number of deaths. A positive residual means
that the model underestimated the number of deaths, while a negative
residual means that the model overestimated the number of deaths.

```{r}
# metrics
summary(mod2)$r.squared
```

The model explains 85% of the variance in the data, which is very good.
The `r.squared` value is close to 1, which means that the model fits the
data well.

Let's see the results in a plot:

```{r}
#| layout-ncol: 2
#| label: fig-covid19_model
#| fig-cap: "COVID-19 data"
#| fig-subcap: 
#|  - "COVID-19 Deaths by Country"
#|  - "Predicted vs. Actual Deaths"
#| fig-alt: "COVID-19 data"
data_pred %>%
  ggplot(aes(Deaths, Pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  geom_segment(aes(xend = Deaths, yend = Pred), alpha = 0.5) +
  labs(title = "Predicted vs. Actual Deaths",
       x = "Actual Deaths", y = "Predicted Deaths") 

data_pred %>%
  ggplot(aes(Confirmed, Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs. Confirmed Cases",
       x = "Confirmed Cases", y = "Residuals") 
```

Finally, we can visualise the results of the model with a scatter plot
and a line plot. The first plot shows the observed vs. predicted values,
while the second plot shows the residuals. The segments represent the
difference between the observed and predicted values.

```{r}
#| label: fig-lm_results
#| layout-ncol: 2
#| fig-cap: "COVID-19 Model Results"
#| fig-subcap: 
#|  - "Observed vs. Predicted" 
#|  - "Segments represent the residuals"
#| fig-alt: "COVID-19 Model Results"
data_pred %>%
  ggplot() +
  geom_point(aes(Confirmed, Deaths)) +
  geom_line(aes(Confirmed, Pred), linetype = "dashed") +
  labs(title = "Observed vs. Predicted",
       x = "Confirmed Cases", y = "Deaths",
       caption = "Source: ECDC") 
  
data_pred %>%
  ggplot(aes(x = Confirmed, y = Deaths)) +
  geom_point() +
  geom_line(aes(Confirmed, Pred), linetype = "dashed") +
  geom_segment(aes(xend = Confirmed, yend = Pred)) +
  labs(title = "Observed vs. Predicted",
       subtitle = "Segments represent the residuals",
       x = "Confirmed Cases", y = "Deaths",
       caption = "Source: ECDC")   
```

#### Exercise: COVID-19 Case Fatality Ratio

Use the `covid19.ALL.agg.cases` data to visualise the **case fatality
ratio**\index{Case fatality ratio} estimated for each country, by last
update and include the latitude and longitude as further information.
Build a suitable model and explain the results with a plot.

```{r}
cfr <- covid19.ALL.agg.cases %>% 
  select(Country_Region, Last_Update, 
         Lat, Long_, Case_Fatality_Ratio)

cfr %>%
  head()
```

### Example: Ischemic Stroke Decision Tree

In this example we have a look at how to visualise the results of a
decision tree\index{Decision tree} model for predicting Ischemic
Stroke\index{Stroke}.

Load necessary libraries:

```{r}
# Ischemic Stroke decision tree
library(tidymodels)
library(rpart)
library(rpart.plot)
```

Load the data for the Ischemic Stroke from the GitHub repository of the
book: "Feature Engineering and Selection: A Practical Approach for
Predictive Models" by Max Kuhn and Kjell Johnson.

url:
<https://github.com/topepo/FES/blob/master/Data_Sets/Ischemic_Stroke/stroke_data.RData?raw=true>

`load(url(url)))`

```{r}
#| echo: false
load(url("https://github.com/topepo/FES/blob/master/Data_Sets/Ischemic_Stroke/stroke_data.RData?raw=true"))
```

Data are already split into training and test sets, we will combine them
for the analysis. We just need to select the variables of interest for
the analysis with `any_of()`, a function to select the variables of
interest for the analysis.

`?any_of()`

```{r}
selected_train <- 
  stroke_train %>%
  dplyr::select(any_of(VC_preds), Stroke)
```

Set up the recipe for the data with the `recipe()` function from the
`tidymodels` package. We will use the `step_corr()` function to remove
highly correlated predictors, `step_center()` and `step_scale()` to
standardise the predictors, `step_YeoJohnson()` to transform the
predictors, and `step_zv()` to remove zero variance predictors.

```{r}
is_recipe <- recipe(Stroke ~ ., data = selected_train) %>%
  #step_interact(int_form)  %>%
  step_corr(all_predictors(), threshold = 0.75) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>%
  step_YeoJohnson(all_predictors()) %>%
  step_zv(all_predictors())

is_recipe%>%prep()%>%bake(new_data=NULL)%>%select(1:5)%>%head(5)
```

Set up the decision tree model with the `decision_tree()` function from
the `tidymodels` package. We will use the `rpart` engine for the
decision tree model and set the mode to classification.

```{r}
class_tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")
```

Finally, we will fit the model with the `fit()` function from the
`tidymodels` package and visualise the results with the `rpart.plot`
package.

```{r}
#| label: fig-dv-tree
#| fig-cap: "Decision Tree for Ischemic Stroke"
#| fig-alt: "Decision Tree for Ischemic Stroke"
is_wfl <- workflow() %>%
  add_model(class_tree_spec) %>%
  add_recipe(is_recipe)

is_dt_fit_wfl <- is_wfl %>%
  fit(data = selected_train,
      control = control_workflow())

is_dt_fit_wfl%>%
  extract_fit_engine() %>%
  rpart.plot::rpart.plot(roundint = FALSE)
```

## Summary

In this chapter, we have learned how to visualise the results of a
model.
