# Spatial Data Modelling and Visualization {#sec-11-spatials}

```{r}
#| echo: false
library(ggplot2)
map_theme <- ggthemes::theme_map() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(map_theme)
```

In the previous chapters, we explored the fundamentals of data
visualization with various techniques for creating static and
interactive plots. In this chapter, we will learn how to model and
visualize spatial data, which involves analyzing and representing data
with spatial components. We will explore the concepts of spatial data,
spatial data models, and spatial models, and learn how to create maps,
simulate infections, and predict the spatial distribution of phenomena
such as disease spread. By combining spatial data with machine learning
techniques, we can model the spatial dynamics of disease transmission
and assess the risks associated with the spread of infections. We will
use various packages such as `sf`, `ggplot2`, and `gstat` to create
spatial models and visualizations, and gain insights into the spatial
distribution of infections in a specific region.

## Spatial Data, Spatial Data Models and Spatial Models

Essential for understanding the spatial distribution of phenomena such
as disease spread, land cover change, or climate patterns, **spatial
data** and **spatial models** are crucial for analyzing and predicting
spatial processes. But, what is the difference between spatial data,
spatial data models, and spatial models?

Spatial data includes geographic coordinates, addresses, or boundaries.
This type of data is most often represented by one of two **data
models**: **vector** or **raster**. Vector data models use geometric
shapes like points, lines, and polygons to represent spatial features,
whereas raster data models use a grid of cells or pixels, where each
cell has a value representing information such as temperature or land
cover.

Another important distinction is that between **spatial data models**
and **spatial models**. Spatial data models are **frameworks** for
organizing and representing spatial data, connecting our understanding
of events to their representation and processing through algorithms as
spatial primitives and relationships.

On the other hand, spatial models are defined as **process models**,
which represent dynamic spatial processes—phenomena that change over
time, such as the spread of a virus, flood formation, or land cover
change. These models are crucial for understanding and predicting how
spatial phenomena evolve, providing a basis for intervention and
management strategies.

To learn more about the differences and applications of spatial data
models and process models, you can refer to the ArcGIS documentation on
solving spatial problems with representation and process models and the
Esri blog on spatial analysis.

Another great source for practical examples and applications of spatial
data in R is <https://r-spatial.org/> which is the main source for the
most recent developments in spatial data analysis in R.

## Making a Map

To learn how to make a map we first need spatial data. There are various
sources and packages that provide spatial data for different regions and
purposes. One such package is `{rnaturalearth}`, this package contains
various types of boundaries of countries in the world. For more
information type `?rnaturalearth`. In particular, we use the
`ne_countries()` function to get the boundaries of Africa, with the
option `returnclass = "sf"` to return the data as **simple features**.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(sf)
library(rnaturalearth)

africa <- ne_countries(continent = "Africa", 
                       returnclass = "sf")

africa %>% 
  select(geometry)%>%
  head()
```

A simple feature object is a type of spatial data that contains the
geometry of the spatial features (e.g., points, lines, polygons) and
additional attributes such as the name of the country, population, and
other information.

This is a simple map of Africa, but we can make it more interesting by
adding some colors to the countries. A better resolution of the colored
map can be found in the online version of the book.

```{r}
#| message: false
#| warning: false
#| layout-ncol: 2
#| label: africa_map
#| fig.cap: 
#|  - "Map of Africa"
#|  - "Map of Africa with colors"
#| fig.alt: "Map of Africa"
ggplot(data = africa) +
  geom_sf() +
  coord_sf()
# Africa with colors
ggplot(data = africa) +
  geom_sf(aes(fill=name), 
          color= "white", 
          linewidth = 1,
          show.legend = F) +
  coord_sf()
```

## Coordinate Reference System (CRS)

A **coordinate reference system** (CRS) is a standardized way of
defining the spatial location of features on the Earth's surface. It
uses a set of coordinates to represent the position of points, lines,
and polygons in space. There are different types of CRS, such as
**geographic** and **projected** CRS, which are used to represent the
Earth's surface in different ways. Geographic CRS (**LatLong**)
represents locations on a curved surface using latitude and longitude
for global mapping, while projected CRS or **Universal Transverse
Mercator (UTM)** projects the curved Earth onto a flat map, using meters
for precise mapping. In particular, the UTM system provides the distance
in meters from the equator and the central meridian of a specific zone.

We use the `{sf}` package to define and transform CRS. The `st_crs()`
function allows you to retrieve the CRS of a spatial object, while the
`st_transform()` function allows you to transform the coordinates of a
spatial object to a different CRS.

```{r}
africa_crs <- st_crs(africa)
africa_crs
```

For example, the CRS of the Africa map is **WGS 84** or *World Geodetic
System 1984* which is a reference system for the Earth's surface that
defines origin and orientation of the coordinate axes, called a
**datum** (a fact known from direct observation). The **EPSG** is a
structured dataset of CRS and Coordinate Transformations, it was
originally compiled by the, now defunct, European Petroleum Survey
Group[^11-spatials-1]. The EPSG code for WGS 84 is 4326.

[^11-spatials-1]: <https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf>

## Simulate Infections in Central Africa Rep.

Synthetic data are created for the Central African Republic. The
simulation of the number of infected individuals, their location, and
temperature levels are obtained with random numbers generating functions
from the `{stats}` package, which is a base package in R.

The Central African Rep. is a landlocked country in Central Africa, with
a population of approximately 5 million people. The country is known for
its diverse wildlife and natural resources, and has faced challenges
such as political instability and armed conflict.

```{r}
ctr_africa <- africa %>%
  filter(name == "Central African Rep.")
```

#### Bounding Box

We can use the `st_bbox()` function to retrieve the **bounding box** of
the `ctr_africa` object, as a matrix with the minimum and maximum values
of the coordinates. A bounding box is a rectangular area defined by two
points: the lower-left corner (minimum values of the coordinates) and
the upper-right corner (maximum values of the coordinates). It provides
a quick way to determine the spatial extent of a region and is often
used in spatial analysis to define the boundaries of a study area.

```{r}
bbox <- ctr_africa %>% st_bbox()

bbox
```

#### Spatial Coordinates

The `st_coordinates()` function allows to extract the coordinates from a
geometry vector and release a matrix with 5 columns, where the first two
columns are the longitude (X) and latitude (Y), and the other are
indexes to group the coordinates into polygons.

The `as.data.frame()` function converts the matrix to a data frame, and
the `dplyr::select()` function selects the longitude and latitude
columns. The `summary()` function provides a summary of the data frame,
including the mean, median, and quartiles of the longitude and latitude
values.

```{r}
ctr_africa_coords <- ctr_africa %>% 
  sf::st_coordinates() %>% 
  as.data.frame() %>% 
  dplyr::select(X, Y) 

ctr_africa_coords %>%
  summary()
```

Synthetic data are created as an image of the spread of infections
observed in Central Africa on a specific point in time. The data is
generated using random numbers for the number of infected individuals,
their location, and temperature levels. The temperature levels consider
a min of 20.3 degrees Celsius (69 degrees Fahrenheit), a maximum of 29.2
°C (85 °F), and a daily average of 24.7 °C (76 °F).

The synthetic data is created using the `rbinom()`, `rnorm()`, and
`rpois()` functions from the `{stats}` package. The data is generated
for 100 locations in Central Africa, with 70% of the locations infected
and 30% non-infected. The latitude and longitude values are generated
using the `rnorm()` function with mean values of 6.294 and 20.70,
respectively. The number of infected individuals is generated using the
`rpois()` function with a mean value of 10, and the temperature levels
are generated using the `rnorm()` function with a mean value of 25.

Set the number of points:

```{r}
num_points <- 100
```

Simulate Data:

```{r}
set.seed(21082024) # set seed for reproducibility 
# simulate the presence of infection 
# 1 = "infected" and 0 = "non_infected"
longitude <- rnorm(n = num_points, mean = 20.70, sd = 1)
latitude <-  rnorm(n = num_points, mean = 6.294, sd = 1)
presence <- rbinom(100, 1, prob = 0.3)  
cases <- ifelse(presence == 1, 
                 rpois(n = num_points*0.7, lambda = 10),0)
temperature <- rnorm(n = num_points, 
                     mean = 24.7, 
                     sd = (29.2 - 20.3) / 4)

# build a dataframe  
df <- data.frame(longitude, latitude, 
                 presence, cases, temperature)  
  
head(df)
```

The `str()` function allows to inspect the structure of the data frame,
showing the type of each column and the first few rows of the data
frame.

```{r}
df%>%
  filter(presence==1)%>%
  head()%>%
  str()
```

```{r}
#| message: false
#| warning: false
#| layout-ncol: 2
#| label: africa_map2
#| fig.cap: 
#|  - "Map of Africa and Central Africa Rep."
#|  - "Map of Central Africa Rep. Simulation of Infections"
#| fig.alt: "Map of Africa"
ggplot() +
  geom_sf(data = africa) +
  geom_sf(data = ctr_africa, fill = "brown") 

ggplot() +
  geom_sf(data = africa) +
  geom_sf(data = ctr_africa, 
          fill = alpha("brown", alpha = 0.2)) +
  geom_point(data = df,
             aes(x = longitude, y = latitude, 
                 color = factor(presence)),
             shape = ".") +
  labs(title = "Synthetic Data Visualization",
       x = "Longitude",
       y = "Latitude",
       color = "Infection Status") +
  theme(legend.position = "top")
```

## Correlation between Response and Predictor

The correlation coefficient between the number of infected individuals
and the temperature can be calculated using the `cor()` function with
the method set to "pearson". This will provide a measure of the strength
and direction of the linear relationship between the two variables. A
correlation coefficient close to 1 indicates a strong positive linear
relationship, while a coefficient close to -1 indicates a strong
negative linear relationship. A coefficient close to 0 indicates no
significant linear relationship between the variables.

```{r}
cor(df$cases, df$temperature, method = "pearson")
```

The correlation coefficient is -0.052, suggesting that there is a weak
negative linear relationship between the number of infected individuals
and the temperature. This indicates that as the temperature increases,
the number of infected individuals decreases slightly, but the
relationship is not significant.

## Histogram and Scatter Plot

To visualize the distribution of the presence of infection in Central
Africa, we can create a histogram of the presence variable using the
`geom_bar()` function from the `{ggplot2}` package. The histogram shows
the frequency of infected and non-infected locations in Central Africa,
with the x-axis representing the presence of infection and the y-axis
representing the count of locations.

```{r}
#| warning: false
#| message: false
#| layout-ncol: 2
#| label: simulated_data
#| fig.cap: 
#|  - "Histogram of Presence of Infection in Central Africa Rep."
#|  - "Scatter Plot of Temperature and Infections"
#| fig.alt: "Histogram and Scatterplot"
ggplot(data = df, 
       aes(x = factor(presence))) +
  geom_bar() +
  labs(title = "Histogram of Presence of Infection",
       x = "Presence",
       y = "Count") +
  theme_minimal()
 
ggplot(data = df %>% filter(presence== 1), 
       aes(x = temperature, y = cases)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Relationship between Temperature and Number of Infected Individuals",
       x = "Temperature",
       y = "Number of Infected Individuals") +
  theme_minimal()
```

The scatter plot shows the temperature on the x-axis and the number of
infected individuals on the y-axis, with each point representing an
infected location in Central Africa. The `geom_point()` function adds
the points to the plot, while the `geom_smooth()` function fits a smooth
curve to the data, showing the trend between independent and dependent
variables, temperature and the number of infected individuals,
respectively. The `labs()` function adds titles and labels to the plot,
specifying the title, x-axis label, and y-axis label. The
`theme_minimal()` function sets the plot theme to a minimal style.

## Grid of Points

To create a grid of points we transform the `df` data as a simple
features object with the `st_as_sf()` function from the `{sf}` package,
specifying the longitude and latitude columns as the coordinates and the
CRS as 4326[^11-spatials-2]. Then, intersect the data with
`st_intersection(ctr_africa)` to keep only the points within the
country. Finally, select the relevant columns for the analysis.

[^11-spatials-2]: More information on coordinate reference system is
    here: <https://epsg.io/>.

    ```{r}
    #| warning: false
    #| message: false

    df_sf <- df %>%
      st_as_sf(coords = c("longitude","latitude"),
               # or use crs = 4326
               crs = "+proj=longlat +datum=WGS84") %>%
      st_intersection(ctr_africa) %>%
      st_make_valid()

    grid <- df_sf %>%
      st_bbox() %>%
      st_as_sfc() %>%
      st_make_grid(what = "centers",
                   cellsize = .2,
                   square = F) 
    ```

To generate the grid of points to cover all Central African Republic we
can also use the `expand_grid()` function from the `{tidyr}` package,
which generates a series of points within specified bounding box, and
the point in polygon (`PtInPoly()`) function from the `{DescTools}`
package. The `pip` vector contains the information if the point is
inside the polygon or not.

```{r}
library(DescTools)
set.seed(240724) # set seed for reproducibility
bbox_grid <- expand_grid(x = seq(from = bbox$xmin,
                                 to = bbox$xmax,
                                 length.out = 100),
                         y = seq(from = bbox$ymin,
                                 to = bbox$ymax,
                                 length.out = 100))

ctr_africa_grid_full <- data.frame(PtInPoly(bbox_grid,
                                            ctr_africa_coords)) 

ctr_africa_grid_full %>% head()
```

Mapping the grid of points on the map of Central African Republic, we
can visualize the spatial distribution of infected locations.

```{r}
#| warning: false
#| message: false
#| layout-ncol: 2
#| label: africa_map3
#| fig.cap: 
#|  - "Map of Central Africa Rep. with a Window Grid"
#|  - "Grid Map of Central Africa Rep."
#| fig.alt: "Grid-Map of Central Africa Rep."
# Window Grid
ggplot() +
  geom_sf(data = ctr_africa)+
  geom_sf(data = grid, shape = ".") +
  geom_sf(data = df_sf %>% filter(presence == 1),
          aes(fill = presence),
          shape = 21, stroke = 0.3) +
  coord_sf() +
  labs(title = "Window Grid") +
  theme(legend.position = "right")

# Central Africa Grid
ggplot(data = ctr_africa_grid_full) +
  geom_point(aes(x, y, color = factor(pip)), shape = ".") +
  geom_sf(data = df_sf %>% filter(presence == 1),
          aes(fill = presence),
          shape = 21, stroke = 0.3,
          show.legend = F,) +
  coord_sf() +
  scale_color_manual(values = c("1" = "grey20", "0" = "grey")) +
  labs(title = "Grid Map",
       color = "PIP") +
  theme(legend.position= "top")
```

## Create a Raster of Temperature

To visualize the temperature data on the map of Central African Republic
we create a raster template. The raster template has the same extent as
the simulated infected locations and will be used to **rasterize** the
**temperature** data. This means that the temperature data will be
converted into a raster format, with each cell in the raster grid
representing a specific temperature value. The rasterized temperature
data will be visualized as a heatmap on the map, with warmer colors
indicating higher temperatures.

The raster template is created using the `rast()` function from the
`{terra}` package, specifying the number of rows and columns, the
minimum and maximum values for the x and y coordinates, and the
coordinate reference system (CRS) of the raster. If you want to know
more about the package, type `?terra`.

The raster template is set to have 18 rows and 36 columns, with the x
and y coordinates ranging from 11 to 28 and 2 to 12, respectively.

```{r}
#| warning: false
#| message: false
library(terra) 
raster_template <- terra::rast(nrows = 18, ncols = 36,
                          xmin = 11, xmax = 28,
                          ymin = 2, ymax = 12,
                          crs = st_crs(df_sf)$proj4string)
```

Rasterize the temperature data using the `rasterize()` function to
convert the temperature data from the data frame to a raster format
based on the raster template. The temperature data is assigned to the
raster cells based on the latitude and longitude coordinates.

```{r}
ctr_africa_raster <- rasterize(df_sf,
                               raster_template,
                               field = "temperature",
                               fun = max)
```

Before plotting the rasterized temperature data on the map of Central
African Republic, we need to convert the rasterized data to a data frame
format. This will allow us to visualize the temperature data as a
heatmap on the map. The rasterized data contains the temperature values
for each cell in the raster grid, which are assigned based on the
latitude and longitude coordinates of the temperature data.

```{r}
ctr_africa_raster_df <- as.data.frame(ctr_africa_raster, 
                                      xy = TRUE)
ctr_africa_raster_df %>% head()
```

```{r}
#| warning: false
#| message: false
#| layout-ncol: 2
#| label: africa_map4
#| fig.cap: 
#|  - "Temperature Raster"
#|  - "Central Africa Rep. Map of Max Teperature Level"
#| fig.alt: "Raster-Map of Central Africa Rep."
# Temperature Raster
ctr_africa_raster %>%
  plot(main = "Rasterized Temperature")
# Central Africa Raster  
ggplot() +
  geom_point(data = ctr_africa_grid_full %>% filter(pip == 1),
             aes(x = x, y = y), 
             shape = ".") +
  geom_raster(data = ctr_africa_raster_df,
              aes(x = x, y = y, fill = max)) +
  geom_sf(data = df_sf %>% filter(presence == 1),
          aes(size = cases),
          shape = 21, stroke = 0.3) +
  scale_fill_gradient(low = "white", high = "grey30",
                      na.value = "transparent") +
  labs(title = "Max Temperature in Central African Rep.\n(Simulated Values)",
       size = "Number of Infected",
       fill = "Max Temperature") +
  theme(legend.position = "right")
```

This plot shows the temperature data as a heatmap on the map of Central
African Republic, with warmer colors indicating higher temperatures. The
infected locations are represented as circles with size indicating the
number of infected individuals at each location.

## The Epicenter of Infection

A central point that represents the spatial distribution of infections
in the region is the center of mass. It is a useful metric for
understanding the spatial dynamics of disease spread and identifying
critical zones for intervention. To calculate the coordinates of the
**center of mass** for the infections, even called **epicenter** of the
infections, we use the `center_of_mass()` function.

The center of mass is the average of the latitude and longitude of the
infected individuals, weighted by the number of infected individuals at
each location.

```{r}
center_of_mass <- function(df) {
  longitude <- sum(df$cases * df$longitude) / sum(df$cases)
  latitude <- sum(df$cases * df$latitude) / sum(df$cases)
  c(longitude, latitude)
  }
```

```{r}
df_com <- tibble(longitude = center_of_mass(df)[1],
                 latitude = center_of_mass(df)[2])
df_com
```

Finally, plot the infections on the map of Central African Republic,
with the temperature data represented as a raster layer. The temperature
data is displayed as a **heatmap**, with warmer colors indicating higher
temperatures. The infections are shown as points on the map, with
different colors representing infected and non-infected locations.

Note, here a new package is used, the `{ggnewscale}` package, which is a
package that allows you to add multiple color scales to a single plot.
This is useful when you want to visualize different variables with
different color scales in the same plot.

```{r}
#| warning: false
#| message: false
#| layout-ncol: 2
#| label: africa_map5
#| fig.cap: 
#|  - "Scatter plot of Simulated Locations"
#|  - "Infections in Central Africa Rep."
#| fig.alt: "Map of Central Africa Rep."
# Location of Infections   
ggplot() +
  geom_sf(data = df_sf, aes(color = factor(presence))) +
  geom_point(data = df_com,
             aes(x = longitude, y = latitude), 
             color = "red", size = 4) +
  labs(title = "Location of Infections in Central Africa Rep.",
       subtitle = "Synthetic Data",
       color = "Infection Status") +
  theme(legend.position = "right")

# Infections Size on Max Temperature  
ggplot() +
  geom_point(data = ctr_africa_grid_full %>% filter(pip == 1),
            aes(x = x, y = y), shape = ".") +
  geom_sf(data = ctr_africa, fill = NA) +
  geom_raster(data = ctr_africa_raster_df,
              aes(x = x, y = y, fill = max)) +
  scale_fill_gradient(low = "white", high = "grey",
              na.value = "transparent") +
  # set a new scale color
  ggnewscale::new_scale_color() +
  geom_sf(data = df_sf  %>% filter(presence == 1),
             mapping = aes(group = presence, 
                           color = factor(presence)),
             alpha = 0.8, show.legend = F) +
  geom_sf(data = df_sf %>% filter(presence == 1),
          aes(size = cases),
          shape = 21, stroke = 0.3) +
  geom_point(data = df_com,
             aes(x = longitude, y = latitude), 
             color = "red", size = 4) +
  labs(title = "Infections Size on Max Temperature",
       subtitle = "Synthetic Data",
       fill = "Temperature", 
       size = "Number") +
  theme(legend.position = "right")
```

::: {.content-visible when-format="html"}
::: solutionbox
**Interactive Maps**

The infections can be visualized with `{mapview}` or `{leaflet}`, type of
interactive maps, here is a code snippet of how to do it.

**Map with Mapview:**

This is the interactive map with the max temperatures for the simulated
data in Central African Republic. The data represented as a raster
points, with warmer colors indicating higher temperatures are shown as
points on the map.

```{r}
#| label: fig-mapview-heatmap
#| fig-cap: "Interactive Temperature Raster Points"
#| fig-alt: "Interactive Temperature"  
library(mapview)

mapview(df_sf,
        zcol = "temperature",
        map.types = "CartoDB.Voyager")
```

**Map with Leaflet:**

The `{leaflet}` and `{leaflet.extras}` packages are widely explained in
the package documentation. Here is shown howe to make a `heatmap` with
the infections and the center of mass.

```{r}
#| label: fig-leaflet-heatmap
#| fig-cap: "Interactive Heatmap"  
#| fig-alt: "Interactive Heatmap"
library(leaflet)
library(leaflet.extras)
# define center of map
lat_center <- as.numeric(df_com[2])
long_center <- as.numeric(df_com[1])
# creating intensity heat map
heatmap <- df %>%
    leaflet() %>%
    addTiles() %>%
    setView(long_center, lat_center, zoom = 4.5) %>%
    addHeatmap(lng = ~longitude, lat = ~latitude,
               intensity = ~cases,
               max = 10, radius = 20, blur = 10) %>%
    # parameter to set
    addMarkers(lng = long_center, lat = lat_center)
heatmap
```

Then, we can save the map as an image with the `mapshot()` function from
the `{mapview}` package.

```{r}
#| eval: false
# webshot::install_phantomjs()
mapview::mapshot(heatmap, file = "images/heatmap.png")
```
:::
:::

## Dynamics of Disease Transmission

The spread of infectious diseases, fundamentally depends on the pattern
of human contact[@volz2011] and the spatial distribution of infected
individuals. Understanding the spatial dynamics of disease transmission
is crucial for predicting the risk of outbreaks and guiding public
health interventions.

To explore the dynamics of disease transmission in Central Africa by
simulating the spread of infections we use a type of network
characterized by a high degree of clustering and short average path
lengths between nodes called **small-world network**. It represents a
realistic model of human contact patterns, where individuals are more
likely to interact with their neighbors but can also have long-range
connections with other individuals. By simulating the spread of
infections in a small-world network, we can investigate how the spatial
structure of the network influences the transmission dynamics and
identify critical nodes for disease control.

We will use the `{igraph}` package to create a small-world network and
simulate the spread of infections through the network. The network will
consist of nodes representing individuals in Central Africa, with
connections between nodes based on a small-world topology. The
`sample_smallworld()` function generates a small-world network with the
specified parameters: the number of nodes `N`, the average degree `k`,
and the rewiring probability `p`. The average degree `k` determines the
number of neighbors each node is connected to, while the rewiring
probability `p` controls the likelihood of rewiring connections to
create long-range connections. By adjusting the parameters `k` and `p`,
we can create small-world networks with different topologies and study
their impact on disease transmission dynamics.

```{r}
#| message: false
#| warning: false
library(igraph)

N <- nrow(df_sf)  # Number of nodes
k <- 3    # Each node connected to k nearest neighbors
p <- 0.1   # Rewiring probability
small_world <- sample_smallworld(dim = 1,
                                 size = N,
                                 nei = k,
                                 p = p)
```

Assign the small_world to a new variable `small_world_dev` and infect 10
random nodes. The status of the nodes is defined as "S" for susceptible
and "I" for infectious. The color of the nodes is assigned based on
their infection status, with infectious nodes shown in red and
susceptible nodes in gold.

```{r}
small_world_dev <- small_world
# Infect 10 random nodes
V(small_world_dev)$status <- "S"  # Susceptible
infected_nodes <- sample(V(small_world_dev), 10)
V(small_world_dev)$status[infected_nodes] <- "I"  # Infectious

# Assign colors based on infection status
V(small_world_dev)$color <- 
           ifelse(V(small_world_dev)$status == "I",
                               "red", "gold")
```

Plot the graphs:

```{r}
#| warning: false
#| message: false
#| label: small_world
#| fig.cap: "Small-World Network"
#| fig.alt: "Small-World Network"  
par(mfrow=c(1,2))

set.seed(06082024)
plot(small_world,
     vertex.size=5,
     vertex.label=NA,
     edge.arrow.size=0.5,
     edge.width=0.5)

plot(small_world_dev,
     vertex.size=5,
     vertex.label=NA,
     edge.arrow.size=0.5,
     edge.width=0.5)
```

We use the Central Africa Rep. grid of points for prediction, and
transform it to a simple features object, with the mean values of the
covariates for simplicity.

```{r}
ctr_africa_grid <- ctr_africa_grid_full %>% 
  filter(pip == 1)

ctr_africa_grid_sf <- ctr_africa_grid %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_make_valid() %>%
  mutate(temperature = mean(df$temperature))

infected_sf <- df_sf  %>%
  filter(presence == 1) %>%
  st_make_valid()
```

Visualize the grid of points and the observed infections on the map of
Central African Republic. To understand the spatial distribution of
infections and the covariates for prediction, we use the small-world
network made above, to simulate the connections between individuals.
Assign the attributes to the nodes of the small-world network based on
the synthetic data (`df_sf`). The attributes include the type of
individual (infected or non-infected), the latitude and longitude
coordinates, and the number of infected individuals. The edges of the
network represent the connections between nodes.

```{r}
coords <- st_coordinates(df_sf$geometry)
# Assign attributes to nodes
V(small_world)$presence <- df_sf$presence
V(small_world)$longitude <- coords[,1]
V(small_world)$latitude <- coords[,2]
V(small_world)$cases <- df_sf$cases
```

```{r}
edges <- igraph::as_data_frame(small_world, what = "edges")

nodes <- coords %>%
     as_tibble() %>%
     rename(longitude = X, latitude = Y)%>%
     mutate(id = 1:nrow(.))

edges <- edges %>%
  left_join(nodes, by = c("from" = "id")) %>%
  rename(lat_from = latitude,
         lon_from = longitude) %>%
  left_join(nodes, by = c("to" = "id")) %>%
  rename(lat_to = latitude,
         lon_to = longitude)
```

```{r}
#| warning: false
#| message: false
#| layout-ncol: 2
#| label: africa_map6
#| fig.cap: 
#|  - "Observed Infections"
#|  - "Map of Central Africa Rep. with Small-World Network"
#| fig.alt: "Map of Central Africa Rep."
ggplot() + 
  geom_sf(data = ctr_africa_grid_sf, 
          shape = 21, stroke = 0.5, 
          fill = NA, size = 0.5, color = "grey") +
  geom_sf(data = infected_sf,
          aes(size = cases),
          shape = 21, stroke = 0.2, color="black") +
  scale_color_continuous(name = "Observed", 
                         low = "white", high = "red") +
  labs(title = "Observed Infections")  +
  theme(legend.position = "right")
  
ggplot() +
  geom_sf(data = ctr_africa, fill = NA) +
  geom_sf(data = ctr_africa_grid_sf, 
          shape = 21, stroke = 0.5, 
          fill = NA, size = 0.5, color = "grey") +
  geom_segment(data = edges,
               aes(x = lon_from, y = lat_from,
                   xend = lon_to, yend = lat_to),
               alpha = 0.3, linewidth = 0.2) +
  geom_sf(data = df_sf,
             aes(shape = factor(presence), 
                 color = factor(presence)),
          size = 0.5) +
  geom_sf(data = infected_sf,
          aes(size = cases),
          shape = 21, stroke = 0.2, color="black") +
  guides(color = "none") +
  scale_color_manual(values = c("0" = "grey20", "1" = "red")) +
  labs(title = "Small-World Network with Infection Status",
       x = "Longitude", y = "Latitude", shape = "Presence") +
  theme(legend.position = "right")
```

The clustering of close-proximity contacts that occurs within
individuals is an important factor in the spread of diseases and subject
of many mathematical models[@volz2011].

## The Euclidean distance

To estimate the likelihood of transmission between individuals based on
their spatial proximity the **Euclidean** distance is used. It is
calculated as the square root of the sum of the squares of the
differences between corresponding coordinates of the two points.

The Euclidean distance formula:

$$d_{i,j} = \sqrt{(x_i - x_j)^2 + (y_i - y_j)^2}$$

Where: $(x_i, y_i)$ are the coordinates (e.g., latitude and longitude)
of the first point (A), e.g., infected individual, and $(x_j, y_j)$ are
the coordinates of the second point (B), e.g., susceptible individual or
location.

It calculates the straight-line distance between two points in a
two-dimensional space, which can be applied to model the spatial spread
of infectious diseases based on the proximity of individuals or
locations.

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig.cap: "Representation of Euclidean Distance between Points A and B"
#| fig.width: 6
#| fig.height: 2
# Define the coordinates for the points A and B
points <- data.frame(x = c(0, 3),
                     y = c(0, 4),
                     label = c("A", "B"))

# Calculate the Euclidean distance between the points
euclidean_distance <- sqrt((points$x[2] - points$x[1])^2 +
                      (points$y[2] - points$y[1])^2)

ggplot(points, aes(x = x, y = y)) +
  geom_point(size = 4) +              
  geom_text(aes(label = label), vjust = -1) + 
  geom_segment(aes(x = x[1], y = y[1], 
               xend = x[2], yend = y[2]), 
               linetype = "solid") + 
  annotate("text", x = 1.5, y = 2, 
            label = paste("d =", 
                    round(euclidean_distance, 2)), 
                    hjust = -0.1) + 
  xlim(-1, 4) + ylim(-1, 5) +
  theme_void() 
```

To calculate the Euclidean distance we make a function that considers
the distance between the center of mass and the infection coordinates.

```{r}
# helper function for calculating euclidean distance metric
euclidean_distance <- function(longitude, latitude, 
                               long_com, lat_com) {
  long_distances <- (longitude - long_com[1])^2
  lat_distances <- (latitude - lat_com[1])^2
  return((long_distances + lat_distances) * 0.5)
}
```

There are other type of distance metrics that can be used, such as the
**Manhattan** distance, which calculates the sum of the absolute
differences between the coordinates of two points, the **Mahlanobis**
distance, which considers the correlation between the variables, and the
**Chebyshev** distance, which calculates the maximum difference between
the coordinates of two points.

## Spatial Autocorrelation

Spatial autocorrelation refers to the principle that spatial data points
close to each other are more likely to have similar values than those
further apart. It is checked on the simulated data to understand how the
location of infections are correlated with each other.

To check the presence of spatial autocorrelation the Global Moran’s law
is used:
$$I = \frac{N \sum_{i=1}^{N} \sum_{j=1}^{N} w_{ij} (x_i - \bar{x})(x_j - \bar{x})}{W \sum_{i=1}^{N} (x_i - \bar{x})^2}$$
Where $x_i$ and $x_j$ are the values of the variable of interest at
location $i$ and $j$, $\bar{x}$ is the mean value and $N$ indicates the
number of spatial units, and $w_{ij}$ are the spatial weight between
locations.

$I$ measures the degree, to which infection is clustered, dispersed or
randomly distributed, with values range from −1 indicating perfect
dispersion to +1 indicating perfect spatial clustering. A value of 0
indicates no spatial autocorrelation, suggesting a random spatial
pattern.

We use the `{spdep}` package, which provide `dnearneigh()`,
`nb2listw()`, and `moran.test()` functions for calculating the
neighbourhood contiguity by distance, the spatial weights for neighbours
lists, and finally the Moran's I test for spatial autocorrelation,
respectively.

```{r}
#| warning: false
#| message: false
library(spdep)
library(sp)

coordinates(df) <- ~longitude + latitude
# Define neighbors (using a distance threshold, for example)
# Neighbors within a distance of 2 units
nb <- dnearneigh(coordinates(df), 0, 2)  

# Create spatial weights
lw <- nb2listw(nb, style="W")

# Calculate Moran's I for infections
moran_result <- moran.test(df$cases, lw)
moran_result
```

The result of the Moran's I test of -0.013 shows a slight negative
spatial autocorrelation, indicating that the location of infected
individuals is slightly dispersed across the region. This suggests that
the spatial distribution of infections is not significantly clustered or
dispersed, but rather randomly distributed. The p-value of 0.5731
indicates that the observed Moran's I value is not statistically
significant, suggesting that the spatial distribution of infections is
consistent with a random spatial pattern. And we know it is as we have
simulated the data.

## Spatial Proximity with Kriging

Although, the Moran's I test is useful for understanding the overall
spatial pattern of infections, it does not provide detailed information
on the spatial distribution of cases across the region. To estimate the
spatial distribution of infections and predict the risk of outbreaks, we
can use spatial interpolation techniques such as
**Kriging**[@kriging2024].

This method used to estimate the spatial distribution of cases by
creating a raster of the predicted the spatial distribution of
infections. It creates a continuous surface that estimates the risk of
infection across the entire study area by interpolating the number of
infected individuals across a region.

For example, the application of Kriging was used to predict the spatial
distribution of Dengue outbreaks in regions like Khyber Pakhtunkhwa,
Pakistan[@ahmad2020]. Similarly, during the COVID-19 pandemic,
**Empirical Bayesian Kriging (EBK)**[@onovo] was used to estimate the
spatial distribution of COVID-19 cases in sub-Saharan Africa. This
approach combined time series data of confirmed cases with
socio-demographic indicators to create detailed spatial risk maps,
aiding in understanding and controlling the outbreak.

Particularly useful for **visualizing the spatial dynamics** of disease
spread, while considering the impact of environmental factors such as
temperature, or humidity on disease transmission; Kriging requires an
estimation of the variance of points at locations.

For estimating the variance, a **variogram model** is fit to the
observed data. The variogram is a function of the distance `h` that
describes the degree of spatial dependence of a spatial random field or
stochastic process. It takes consideration of the spatial
auto-correlation of the data. This means, for instance, the number of
cases observed in one location are supposed to be correlated with the
number of cases observed in nearby locations. This also considers the
enverinmental variables that interact with the disease transmission.

The variogram model formula:

$$\gamma(h)= \frac{1}{2N(h)} \sum_{i=1}^{N(h)} (Z(x_i) - Z(x_i + h))^2$$

Where: $\gamma(h)$ is the semivariance, $N(h)$ is the number of pairs of
points separated by a distance $h$, $Z(x_i)$ is the value of the
variable at location $x_i$, and $Z(x_i + h)$ is the value of the
variable at location $x_i + h$.

We use the `{gstat}` package to perform Kriging with
`variogram()`[@pebesma2004] [@wackernagel2003], `fit.variogram()`, and
`gstat()` functions to perform **Universal Kriging**, a kriging method
that allows to incorporate an external drift, such as temperature. The
predictions are visualized on the grid of points generated earlier, with
warmer colors indicating higher predicted values.

```{r}
library(gstat)
```

```{r}
v <-  variogram(object = cases ~  temperature, data = df_sf)

v_model <- fit.variogram(v, model = vgm("Sph"))
  
plot(v, model = v_model)
```

Or, just use the `kbfit()` function[@kriging], which tries different
initial values and models to find the best fit for the variogram, the
function can be used directly from the book package `hmsidwR::kbfit()`.

Then, the Kriging technique is applied by using the `gstat::gstat()`
function[@moraga], to predict the number of infected individuals at
unobserved locations, based on the best variogram model. The predictions
are visualized on the grid of points generated earlier, with warmer
colors indicating higher predicted values.

Perform Kriging with the best variogram model and predict on the grid of
points generated earlier. The predictions are stored in the `kpred`
object, which contains the predicted values and the variance of the
predictions.

```{r}
# Perform Kriging 
k <- gstat::gstat(formula = presence ~ temperature, 
                  data = df_sf, 
                  model = v_model)

kpred <- predict(k, newdata = ctr_africa_grid_sf)

kpred %>% head()
```

Visualize predictions:

```{r}
#| warning: false
#| message: false
#| layout-ncol: 2
#| label: africa_map7
#| fig.cap: 
#|  - "Kriging Prediction in Central Africa Rep."
#|  - "Kriging Variance in Central Africa Rep."
#| fig.alt: "Kriging Map of Central Africa Rep."
ggplot() +
  geom_sf(data = kpred, 
          aes(fill = var1.pred),
          shape=21, stroke=0.5) +
  geom_sf(data = infected_sf) +
  scale_fill_viridis_c() +
  labs(title = "Kriging Prediction in Central Africa Rep.") +
  theme(legend.position = "right")

ggplot() +
  geom_sf(data = kpred,           
          aes(fill = var1.var),
          shape=21, stroke=0.5) +
  geom_sf(data = infected_sf) +
  scale_fill_viridis_c() +
  labs(title = "Kriging Variance in Central Africa Rep.") +
  theme(legend.position = "right")
```

## Mapping Risk of Infections

To map the risk of infections in Central Africa, we can combine the
Kriging predictions with the spatial distribution of infections and the
small-world network. The Kriging predictions provide an estimate of the
risk of infections across the region, while the spatial distribution of
infections and the small-world network represent the spatial dynamics of
disease transmission.

The risk of infections is visualized on the map of Central African
Republic, with warmer colors indicating higher predicted values. The
spatial distribution of infections is shown as points on the map, with
different colors representing infected and non-infected locations. The
small-world network is overlaid on the map, showing the connections
between nodes and the spatial dynamics of disease transmission.

This is a different way to visualize the Kriging predictions, first a
data frame is created with the coordinates and the predicted values, and
then the predictions plotted on the map of Central African Republic. The
predictions are shown as a **raster layer**.

```{r}
kriging_df <- kpred %>%
  sf::st_coordinates() %>% 
  cbind(as.data.frame(kpred) %>% select(var1.pred)) %>%
  rename(longitude = X, latitude = Y) 
```

```{r}
#| warning: false
#| message: false
#| label: africa_map8
#| fig.cap: "Kriging Prediction of Infections Size on Max Temperature"
#| fig.alt: "Kriging Map of Central Africa Rep."
ggplot() +
  geom_sf(data = ctr_africa, fill = "grey") +
  geom_raster(data =  kriging_df,
              aes(x = longitude, y = latitude, 
                  fill = var1.pred)) +
  scale_fill_viridis_c() +
  geom_segment(data = edges,
               aes(x = lon_from, y = lat_from,
                   xend = lon_to, yend = lat_to),
               color = "grey", alpha = 0.2, linewidth = 0.2) +
  geom_sf(data = df_sf %>% filter(presence == 1),
             aes(size = cases),
         # show.legend = F,
             shape = 21, stroke = 0.2) +
  geom_sf(data = df_sf,
          aes(color = factor(presence)),
          shape = ".") +
  scale_color_manual(values = c("0" = "grey", "1" = "red")) +
  guides(size = guide_legend(order = 1), 
         color = guide_legend(order = 2), 
         fill = guide_colorbar(order = 3)) +
  labs(title = "Kriging Prediction of Infections Size on Max Temperature",
       fill = "Prediction of Infections",
       size = "Infections Size",
       color = "Presence") +
  ggthemes::theme_map() +
  theme(legend.position = "bottom", 
        legend.box = "vertical",
        legend.key.size = unit(0.5, "cm"),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6))
```

Finally, the last plot is made with the `spplot()` function from the
`{sp}` package, which creates a **spatial plot** of the Kriging
predictions. The predictions are visualized as a **heatmap** on the map
of Central African Republic, with warmer colors indicating higher
predicted values. The plot provides a detailed view of the spatial
distribution of infections and the risk of outbreaks across the region.

```{r}
#| warning: false
#| message: false
#| label: africa_map9
#| fig.cap: "Kriging Prediction of Infections with a Spplot"
#| fig.alt: "Kriging Map of Central Africa Rep."
kriging_sp <- kriging_df
coordinates(kriging_sp) <- ~longitude+latitude
spplot(kriging_sp, "var1.pred", main = "Kriging Prediction of Infections")
```

## Summary

In this chapter, we have explored the spatial dynamics of disease
transmission in Central Africa using a combination of spatial analysis
techniques and visualization tools. We have simulated the spatial
distribution of infections, created a grid of points to cover the
region, and visualized the spatial distribution of infections on the map
of Central African Republic. We have used a small-world network to model
the spatial connections between individuals and simulated the spread of
infections through the network. We have shown how the Euclidean distance
between points is calculated, to estimate the likelihood of transmission
based on spatial proximity and performed spatial autocorrelation
analysis to understand the spatial pattern of infections. We have used
Kriging to estimate the spatial distribution of infections and predict
the risk of outbreaks across the region. The results provide valuable
insights into the spatial dynamics of disease transmission and the
spatial distribution of infections in Central Africa, which can inform
public health interventions and guide efforts to control the spread of
infectious diseases.
