# Spatial Data Modelling and Visualization {#sec-11-spatials}

```{r}
#| echo: false
library(ggplot2)
map_theme <- ggthemes::theme_map() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(map_theme)
```

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   How to make a map
-   Identify the missing pieces
-   Map the results
:::
:::

This chapter is dedicated to spatial modelling and visualization. All
that we saw in the previous chapters is now applied to spatial data. We
will use the `{sf}` package to handle spatial data and the `{ggplot2}`
package to make maps, in addition to other interesting packages. Also,
we will see how to make spatial models and how to visualize them.

## Making a Map

In this section, we will learn how to make a map using spatial data from
the `{rnaturalearth}` package. This package contains various types of
boundaries of countries in the world. For more information type
`?rnaturalearth`. Also for surface meteorological data we use the
`{worldmet}` package, which import data from NOAA Integrated Surface
Database (ISD).

We use the `rnaturalearth::ne_countries()` function to get the
boundaries of Africa, with the option `returnclass = "sf"` to return the
data as **simple features**. The other option is to return the data as
**spatial data** with the option `returnclass = "sp"`. This means that
with an `sf` object we obtaining a data frame with the geometry of the
countries, while with an `sp` object we obtain the longitude and
latitude of the countries.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(sf)
library(rnaturalearth)

africa <- ne_countries(continent = "Africa", returnclass = "sf")

africa %>% class()
```

The `africa` dataset contains the boundaries of African countries as
simple features, and other information such as the country name,
continent, population, and geometry of the country. This is a simple map
of Africa, but we can make it more interesting by adding some colors to
the countries. A better resolution of the colored map can be found in
the online version of the book.

```{r}
#| layout-ncol: 2
#| label: fig-simple-africa
#| fig-cap:
#|  - "Simple Africa"
#|  - "Africa with colors"
#| fig-alt: "Africa Country"
# Simple Africa  
ggplot(data = africa) +
  geom_sf() 
# Africa with colors
ggplot(data = africa) +
  geom_sf(aes(fill=name), 
          color= "white", 
          linewidth = 1,
          show.legend = F) 
```

## Locate Infections with Montecarlo Simulations

Now that the map of Africa is ready, we can simulate infections in a
specific country, such as the Central African Republic. We will create
synthetic data for the infections, including the number of infected
people, their location, and other variables such as wind speed,
temperature, precipitation, and humidity. This data will be used to
visualize the spatial distribution of infections and assess the risks
associated with the spread of the disease.

```{r}
ctr_africa <- africa %>%
  filter(name == "Central African Rep.")
```

With the `sf::st_bbox()` we can retrieve the bounding box of the
`ctr_africa` object, which is a matrix with the minimum and maximum
values of the coordinates.

```{r}
bbox <- ctr_africa %>% st_bbox()

bbox
```

And retrieve the mean values for the longitude and latitude with the
`sf::st_coordinates()` function to extract the coordinates and then,
extract the information with the `summary()` function.

```{r}
ctr_africa_coords <- ctr_africa %>% 
  sf::st_coordinates() %>% 
  as.data.frame() %>% 
  dplyr::select(X, Y) 

ctr_africa_coords %>%
  summary()
```

Then, create a grid of points within the bounding box of the Central
African Republic, with the `DescTools::PtInPoly()` function. The
function checks if a point is inside a polygon, which in this case is
the Central African Republic.

```{r}
library(DescTools)
set.seed(240724) # set seed for reproducibility
bbox_grid <- expand_grid(x = seq(from = bbox$xmin,
                                       to = bbox$xmax,
                                       length.out = 100),
                               y = seq(from = bbox$ymin,
                                       to = bbox$ymax,
                                       length.out = 100))
ctr_africa_grid <- data.frame(DescTools::PtInPoly(bbox_grid,
                               ctr_africa_coords)) 

ctr_africa_grid %>% head()
```

```{r}
#| layout-ncol: 2
#| label: fig-grid-africa
#| fig-cap:
#|  - "Central Africa"
#|  - "Central Africa Grid"
#| fig-alt: "Central Africa Grid"
# Central Africa
ggplot() +
  geom_sf(data = africa) +
  geom_sf(data = ctr_africa, fill = "grey20") +
  labs(title = "Central Africa Region")
# Central Africa Grid
ggplot(data = ctr_africa_grid) +
  geom_point(aes(x,y, color = factor(pip)), shape = ".") +
  scale_color_manual(values = c("1" = "red", "0" = "grey")) +
  labs(title = "Central Africa Grid",
       color = "PIP") +
  theme(legend.position= "top")
```

Let's now create the synthetic data containing not only the number of
infections in different locations in Central Africa but also other
variables such as wind speed, temperature, precipitation, and humidity.
The data frame is made of 100 point locations, with 70% of infected
locations and 30% of non-infected locations. The latitude and longitude
are generated from a normal distribution with mean values of 6.294 and
20.70, respectively. The number of infected people is generated from a
Poisson distribution with a mean value of 10. The wind speed is
generated from a Weibull distribution, the temperature from a normal
distribution, the precipitation from a gamma distribution, and the
humidity from a beta distribution.

Set the number of points (simulations):

```{r}
num_points <- 100
```

In particular, for generating random data for the different variables in
the synthetic data, we use the following functions:

-   The `rnorm()` function generates random numbers from a **Normal**
    distribution for the temperature, and latitude and longitude,
    considering the mean values of the Central African Republic.

-   The `rpois()` function generates random numbers from a **Poisson**
    distribution for the number of infected people, with a mean value of
    lambda equal to 10.

-   The `rweibull()` function generates random numbers from a
    **Weibull** distribution for the wind speed.

-   The `rgamma()` function generates random numbers from a **Gamma**
    distribution for the precipitation.

-   The `rbeta()` function generates random numbers from a **Beta**
    distribution for the humidity.

```{r}
set.seed(200724) # set seed for reproducibility 
  
type <- ifelse(1:num_points > num_points * 0.7, "infected", "non_infected")
latitude <-  rnorm(n = num_points, mean = 6.294, sd = 1)
longitude <- rnorm(n = num_points, mean = 20.70, sd = 1)
number <- ifelse(type == "infected", rpois(n = num_points*0.7, lambda = 10),0)
temperature <- rnorm(n = num_points, mean = 25, sd = 5)
wind_speed <- rweibull(n = num_points, shape = 2, scale = 10)
precipitation <- rgamma(n = num_points, shape = 2, scale = 5)
humidity <- rbeta(n = num_points, shape1 = 2, shape2 = 5) * 100 
sunlight_hours <- rnorm(n = num_points, mean = 8, sd = 2)
  
df <- data.frame(type, latitude, longitude, number, 
                 wind_speed, temperature, precipitation, 
                 humidity, sunlight_hours)  
  
head(df[c(1:5)])
```

Then, we create a raster template to plot the temperature data on the
map of Central African Republic. The raster template has the same extent
as the simulated infected locations and will be used to **rasterize**
the **temperature** data. This means that the temperature data will be
converted into a raster format, with each cell in the raster grid
representing a specific temperature value. The rasterized temperature
data will be visualized as a heatmap on the map, with warmer colors
indicating higher temperatures.

The raster template is created using the `rast()` function from the
`{terra}` package, specifying the number of rows and columns, the
minimum and maximum values for the x and y coordinates, and the
coordinate reference system (CRS) of the raster. If you want to know
more about the raster package, type `?terra`.

Transform the `df` data as a simple features object:

```{r}
df_sf <- st_as_sf(df,
                  coords = c("longitude", "latitude"),
                  crs = 4326)

df_sf %>% st_bbox()
# xmin      ymin      xmax      ymax
# 17.628586  3.676935 23.087899  8.487271
```

The raster template is set to have 18 rows and 36 columns, with the x
and y coordinates ranging from 11 to 28 and 2 to 12, respectively.

```{r}
#| warning: false
#| message: false

library(terra) # load the terra package 
raster_template <- terra::rast(nrows = 18, ncols = 36,
                          xmin = 11, xmax = 28,
                          ymin = 2, ymax = 12,
                          crs = st_crs(df_sf)$proj4string)
```

Rasterize the temperature data using the `rasterize()` function from the
`{raster}` package. The rasterize function converts the temperature data
from the data frame to a raster format based on the raster template. The
temperature data is assigned to the raster cells based on the latitude
and longitude coordinates.

```{r}
ctr_africa_raster <- terra::rasterize(df_sf,
                          raster_template,
                          field = "temperature",
                          fun = max)
```

Before plotting the rasterized temperature data on the map of Central
African Republic, we need to convert the rasterized data to a data frame
format. This will allow us to visualize the temperature data as a
heatmap on the map. The rasterized data contains the temperature values
for each cell in the raster grid, which are assigned based on the
latitude and longitude coordinates of the temperature data.

```{r}
ctr_africa_raster_df <- as.data.frame(ctr_africa_raster, 
                                      xy = TRUE)
ctr_africa_raster_df %>% head()
```

```{r}
#| layout-ncol: 2
#| label: fig-raster-africa
#| fig-cap:
#|  - "Temperature Raster"
#|  - "Central Africa Raster"
#| fig-alt: "Central Africa Raster"
# Temperature Raster
ctr_africa_raster %>%
  plot(main = "Simulated Temperature in Central African Rep.")
# Central Africa Raster  
ggplot() +
 geom_point(data = ctr_africa_grid %>% 
                      filter(pip == 1),
            aes(x,y), shape = ".") +
 geom_raster(data = ctr_africa_raster_df,
              aes(x = x, y = y, fill = max)) +
  scale_fill_gradient(low = "white", high = "black",
              na.value = "transparent") +
 labs(title = "Max Temperature in Central African Rep.(Simulated Values)",
      fill = "Max Temperature") 
```

In addition, we add the epicenter of the infections to the map to
visualize the center of mass of the infections.

## The epicenter of infection

To calculate the coordinates of the **center of mass** for the infections, even
called **epicenter** of the infections, we use the `center_of_mass()` function.
The center of mass is the average of the latitude and longitude of the infected individuals,
weighted by the number of infected individuals at each location. This
provides a central point that represents the spatial distribution of
infections in the region. The center of mass is a useful metric for
understanding the spatial dynamics of disease spread and identifying
critical zones for intervention.

```{r}
center_of_mass <- function(df) {
  longitude <- sum(df$number * df$longitude) / sum(df$number)
  latitude <- sum(df$number * df$latitude) / sum(df$number)
  return(c(as.numeric(longitude), as.numeric(latitude)))
  }
```

```{r}
# coordinates of the center of mass
com <- center_of_mass(df)
com <- as.double(com)
com
```

Then we create a dataframe `df_com`
```{r}
df_com <- tibble(longitude = com[1],
                 latitude = com[2])
df_com
```

Finally, plot the infections on the map of Central African Republic,
with the temperature data represented as a raster layer. The temperature
data is displayed as a **heatmap**, with warmer colors indicating higher
temperatures. The infections are shown as points on the map, with
different colors representing infected and non-infected locations.

Note, here a new package is used, the `{ggnewscale}` package, which is a package that allows you to add multiple color scales to a single plot. This is useful when you want to visualize different variables with different color scales in the same plot.

```{r}
#| layout-ncol: 2
#| label: fig-synthetic-data 
#| fig-cap:
#|  - "Location of Infections"
#|  - "Infections Size on Max Temperature"
#| fig-alt: "Synthetic Data Visualization"
# Location of Infections   
ggplot() +
  geom_point(data = df, 
             aes(x = longitude, y = latitude, color = type)) +
  scale_color_manual(values = c("infected" = "#a60845",
                                "non_infected" = "grey")) +
  geom_point(data = df_com,
             aes(x = longitude, y = latitude), 
             color = "red", size = 4) +
  labs(title = "Location of Infections",
       subtitle = "Synthetic Data",
       color = "Infection Status") 
# Infections Size on Max Temperature  
ggplot() +
  geom_sf(data = ctr_africa,
          fill = alpha("pink",alpha=0.3)) +
  geom_raster(data = ctr_africa_raster_df,
              aes(x = x, y = y, fill = max)) +
  scale_fill_gradient(low = "white", high = "black",
              na.value = "transparent") +
  # set a new scale color
  ggnewscale::new_scale_color() +
  geom_point(data = df,
             mapping = aes(x = longitude,
                           y = latitude,
                           group = type,
                           color = factor(type),
                           size = number),
             alpha = 0.8, show.legend = F) +
  scale_color_manual(values = c("infected" = "#a60845",
                                "non_infected" = "grey")) +
  geom_point(data = df_com,
             aes(x = longitude, y = latitude), 
             color = "red", size = 4) +
  labs(title = "Infections Size on Max Temperature",
       subtitle = "Synthetic Data",
       fill = "Temperature") +
  theme(plot.title = element_text(face = "bold"))
```

:::: {.content-visible when-format="html"}
::: solutionbox
**Interactive Maps**

The infections can be visualized with {mapview} or {leaflet}, type of interactive maps, here is a code snippet of how to do it. 

**Map with Mapview:**

This is the interactive map with the max temperatures for the simulated data in Central African Republic. The data represented as a raster points, with warmer colors indicating higher temperatures are shown as points on the map.

```{r}
#| label: fig-mapview-heatmap
#| fig-cap: "Interactive Temperature Raster Points"
#| fig-alt: "Interactive Temperature"  
library(mapview)

mapview(df_sf,
        zcol = "temperature",
        map.types = "CartoDB.Voyager")
```

**Map with Leaflet:**

The `{leaflet}` and `{leaflet.extras}` packages are widely explained in the package documentation. Here is shown howe to make a `heatmap` with the infections and the center of mass.

```{r}
#| label: fig-leaflet-heatmap
#| fig-cap: "Interactive Heatmap"  
#| fig-alt: "Interactive Heatmap"
library(leaflet)
library(leaflet.extras)
# define center of map
lat_center <- com[2]
long_center <- com[1]
# creating intensity heat map
heatmap <- df %>%
    leaflet() %>%
    addTiles() %>%
    setView(long_center, lat_center, 4.5) %>%
    addHeatmap(lng = ~longitude, lat = ~latitude,
               intensity = ~number,
               max = 10, radius = 20, blur = 10) %>%
    # parameter to set
    addMarkers(lng = as.numeric(com[1]), 
               lat = as.numeric(com[2]))
heatmap
```

Then, we can save the map as an image with the `mapshot()` function from the `{mapview}` package.
```{r}
#| eval: false
# webshot::install_phantomjs()
mapview::mapshot(heatmap, file = "images/heatmap.png")
```

:::
::::

### The Euclidean distance

In infectious disease modelling is very helpful to understand how
infections spread through populations for making strategic plans for
disease control and prevention. The Euclidean distance serves as a
fundamental tool in spatial epidemiology, it represents the shortest
distance between two points in a plane.

It is calculated as the square root of the sum of the squares of the
differences between corresponding coordinates of the two points.

The Euclidean distance formula:

$$\text{Euclidean Distance} = \sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2}$$

Where: $(x_1, y_1)$ are the coordinates (e.g., latitude and longitude)
of the first point (A), e.g., infected individual, and $(x_2, y_2)$ are the coordinates of the second point (B), e.g., susceptible individual or
location.

It calculates the straight-line distance between two points in a
two-dimensional space, which can be applied to model the spatial spread
of infectious diseases based on the proximity of individuals or
locations.

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 4
#| fig-height: 2
#| fig-align: center
#| fig-cap: "Euclidean Distance representation"

library(ggdag)
dagify(B ~ A,
       exposure = "B",
       outcome = "A",
       coords = time_ordered_coords()) %>%
  ggdag() + 
  theme_dag_blank()
```

To calculate the Euclidean distance we make a function that considers
the distance between the center of mass and the infection coordinates.

```{r}
# helper function for calculating euclidean distance metric
euclidean_distance <- function(longitude, latitude, long_com, lat_com) {
  long_distances <- (longitude - long_com[1])^2
  lat_distances <- (latitude - lat_com[1])^2
  return((long_distances + lat_distances) * 0.5)
}
```

The Euclidean distance\index{euclidean distance} is used to estimate the likelihood of
transmission between individuals based on their spatial proximity.
Individuals within a certain distance of an infected individual may have
a higher risk of contracting the infection compared to those farther
away.

### Cost and Loss functions

The simulation of the spread of infectious diseases requires the
calculation of a **loss function**\index{loss function}, which serves as a crucial tool for
quantifying the negative consequences associated with the spread of the
disease. Specifically, the loss function helps delineate the
**threshold**\index{threshold} beyond which adverse outcomes occur, and essentially
assigns a value to different distances, reflecting the **severity** of
the consequences when individuals are within certain proximity to each
other. For instance, when the Euclidean distance exceeds a predefined
threshold, denoting that susceptible individuals are dangerously close
to infected ones, the loss function assigns a higher loss value.
Conversely, when individuals are at a safe distance from each other, the
loss value may be minimal or even zero.

By incorporating the Euclidean distance into the loss function, we can
effectively model the spatial dynamics of disease transmission and
assess the associated risks. This approach enables us to identify
critical zones where the likelihood of disease spread is elevated,
guiding public health interventions and mitigating potential outbreaks.

For example, if the threshold is set to a certain distance (e.g., 5
meters), and the Euclidean distance between an infected individual and a
susceptible individual exceeds this threshold (e.g., 6 meters), it
indicates that the susceptible individual is too close to the infected
individual and is at risk of contracting the disease. This situation
would **trigger a loss** according to the loss function.

The threshold value is crucial as it determines the sensitivity of the
loss function to the distance between infected and susceptible
individuals. A lower threshold implies that even relatively small
distances can lead to losses, while a higher threshold indicates that
losses occur only when individuals are in closer proximity to each
other. The choice of threshold depends on various factors such as the
**infectiousness of the disease**, the **mode of transmission**, and the
desired **level of risk tolerance**. Adjusting the threshold allows for
fine-tuning the model to better capture the dynamics of disease spread
and its impact on the population, and this is something that is related
to model calibration in machine learning.

A simple example would be considering:

-   Euclidean distance between infected and susceptible individuals
-   Threshold distance beyond which loss occurs

```{r}
# Distance in meters
distance <- 5
threshold <- 3
```

#### Define the loss function

The loss function calculates the loss incurred when the Euclidean distance between infected and susceptible individuals exceeds the threshold. The loss value increases linearly with the distance beyond the threshold, reflecting the escalating risk of disease transmission as individuals come into closer proximity to each other.
```{r}
loss_function <- function(distance, threshold) {
  if (distance <= threshold) {
    # No loss if distance is within threshold
    return(0)
  } else {
    # Loss increases linearly with distance beyond threshold
    return(distance - threshold)
  }
}
```

```{r}
loss <- loss_function(distance, threshold)
loss
```

#### Define the cost function

The **cost function**\index{cost function} calculates the total cost incurred due to the spread of infections, considering the Euclidean distance between infected and susceptible individuals and the number of infected individuals at each location. The cost value is derived from the product of the Euclidean distance and the number of infected individuals, reflecting the **cumulative impact of disease transmission across different locations**. The cost function provides a comprehensive assessment of the economic and social consequences associated with the spread of infections, guiding decision-making on public health interventions and resource allocation.
```{r}
# cost calculation function
cost_function <- function(df) {
  long_com <- df$longitude[nrow(df)]
  lat_com <- df$latitude[nrow(df)]
  euclideans <- euclidean_distance(df$latitude[-nrow(df)],
                                   df$latitude[-nrow(df)],
                                   long_com, lat_com)
  # implementation of the cost function
  costs <- euclideans * df$number[-nrow(df)]
  return(sum(costs))
}
```

Here we calculate the cost of the spread of infections in Central Africa, considering the Euclidean distance between infected and susceptible individuals and the number of infected individuals at each location. The cost value is expressed in millions, reflecting the economic impact of disease transmission in the region.
```{r}
cost_function(df) / 1000000
```

## Applying Monte Carlo simulation 

In the context of infectious diseases, to assess the cost and loss associated
with the spread of infections, data on transmission rates, infection severity, and other factors are subject to volatility, seasonality, and uncertainty. Monte Carlo
simulation, a widely-used technique in epidemiological analysis, is an
appropriate method for assessing the risks associated with the spread of
infections. In this case, there is a risk involved in selecting
strategies to mitigate the spread of infections, such as implementing
public health measures or vaccination campaigns.

-   How sensitive are the costs and losses incurred if we choose a
    specific intervention strategy?
-   How much higher could the costs and losses be, compared to the
    baseline scenario?

We can apply Monte Carlo simulation by repeating the experiment,
modelling the distribution of transmission rates and infection outcomes. We will repeat the scenario of infection spread (i.e., the distribution
of transmission rates and infection outcomes) 1,000 times, and for each run, calculate the resulting costs and losses associated with the spread of infections.

```{r}
# dataframe for storing Monte-carlo simulation results
results <- as.data.frame(matrix(nrow = 10000, ncol = 2))
colnames(results) <- c("iteration", "costs")
```

Execute 10,000 iterations of the Monte-carlo simulation to assess cost
uncertainty / cost sensitivity.

```{r}
for (i in 1:10000) {
  df$number[-101] <- rpois(n = num_points * 0.7, lambda = 10)
  results$iteration[i] <- i
  results$costs[i] <- cost_function(df) / 1000000
  }
```

This is the `results` dataframe with the costs of the Monte Carlo simulation.

```{r}
results %>% dim()
results %>% head()
```

Plot results of **Montecarlo simulation** in the form of a histogram:

```{r}
#| label: fig-monte-carlo
#| fig-cap: "Simulated infection sensitivity"
#| fig-alt: "Simulated infection sensitivity"
ggplot() +
  geom_histogram(data = results,
                 mapping = aes(x = costs),
                 fill = alpha("#a60845", alpha = 0.5),
                 bins = 25) +
  labs(title = "Simulated infection sensitivity",
       subtitle = "at center of gravity (center of mass)") +
  xlab("Simulated loss") +
  ylab("Absolute frequency") +
  theme_minimal()
```
