# Advanced Data Visualisation Techniques {#sec-12-examples}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title = element_text(face="bold", size = 11))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   Learn advanced data visualisation techniques
-   Create contour plots to visualise interaction effects
-   Create a pyramid plot to visualise population distribution
:::
::::::

\
In this chapter, we will learn how to create advanced data visualisation
techniques using `ggplot2`. We will create contour plots to visualise
interaction effects and pyramid plots to visualise population
distribution.

## Example: Detecting Interaction Effects with Contour Plots

Interaction effects can cause predictors to act together on the response
variable, leading to additional variation in the response. Interaction
effects occur when the combined effect of two or more predictors is
different from what would be expected if the impact of each predictor
were considered alone. For instance, in the case of **cardiovascular
disease (CVD)** risk prediction, the interaction between age and
cholesterol levels can have a significant impact on the risk of
developing heart disease. In addition, if smoke is added to the model,
the effect of cholesterol levels on CVD risk may change depending on
whether the individual is a smoker or non-smoker. So, smoke is said to
interact with cholesterol levels in predicting CVD risk, as well as age,
but in what way?

$$y=\beta_0 +\beta_1 x_1 + \beta_2 x_2 +\beta_3 x_3 + \epsilon$$ {#eq-interaction}

where $y$ is the risk of developing heart disease, $\beta_0$ is the
intercept, $\beta_1$ and $\beta_2$ are the coefficients for the
predictors $x_1$ and $x_2$, respectively, and $\beta_3$ is the
coefficient for the **interaction term** $x_1*x_2$. The error term
$\epsilon$ accounts for the variability in the response variable that is
not explained by the predictors.

There are four type of interactions:

1.  additive is when $\beta_3 \approx{0}$
2.  antagonistic is when $\beta_3 < 0$
3.  synergistic is when $\beta_3 > 0$
4.  atypical is when $\beta_3 \neq 0$

In order to simulate the interaction effects for each type of
interaction, we cannot just simulate random values for $y$, we need to
simulate different levels of interactions between predictors setting up
the values for the coefficients $\beta_0$, $\beta_1$, $\beta_2$, and
$\beta_3$, and the predictors $x_1$ and $x_2$ and the error term
$\epsilon$.

Load the necessary libraries and set the seed for reproducibility.

```{r}
library(tidyverse)
set.seed(123)
```

Simulate the data ready for different interaction effects.

```{r}
beta0<- rep(0,200)
beta1<- rep(1,200)
beta2<- rep(1,200)
```

Then, simulate the predictors $x_1$ and $x_2$ and the error term
$\epsilon$ for 200 observations.

```{r}
x1<- runif(200, min = 0, max = 1)
x2 <- runif(200, min = 0, max = 1)
e <- rnorm(200)
```

Then for each of the three cases we simulate a $\beta_3$ based on
antagonism, no interaction, or synergism with values -10, 0, and 10
respectively.

**Case 1: Antagonism**

```{r}
beta3 <- rep(-10,200) 
```

Assemble the values for the coefficients and predictors into the
equation, to obtaining the response variable $y$. This will be our
synthetic "observed" data.

```{r}
y1 = beta0 + beta1*x1 + beta2*x2 + beta3*(x1*x2) + e
```

Apply a linear model to the observed data to predict the response
variable $y$ based on interaction between $x_1$ and $x_2$.

```{r}
observed1 <- tibble(y1, x1, x2)

mod1 <- lm(y1 ~ x1*x2, data = observed1)

observed1$z <- predict(mod1, observed1)
```

Then create a grid of values for $x_1$ and $x_2$ to predict the response
variable $y$ based on the interaction between $x_1$ and $x_2$.

```{r}
grid <- with(observed1, 
             interp::interp(x = x1, y = x2, z))
griddf <- subset(data.frame(x = rep(grid$x, nrow(grid$z)),
                            y = rep(grid$y, 
                                    each = ncol(grid$z)),
                            z = as.numeric(grid$z)),!is.na(z))
```

```{r}
p1 <- ggplot(griddf, aes(x, y, z = z)) +
  geom_contour(aes(colour = after_stat(level),
                   linetype = factor(after_stat(level))),
                   linewidth = 2) +
  scale_color_viridis_c()+
  guides(linetype = "none")+
  labs(title="Antagonism",
       color="Prediction", x = "x1", y = "x2")+
  theme(legend.position = "top")
```

**Case 2: Additive (no interaction)**

```{r}
beta3 <- rep(0,200) 
```

```{r}
y2 = beta0 + beta1*x1 + beta2*x2 + beta3*(x1*x2) + e

observed2 <- tibble(y2, x1, x2)

mod2 <- lm(y2 ~ x1*x2, data = observed2)
observed2$z <- predict(mod2, observed2)

grid <- with(observed2, interp::interp(x = x1, y = x2, z))
griddf <- subset(data.frame(x = rep(grid$x, nrow(grid$z)),
                            y = rep(grid$y, 
                                    each = ncol(grid$z)),
                            z = as.numeric(grid$z)),!is.na(z))
```

```{r}
p2 <- ggplot(griddf, aes(x, y, z = z)) +
  geom_contour(aes(colour = after_stat(level),
                   linetype = factor(after_stat(level))),
                   linewidth = 2) +
  scale_color_viridis_c()+
  guides(linetype = "none")+
  labs(title="Antagonism",
       color="Prediction", x = "x1", y = "x2")+
  theme(legend.position = "top")
```

**Case 3: Synergism**

```{r}
beta3<- rep(10,200) 
```

```{r}
y3 = beta0 + beta1*x1 + beta2*x2 + beta3*(x1*x2) + e

observed3 <- tibble(y3, x1, x2)

mod3 <- lm(y3 ~ x1 * x2 , data = observed3)  
observed3$z <- predict(mod3, observed3)

grid <- with(observed3, interp::interp(x=x1,y=x2,z))
griddf <- subset(data.frame(x = rep(grid$x, nrow(grid$z)),
                            y = rep(grid$y, each = ncol(grid$z)),
                            z = as.numeric(grid$z)),!is.na(z))
```

```{r}
p3 <- ggplot(griddf, aes(x, y, z = z)) +
  geom_contour(aes(colour = after_stat(level),
                   linetype = factor(after_stat(level))),
                   linewidth = 2) +
  scale_color_viridis_c()+
  guides(linetype = "none")+
  labs(title="Antagonism",
       color="Prediction", x = "x1", y = "x2")+
  theme(legend.position = "top")
```

```{r}
#| eval: false
library(patchwork)
p1|p2|p3
```

```{r}
#| layout-ncol: 3
#| label: fig-scatter
#| fig-cap: "Scatterplots"
#| fig-subcap: 
#|  - "Scatterplot for Antagonistic"
#|  - "Scatterplot for Additive"
#|  - "Scatterplot for Synergistic"
#| fig-alt: "Scatterplots for Antagonistic, Additive, and Synergistic interaction effects"
#| echo: false
observed1 %>%
   ggplot(aes(x1, y1, color = factor(x2))) +
   geom_point(show.legend = FALSE) +
   labs(title = "Antagonist")
observed2 %>%
   ggplot(aes(x1, y2, color = factor(x2))) +
   geom_point(show.legend = FALSE) +
   labs(title = "Additive")
observed3 %>%
   ggplot(aes(x1, y3, color = factor(x2))) +
   geom_point(show.legend = FALSE) +
   labs(title = "Synergist")
```

```{r}
#| layout-ncol: 3
#| label: fig-contour
#| fig-cap: "Contour Plots"
#| fig-subcap: 
#|  - "Contour plot for Antagonistic"
#|  - "Contour plot for Additive"
#|  - "Contour plot for Synergistic"
#| fig-alt: "Contour plots for Antagonistic, Additive, and Synergistic interaction effects"
#| echo: false
p1
p2
p3
```

## Example: Pyramid Plot

A pyramid plot is a type of bar chart that displays the distribution of
a population by age. We use the data from the `{wpp2022}` package,
`data(package = "wpp2022")` which is a collection of population
datasets, with population estimates. The data is from the United Nations
World Population Prospects 2022.

```{r}
library(wpp2022)
```

The data contains the following columns:

```{r}
# load the data on the population
data(popAge1dt)
popAge1dt %>% head()
```

Check all Countries in the dataset with `View()`:

```{r}
#| eval: false
popAge1dt %>% count(name) %>% View
```

The data is then aggregated to show the total population for all ages by
year. Here, in particular, we show the use of
`geomtextpath::geom_textline()` to add a label to the World line plot.
`{geomtextpath}` is a nice R-package which allows you to add text along
a path.

```{r}
#| label: fig-wld-pop
#| fig-cap: "Population for all ages from 1949 to 2021"
#| fig-alt: "Population for all ages from 1949 to 2021"
all_ages <- popAge1dt %>%
  group_by(year, name) %>%
  reframe(tot_pop=sum(pop)) 

all_ages %>% 
  filter(!name == "World")%>%
  ggplot(aes(x = year, y = tot_pop, group = name)) +
  geom_line(color= "grey", linewidth =0.2) +
  geomtextpath::geom_textline(data = all_ages %>% filter(name == "World"), 
                              aes(label=name),
                              color = "red", linewidth = 1) +
  labs(title = "United Nations World Population Prospects 2022 - [1949 - 2021]",
       x = "Year", y = "Population in thousands", 
       caption = "Data Source: UN World Pop | Graphic: @fgazzelloni") 
```

Data is then transformed to a long format to create a pyramid plot. We
select the columns `name`, `year`, `age`, `popF`, and `popM` and pivot
the data to a long format using the `pivot_longer()` function. We then
create a new column `value` that contains positive values for male
population and negative values for females.

```{r}
data <- popAge1dt %>%
   select(name, year, age, popF, popM) %>%
   pivot_longer(cols = c(popM, popF), 
                names_to = "sex", 
                values_to = "population") %>%
   mutate(value = ifelse(sex == "popF",
                         as.integer(population * -1),
                         as.integer(population)))
  
data %>% head()  
```

Then, we can create pyramid plots that show the distribution of the
population by age for **High-income countries**, **Lower-middle-income
countries**, and **Low-income countries**, and put them in one chunk
with the option `layout-ncol: 3`.

```{r}
#| layout-ncol: 3
#| label: fig-pyramids
#| fig-cap: "Pyramid Plots"
#| fig-subcap:
#|  - "High-income countries"
#|  - "Lower-middle-income countries"
#|  - "Low-income countries"
#| fig-alt: "Pyramid plots for High-income, Lower-middle-income, and Low-income countries"  

data %>%
  filter(name == "High-income countries") %>% 
  ggplot(aes(x = age, y = value, fill = sex)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#CC6666", "#9999CC")) +
  coord_flip() +
  labs(title = "United Nations World Population Prospects 2022", 
       subtitle = "High-income countries",
       x= "Age", y = "Population in thousands", fill = "",
       caption = "Data Source: UN World Pop | Graphic: @fgazzelloni") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", 
                                  hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text = element_text(size = 10))

data %>%
  filter(name == "Lower-middle-income countries") %>% 
  ggplot(aes(x = age, y = value, fill = sex)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#CC6666", "#9999CC")) +
  coord_flip() +
  labs(title = "United Nations World Population Prospects 2022", 
       subtitle = "Lower-middle-income countries",
       x= "Age", y = "Population in thousands", fill = "",
       caption = "Data Source: UN World Pop | Graphic: @fgazzelloni") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", 
                                  hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text = element_text(size = 10))

data %>%
  filter(name == "Low-income countries") %>% 
  ggplot(aes(x = age, y = value, fill = sex)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#CC6666", "#9999CC")) +
  coord_flip() +
  labs(title = "United Nations World Population Prospects 2022", 
       subtitle = "Low-income countries",
       x= "Age", y = "Population in thousands", fill = "",
       caption = "Data Source: UN World Pop | Graphic: @fgazzelloni") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", 
                                  hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text = element_text(size = 10))
```

```{r}
pyramid <- 
  data %>%
  filter(name == "World") %>% 
  ggplot(aes(x = age, y = value, fill = sex)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#CC6666", "#9999CC")) +
  coord_flip() +
  labs(title = "United Nations World Population Prospects 2022 - {closest_state}", 
       x= "Age", y = "Population in thousands", fill = "",
       caption = "Data Source: UN World Pop | Graphic: @fgazzelloni") +
       theme_minimal() +
       theme(plot.title = element_text(face = "bold", 
                                           hjust = 0.5),
            axis.text = element_text(size = 10))
```

```{r}
#| eval: false
library(gganimate)
pyramid_gif <- pyramid +
  transition_states(year, 
                    transition_length = 1, 
                    state_length = 2) +
  enter_fade() +
  exit_fade() +
  ease_aes("cubic-in-out")

animate(pyramid_gif,
        fps = 72, duration = 6, 
        width = 1200, height = 1400, res = 180,
        renderer = gifski_renderer("images/12_unp_pyramid.gif"))
```

::: {.content-visible when-format="html"}
![United Nations World Population Prospects 2022 \[1949 -
2021\]](images/12_unp_pyramid.gif){#fig-pyramid-animate
fig-align="center" width="375"
fig-alt="Pyramid plot for the United Nations"}
:::
