# COVID-19 Outbreaks {#sec-14-covid19}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title = element_text(face = "bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   What is Covid-19
-   How does it spread
-   Map Covid-19 outbreaks
:::
::::::

\
In this chapter, we explore the dynamics of **COVID-19** disease
outbreaks in more detail. We examine the results of various model's
applications that simulate the spread of the virus.

## Epidemiology

COVID-19, short for **"Coronavirus Disease 2019,"** is an infectious
disease caused by the novel coronavirus SARS-CoV-2, a type of virus that
is part of the **Coronaviridae**[@coronavi] family of viruses that can
cause illness in animals and humans alike. The disease was first
identified in December 2019 in Wuhan, Hubei province, China, and has
since then spread globally, leading to a **pandemic**[@pandemic].

The origin of the virus is still under investigation[@who-conv], but it
is believed to have **zoonotic**[@zoonosis]\index{zoonotic} origins,
with **bats** being the most likely reservoir. The virus has also been
linked to **pangolins**[@huang2023], mammals that are illegally
trafficked for their scales and meat. The virus is thought to have
jumped from animals to humans, possibly through a wet market in Wuhan
where live animals were sold.

::: {.callout .callout-info}
Examples of Zoonotic Viruses:

1.  Influenza (Flu) Viruses: Avian (bird) flu and swine (pig) flu.
2.  Coronavirus: SARS, MERS, and COVID-19 from fruit bats, pangolins.
3.  Ebola Virus: Likely originated from fruit bats.
4.  HIV/AIDS: Believed to have originated from non-human primates
    (chimpanzees).
5.  Rabies: Spread from infected animals like dogs, bats, and raccoons.
:::

On January 30, 2020, The **World Health Organization (WHO)** declared
the **outbreak** a Public Health Emergency of International Concern and
a worldwide investigation started to understand more about the virus and
its effects on humans. This challenge involved the unprecedented
identification of this type of virus, requiring thorough testing and
confirmation of the incubation period, recovery rate, and infection rate
through the collection of cases globally.

The virus spreads primarily through respiratory droplets. Symptoms can
include loss of taste and smell, fatigue, muscle aches, and
gastrointestinal issues, with fever, cough, and shortness of breath.
Severe cases can lead to **pneumonia**, **acute respiratory distress
syndrome (ARDS)**, and **death**. The emergency was primarily due to the
high number of deaths occurring without a known pharmacological
intervention to contain it.

Various preventive measures were established, but the presence of
asymptomatic and symptomatic cases created challenges in identifying the
incubation time, leading governments to impose restrictions through
lockdowns of entire cities and regions in affected countries. These
measures restricted movement and activities to curb the virus’s
transmission.

::: {.callout .callout-info}
Some notable examples of lockdown measures include:

-   **China**: The initial epicentre in Wuhan experienced strict
    lockdowns, with entire cities quarantined and movement heavily
    restricted. These measures were effective in curbing the initial
    spread of the virus.
-   **Italy**: One of the first European countries severely impacted,
    Italy imposed nationwide lockdowns in March 2020, closing schools,
    businesses, and restricting travel.
-   **United States**: Lockdown measures varied by state, with
    significant restrictions during the early waves of the pandemic.
    These included stay-at-home orders, business closures, and mask
    mandates.
-   **India**: Implemented one of the world's largest lockdowns in March
    2020, affecting millions of people and including strict travel bans
    and business closures.
-   **Australia**: Used localised lockdowns effectively, particularly in
    cities like Melbourne, which saw extended periods of strict
    restrictions.
-   **United Kingdom**: Imposed several lockdowns, including a strict
    nationwide lockdown in early 2021, along with tiered restrictions
    based on regional infection rates.
:::

The pandemic has led to widespread healthcare challenges, economic
disruptions, and significant changes to daily life, including social
distancing measures, lockdowns, and travel restrictions.

Since its identification, COVID-19 has caused multiple outbreaks
globally. The incubation period ranges from 2 to 14 days, during which
an infected person may be asymptomatic but still contagious. On March
11, 2020, it was declared a pandemic. As of May 2024, there have been
over 775 million confirmed cases and more than seven million deaths
worldwide. The impact of the pandemic has varied across regions,
influenced by factors such as virus variants, public health responses,
and vaccination rates.

::: {.callout .callout-warning}
Public Health COVID-19 Prevention Measures:

-   **Diagnostic Tests**: PCR (polymerase chain reaction) tests and
    rapid antigen tests are used to detect active infections, while
    antibody tests determine past infections.
-   **Treatment**: While there is no specific antiviral treatment for
    COVID-19, supportive care, including oxygen therapy and mechanical
    ventilation, is used for severe cases. Some antiviral medications
    and therapies have been repurposed or developed specifically for
    COVID-19.
-   **Vaccination**: Vaccines have been developed and distributed
    globally to prevent COVID-19. These vaccines, including mRNA
    vaccines (Pfizer-BioNTech and Moderna), viral vector vaccines
    (AstraZeneca and Johnson & Johnson), and inactivated virus vaccines
    (Sinovac and Sinopharm), have shown high efficacy in preventing
    severe disease and reducing transmission.
:::

COVID-19 has had a profound impact on global health, economies, and
daily life. Efforts to combat the pandemic have involved unprecedented
levels of international cooperation, scientific research, and public
health initiatives. Ongoing vaccination campaigns and adherence to
public health guidelines are critical in controlling the spread of the
virus and preventing further outbreaks. However, vaccine hesitancy and
resistance, driven by concerns about the speed of development and
potential long-term effects, remain significant challenges in achieving
widespread vaccination coverage.

## Mapping COVID-19 Outbreaks

Mapping COVID-19 outbreaks helps identify infection hotspots, track the
virus’s spread, and inform public health interventions. Geographic
Information Systems (GIS) and spatial analysis techniques can visualise
and analyse COVID-19 data, such as the number of cases, deaths, and
recoveries, at local, regional, and global levels.

## Example: Modelling the Spread of COVID-19 {#sec-14-modelling-covid19}

Spatiotemporal modelling of COVID-19 outbreaks can provide valuable
insights into the pandemic's dynamics and help guide public health
responses. In this example, we'll use the
**Susceptible-Exposed-Infected-Recovered (SEIR)** model to simulate the
spread of COVID-19 infections, as we saw in @sec-05-epidemic-x. A
**Bayesian machine learning approach** is used to predict the spread of
the virus over time and across different regions. Bayesian methods
handle uncertainties and allow for the integration of prior information
into the model.

### SEIR model

First, we use the **SEIR model** function to simulate the spread of the
disease. The model compartments are: the number of susceptible (S), the
exposed (E), the infected (I), and R represents the number of recovered
individuals. The parameters of the model are the transmission rate
$(\beta)$ and the recovery rate $(\gamma)$, $(t_{exp})$ the latent
period, and $(N)$ is the total population.

```{r}
library(deSolve)
library(tidyverse)
# Define parameters
N <- 1e6 # Total population
beta <- 0.5 # Transmission rate
gamma <- 0.1 # Recovery rate
t_exp <- 5 # Latent period - or incubation period

# Initial conditions
init <- c(S = N - 1, E = 1, I = 0, R = 0)

# Define the SEIR model
seir_model <- function(t, y, parameters) {
  with(as.list(y), {
    dS <- -beta * S * I / N
    dE <- beta * S * I / N - (1 / t_exp) * E
    dI <- (1 / t_exp) * E - gamma * I
    dR <- gamma * I
    return(list(c(dS, dE, dI, dR)))
  })
}
```

As seen previously, we'll use the `ode()` function from the `{deSolve}`
package to solve the ordinary differential equations (ODEs) and simulate
the spread of the virus over a period of 365 days:

```{r}
times <- seq(0, 365, by = 1)
result <- ode(y = init, times = times, func = seir_model)
result %>% head()
```

```{r}
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "Simulation of the SEIR model for COVID-19"
#| fig-alt: "Simulation of the SEIR model for COVID-19"
#| label: fig-seir-simulation

ggplot(as.data.frame(result), aes(time, I, color = "Infected")) +
  geom_line() +
  geom_line(aes(time, R, color = "Recovered")) +
  geom_line(aes(time, S, color = "Susceptible")) +
  geom_line(aes(time, E, color = "Exposed")) +
  scale_color_discrete(name = "Compartment") +
  labs(x = "Time (days)", y = "Number of individuals")
```

In reality, the spread of a virus is much more complex and influenced by
many factors such as human behaviour, government policies, healthcare
systems, vaccination campaigns, and human contact.

### Bayesian Analysis

The use of **Bayesian analysis**\index{Bayesian analysis} in modelling
infectious diseases offers several advantages in understanding and
predicting the dynamics of the spread. We implement a Bayesian
regression model for COVID-19 cases using the `{brms}` package, which is
an interface of **Stan** for Bayesian analysis. The source for **Stan**
is available at <https://mc-stan.org/>.

The application of Bayesian analysis allows us to estimate the
regression coefficients for the lag-1 and lag-7 cases, which can help
predict the number of infected individuals over time, with the
flexibility to incorporate prior knowledge and uncertainties into the
model.

The model is specified as follows:

$$I_t \sim \text{Normal}(\beta \times I_{t-1} + \gamma \times I_{t-7}, \sigma)$$

where $I_t$ is the number of infected individuals at time $t$, $\beta$
is the regression coefficient for the lag-1 cases, $\gamma$ is the
regression coefficient for the lag-7 cases, and $\sigma$ is the error
term. The model assumes that the number of infected individuals at time
$t$ is normally distributed around the predicted value based on the
number of cases at times $t-1$ and $t-7$.

We can specify the prior distributions for $\beta$ and $\gamma$ using
expert knowledge and available data. For example, we might specify a
gamma distribution with a mean of 0.1 and a standard deviation of 0.05
for $\beta$, and a gamma distribution with a mean of 0.05 and a standard
deviation of 0.02 for \$\gamma.

```{r}
result <- as.data.frame(result)

cv19_data <- result %>%
  mutate(date = as.Date(time, origin = "2022-01-01"),
         day_of_week = wday(date, label = TRUE),
         week_of_year = week(date),
         month = month(date),
         lag_1_cases = lag(I, 1),
         lag_7_cases = lag(I, 7)) %>%
  drop_na()

cv19_data %>% head(n = 3) %>% glimpse()
```

In this example we do not split the data into training and testing sets,
but in a real-world scenario, it is essential to validate the model's
performance on unseen data. Our data is made of 365 observations and 7
variables.

```{r}
cv19_data %>% dim()
```

The `brm()` function from the `{brms}` package, fits the Bayesian
regression model to the data by specifying the outcome variable `I`
(number of infected individuals) and the predictors `day_of_week`,
`week_of_year`, `month`, `lag_1_cases`, and `lag_7_cases`. We use a
Gaussian likelihood function and specify prior distributions for the
regression coefficients and the error term. The **Stan algorithm** is
used to sample from the posterior distribution of the model parameters.
In particular, we use a **normal prior distribution** with a mean of 0
and standard deviation of 10 for the regression coefficients and a
**Cauchy prior distribution** with a scale of 5 for the error term.

The decision to use a normal and a Cauchy prior is based on the
properties of the data and the model assumptions. In this case, we
assume that the regression coefficients are normally distributed around
0 with a standard deviation of 10, and the error term is distributed
according to a Cauchy distribution with a scale of 5.

The model is run for 2000 iterations with a `warmup` of 1000 iterations
and 4 chains. The reason for the warmup is to allow the **Markov Chain
Monte Carlo (MCMC) algorithm** to converge to the posterior
distribution. The chains are run in parallel to explore the parameter
space and estimate the posterior distribution of the model parameters.

```{r}
library(brms)
```

```{r}
#| eval: false
# Define the Bayesian regression model
bayesian_model <- brm(
  I ~ day_of_week + week_of_year + month + lag_1_cases + lag_7_cases,
  data = cv19_data,
  family = gaussian(),
  prior = c(set_prior("normal(0, 10)", class = "b"),
            set_prior("cauchy(0, 5)", class = "sigma")),
  iter = 2000,
  warmup = 1000,
  chains = 4,
  seed = 42
  )
```

```{r}
#| echo: false
bayesian_model <- readRDS("data/model-data/bayesian_model.rds")
```

The `bayesian_model` object contains the fitted model, including the
posterior samples of the model parameters. We can inspect the summary of
the model to assess the convergence of the chains, the posterior
distribution of the parameters, and the model fit statistics.

```{r}
summary(bayesian_model)$fixed %>% 
  rownames_to_column(var = "parameter") %>%
  head(n = 3) %>%
  glimpse()
```

The model output tells us about the convergence of the chains, the
posterior distribution of the parameters, and the model fit statistics.
We can see the mean, median, and 95% credible intervals for the
regression coefficients, as well as the standard deviation of the error
term. The `Rhat` statistic measures the convergence of the chains, with
values close to 1 indicating convergence. The `n_eff` statistic measures
the effective sample size of the chains, with higher values indicating
more reliable estimates. The `looic` statistic is the leave-one-out
cross-validation information criterion, which measures the model's
predictive performance.

Or access the formula used in the model:

```{r}
bayesian_model$formula
```

After fitting the model, we can plot the model diagnostics to assess its
performance with the `plot()` function. The diagnostics include trace
plots of the chains, density plots of the posterior distributions, and
the `Gelman-Rubin statistic`, which measures the convergence of the
chains. The Gelman-Rubin statistic should be close to 1 for the model to
have converged.

```{r}
#| eval: false
# Plot the model diagnostics
plot(bayesian_model)
```

<center>

![Bayesian Model
diagnostics](images/14_bayesian_model_diagnostics.png){#fig-bayes-covid19
fig-alt="Bayesian Model diagnostics"}

</center>

Then, finally we can make predictions using the fitted model and
visualise the actual vs. predicted COVID-19 cases over time. The
`predictions` object is type matrix/array.

```{r}
predictions <- predict(bayesian_model, newdata = cv19_data)
predictions %>% head()
```

Create a new data frame with the actual and predicted values for
COVID-19 cases, and then plot the actual vs. predicted values over time.

```{r}
#| label: fig-bayes-predictions
#| layout-ncol: 2
#| fig-cap: "Actual vs. Predicted COVID-19 Cases"
#| fig-alt: "Actual vs. Predicted COVID-19 Cases"

predictions <- as_tibble(predictions)
cv19_data_pred <- cv19_data %>%
  mutate(predicted_cases = predictions$Estimate)

# Plot actual vs. predicted values
ggplot(data = cv19_data_pred, 
       aes(x = I, y = predicted_cases)) +
  geom_point() +
  geom_abline() +
  labs(title = "Actual vs. Predicted COVID-19 Cases",
       x = "Date", y = "Cases", color = "Legend") 

ggplot(data = cv19_data_pred, aes(x = date)) +
  geom_point(aes(y = I, 
                 color = "Actual")) +
  geom_line(aes(y = predicted_cases,
                color = "Predicted")) +
  labs(title = "Actual vs. Predicted COVID-19 Cases",
       x = "Date", y = "Cases", color = "Legend") 
```

### Ensemble Modelling - Combining Multiple Models

We have been able to simulate the spread of COVID-19 using the SEIR and
predict the spread using the Bayesian regression model. However,
combining the predictions of multiple models can provide more accurate
and robust forecasts. This technique is known as **ensemble modelling**.

Ensemble modelling combines the predictions of multiple models to
improve the overall accuracy and robustness of predictions. This
technique is particularly useful in infectious disease modelling to
account for uncertainties and variability in data, providing more
reliable estimates of disease spread.

This method combines the strengths of different models to produce more
accurate and stable predictions by averaging the results of individual
models or using more sophisticated techniques such as **stacking** or
**boosting**.

For example, ensemble models have been used to predict confirmed
COVID-19 cases and provide short-term forecasts by integrating multiple
model types. Other methods combine sub-epidemic models over various
forecasting periods to enhance prediction accuracy[@chowell2022].

In this example, we use the `{tidymodels}` meta-package and other key
packages such as `{modeltimes}`, and `{stacks}` to combine different
types of models. We will use: Decision Tree, Random Forest, K-Nearest
Neighbors (KNN), and Support Vector Machines (SVM) models to combine and
predict the spread of COVID-19 cases.

```{r}
library(tidymodels)
library(modeltime)
library(stacks)
```

We start by splitting the data into training and testing sets using the
`initial_split()` function from the `{rsample}` package. The training
set will be used to train the models, while the testing set will be used
to evaluate the models' performance.

```{r}
# Split the data into training and testing sets
cv_split <- initial_split(cv19_data, prop = 0.8)
cv_train <- training(cv_split)
cv_test <- testing(cv_split)
```

Create a `vfold_cv()` object for cross-validation with 10 folds. This
object will be used to evaluate the models' performance during training.

```{r}
set.seed(1001)
cv_folds <- vfold_cv(cv_train, v = 10)
```

We also define a recipe to preprocess the data, including log
transformation and dummy encoding of categorical variables. The recipe
is then used to create a workflow that combines the preprocessing steps
with the model specifications.

```{r}
cv_recipe <- recipe(I ~ ., data = cv_train) %>%
  step_scale(all_numeric(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes()) 
```

To check the output of the recipe transformation:
`cv_recipe %>% prep() %>% juice()`

Define the models' specifications with the parameters to tune, and the
engine to use:

```{r}
# Decision Tree
cart <- decision_tree(cost_complexity = tune(), 
                      tree_depth = tune(), 
                      min_n = tune()) %>%
  set_engine("rpart") %>%
  set_mode("regression")

# Random Forest
rand_forest <- rand_forest(mtry = tune(), 
                           min_n = tune()) %>%
  set_engine('ranger') %>%
  set_mode('regression')

# K-Nearest Neighbors (KNN)
knn <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn") %>%
  set_mode("regression")

# Support Vector Machine (SVM)
svm <- svm_rbf(cost = tune(), 
               rbf_sigma = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("regression")
```

Then, create a `workflow_set()` that combines the predictions of
different models:

```{r}
cv_wkf <- workflow_set(
  preproc = list(id = cv_recipe),
  models = list(Decision_tree = cart,
                Random_Forest = rand_forest,
                Knn = knn,
                SVM = svm)
  )

cv_wkf
```

Create a `workflow_map()` to fit the models to the training data. We use
the `{finetune}` package to use the `tune_race_anova` function to tune
the hyperparameters of the models. The `race_ctrl` object specifies the
control parameters for the tuning process, such as the number of
resamples and the grid size. The `parallel_over` argument allows the
tuning process to run in parallel over multiple cores for faster
computation.

```{r}
# Fit the models to the training data
library(finetune)
set.seed(112233)
race_ctrl <-
   control_race(save_pred = TRUE,
                parallel_over = "everything",
                save_workflow = TRUE)

race_results <- cv_wkf %>%
  workflow_map("tune_race_anova",
               seed = 1503,
               resamples = cv_folds,
               grid = 3, 
               control = race_ctrl)

race_results
```

Then we collect the metrics for the models to evaluate their
performance. We filter the results for the root mean squared error
(`rmse`) and arrange them in ascending order to identify the
best-performing model.

```{r}
rmse <- collect_metrics(race_results) %>% 
  filter(.metric == "rmse") %>%
  select(wflow_id, .config, mean) %>%
  arrange(mean)

rmse
```

Plot the root mean squared error (`rmse`) values for the models to
visualise their performance.

```{r}
#| label: fig-rmse
#| fig-cap: "Root Mean Squared Error (RMSE) for Models"
#| fig-alt: "Root Mean Squared Error (RMSE) for Models"
rmse %>%
  ggplot(aes(x = fct_reorder(wflow_id, mean), y = mean)) +
  geom_col(fill = "navy") +
  labs(title = "Root Mean Squared Error (RMSE) for Models",
       x = "Model", y = "RMSE") 
```

The results of the race show that the Random Forest model has the lowest
`rmse` value, indicating better performance compared to the other
models.

Extract the model's estimations:

```{r}
#| label: fig-race-predictions
#| fig-cap: "Actual vs. Predicted COVID-19 Cases"
#| fig-alt: "Actual vs. Predicted COVID"
race_results %>%
  collect_predictions() %>%
  ggplot(aes(x = I, y = .pred, 
             group= model,
             shape= model,
             color = model)) +
  geom_point() +
  geom_abline() 
```

The plot shows the actual vs. predicted values for COVID-19 cases using
the different models. The Random Forest model appears to have the best
fit to the data, closely following the 45-degree line, indicating
accurate predictions.

So, let's see how combined predictions look like:

```{r}
#| label: fig-ensemble-predictions
#| fig-cap: "Ensemble Predictions of COVID-19 Cases"
#| fig-alt: "Ensemble Predictions of COVID"
race_results %>%
  collect_predictions() %>%
  group_by(I) %>%
  summarise(mean_pred = mean(.pred)) %>%
  ggplot(aes(x = I, y = mean_pred)) +
  geom_point() +
  geom_abline()
```

Finally, we can combine the predictions of the models using the
`{stacks}` package to create an ensemble model. The `stack()` function
combines the predictions of the models using a meta-learner, such as a
linear regression model, to produce a final prediction.

```{r}
cv_stack <- stacks() %>% 
  add_candidates(race_results)
cv_stack
```

Blend the predictions of the models:

```{r}
set.seed(2001)
cv_ens <- blend_predictions(cv_stack)
cv_ens
```

```{r}
#| label: fig-ensemble-weights
#| fig-cap: "Ensemble Model Weights"
#| fig-alt: "Ensemble Model Weights"
autoplot(cv_ens, "weights") 
```

```{r}
fit_cv_ens <- fit_members(cv_ens)
fit_cv_ens
```

```{r}
cv_ens_test_pred <- 
  predict(fit_cv_ens, new_data = cv_test) %>% 
  bind_cols(cv_test)
```

```{r}
#| label: fig-ensemble-predictions2
#| fig-cap: "Ensemble Predictions of COVID-19 Cases"
#| fig-alt: "Ensemble Predictions of COVID"
cv_ens_test_pred %>%
  ggplot(aes(x = I, y = .pred)) +
  geom_point(alpha=0.5) +
  geom_abline()
```

In summary, integrating machine learning models and advanced statistical
techniques such as Bayesian analysis and ensemble modelling can
significantly enhance our ability to predict and understand the spread
of infectious diseases like COVID-19. These models provide critical
insights that can inform public health responses and help mitigate the
impact of pandemics.

## COVID-19 & DALYs

The influence of COVID-19 on global health can be measured using the
DALYs. We use the data from the JHU official GitHub repository
**CSSEGISandData**[^14-covid19-1] to calculate the DALYs for COVID-19.
The data includes the number of confirmed cases, deaths, and recovered
cases for each country and region. We will calculate the DALYs for the
US, China, United Kingdom, and Canada.

[^14-covid19-1]: cases \<-
    read.csv("[https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")](https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv%22))
    deaths \<-
    read.csv("[https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")](https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv%22))

```{r}
#| echo: false
library(tidyverse)
cases <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
deaths <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")
```

```{r}
cases1 <- cases %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "cases") %>%
  mutate(date = gsub("X", "", date),
         date = as.Date(date, format = "%m.%d.%y")) %>%
  janitor::clean_names()

cases1 %>% head()
```

```{r}
deaths1 <- deaths %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "deaths") %>%
  mutate(date = gsub("X", "", date),
         date = as.Date(date, format = "%m.%d.%y")) %>%
  janitor::clean_names()

deaths1 %>% head()
```

```{r}
COVID19 <- cases1 %>%
  left_join(deaths1, 
            by = c("province_state", 
                   "country_region", 
                   "lat", "long", "date")) %>%
  mutate(cases = ifelse(is.na(cases), 0, cases),
         deaths = ifelse(is.na(deaths), 0, deaths)) %>%
  group_by(country_region, date) %>%
  mutate(cases = sum(cases),
         deaths = sum(deaths),
         cfr = round(deaths / cases, 3)) %>%
  filter(cases > 0) %>%
  arrange(country_region, date)

COVID19 %>%
  arrange(desc(date)) %>%
  head()
```

```{r}
#| echo: false
#| eval: false
# saveRDS(COVID-19, "COVID-19.rds")
```

```{r}
COVID19_4countries <- COVID19 %>%
  filter(country_region %in%
    c("US", "China", "United Kingdom", "Canada"))

COVID19_4countries %>% head()
```

Selected countries are the US, China, United Kingdom, and Canada.

```{r}
#| layout-ncol: 2
#| fig-cap: "COVID-19 Cases and Deaths by Country"
#| fig-alt: "COVID-19 Cases and Deaths by Country"
#| fig-subcap: 
#|  - "COVID-19 Cases by Country"
#|  - "COVID-19 Deaths by Country"
#| label: fig-cases-deaths
COVID19 %>%
  ggplot(aes(x = date, y = cases,
  group = country_region)) +
  geom_line(color = "grey", 
            linewidth = 0.2) +
  geomtextpath::geom_textline(
  data = COVID19_4countries,
  aes(label = country_region),
      show.legend = F) +
  scale_y_log10() +
  labs(title = "COVID-19 Cases by Country",
       x = "Date",
       y = "Cases") 

COVID19 %>%
  ggplot(aes(x = date, y = deaths,
             group = country_region)) +
  geom_line(color = "grey", 
            linewidth = 0.2) +
  geomtextpath::geom_textline(
    data = COVID19_4countries,
    aes(label = country_region),
        show.legend = F) +
  scale_y_log10() +
  labs(title = "COVID-19 Deaths by Country",
       x = "Date",
       y = "Deaths") 
```

```{r}
worldmap <- map_data("world") %>%
  filter(!region == "Antarctica")

centroids <- COVID19 %>%
  group_by(country_region, lat, long) %>%
  reframe(tot_cases = sum(cases),
          tot_deaths = sum(deaths),
          avg_cfr = round(mean(cfr), 3)) %>%
  distinct()
```

```{r}
options(scipen = 999)
```

```{r}
#| label: fig-covid-cases
#| fig-cap: "COVID-19 Cases by Country"
#| fig-alt: "COVID-19 Cases by Country"  
ggplot() +
  geom_polygon(data = worldmap,
               aes(x = long, y = lat, group = group),
               fill = "grey",
               color = "white") +
  geom_point(data = centroids,
             aes(x = long, y = lat,
             size = tot_cases,
             color = log(avg_cfr)), 
             alpha = 0.5) +
  scale_size_continuous(
        labels = scales::unit_format(unit = "M", 
                                     scale = 1e-6)) +
  scale_color_viridis_c() +
  coord_quickmap() +
  labs(title = "COVID-19 Cases by Country",
       caption = "Circle size: total cases, Color: average CFR", 
      size= "Total cases (M)",
      color= "Average CFR (log scale)",
      x = "", y = "") +
  theme(axis.text = element_blank(),
        plot.subtitle = element_text(size = 8))
```

The same graph is made with selected countries only.

```{r}
#| label: fig-covid-cases2
#| fig-cap: "COVID-19 Cases by Selected Countries"
#| fig-alt: "COVID-19 Cases by Country"   
ggplot() +
  geom_polygon(data = worldmap,
               aes(x = long, y = lat, group = group),
               fill = "grey",
               color = "white") +
  geom_point(data = centroids %>%
             filter(country_region %in%
             c("US", "China","United Kingdom", "Canada")),
             aes(x = long, y = lat,
                 size = tot_cases,
                 color = log(avg_cfr)), 
                 alpha = 0.5) +
  scale_size_continuous(
            labels = scales::unit_format(unit = "M", 
                                         scale = 1e-6)) +
  scale_color_viridis_c() +
  coord_quickmap() +
  labs(title = "COVID-19 Cases by Country",
       caption = "Circle size: total cases, Color: average CFR",
       subtitle = "US, China, United Kingdom, Canada",
       size= "Total cases (M)",
       color= "Average CFR (log scale)",
       x = "", y = "") +
  theme(axis.text = element_blank(),
        plot.subtitle = element_text(size = 8))
```

The difference in the number of cases and deaths between the US and
China is striking. The US has the highest number of cases and deaths,
while China has the lowest number of cases and deaths. The United
Kingdom and Canada have a similar number of cases and deaths, but the
United Kingdom has a higher **case fatality rate (CFR)**\index{CFR} than
Canada.

Let's now calculate the DALYs for COVID-19, starting with the YLLs. We
know that the simplified formula for YLLs is the number of deaths
multiplied by the standard life expectancy at the age of death. The
standard life expectancy for these countries varies, but we can use the
global average of 72.6 years.

Also, we group the date by 4 months cycles to calculate the YLLs.

```{r}
COVID19_yll_4months <- COVID19_4countries %>%
  mutate(month = as.numeric(format(date, "%m")),
         year = as.numeric(format(date, "%Y")),
         month_cycle = case_when(month %in% 1:4 ~ "Jan-Apr",
                                 month %in% 5:8 ~ "May-Aug",
                                 month %in% 9:12 ~ "Sep-Dec")) %>%
  group_by(country_region, year, month_cycle) %>%
  summarise(cases = sum(cases),
            deaths = sum(deaths),
            cfr = round(deaths / cases, 3)) %>%
  mutate(ylls = deaths * 72.6) %>%
  arrange(year, month_cycle, ylls)

COVID19_yll_4months %>%
  select(year, month_cycle, ylls) %>%
  head()
```

And then plot the YLLs by country and month cycle, with the y axis
representing the YLLs formatted in millions.

```{r}
#| label: fig-covid-yll
#| fig-cap: "YLLs by Country and Month Cycle"
#| fig-alt: "YLLs by Country and Month Cycle"   
COVID19_yll_4months %>%
  ggplot() +
  geom_col(aes(x = month_cycle, y = ylls, 
               fill = country_region)) +
  scale_y_continuous(
         labels = scales::unit_format(unit = "M", 
                                      scale = 1e-6)) +
  labs(title = "YLLs by Country and Month Cycle",
       fill = "Country",
       x = "4-Month cycle",
       y = "YLLs") +
  facet_wrap(~year) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The majority of YLLs are in the United Kingdom, followed by the US,
Canada, and China. The YLLs are highest in the first 4-month cycle
(Jan-Apr) and decrease in the following cycles.

Let's now calculate the YLDs for COVID-19. The YLDs are calculated by
multiplying the number of cases by the disability weight. The Disability
Weight applied is that for low respiratory diseases which is 0.133. As
of 2020, the Global Burden of Disease Study 2019 estimated that the
DALYs for lower respiratory diseases was 1,000,000. We can use the
`{hmsidwR}` package to get the disability weight for lower respiratory
diseases.

```{r}
hmsidwR::disweights %>%
  filter(str_detect(sequela, "lower respiratory")) %>%
  select(sequela, dw)
```

We know that the simplified formula for YLDs is the prevalence of the
disease multiplied by the disability weight. To calculate the
prevalence, we need the population of each country. For this task we use
the `{wpp2022}` package, which provides the population data from the
2022 revision of the United Nations World Population Prospects (WPP
2022). It includes both historical estimates and projections of
demographic indicators such as population counts and fertility rates,
accommodating the challenges brought on by events like the COVID-19
pandemic and transitioning from five-year to one-year data intervals.

```{r}
#| eval: false
# library(devtools)
# options(timeout = 600)
# install_github("PPgp/wpp2022")
```

```{r}
library(wpp2022)
# data(package = "wpp2022")
```

```{r}
data(pop1dt)
pop_4countries <- pop1dt %>%
  filter(
    year %in% c(2020, 2021),
    name %in% c(
      "United States of America",
      "China",
      "United Kingdom",
      "Canada")) %>%
  mutate(name = ifelse(name == "United States of America", "US", name)) %>%
  select(name, year, pop) %>%
  arrange(year, name)

pop_4countries %>% head()
```

Calculate the prevalence of COVID-19 cases per 1,000,000 people.

$$prevalence = \frac{cases}{population} * 1,000,000$$

```{r}
COVID19_dalys <- COVID19_yll_4months %>%
  full_join(pop_4countries, 
            by = c("country_region" = "name", "year")) %>%
  mutate(prevalence = cases / pop * 1000000,
         dw = 0.133,
         ylds = prevalence * dw,
         dalys = ylls + ylds) %>%
  arrange(year, month_cycle, prevalence) %>%
  select(country_region, year, 
         month_cycle, ylls, ylds, dalys)


COVID19_dalys %>%
  head()
```

If we calculate the YLDs with the incidence instead of prevalence, we
get the same results, because the incidence is the number of new cases
in a given period, while the prevalence is the number of existing cases
in a given period. So, we should consider the cumulative number of cases
to calculate the prevalence. In order to calculate the cumulative number
of cases, we need to group the data by year and month cycle.

And then plot the YLDs by country and month cycle, with the y axis
representing the YLDs formatted in millions.

```{r}
#| label: fig-yld
#| fig-cap: "YLDs by Country and Month Cycle"
#| fig-alt: "YLDs by Country and Month Cycle" 
COVID19_dalys %>%
  filter(year %in% c(2020, 2021)) %>%
  ggplot() +
  geom_col(aes(x = month_cycle, y = ylds, 
               fill = country_region)) +
  scale_y_continuous(
            labels = scales::unit_format(unit = "M", 
                                         scale = 1e-6)) +
  labs(title = "YLDs by Country and Month Cycle",
       fill = "Country",
       x = "4-Month Cycle", y = "YLDs") +
  facet_wrap(~year) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We can see that the YLDs are highest in the United Kingdom, followed by
the US, Canada, and China. The YLDs are highest in the first 4-month
cycle (Jan-Apr) and decrease in the following cycles.

Finally, let's plot the DALYs by country and month cycle, with the y
axis representing the DALYs formatted in millions.

```{r}
#| label: fig-dalys
#| fig-cap: "DALYs by Country and Month Cycle"
#| fig-alt: "DALYs by Country and Month Cycle"   
COVID19_dalys %>%
  filter(year %in% c(2020, 2021)) %>%
  ggplot() +
  geom_col(aes(x = month_cycle, y = dalys, 
               fill = country_region)) +
  scale_y_continuous(
            labels = scales::unit_format(unit = "M", 
                                         scale = 1e-6)) +
  labs(title = "DALYs by Country and Month Cycle",
       fill = "Country",
       x = "4-Month Cycle", y = "DALYs") +
  facet_wrap(~year) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

To explain the magnitude of difference between ylls and ylds,

prevalence as the proportion of cases at baseline among the total number
of enrolled participants who completed baseline visits.
(https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8992269/)

cumulative incidence as the number of incident cases divided by the
total follow-up time per 100 person-years and assumed a uniform
incidence distribution across the 6-month follow-up time.
(https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8992269/)
