# The Case of Malaria {#sec-15-malaria}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title = element_text(face = "bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   What is Malaria
-   How does it spread
-   Map Malaria outbreaks
:::
::::::

\
In this chapter, we explore the dynamics of Malaria transmission in more
detail. We examine the results of various model’s applications that
simulate the spread of the virus.

## Epidemiology

**Malaria** is a mosquito-borne infectious disease that affects humans
and other animals. It is caused by parasitic protozoans belonging to the
Plasmodium type. The disease is **transmitted through the bites of
Anopheles mosquitoes**. The symptoms of malaria typically include fever,
fatigue, vomiting, and headaches. If left untreated, malaria can be
fatal. Malaria is a major public health concern in many tropical and
subtropical regions, particularly in Africa.

Malaria transmission dynamics are influenced by various factors,
including the prevalence of infected individuals, the density of
mosquito vectors, and environmental conditions. The transmission of
malaria occurs when an infected mosquito bites a human host, injecting
the Plasmodium parasites into the bloodstream. The parasites then
multiply within the host's red blood cells, leading to the
characteristic symptoms of malaria. The parasites can be transmitted
back to mosquitoes when they feed on infected individuals, completing
the transmission cycle.

## Mapping Malaria Outbreaks

Mapping malaria outbreaks is essential for locating high-risk areas and
guiding effective public health interventions. Geographic Information
Systems (GIS) enable the visualisation of malaria’s spatial
distribution, revealing transmission patterns and hotspots. The analysis
of malaria cases, alongside environmental factors, such as temperature,
humidity, and vegetation cover, allows the identification of conditions
that facilitate transmission and support the development of targeted
control strategies.

Spatial data on climate, population density, and mosquito habitats are
instrumental in predicting high-risk areas for malaria, aiding
preventive measures like bed net distribution and indoor residual
spraying. In Tanzania, for example, spatial models identified villages
with the highest malaria incidence, allowing the government and NGOs to
prioritize these areas for intervention[@selemani2016]. This targeted
approach enhances the cost-effectiveness of malaria control programs,
reducing the disease burden by directing resources to areas with the
greatest need.

## Example: Simulating Malaria Transmission Dynamics

In this example, we will simulate Malaria transmission dynamics using a
simple mathematical model. We will generate synthetic data representing
the number of infected individuals over time and use machine learning to
predict future trends. We will demonstrate the process of data
preparation, feature engineering, model selection, training, evaluation,
and iterative improvement.

Synthetic data are used in this example to illustrate the modelling and
prediction process. In practice, real-world data on Malaria cases,
environmental factors, and other relevant variables would be used to
develop more accurate models.

```{r}
#| label: fig-malaria-sim
#| fig-cap: "Simulated Malaria Data"
#| fig-alt: "Line plot showing the number of infected individuals over time." 
# Set seed for reproducibility
set.seed(123)

# Number of time steps (e.g., months)
n_steps <- 12

# Initial number of infected individuals
initial_infected <- 10

# Parameters for malaria transmission dynamics
transmission_rate <- 0.2 # Rate of malaria transmission
recovery_rate <- 0.1 # Rate of recovery from malaria

# Initialize vectors to store data
time <- 1:n_steps
infected <- numeric(n_steps)

# Initialize the number of infected individuals
infected[1] <- initial_infected

# Simulate malaria transmission dynamics
for (i in 2:n_steps) {
  # Calculate the number of new infections
  new_infections <- rbinom(1, infected[i - 1], transmission_rate)

  # Calculate the number of recoveries
  recoveries <- rbinom(1, infected[i - 1], recovery_rate)

  # Update the number of infected individuals
  infected[i] <- infected[i - 1] + new_infections - recoveries

  # Ensure that the number of infected individuals is non-negative
  if (infected[i] < 0) {
    infected[i] <- 0
  }
}

# Plot the simulated malaria data
plot(time, infected,
  type = "l",
  xlab = "Time",
  ylab = "Number of Infected Individuals",
  main = "Simulated Malaria Data"
)
```

### Modelling with caret

```{r}
#| eval: true
# Load necessary libraries
library(caret)
```

1.  Data Preparation

Convert 'time' column to Date format

```{r}
malaria_data <- data.frame(time = time, infected_cases = infected)
# malaria_data$time <- as.Date(malaria_data$time)
```

2.  Feature Engineering

Create lagged variables if needed

```{r}
malaria_data$lagged_cases <- lag(malaria_data$infected_cases, 1)
malaria_data <- na.omit(malaria_data)
```

3.  Model Selection

Define the machine learning model (e.g., Random Forest)

```{r}
model <- train(
  infected_cases ~ lagged_cases, # Specify the formula
  data = malaria_data, # Specify the data
  # Specify the machine learning algorithm (Random Forest)
  method = "rf",
  # Cross-validation for hyperparameter tuning
  trControl = trainControl(method = "cv"),
  # Hyperparameter grid for tuning
  tuneGrid = expand.grid(mtry = c(2, 3, 4))
)
```

4.  Model Training (Parameter Calibration)

The 'train' function automatically performs parameter calibration
(hyperparameter tuning)

5.  Evaluation

Predict future trends using the trained model

```{r}
predictions <- predict(model, newdata = malaria_data)
```

Evaluate the model's performance (e.g., using RMSE)

```{r}
rmse <- sqrt(mean((predictions - malaria_data$infected_cases)^2))
cat("Root Mean Squared Error (RMSE):", rmse, "\n")
```

6.  Iterative Improvement

Refine your model by adjusting parameters, feature engineering, etc.

-   Adjusting Parameters, and fine-tune the model by adjusting
    parameters such as 'mtry' in Random Forest. For example, try
    different values of 'mtry' and see how it affects model performance.
    Define tuning grid for mtry:

```{r}
tuneGrid <- expand.grid(mtry = seq(1, ncol(malaria_data) - 1))
model <- train(
  infected_cases ~ lagged_cases,
  data = malaria_data,
  method = "rf",
  trControl = trainControl(method = "cv"),
  tuneGrid = tuneGrid # Adjust the range of 'mtry'
)
```

-   Feature Engineering by exploring additional features that may
    improve model performance. For example, create lagged variables for
    other relevant features or transform existing features to capture
    non-linear relationships. Add a new feature (e.g., log-transformed
    feature) and re-train the model:

```{r}
malaria_data$feature <- log(malaria_data$infected_cases)
malaria_data$lagged_feature <- lag(malaria_data$feature, 1)
malaria_data <- na.omit(malaria_data)

# Re-train the model with the new feature
model <- train(
  # Include the new feature
  infected_cases ~ lagged_cases + lagged_feature,
  data = malaria_data,
  method = "rf",
  trControl = trainControl(method = "cv"),
  tuneGrid = expand.grid(mtry = c(2, 3, 4))
)
```

-   Evaluate Model Performance, after making adjustments, evaluate the
    model's performance again.

```{r}
predictions <- predict(model, newdata = malaria_data)
rmse <- sqrt(mean((predictions - malaria_data$infected_cases)^2))
cat("RMSE after Iterative Improvement:", rmse, "\n")
```
