# The Case of Malaria {#sec-15-malaria}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title = element_text(face = "bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   What is Malaria
-   How does it spread
-   Map Malaria outbreaks
:::
::::::

\
In this chapter, we explore the dynamics of Malaria transmission in more
detail. We examine the results of various model’s applications that
simulate the spread of the virus.

## Epidemiology

**Malaria** is a mosquito-borne infectious disease that affects humans
and other animals. It is caused by parasitic protozoans belonging to the
Plasmodium type. The disease is **transmitted through the bites of
Anopheles mosquitoes**. The symptoms of malaria typically include fever,
fatigue, vomiting, and headaches. If left untreated, malaria can be
fatal. Malaria is a major public health concern in many tropical and
subtropical regions, particularly in Africa.

Malaria transmission dynamics are influenced by various factors,
including the prevalence of infected individuals, the density of
mosquito vectors, and environmental conditions. The transmission of
malaria occurs when an infected mosquito bites a human host, injecting
the Plasmodium parasites into the bloodstream. The parasites then
multiply within the host's red blood cells, leading to the
characteristic symptoms of malaria. The parasites can be transmitted
back to mosquitoes when they feed on infected individuals, completing
the transmission cycle.

## Mapping Malaria Outbreaks

Mapping malaria outbreaks is essential for locating high-risk areas and
guiding effective public health interventions. Geographic Information
Systems (GIS) enable the visualisation of malaria’s spatial
distribution, revealing transmission patterns and hotspots. The analysis
of malaria cases, alongside environmental factors, such as temperature,
humidity, and vegetation cover, allows the identification of conditions
that facilitate transmission and support the development of targeted
control strategies.

In this example, we will use the `malariaAtlas` package to download
malaria data for Nigeria and visualise the distribution of malaria cases
in the country. We will plot the malaria hotspots on a map of Nigeria to
identify regions with the highest incidence of the disease.

```{r}
#| message: false
#| warning: false
# Load necessary libraries
library(malariaAtlas)
library(tidyverse)
library(sf)
library(rnaturalearth)
```

To download malaria data we can use the `getPR()` function, it releases
the data from the Malaria Atlas Project API. The function requires the
country and species of Plasmodium to be specified. In this example, we
will download data for Nigeria and the Plasmodium falciparum species.

```{r}
#| eval: false
nigeria_data <- getPR(country = "Nigeria", species = "Pf")
```

```{r}
#| eval: false
#| echo: false
# saveRDS(nigeria_data, "data/inst/extdata/nigeria_data.rds")
```

```{r}
#| echo: false
nigeria_data <- readRDS("data/inst/extdata/nigeria_data.rds")
nigeria_data %>% pull(year_start) %>% range();
# 23 years spanning from 1985 to 2018
nigeria_data %>% count(year_start) %>% dim()
```

Data contains cases for 23 years spanning from 1985 to 2018. We can
extract the relevant information (year_start (of the survey), longitude,
latitude, and number of positive cases) and convert the data to a
spatial object using the `st_as_sf()` function from the `sf` package.

```{r}
nigeria_data <- nigeria_data %>%
  arrange(year_start) %>%
  select(year_start,longitude,latitude,positive) %>%
  filter(!is.na(longitude) & !is.na(year_start)) 

nigeria_data_sf <- nigeria_data %>%
  st_as_sf(coords = c("longitude", "latitude"), 
                       crs = 4326)

head(nigeria_data_sf)
```

To get the administrative boundaries of Nigeria we use the
{rnaturlaearth} package:

```{r}
nigeria_sf <- ne_countries(country = "Nigeria", returnclass = "sf")
```

And finally plot the malaria hotspots on a map of Nigeria to visualise
the distribution of malaria cases in the country.

```{r}
#| label: fig-malaria-map
#| fig-cap: "The map shows the distribution of malaria cases in Nigeria, with the size and color of the points representing the number of positive cases. The map highlights regions with a high incidence of malaria, providing valuable information for public health interventions."
#| fig-alt: "Map showing the distribution of malaria cases in Nigeria."
# Plot Malaria Hotspots with expanded spatial data range
ggplot() +
  geom_sf(data = nigeria_sf, 
          fill = "gray90", 
          color = "white")+
geom_sf(data = nigeria_data_sf, 
        aes(size=positive, color = positive),
        alpha=0.5)+
  scale_color_viridis_c(option = "plasma", name = "Malaria Cases") +
  guides(size = "none") +
  labs(title = "Malaria Hotspots in Nigeria",
       subtitle = "1985-2018",
       caption = "Source: Malaria Atlas Project | @fgazzelloni") +
  theme(legend.position = "right")
```

Spatial data on climate, population density, and mosquito habitats are
instrumental in predicting high-risk areas for malaria, aiding
preventive measures like bed net distribution and indoor residual
spraying. In Tanzania, for example, spatial models identified villages
with the highest malaria incidence, allowing the government and NGOs to
prioritize these areas for intervention[@selemani2016]. This targeted
approach enhances the cost-effectiveness of malaria control programs,
reducing the disease burden by directing resources to areas with the
greatest need.

```{r}
#| echo: false
#| label: fig-malaria-tanzania
#| fig-cap: "The map shows the distribution of malaria cases in Tanzania, with the size and color of the points representing the number of positive cases. The map highlights regions with a high incidence of malaria, providing valuable information for public health interventions."
#| fig-alt: "Map showing the distribution of malaria cases in Tanzania."  
#tanzania_data <- getPR(country = "Tanzania", species = "Pf")
#saveRDS(tanzania_data, "data/inst/extdata/tanzania_data.rds")  
tanzania_data <- readRDS("data/inst/extdata/tanzania_data.rds")
tanzania_data_sf <- tanzania_data %>%
  arrange(year_start) %>%
  select(year_start,longitude,latitude,positive) %>%
  filter(!is.na(longitude) & !is.na(year_start)) %>%
  st_as_sf(coords = c("longitude", "latitude"), 
                       crs = 4326)

tanzania_sf <- ne_countries(country = "United Republic of Tanzania", 
                                        returnclass = "sf")
ggplot() +
  geom_sf(data = tanzania_sf, 
          fill = "gray90", 
          color = "white")+
geom_sf(data = tanzania_data_sf, 
        aes(size=positive, color = positive),
        alpha=0.5)+
  scale_color_viridis_c(option = "plasma", name = "Malaria Cases") +
  guides(size = "none") +
  labs(title = "Malaria Hotspots in Tanzania",
       subtitle = "1985-1994",
       caption = "Source: Malaria Atlas Project | @fgazzelloni") +
  theme(legend.position = "right")
```

## Example: Simulating Malaria Transmission Dynamics

In this example, we will use data for Malaria positive cases in Nigeria
from the Malaria Atlas Project to evaluate the transmission dynamics
using a simple mathematical model. We will use machine learning to
predict future trends.

To illustrate the modelling and prediction process we will use the
`{caret}` package, which provides a unified interface for training and
evaluating machine learning models in R. We will train a Random Forest
model to predict future trends in malaria transmission based on
historical data.

```{r}
nigeria_cases <- nigeria_data %>% 
  group_by(year_start) %>%
  reframe(positive = sum(positive)) %>%
  rename(year = year_start) %>%
  drop_na() %>%
  select(year, positive)
head(nigeria_cases)
```

Visualize the dynamics of malaria transmission in Nigeria using a line
plot to show the number of infected cases over time.

```{r}
nigeria_cases %>%
  ggplot() +
  geom_line(aes(x = year, y = positive), color = "navy", linetype = "solid") +
  labs(title = "Malaria Transmission Dynamics in Nigeria",
       subtitle = "23 years spanning from 1985 to 2018",
       x = "Time(Year)",
       y = "Number of Infected Cases") +
  theme_minimal()
```

### Modelling with caret

To train a machine learning model to predict future trends in malaria
transmission, we will be using 80% of the original data to train the
model and 20% to test it. Then, we will evaluate its performance using
the Root Mean Squared Error (RMSE).

The `{caret}` package provides a unified interface for training and
evaluating machine learning models in R (@sec-07-packages). It supports
a wide range of algorithms and provides tools for hyperparameter tuning,
cross-validation, and model evaluation.

```{r}
#| eval: true
# Load necessary libraries
library(caret)
```

#### Feature Engineering

Create lagged variables to capture the temporal dynamics of malaria
transmission. For time series forecasting, it’s helpful to create
variables, which represent past values of the target variable. This
allows the model to learn trends from previous values.

```{r}
# Create lagged features (e.g., previous year's cases)
nigeria_cases <- nigeria_cases %>%
  arrange(year) %>%
  mutate(lag_1 = lag(positive, 1),
         lag_2 = lag(positive, 2),
         lag_3 = lag(positive, 3)) %>%
  drop_na()
head(nigeria_cases)
```

Create partition for training and testing data using the
`createDataPartition()` function from the `caret` package.

```{r}
set.seed(123)
train_index <- createDataPartition(nigeria_cases$positive, 
                                   p = 0.8, list = FALSE)
train <- nigeria_cases[train_index, ]
test <- nigeria_cases[-train_index, ]
```

#### Model Selection and Training (Parameter Calibration)

Define the machine learning model specific for investigating Malaria
dynamics. In this case, we will use the time and number of lagged cases
to predict the number of infected cases. We will use the `train()`
function from the `{caret}` package to train the model. A suitable model
would be? List the models that can be used for this task: - Random
Forest ("rf") - Gradient Boosting ("gbm") - Support Vector Machines
("svm") - Neural Networks ("nnet") - etc.

In the `method` argument, specify the model to be used (e.g., "rf" for
Random Forest). The `trControl` argument specifies the cross-validation
method for hyperparameter tuning, and the `tuneGrid` argument defines
the hyperparameter grid for tuning.

```{r}
# Define the training control with 5-fold cross-validation
train_control <- trainControl(method = "cv", number = 5)

# Define a tuning grid for mtry (number of variables sampled at each split)
tuning_grid <- expand.grid(mtry = c(1, 2, 3))

# Train the Random Forest model
set.seed(123)
rf_model <- train(
  positive ~ lag_1 + lag_2 + lag_3,
  data = train,
  method = "rf",
  trControl = train_control,
  tuneGrid = tuning_grid,
  ntree = 500  # Increase the number of trees if desired
)
```

The parameter calibration (hyperparameter tuning) is performed by the
`train` function automatically using cross-validation to select the
optimal hyperparameters for the model.

#### Model Evaluation

Testing the model on observed data will show how the model performs in
estimating the number of infected cases.

```{r}
# Predict on the test data
nigeria_cases$pred <- predict(rf_model, newdata = nigeria_cases)

nigeria_cases %>%
 ggplot(aes(x = year, y = positive)) +
  geom_line(color = "navy", linetype = "solid") +
  geom_line(aes(y = pred), color = "darkred", linetype = "dashed") +
  labs(title = "Observed vs Predicted Malaria Positive Cases in Nigeria",
       x = "Year",
       y = "Number of Positive Cases") +
  theme_minimal()
```

To evaluate the model's ability in predicting future trends, we predict
the number of infected cases on the test data and calculate the Root
Mean Squared Error (RMSE). The test data is a subset of the original
data that was not used for training the model, and it provides an
independent evaluation of the model's performance.

```{r}
# Predict on the test data
test$pred <- predict(rf_model, newdata = test)

# View predictions alongside actual values
head(test[, c("year", "positive", "pred")])
```

Visualize the predicted vs actual infected cases using a line plot to
compare the model's predictions with the actual data.

```{r}
#| label: fig-rf-malaria-predictions
#| fig-cap: "The plot shows the observed vs predicted malaria positive cases in Nigeria made with a Random Forest model. The blue line represents the observed cases, while the red line represents the predicted cases. The model's predictions are not closely aligned with the observed values, indicating potential limitations in capturing the dynamics of malaria transmission."
#| fig-alt: "Line plot showing the observed vs predicted malaria positive cases in Nigeria."
# Reshape data for plotting
plot_data <- test %>%
  select(year, positive, pred) %>%
  rename(Observed = positive, Predicted = pred) %>%
  pivot_longer(cols = c("Observed", "Predicted"), 
               names_to = "Type", values_to = "Cases")

# Plot observed vs predicted cases
ggplot(plot_data, aes(x = year, y = Cases, 
                      color = Type, linetype = Type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("navy", "darkred")) +
  labs(title = "Observed vs Predicted Malaria Positive Cases in Nigeria",
       x = "Year",
       y = "Number of Positive Cases") +
  theme_minimal() +
  theme(legend.position = "top")
```

Evaluate the model's performance using RMSE:

```{r}
# Calculate RMSE
rmse <- sqrt(mean((test$positive - test$pred)^2))
cat("RMSE:", rmse, "\n")
```

A Root Mean Squared Error (RMSE) of 247.3936 indicates the average
difference between the predicted and actual values of infected cases.
Lower RMSE values indicate better model performance.

The output shows that the model’s predictions are not closely aligned
with the observed values, the predicted trend seems to be relatively
flat or declining, while the observed cases show significant
fluctuations. This mismatch suggests that the current model setup may
not be effectively capturing the dynamics of malaria transmission in
this dataset.

## Model Refinement

To improve the model's performance, we can refine the model by adjusting
parameters such as 'mtry' in Random Forest. For example, increase the
number of folds in cross-validation, adjust the tuning grid, or add
additional features to improve the model's performance.

```{r}
#| output: false
# Increase the number of folds in cross-validation
train_control <- trainControl(method = "cv", number = 10)  

# Scaling data in caret
rf_model2 <- train(
  positive ~ year + lag_1 + lag_2 + lag_3,
  data = train,
  method = "rf",
  trControl = train_control,
  tuneGrid = expand.grid(mtry = 2:4),
  preProcess = c("center", "scale")
)
```

Predict future trends using the trained adjusted Random Forest model on
the test data.

```{r}
# Predict on the test data
test$pred_rf2 <- predict(rf_model2, newdata = test)
```

Visualize the predicted vs actual infected cases using a line plot to
compare the adjusted Random Forest model's predictions with the actual
data.

```{r}
#| label: fig-xgb-malaria-predictions
#| fig-cap: "The plot shows the observed vs predicted malaria positive cases in Nigeria made with an adjusted Random Forest model. The blue line represents the observed cases, while the red line represents the predicted cases. The model's predictions show a closer alignment with the observed values compared to the first Random Forest model, indicating improved performance."
#| fig-alt: "Line plot showing the observed vs predicted malaria positive cases in Nigeria using XGBoost."  
# Reshape data for plotting
plot_data_xgb <- test %>%
  select(year, positive, pred_rf2) %>%
  rename(Observed = positive, Predicted = pred_rf2) %>%
  pivot_longer(cols = c("Observed", "Predicted"), 
               names_to = "Type", values_to = "Cases")

# Plot observed vs predicted cases
ggplot(plot_data_xgb, aes(x = year, y = Cases, 
                      color = Type, linetype = Type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("navy", "darkred")) +
  labs(title = "Observed vs Predicted Malaria Positive Cases in Nigeria (XGBoost)",
       x = "Year",
       y = "Number of Positive Cases") +
  theme_minimal() +
  theme(legend.position = "top")
```

Evaluate the model's performance improvement using RMSE:

```{r}
# Calculate RMSE
rmse <- sqrt(mean((test$positive - test$pred_rf2)^2))
cat("RMSE:", rmse, "\n")
```

In conclusion, the model's performance can be further improved by
adjusting hyperparameters, adding additional features, or using more
advanced algorithms. The iterative process of model refinement and
evaluation is essential for developing accurate predictions of malaria
transmission dynamics.

## Summary

In this chapter, we explored the dynamics of Malaria transmission and
the importance of mapping outbreaks to guide public health
interventions. We used the `malariaAtlas` package to download malaria
data for Nigeria and visualised the distribution of malaria cases in the
country. We also demonstrated how machine learning can be used to
predict future trends in malaria transmission using historical data. The
iterative process of model refinement and evaluation is crucial for
developing accurate predictions and guiding effective public health
interventions.


