# Summary: The State of Health {#sec-16-summary}

```{r}
#| echo: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

::: solutionbox
::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
:::

::: solutionbox-body
-   Estimating the burden of rabies
-   Health metrics countries comparison
:::
:::

In this chapter, we delve into two primary learning objectives aimed at
enhancing our understanding of public health dynamics: **estimating the
burden of rabies** and **comparing health metric levels across different
countries**. Rabies, a deadly viral disease transmitted through the
saliva of infected animals, poses a significant threat to human health
globally. Understanding its burden is crucial for effective control and
prevention strategies. Additionally, we explore the variations in health
metrics due to rabies. In conclusion, an overview of DALYs levels among
different countries will provide insights into the disparities in health
outcomes and the factors influencing them. Through these learning
objectives, we aim to equip readers with the knowledge and tools
necessary for a comprehensive understanding of public health challenges.

## Estimating the burden of rabies

The disability-adjusted life–year (DALY), as standardized metrics
incorporate premature mortality and disability due to disease; as rabies
is rapidly fatal, disability accounts for a minimal part of the disease
burden[@whoexpe].

Let's have a look at the data for Rabies in Asia, Africa and Middle East
from to 2020

```{r}
suppressPackageStartupMessages({
  library(hmsidwR)
  library(tidyverse)
  library(fpp3)
  library(tsibble)
})

data(rabies)
```

```{r}
hmsidwR::rabies %>%
  head() %>%
  knitr::kable()
```

```{r}
rabies %>%
  count(location)
```

```{r}
ts_rabies <- rabies %>%
  filter(
    measure == "DALYs",
    location == "Asia",
    cause == "Rabies"
  ) %>%
  select(year, val) %>%
  as_tsibble(index = "year")


ts_rabies %>%
  autoplot(.vars = val) +
  labs(
    title = "DALYs trend of Rabies in Asia",
    subtitle = "1990 to 2020", y = "DALYs"
  ) #+
# theme_hmsid()
```

```{r}
ts_rabies %>%
  model(stl = STL(val)) %>%
  components() %>%
  autoplot()
```

```{r}
ts_rabies %>%
  model(stl = STL(val)) %>%
  components() %>%
  as_tsibble() |>
  autoplot(val, colour = "gray") +
  geom_line(aes(y = trend), colour = "#D55E00")
```

```{r}
ts_rabies %>%
  gg_tsdisplay(
    y = val,
    plot_type = "partial"
  )
```

```{r}
rabies %>%
  filter(
    measure == "DALYs",
    # location=="Asia",
    cause == "Rabies"
  ) %>%
  select(location, year, val) %>%
  ggplot(aes(x = year, y = val, group = location, color = location)) +
  geom_line() +
  labs(
    title = "DALYs trend of Rabies",
    subtitle = "1990 to 2020", y = "DALYs"
  )
```

Assume the estimates made by the IHME are now our observed values, apply
a model to see how would the number of DALYs due to rabies would develop
over time. We can safety say it is a decreasing trend, but as one of the
deadliest neglected disease the objective is its eradication.

The only covariate (predictor) used is `year`, as variable that is
potentially predictive of the outcome number of DALYs (`val`). It might
be worthy to have a look at how it could vary if other covariates are
added in the model.

The relationship is clearly linear, so we apply a linear model.

```{r}
dat <- rabies %>%
  filter(
    measure == "DALYs",
    location == "Asia",
    cause == "Rabies"
  ) %>%
  select(year, val)

mod1 <- lm(val ~ ., data = dat)
summary(mod1)
```

```{r}
cbind(
  dat,
  pred = predict(mod1, dat)
) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = val, ), color = "grey") +
  geom_line(aes(y = pred), color = "red")
```

Our basic regression model explains \~ 90% of the variability in the
data. This can be seen looking at the R-squared, we would have looked at
the adjusted R-squared in case of more than one predictor.

Let's make a random forest model

```{r}
#| message: false
#| warning: false
library(randomForest)
myrf <- randomForest::randomForest(formula = val ~ ., data = dat)

cbind(
  dat,
  pred = predict(myrf, dat)
) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = val, ), color = "grey") +
  geom_line(aes(y = pred), color = "red")
```

Now we add some other predictors and parameters, death rates (`deaths`)
due to rabies and location.

```{r}
dat2 <- rabies %>%
  filter(cause == "Rabies") %>%
  select(-upper, -lower, -cause) %>%
  pivot_wider(names_from = measure, values_from = val) %>%
  rename(deaths = Deaths, dalys = DALYs)

dat2 %>% head()
```

```{r}
#| fig.align: center

mod_rf2 <- randomForest::randomForest(formula = dalys ~ ., data = dat2)

cbind(
  dat2,
  pred = predict(mod_rf2, dat2)
) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = dalys, group = location, color = location)) +
  geom_line(aes(y = pred, group = location), color = "red") +
  scale_color_manual(values = c("#29306c", "#f8bc05"))
```

### Machine Learning

Parameters estimation with `{tidymodels}`, we use techniques such as
`cross-validation` to tune the hyperparameters of the model. Here's how
you can do it:

```{r}
suppressPackageStartupMessages({
  library(tidymodels)
})
tidymodels_prefer()
```

Define the recipe for data preprocessing

```{r}
data_recipe <- recipe(dalys ~ ., data = dat2) %>%
  step_dummy(location) %>%
  step_scale(all_predictors()) %>%
  step_center(all_predictors())

data_recipe %>%
  prep() %>%
  juice() %>%
  head()
```

Define the model specification

```{r}
rf_spec <- rand_forest(mtry = tune(), trees = tune()) %>%
  set_mode("regression") %>%
  set_engine("ranger") # Use the ranger engine for faster computation
```

Define the tuning grid The parameters to be tuned for a `randomforest`
model are:

-   mtry: the number of predictors that will be randomly sampled at each
    split when creating the tree models
-   trees: the number of trees contained in the ensemble
-   min_n: the minimum number of data points in a node that are required
    for the node to be split further

```{r}
rf_grid <- grid_regular(
  # Vary the number of variables randomly sampled as candidates at each split
  mtry(range = c(1, ncol(dat2) - 1)),
  # Vary the number of trees in the forest
  trees(range = c(50, 200))
)
```

Define the resampling strategy IN this case we use a cross validation
resampling methods performing 10 folds.

```{r}
set.seed(123)
cv <- vfold_cv(dat2, v = 10)
```

Define the model workflow

```{r}
rf_wf <- workflow() %>%
  add_recipe(data_recipe) %>%
  add_model(rf_spec)
```

Machine Learning: Perform parameter tuning

```{r}
rf_tune <- tune_grid(
  object = rf_wf,
  resamples = cv,
  grid = rf_grid,
  metrics = metric_set(rmse)
)
```

Model selection

```{r}
best_model <- select_best(rf_tune)
best_model
```

```{r}
# Fit the best model to the entire dataset
final_rf_model <- rf_wf %>%
  finalize_workflow(best_model) %>%
  fit(dat2)
```

Make predictions

```{r}
predictions <- predict(final_rf_model, dat2)
```

Visualize the results

```{r}
cbind(dat2, predictions) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = dalys, group = location, color = location)) +
  geom_line(aes(y = .pred, group = location), color = "red") +
  scale_color_manual(values = c("#29306c", "#f8bc05"))
```

## Health metrics countries comparison

The Global Burden of Disease (GBD) study provides a comprehensive and up
to date assessment of health metrics across countries. Let's compare the
levels of health metrics in different countries.

```{r}
```
