# Summary: The State of Health {#sec-16-summary}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title = element_text(face = "bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   How health metrics vary across different countries.
-   Gain insights from the Global Burden of Disease (GBD) study on
    health metrics at the country level.
-   Evaluate the implications of health metrics for global public health
    policy and interventions.
:::
::::::

\
In the previous sections, we explored various aspects of health metrics,
including tailored case studies, data sources, modelling, and
visualisation techniques. We discussed key health metrics such as
mortality rates, life expectancy, Disability-Adjusted Life Years
(DALYs), Years of Life Lost (YLLs), and Years Lived with Disability
(YLDs). We also examined the role of risk factors, disease burdens, and
health outcomes in shaping population health profiles.

In this final chapter, we will focus on visualising health metrics
across countries to understand the state of health globally. We will
explore the implications of cross-country comparisons of health metrics
for public health policy and interventions. We will use data from the
**Global Burden of Disease (GBD)** study and the **OECD’s Health at a
Glance** report to compare health indicators, disease burdens, and risk
factors across different countries. By analysing health metrics at the
country level, we can gain insights into public health priorities,
identify areas for improvement, and inform global public health policy.

## Data Sources for Health Metrics Comparison

Data availability is crucial for comparing health metrics across
countries and regions. The **OECD’s Health at a Glance** report and the
**Global Burden of Disease (GBD)** study offer comprehensive data on
health indicators, disease burdens, and risk factors, enabling
cross-country comparisons and insights into public health priorities.

### Example of OECD Health at a Glance Data

The **OECD’s Health at a Glance** report offers a comparative overview
of health indicators among member countries. The report examines metrics
such as life expectancy, health expenditures, and major health trends.
It also provides accessible insights into public health priorities and
performance indicators, facilitating comparisons and helping to identify
areas for improvement ([www.oecd-ilibrary.org](www.oecd-ilibrary.org)).

In this example, we will download data from the OECD's Health at a
Glance report website (<https://data-explorer.oecd.org/>) to compare
**Disability-adjusted life years (DALYs) due to Ambient Particulate
Matter** as a risk exposure per 1 000 inhabitants (accessed Nov 2024).

We can use the `url` for DALYs due to Ambient Particulate Matter found
by searching for the keyword `DALYs` in the search box of the OECD
report website.

The url is broken down into the following components:

-   sdmx.oecd.org/public/rest/data/
-   OECD.ENV.EPI,DSD_EXP_MORSC\@DF_EXP_MORSC,1.0/
-   .A.DALY.10P3HB.PM_2_5_OUT.\_T.\_T?startPeriod=2010

Data are available to download in the `SDMX` format, which is a standard
for exchanging statistical data. We will use the `{rsdmx}` package to
download the data and convert it to a data frame for analysis.

```{r}
#| eval: false
install.packages("rsdmx")
library(rsdmx)
```

```{r}
library(tidyverse)
```

```{r}
#| eval: false
# Read the data
sdmx_data <- readSDMX(url)

# Convert SDMX data to a data frame
df <- as.data.frame(sdmx_data)
```

```{r}
#| eval: false
#| echo: false
saveRDS(df, "data/inst/extdata/16-oecd_dalys.rds")
```

```{r}
#| echo: false
df <- readRDS("data/inst/extdata/16-oecd_dalys.rds")
```

A glimpse of the data shows the structure of the data frame, including
the column names and data types.

```{r}
df %>%
  janitor::clean_names() %>%
  head(3) %>%
  glimpse()
```

Observation time spans from 2010 to 2019 for 212 countries. We filter
the data to include only the relevant columns and arrange it by
observation time to get a better understanding of the data.

```{r}
dalys_pm25 <- df %>% 
  janitor::clean_names() %>%
  select(ref_area, measure, risk, obs_time, obs_value) %>%
  arrange(obs_time) 

dalys_pm25 %>%
  head()
```

Then, to make the data more interpretable, we convert the `ref_area`
column from `ISO3` codes to `country names` using the `{countrycode}`
package. Data provides information for various regions and groups, which
we exclude from the analysis.

```{r}
library(countrycode)
dalys_pm25 <- dalys_pm25 %>%
   filter(!ref_area %in% c("A9", "ASEAN", "EA19", "EU27",
                         "EU28", "F98", "G20", "OECD", 
                         "OECDA", "OECDE", "OECDSO", "W")) %>%
  mutate(ref_area = countrycode::countrycode(ref_area, 
                                             origin = "iso3c", 
                                             destination = "country.name"))
```

Finally, to visualise the data we use a line plot to compare
Disability-adjusted life years (DALYs) due to Ambient Particulate Matter
as a risk exposure per 1 000 inhabitants for Egypt, Serbia, Morocco,
Italy, and Iceland. These countries were selected based on their
geographical location and varying levels of DALYs due to Ambient
Particulate Matter.

```{r}
#| echo: false
#| label: fig-oecd-dalys
#| fig.cap: "Disability-adjusted life years (DALYs) due to Ambient Particulate Matter as a risk exposure per 1 000 inhabitants in selected countries from 2010 to 2019."
#| fig.alt: "A line plot showing Disability-adjusted life years (DALYs) due to Ambient Particulate Matter as a risk exposure per 1 000 inhabitants."

dalys_pm25 %>%
  filter(ref_area %in% c("Egypt","Serbia","Morocco","Italy","Iceland")) %>%
  ggplot(aes(x = obs_time, y = obs_value, 
             group = ref_area, 
             color = ref_area)) +
  geomtextpath::geom_textline(aes(label = ref_area), 
                              linewidth = 2,
                              hjust = 0.5, 
                              vjust = 0.5, 
                              size = 5) +
  labs(title = "Disability-Adjusted Life Years (DALYs)",
       subtitle = "Risk Exposure: Ambient Particulate Matter (per 1 000 inhabitants)",
       x = "Time(Year)", y = "DALYs per 1 000 inhabitants") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

We can conclude that the Disability-adjusted life years (DALYs) due to
Ambient Particulate Matter varied across countries from 2010 to 2019.
This variation may be due to differences in environmental policies, air
quality, and public health interventions aimed at reducing exposure to
particulate matter. In particular, countries like Egypt and Serbia
experienced higher DALYs due to Ambient Particulate Matter compared to
countries like Iceland and Italy.

In general, the data shows that regions with high air pollution, such as
parts of East Asia, see increased DALYs from respiratory conditions.
This is due to the high levels of particulate matter in the air, which
can lead to respiratory diseases like asthma and chronic obstructive
pulmonary disease (COPD). Countries with stringent air quality
regulations, such as those in Western Europe, tend to have lower DALYs
from air pollution-related conditions.

### Example of GBD Data Cross-Country Comparisons

The **Global Burden of Disease GBD** study by the **Institute for Health
Metrics and Evaluation (IHME)** provides comprehensive data on various
health metrics covering a vast array of diseases, injuries, and risk
factors across 204 countries. To access the GBD data
(\<www.healthdata.org\>), we can use the **Sustainable Development Goals
(SDG) API** provided by the IHME. The API allows users to access GBD
data for specific indicators, locations, and years, enabling
cross-country comparisons. In @sec-collecting-data is provided a
detailed guide on how to access and use the **SDG API**.

In this example, we will compare the **Tuberculosis Incidence Rate in
2019** for selected countries using the Sustainable Development Goals
(SDG) API data. We will use the `{hmsidwR}` package to download the data
and visualise the results using a bar plot.

```{r}
#| eval: false
# Install the hmsidwR package
install.packages("hmsidwR")
# or the development version
devtools::install_github("fgazzelloni/hmsidwR")
```

```{r}
# Load the libraries
library(hmsidwR)
```

We download the data using the `gbd_get_data` function from the
`{hmsidwR}` package. The function requires the URL, API key, endpoint,
indicator ID, location ID, and year as input parameters, as shown below:

```{r}
#| eval: false
data_1001 <- hmsidwR::gbd_get_data(
  url = "https://api.healthdata.org/sdg/v1",
  key = "YOUR-KEY",
  endpoint = "GetResultsByIndicator",
  indicator_id = "1001",
  location_id = c("29","86","102"),
  year = "2019"
)
```

```{r}
#| eval: false
#| echo: false
saveRDS(data_1001, "data/inst/extdata/16-data_1001.rds")
```

```{r}
#| echo: false
data_1001 <- readRDS("data/inst/extdata/16-data_1001.rds")
```

Filter the data by both gender and subset to include only the relevant
columns:

```{r}
data_1001 %>%
  filter(sex_label=="Both sexes") %>%
  select(location_name, indicator_short, mean_estimate) 
```

To compare the cross-countries results to identify disparities in
Tuberculosis Incidence Rate in 2019 for selected countries, we can use a
bar plot.

```{r}
#| echo: false
#| label: fig-tbc-comparison
#| fig.cap: "A bar plot showing Tuberculosis Incidence Rate in 2019 for selected countries. Tonga is a country located in the South Pacific, east of Australia, and north of New Zealand. It is part of the region known as Oceania, being situated east of Fiji, south of Samoa, and west of the Cook Islands. Tonga is made up of 170 islands, where four-fifths of them are uninhabited."
#| fig.alt: "A bar plot showing Tuberculosis Incidence Rate in 2019 for selected countries."

data_1001 %>%
  filter(sex_label == "Both sexes") %>%
  mutate(mean_estimate = round(as.double(mean_estimate),2)) %>%
  select(location_name, indicator_short, sex_label, mean_estimate) %>%
  ggplot(aes(x = fct_reorder(location_name,mean_estimate),
             y = mean_estimate)) +
  geom_bar(stat = "identity", 
           position = "stack", 
           fill = "navy") +
  coord_flip() +
  labs(title = "Tuberculosis Incidence Rate in 2019",
       subtitle = "Italy, Tonga, and United States of America",
       x = "Country", y = "Tuberculosis Incidence Rate per capita") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Tuberculosis can still be a significant concern, even in smaller island
nations, due to factors like limited healthcare infrastructure, high
prevalence of co-morbidities, and socioeconomic conditions. Tonga, a
country located in the South Pacific, east of Australia, and north of
New Zealand, part of the region known as Oceania, is made up of 170
islands, where four-fifths of them are uninhabited. The country faces
challenges in managing infectious diseases like tuberculosis due to its
geographical location and limited resources. The comparison of
Tuberculosis Incidence Rate in 2019 for Italy, Tonga, and the United
States of America highlights the disparities in disease burden and
health outcomes across countries.

## Key Determinants of Health Metrics Variation

Health metrics can vary significantly across countries due to a
combination of factors that influence population health outcomes. Some
of the key determinants of health metrics variation include:

-   Healthcare Infrastructure and Access
-   Socioeconomic Status
-   Environmental Factors
-   Lifestyle and Behavioral Factors
-   Public Health Policies and Interventions
-   Cultural and Social Norms

These determinants influence health outcomes and disease burdens in
different populations, shaping the overall health profile of a country.
The availability and quality of healthcare services, income level,
employment status, and education, are strongly linked to health
outcomes. Geographic factors, such as climate, also influence disease
patterns; tropical regions may have higher rates of vector-borne
diseases like malaria and for this reason have higher DALYs due to these
diseases. Countries that prioritize preventive health measures often see
long-term benefits in population health metrics.

## Example: Years Lived with Disability (YLDs) due to Injuries

The **Global Burden of Disease (GBD)** study provides detailed data on a
wide range of health metrics, including Years Lived with Disability
(YLDs) due to injuries. Considered non fatal outcomes, YLDs measure the
impact of injuries on population health by quantifying the years lived
with a disability caused by an injury. By comparing YLDs due to injuries
across countries with different Sustainable Development Index (SDI)
levels, we can identify disparities in injury-related health outcomes
and inform public health interventions to reduce the burden of injuries.

In this example we will use the GBD data to compare the **Years Lived
with Disability (YLDs) due to Injuries** for high, middle, and low
**Sustainable Development Index (SDI)** countries from 1990 to 2021.
Data are available on the GBD website at
[www.healthdata.org](www.healthdata.org)[^16-summary-1], and we will use
the `{ggplot2}` package to create a line plot visualising the YLDs due
to Injuries by country and SDI level.

[^16-summary-1]: For reproducibility a copy of the "injuries" data are
    stored on the GitHub Repository of this book
    (<https://github.com/Fgazzelloni/hmsidR>) in the data/inst/extdata
    folder.

```{r}
#| echo: false
ylds_injuries <- readRDS("data/inst/extdata/16-ylds_injuries.rds")
```

```{r}
ylds_injuries %>% head(3) %>% glimpse()
```

```{r}
#| label: fig-ylds-injuries
#| fig.cap: "GBD-YLDs due to Injuries (1990 - 2021) by country and SDI level."
#| fig.alt: "A line plot showing GBD-YLDs due to Injuries (1990 - 2021)"
ylds_injuries %>%
  filter(!str_detect(location_name, "-middle")) %>%
  ggplot(aes(x = year, y = avg, 
             group = location_name)) +
  geomtextpath::geom_textline(aes(label = location_name), 
                                   hjust = 0.5, 
                                   vjust = 0.5, 
                                   size = 2) +
  facet_wrap(~ sex_name, scales = "free") +
  labs(title = "GBD-YLDs due to Injuries (1990 - 2021)",
       caption = "IHME Data | Graphic: @fgazzelloni",
       x = "Time (Year)", y = "Average Value") +
  theme_minimal()
```

We can conclude that the Years Lived with Disability (YLDs) due to
Injuries varied across countries with different Sustainable Development
Index (SDI) levels from 1990 to 2021. High SDI countries generally had
lower YLDs due to Injuries compared to middle and low SDI countries.
This variation may be due to differences in healthcare infrastructure,
access to preventive services, and public health interventions aimed at
reducing injuries and their impact on population health.

Additionally, the data reveals a consistent decline in YLD rates due to
injuries across all SDI levels (High, Middle, and Low) from 1990 to
2021. This trend reflects advancements in injury prevention, improved
healthcare access, and enhanced living conditions, which have
collectively reduced the disability burden of injuries among females.
However, there is a notable crossover in injury-related YLD rates for
females between middle and low SDI countries over time. This shift
suggests changing dynamics in injury risk and healthcare access, with
middle SDI countries achieving greater reductions in disability from
injuries, while low SDI countries experience a rising burden.

## Example: Injuries Cross-Country Variation by Type

In this example, we will use the GBD data to compare the types of
injuries contributing to the burden of disease in high, middle, and low
SDI countries from 1990 to 2021[^16-summary-2]. For example, we can
compare the average number of injuries by type (e.g., road injuries,
falls, self-harm) in high, middle, and low SDI countries. This
comparison can help identify the most common types of injuries and
inform targeted public health interventions to reduce the burden of
injuries in different populations.

[^16-summary-2]: For reproducibility a copy of the "injuries" data are
    stored on the GitHub Repository of this book
    (<https://github.com/Fgazzelloni/hmsidR>) in the data/inst/extdata
    folder.

```{r}
#| echo: false
injuries_types <- readRDS("data/inst/extdata/16-injuries_types.rds")
```

```{r}
injuries_types %>% head(2) %>% glimpse()
```

To compare the types of injuries contributing to the burden of disease
in different countries, we can use a **treemap visualisation**. A
treemap is a hierarchical chart that displays data in nested rectangles,
with each rectangle representing a different category or subcategory. In
this example, we will create a treemap to visualise the average number
of injuries by type in high, middle, and low SDI countries.

```{r}
library(treemapify)
# ?geom_treemap
```

```{r}
#| label: fig-injuries-treemap
#| fig.cap: "GBD-Injuries Treemap (1990 - 2021)"
#| fig.alt: "A treemap showing GBD-Injuries by type (1990 - 2021)."
injuries_types %>%
  filter(!str_detect(location_name, "-middle")) %>%
  ggplot(aes(area = avg, 
             label = cause_name,
             fill = cause_name)) +
  geom_treemap(show.legend = F) +
  geom_treemap_text(fontface = "bold",
                    reflow = T,
                    min.size = 1,
                    color = "white", 
                    place = "centre", 
                    grow = TRUE) +
  facet_wrap(~location_name, scales = "free") +
  labs(title = "GBD-YLDs due to Injuries types - Treemap (1990 - 2021)",
       subtitle = "Average values",
       caption = "IHME Data | Graphic: @fgazzelloni")
```

We can conclude that the types of injuries contributing to the burden of
disease varied across high, middle, and low SDI countries from 1990 to
2021. Road injuries, falls, and self-harm were common types of injuries
in all countries, with variations in the average number of injuries by
type. This information can help inform targeted public health
interventions to reduce the burden of injuries and improve population
health outcomes.

## Example: All Causes Disability Adjusted Life Years (DALYs) Rate by Country

In this example, we will visualise the **Disability Adjusted Life Years
(DALYs) Rate by Country** for all causes using the GBD data. We will use
data from the `{hmsidwR}` package, `spatialdalys2021`, to create a
choropleth map showing the DALYs rate per 100,000 population for each
country. The map will provide a visual representation of the burden of
disease across different countries, highlighting disparities in health
outcomes and disease burdens.

```{r}
#| label: fig-dalys-choropleth
#| fig.cap: "All Causes | DALYs Rate by Country (per 100,000)"
#| fig.alt: "A choropleth map showing DALYs Rate by Country (per 100,000)."
hmsidwR::spatialdalys2021 %>%
  ggplot(mapping=aes(x=long, y=lat,
                     fill = value, map_id = location)) +
  geom_polygon(aes(group=group)) + 
  # change the values of the fill scale
  scale_fill_viridis_c(option = "plasma", 
                       labels = scales::number_format(scale = 1e-3),
                       name = "DALYs") +
  coord_sf() +
  labs(title = "All Causes | DALYs Rate by Country (per 100,000)",
       subtitle = "Values in thousands",
       caption = "IHME Data 2021 | Graphic: @fgazzelloni") +
  theme_void()
```

We can conclude that the Disability Adjusted Life Years (DALYs) Rate
varied across countries, with some countries experiencing higher DALYs
rates compared to others. This variation reflects differences in disease
burdens, health outcomes, and the impact of different health conditions
on population health. The choropleth map provides a visual
representation of the burden of disease, highlighting disparities in
health outcomes and informing public health interventions to improve
population health.

In general, what emerges from the GBD 2021 data is that low-income
countries, particularly in sub-Saharan Africa, often exhibit the highest
DALYs per capita. This is primarily due to a heavy burden of infectious
diseases (e.g., malaria, HIV/AIDS, tuberculosis), maternal and neonatal
health issues, and malnutrition-related conditions. Countries like the
Central African Republic, South Sudan, and Lesotho consistently report
some of the highest DALYs per capita.

On the other hand, high-income countries, particularly in Western
Europe, North America, and parts of East Asia (e.g., Japan, Singapore),
tend to have the lowest DALYs per capita. These countries benefit from
well-developed healthcare systems, lower prevalence of infectious
diseases, and greater control over non-communicable diseases through
early detection and lifestyle interventions. Countries like Iceland,
Switzerland, and Singapore consistently report some of the lowest DALYs
per capita.

Additionally, lifestyle-related factors like smoking and dietary habits
play a significant role in the burden of cardiovascular diseases and
cancers globally.

## Implications for Global Public Health Policy

In summary, comparing health metrics across countries is essential for
understanding the state of health globally. Resources like the GBD study
and the OECD’s Health at a Glance report provide valuable insights into
health metrics, enabling comparisons and identifying areas for
improvement. By analysing health metrics across countries, we can gain
insights into public health priorities, disease burdens, and health
outcomes, which are crucial for informing global public health policy
and interventions.

Analysing health metrics at a global level reveals key patterns in
disease prevalence, mortality rates, and risk factors, offering insights
into how socioeconomic, environmental, and policy variables influence
health outcomes. Such comparative analyses are critical for setting
international health priorities and targeting interventions in regions
with the greatest need.

Regional studies play an equally significant role in contextualizing
global insights. For instance, **The State of Health in the European
Union in 2019 (Santos et al., 2019)**[@vascosantos2021] highlights the
importance of tailoring health policies to regional contexts. By
focusing on EU countries, the study identifies unique challenges and
opportunities, providing a blueprint for targeted interventions and
efficient resource allocation. These findings underscore that while
global frameworks are valuable, effective public health strategies often
require adaptation to local circumstances, cultures, and health systems.
Investing in health metric comparisons strengthens global collaboration
by fostering data sharing, capacity building, and the exchange of best
practices among nations.
