# Summary: The State of Health {#sec-16-summary}

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title = element_text(face = "bold"))
ggplot2::theme_set(book_theme)
```

:::::: solutionbox
:::: solutionbox-header
::: solutionbox-icon
:::

Learning Objectives
::::

::: solutionbox-body
-   How health metrics vary across different countries.
-   Gain insights from the Global Burden of Disease (GBD) study on
    health metrics at the country level.
-   Evaluate the implications of health metrics for global public health
    policy and interventions.
:::
::::::

\
In this final chapter, we synthesise the various elements covered
throughout the book, focusing on the *comparison of health metrics
across different countries*. For learning and comparing health metrics
across different countries, resources like the **Global Burden of
Disease (GBD)** study by the **Institute for Health Metrics and
Evaluation (IHME)**, and the **OECD’s Health at a Glance report** are
incredibly valuable.

The GBD provides comprehensive data on various health metrics, including
disability-adjusted life years (DALYs), years lived with disability
(YLDs), and years of life lost (YLLs), covering a vast array of
diseases, injuries, and risk factors across 204 countries. It is a
valuable resource for understanding the state of health across different
countries and regions. You can explore GBD data on metrics like Years of
Life Lost (YLLs), Disability-Adjusted Life Years (DALYs), and more by
country, which is instrumental for cross-country comparisons on health
outcomes and disease impacts ([www.healthdata.org](www.healthdata.org)).

The OECD’s Health at a Glance report offers a comparative overview of
health indicators among member countries. This report examines metrics
such as life expectancy, health expenditures, and major health trends,
making it a useful resource for understanding how health metrics vary
across different countries. The report provides accessible insights into
public health priorities and performance indicators, facilitating
comparisons and helping to identify areas for improvement
([www.oecd-ilibrary.org](www.oecd-ilibrary.org)).

## Health Metrics Countries Comparison

Metrics like **life expectancy**, **Disability-Adjusted Life Years (DALYs)**, and **Years Lived with Disability (YLDs)** are chosen as primary comparators; they provide a comprehensive view of population health, and able to capture both longevity and quality of life. In real-world applications, to compare the relative impact of different health issues across regions and populations, and identify which diseases contribute the most to health burdens in different countries, these metrics are essential.

### Example: Life Expectancy Comparison

Let's compare the levels of health metrics in different countries
starting with a simple example. We will compare the life expectancy at
birth in 2019 for a few countries.

```{r}
#| label: fig-life-expectancy
#| fig.cap: "Life expectancy at birth in 2019 for selected countries."
#| fig.alt: "A bar plot showing life expectancy at birth in 2019 for selected countries."

library(tidyverse)

# Data
life_expectancy <- data.frame(
  country = c("United States", "China", "India", "Brazil", "Russia"),
  life_expectancy = c(78.9, 76.7, 69.7, 75.9, 72.6))

# Plot
ggplot(life_expectancy, aes(x = country, y = life_expectancy)) +
  geom_col(fill = "skyblue") +
  geom_text(aes(label = life_expectancy), 
            vjust = -0.1, size = 3) +
  labs(title = "Life expectancy at birth in 2019 for selected countries",
       x = "Country", 
       y = "Life expectancy at birth (years)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Example: DALYs Comparison by Country

Then, we can use the GBD data to compare health metrics across
countries. For example, we can compare the burden of disease in terms of
Disability-Adjusted Life Years (DALYs) for selected countries.

```{r}
#| label: fig-dalys
#| fig.cap: "Disability-Adjusted Life Years (DALYs) in 2019 for selected countries."
#| fig.alt: "A bar plot showing Disability-Adjusted Life Years (DALYs) in 2019 for selected countries."

# Data
dalys <- data.frame(
  country = c("United States", "China", "India", "Brazil", "Russia"),
  dalys = c(0.03, 0.04, 0.06, 0.04, 0.05))

# Plot
ggplot(dalys, aes(x = country, y = dalys)) +
  geom_col(fill = "skyblue") +
  geom_text(aes(label = dalys), 
            vjust = -0.1, size = 3) +
  labs(title = "Disability-Adjusted Life Years (DALYs) in 2019 for selected countries",
       x = "Country", y = "DALYs per capita") +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1))
```
Differences in Disability-Adjusted Life Years (DALYs) across countries are often influenced by multiple interrelated factors, including **healthcare infrastructure**, **socioeconomic status**, and the **prevalence of specific diseases**. In countries with well-developed healthcare systems, accessible preventive services, early diagnosis, and quality treatment options contribute to lower DALY rates by reducing both mortality and disability from various health conditions. Conversely, countries with limited healthcare resources may experience higher DALY rates, as inadequate access to treatment and preventive care can lead to increased disease burden and complications.

### Example: Top Diseases Causing DALYs in the United States

Now, looking at the 5 top diseases causing DALYs in the United States in
2019:

```{r}
#| label: fig-top-diseases-us
#| fig.cap: "Top 5 diseases causing DALYs in the United States in 2019."
#| fig.alt: "A bar plot showing the top 5 diseases causing DALYs in the United States in 2019."

# Data
top_diseases_us <- data.frame(
  disease = c("Ischemic\nheart disease", "Low back pain", "Migraine", "Chronic obstructive\npulmonary disease", "Diabetes"),
  dalys = c(0.06, 0.05, 0.04, 0.03, 0.03))

# Plot
ggplot(top_diseases_us, aes(x = disease, y = dalys)) +
  geom_col(fill = "skyblue") +
  geom_text(aes(label = dalys), 
            vjust = -0.1, size = 3) +
  labs(title = "Top 5 diseases causing DALYs in the United States in 2019",
       x = "Disease", y = "DALYs per capita") +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1))
```

### Example: DALYs Map Comparison by Country

Finally, we can visualise the distribution of DALYs across countries on a world map. This type of visualisation provides a comprehensive view of the global burden of disease, highlighting the disparities in health outcomes and disease prevalence across different regions.

```{r}
#| label: fig-dalys-map
#| fig.cap: "Disability-Adjusted Life Years (DALYs) per 100,000 population in 2021 by country."
#| fig.alt: "A world map showing Disability-Adjusted Life Years (DALYs) per 100,000 population in 2021 by country."
```


**Exercise**: What type of plots can we draw to compare health metrics
across countries?

1.  Bar plots
2.  Line plots
3.  Scatter plots
4.  All of the above

The correct answer is: All of the above.

## Implications for Global Public Health Policy

In summary, comparing health metrics across countries is essential for
understanding the state of health globally. Resources like the GBD study
and the OECD’s Health at a Glance report provide valuable insights into
health metrics, enabling comparisons and identifying areas for
improvement. By analysing health metrics across countries, we can gain
insights into public health priorities, disease burdens, and health
outcomes, which are crucial for informing global public health policy
and interventions.
