# Markov Chain {#sec-appendixB}

```{r}
#| echo: false
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_minimal())
```

## Overview

A Markov chain is a stochastic model describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. The state space of a Markov chain is the set of all possible states of the system. The transition probabilities are the probabilities of moving from one state to another. The transition matrix is a square matrix that describes the transition probabilities between states.

The code for this replication of the **Markov Chain** is from Dobrow - **Bayesian analysis of infectious diseases** book (chapter4). It shows clearly how to create a Markov chain and calculate the transition probabilities.

```{r}
set.seed(000)
markov <- function(init, mat, n, labels) {
  if (missing(labels)) labels <- 1:length(init)
  simlist <- numeric(n + 1)
  states <- 1:length(init)
  simlist[1] <- sample(states, 1, prob = init)

  for (i in 2:(n + 1)) {
    simlist[i] <- sample(states, 1, prob = mat[simlist[i - 1], ])
  }

  labels[simlist]
}
```

```{r}
P <- matrix(c(.51, .49, .49, .51),
  nrow = 2, ncol = 2, byrow = TRUE
)
init <- c(1, 0)
```

```{r}
markov(init = init, mat = P, n = 100)
```

```{r}
# Define the number of transitions
n_transitions <- 10000

# Simulate Markov chain
simulated_chain <- markov(init, P, n_transitions)

# Calculate the transition probabilities
transition_counts <- table(
  simulated_chain[-1],
  simulated_chain[-(n_transitions + 1)]
)
transition_probabilities <- transition_counts / rowSums(transition_counts)
transition_probabilities
```

```{r}
# Calculate prior probabilities
prior_probabilities <- table(simulated_chain[-n_transitions]) / n_transitions

# Calculate posterior probabilities
posterior_probabilities <- transition_probabilities %*% prior_probabilities

# Print prior probabilities
print("Prior probabilities:")
print(prior_probabilities)

# Print posterior probabilities
print("Posterior probabilities:")
print(posterior_probabilities)
```
