# Collecting Data with APIs {#sec-appendixE}

Collecting data to use in a research analysis involves a selection of
sources and methods to use for optimizing computational time when
downloading and reading the files. Once data is set and ready to use a
further step is required to make the data suitable for the selected
model. The source of data is an important variable. Generally, data can
be downloaded by using an API (application programming interface) which
allow the user to get access to data directly from source, with the use
of specified back-end computations. There are alternatives at using an
API; data can be obtained by downloading it directly into the
computer,or loaded through library packages. Usually, available files
are provided under various forms such as delimited type of files, .csv,
.xls, .json, and other types. Here is an example of how to use an API
for downloading a file directly onto your computer.

```{r}
#| eval: false
library(httr)
url <- ""
httr::GET(url = url)
```

### Download IHME Data with APIs

```{r}
#| warning: false
#| message: false
library(httr)
library(jsonlite)
library(tidyverse)

# `SDG query input:- GetGoal - GetIndicator - GetTarget - GetLocation - GetAgeGroup - GetScenario - GetSexSDG 
# query input for results:- GetResultsByTarget - GetResultsByIndicator - GetResultsByLocation - 
# GetResultsByYear`


# url="https://api.healthdata.org/sdg/v1/"key="YOUR-KEY-HERE"

get_data <- function(url,key,sdg) {  
library(httr)  
library(jsonlite)  
url <- paste0(url,"/",sdg)  
headers <- c('Authorization'=key)  
res <- GET(url, add_headers(headers))  
data = fromJSON(rawToChar(res$content))  
data = data$results
}
```

Build the query to download results by adding specification of the data.

`GetResultsByLocation?location_id=86&indicator_id=1002&year=2019? location_id & indicator_id ...`

```{r}
#| echo: false
key="8ym9ht687svohiwtqiinfjd3gxq7vihc"
```

```{r}
#| eval: false
data <- get_data(url="https://api.healthdata.org/sdg/v1",
         key=key,
         sdg="GetResultsByLocation?location_id=86&indicator_id=1002&year=2019")
data
```
