# Life Tables, Markov Chain and APIs {#sec-appendixA}

```{r}
#| echo: false
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_minimal())
```

## Life Tables and Life Expectancy

Back in the 1700s the Swiss mathematician and physicist Daniel Bernoulli
(1700 - 1782) developed the use of a life table model by differentiating
life tables based on specific causes of death [see @kovacheva2017, and
@lifetab].

Originally made by the English scientist **John Graunt** (1620-1674),
for the analysis of the mortality of the population of London and the
impact of different diseases. Life tables contain fundamental statistics
for the calculation of probabilities of deaths and the computation of
life expectancy at birth and at different ages.

[!["Life tables", William Farr (England
1859)](images/appendixA_life_tables.png){fig-align="center"
width="55%"}](https://www.jstor.org/stable/community.24736608?searchText=life+tables&searchUri=%2Faction%2FdoBasicSearch%3FQuery%3Dlife%2Btables%26efqs%3DeyJjdHkiOlsiWTI5dWRISnBZblYwWldSZmFXMWhaMlZ6Il19%26image_search_referrer%3Dglobal&ab_segments=0%2FSYC-6744_basic_search%2Ftest-1&refreqid=fastly-default%3Ac770f9049f8754dad03896e1b52df20c&searchkey=1675639441064)

### Life tables components

More recent life tables are standardized to be used for a population of
100 000 at age 0.

::: callout
$l_x$ survivors at age $x$ it starts with a value of 100 000

$d_x$ deceased at age $x$

$q_x$ probability of deaths

$p_x$ probability of survival
:::

The probability of survival is given by: $$p_x=1-q_x$$

**Let's start constructing a life table**

The **Global Life Tables** are included in the `{hmsidwR}` package as
`gho_lifetables` dataset. This dataset has been released by the WHO, and
contains various indicators.

The construction of the Global Life Tables takes consideration of
age-specific mortality patterns, which is the main improvements made on
life tables construction since the first set of model life tables
published by the United Nations in 1955, see [@modified] for more
information about a detailed procedure.

To have a look at the package documentation for this dataset, use:

```{r}
#| eval: false
?hmsidwR::gho_lifetables
```

```{r}
library(tidyverse)
hmsidwR::gho_lifetables %>%
  count(indicator) %>%
  select(-n)
```

The indicator of interest for re-building a life table are:

-   `lx` - number of people left alive at age x
-   `age`

These two key elements are crucial for building the life tables.

```{r}
lx <- hmsidwR::gho_lifetables %>%
  distinct() %>%
  filter(
    indicator == "lx",
    year == "2019"
  )
```

```{r}
x <- lx %>%
  filter(sex == "female") %>%
  select(x = age)

lx_f <- lx %>%
  filter(sex == "female") %>%
  select(lx = value)

lx_m <- lx %>%
  filter(sex == "male") %>%
  select(lx = value)
```

The probability of survival is calculated as follow:

$$p_x = \frac{l_x}{l_{x+1}}$$

```{r}
px <- lx_f$lx / lag(lx_f$lx)
```

```{r}
data.frame(
  x,
  lx = round(lx_f),
  dx = round(c(-diff(lx_f$lx), 0)),
  px,
  qx = 1 - lead(px),
  Lx = c(
    (lx_f$lx[1] + (lx_f$lx[2])) / 2,
    5 * (lx_f$lx[-1] + lead(lx_f$lx[-1])) / 2
  )
) %>%
  head()
```

```{r}
hmsidwR::gho_lifetables %>%
  distinct() %>%
  mutate(indicator = gsub(" .*", "", indicator)) %>%
  filter(
    indicator == "nLx",
    year == "2019",
    sex == "female"
  ) %>%
  head()
```

### Life expectancy

Life expectancy is the expected number of years a person will live,
based on current age and prevailing mortality rates. There are several
methods to calculate life expectancy, but one common approach is to use
the actuarial life table, which is a statistical table that provides the
mortality rates for a population at different ages. The following steps
can be used to calculate life expectancy using a life table:

Identify the relevant mortality rates for the population and time period
of interest. Calculate the probability of surviving to each age, given
the mortality rates. Multiply the probability of surviving to each age
by the remaining life expectancy at that age to obtain the expected
number of years of life remaining at each age. Sum the expected number
of years of life remaining at each age to obtain the total life
expectancy. Note that life expectancy is a statistical estimate and can
be influenced by many factors, such as lifestyle, health, and
environmental factors, so actual individual life expectancy can vary
widely.

Here are some key references for calculating life expectancy:

1.  United Nations World Population Prospects - The UN provides detailed
    life tables and population data, including life expectancy, for
    countries and regions around the world.

2.  Centers for Disease Control and Prevention (CDC) - The CDC provides
    life tables for the United States, as well as information on how
    life expectancy is calculated and factors that affect it.

3.  World Health Organization (WHO) - The WHO provides information on
    global health and life expectancy, including data and reports on
    trends in life expectancy and mortality.

4.  Actuarial Science textbooks - Books such as "Actuarial Mathematics"
    by Bowers, Gerber, Hickman, Jones, and Nesbitt, or "An Introduction
    to Actuarial Mathematics" by Michel Millar, provide comprehensive
    coverage of the methods and mathematics used in calculating life
    expectancy.

5.  Journal articles - Articles in actuarial and demographic journals,
    such as the North American Actuarial Journal or Demographic
    Research, often provide in-depth coverage of the latest research and
    methods for calculating life expectancy.

## Markov Chain

```{r}
#| echo: false
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_minimal())
```

A Markov chain is a stochastic model describing a sequence of possible
events in which the probability of each event depends only on the state
attained in the previous event. The state space of a Markov chain is the
set of all possible states of the system. The transition probabilities
are the probabilities of moving from one state to another. The
transition matrix is a square matrix that describes the transition
probabilities between states.

The code for this replication of the **Markov Chain** is from Dobrow -
**Bayesian analysis of infectious diseases** book (chapter4). It shows
clearly how to create a Markov chain and calculate the transition
probabilities.

```{r}
set.seed(000)
markov <- function(init, mat, n, labels) {
  if (missing(labels)) labels <- 1:length(init)
  simlist <- numeric(n + 1)
  states <- 1:length(init)
  simlist[1] <- sample(states, 1, prob = init)

  for (i in 2:(n + 1)) {
    simlist[i] <- sample(states, 1, prob = mat[simlist[i - 1], ])
  }

  labels[simlist]
}
```

```{r}
P <- matrix(c(.51, .49, .49, .51),
  nrow = 2, ncol = 2, byrow = TRUE
)
init <- c(1, 0)
```

```{r}
markov(init = init, mat = P, n = 100)
```

```{r}
# Define the number of transitions
n_transitions <- 10000

# Simulate Markov chain
simulated_chain <- markov(init, P, n_transitions)

# Calculate the transition probabilities
transition_counts <- table(
  simulated_chain[-1],
  simulated_chain[-(n_transitions + 1)]
)
transition_probabilities <- transition_counts / rowSums(transition_counts)
transition_probabilities
```

```{r}
# Calculate prior probabilities
prior_probabilities <- table(simulated_chain[-n_transitions]) / n_transitions

# Calculate posterior probabilities
posterior_probabilities <- transition_probabilities %*% prior_probabilities

# Print prior probabilities
print("Prior probabilities:")
print(prior_probabilities)

# Print posterior probabilities
print("Posterior probabilities:")
print(posterior_probabilities)
```

## Collecting Data with APIs

Collecting data to use in a research analysis involves a selection of
sources and methods to use for optimizing computational time when
downloading and reading the files. Once data is set and ready to use a
further step is required to make the data suitable for the selected
model. The source of data is an important variable. Generally, data can
be downloaded by using an API (application programming interface) which
allow the user to get access to data directly from source, with the use
of specified back-end computations. There are alternatives at using an
API; data can be obtained by downloading it directly into the
computer,or loaded through library packages. Usually, available files
are provided under various forms such as delimited type of files, .csv,
.xls, .json, and other types. Here is an example of how to use an API
for downloading a file directly onto your computer.

```{r}
#| eval: false
library(httr)
url <- ""
httr::GET(url = url)
```

### Download Data with APIs

#### IHME Data APIs

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
```

-   SDG Query Input:

    -   GetGoal
    -   GetIndicator
    -   GetTarget
    -   GetLocation
    -   GetAgeGroup
    -   GetScenario
    -   GetSexSDG

-   Query Input for Results:

    -   GetResultsByTarget
    -   GetResultsByIndicator
    -   GetResultsByLocation
    -   GetResultsByYear

Use the function `get_data` to download data from the IHME API. The
function requires the following arguments:

```         
url = "https://api.healthdata.org/sdg/v1
key = "YOUR-KEY"
sdg = "GetResultsByLocation?location_id=86&indicator_id=1002&year=2019"
```

```{r}
get_data <- function(url, key, sdg) {
  library(httr)
  library(jsonlite)
  url <- paste0(url, "/", sdg)
  headers <- c("Authorization" = key)
  res <- GET(url, add_headers(headers))
  data <- fromJSON(rawToChar(res$content))
  data <- data$results
}
```

Build the query to download results by adding specification of the data.

`GetResultsByLocation?location_id=86&indicator_id=1002&year=2019? location_id & indicator_id ...`

```{r}
#| eval: false
data <- get_data(
  url = "https://api.healthdata.org/sdg/v1",
  key = YourKeyHere,
  sdg = "GetResultsByLocation?location_id=86&indicator_id=1002&year=2019"
)
data
```
