---
title: "Example rabies outside the library boxes"
---

### Example

Working outside the library boxes is a good way to start.

This is a study that uses a probability decision-tree framework to
estimate the burden of rabies in Tanzania and assess the
cost-effectiveness of different control strategies [@hampson2015a]. The
model incorporates various parameters, such as bite incidence, the
probability of a biting animal being rabid, the probability of a bite
victim receiving post-exposure prophylaxis, and the probability of
developing rabies in the absence of prophylaxis. It also includes an
economic component to calculate the costs of rabies prevention and
control measures like post-exposure prophylaxis (PEP).

Probability of Progression: $PP=RP*DP/((RP*DP)*D/T)$

Incidence: $I = I_{max}*(1-VC)^s$

DALYs calculations involve two components:

-   Years of Life Lost (YLL), which include deaths due to infection

-   Years of Life lived with Disability (YLD), which include disability
    following AEs due to use of nerve tissue vaccines, still used in
    some parts of the world.

**Parameters derived from the data and literature**

```{r}
bite_incidence <- 10000      # Example number of bites
RP_max <- 0.74               # Max probability of bite being from a rabid animal
PEP_probability <- 0.5       # Probability of receiving PEP
DP <- 0.19                   # Probability of developing rabies in the absence of PEP
GDP_per_capita <- 2000       # Example GDP per capita in USD
DALY_discount_rate <- 0.03   # Discount rate for DALY calculation
average_age_of_death <- 30   # Average age of rabies death
life_expectancy <- 70        # Average life expectancy

# Dog vaccination coverage
dog_vaccination_coverage <- 0.3

# Function to estimate the probability of a bite being rabid
estimate_RP <- function(vaccination_coverage, RP_max) {
  return(RP_max * (1 - vaccination_coverage))
}

# Function to calculate the economic burden of rabies deaths
calculate_economic_burden <- function(deaths, GDP_per_capita) {
  return(deaths * GDP_per_capita)
}

# Function to estimate DALYs
estimate_DALYs <- function(deaths, 
                           average_age_of_death, 
                           life_expectancy, 
                           discount_rate) {
  YLL <- deaths * (life_expectancy - average_age_of_death)
  discounted_YLL <- YLL / (1 + discount_rate)^(life_expectancy - average_age_of_death)
  return(discounted_YLL)
}

# Estimate the probability that a bite is from a rabid animal
RP <- estimate_RP(dog_vaccination_coverage, RP_max)

# Estimate the number of rabies deaths
rabies_deaths <- bite_incidence * RP * (1 - PEP_probability) * DP

# Calculate the economic burden
economic_burden <- calculate_economic_burden(rabies_deaths, GDP_per_capita)

# Estimate the DALYs
DALYs <- estimate_DALYs(rabies_deaths, 
                        average_age_of_death, 
                        life_expectancy, 
                        DALY_discount_rate)

# Print the results
cat("Estimated Rabies Deaths:", rabies_deaths, "\n")
cat("Estimated Economic Burden (USD):", economic_burden, "\n")
cat("Estimated DALYs:", DALYs, "\n")

```

Plot the relationship between dog vaccination coverage and rabies
incidence

```{r}
coverage_seq <- seq(0, 1, by = 0.01)
incidence_seq <- RP_max * (1 - coverage_seq)
data <- data.frame(Coverage = coverage_seq, Incidence = incidence_seq)
head(data)
```

```{r}
library(ggplot2)
ggplot(data, aes(x = Coverage, y = Incidence)) +
  geom_line(color = "blue") +
  labs(title = "Rabies Incidence vs Dog Vaccination Coverage",
       x = "Dog Vaccination Coverage",
       y = "Probability of Bite being Rabid") +
  theme_minimal()

```
