---
title: "The case of Malaria"
format: html
editor: visual
---

## Overview

> August 18, 2023 "A case of locally acquired #malaria has been confirmed in Maryland, Washington, D.C., area. Nine cases have been reported this summer in Florida and Texas, the first in the US in 20 years, according to the US Centers for Disease Control and Prevention. #epidemics" [^1]

[^1]: worldbank_data source: <https://databank.worldbank.org/reports.aspx?source=2&series=SH.MLR.INCD.P3&country=#>

The Malaria case is somehow a concerning case, eradicated all over the World except for some areas in the Africa's continent, the highlight of cases of domestic origins are considered epidemics.

Let's dig on some data about Malaria.

## WHO Malaria Map

The first source of data is the [WHO Malaria Map](https://bit.ly/3OUhT1x), a collection of information about location of cases, the vector species, their invasive status, and other variables of interest such as temporal of the study, the sampling methods, and so on.

Load necessary libraries

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(readxl)
```

```{r}
malaria_who <- read_excel(here::here("R_general_resources/Spatials/infectious_container/malaria/who_data.xlsx"), 
    sheet = "Data")
malaria_who %>% head()
```

```{r}
dim(malaria_who)
```

```{r}
malaria_who%>%glimpse
```

```{r}
malaria_who <- malaria_who %>%
  janitor::clean_names() %>%
  mutate(longitude=as.double(longitude),latitude=as.double(latitude))
```

```{r}
#| echo: false
#| include: false
malaria_who %>% DataExplorer::profile_missing()%>%arrange(-pct_missing)
```

Historical collection of data about mosquito number shows data is collected from 1985 and updated to 2022.

```{r}
#| echo: false
malaria_who%>%
  count(year_start,year_end)%>%arrange(year_start)
```

Let's omit missing values for the `year_start` variable and create a new variable `midyear` which is the middle year between the `year_start` and `year_end` of each study.

```{r}
malaria_who_mid_year <- malaria_who %>%
  filter(!is.na(year_start),
         !is.na(year_end),
         !is.na(mosquito_number),
         !mosquito_number=="NR") %>%
  mutate(mosquito_number=as.numeric(mosquito_number),
         year_end=as.double(year_end),
         midyear=round((year_end+year_start)/2,0))
```

Along the time, the trend of the number of mosquito varied with ups and downs.

```{r}
malaria_who_mid_year %>%
group_by(midyear)%>%# count(midyear)
  mutate(avg_n_mosquito=mean(mosquito_number))%>%
  ungroup()%>%
  count(country_name,midyear,avg_n_mosquito,invasive_status)%>%
  ggplot(aes(x=midyear,y=log10(avg_n_mosquito),
             group=invasive_status,color=invasive_status))+
  geom_line(linewidth=2)+
  scale_color_viridis_d(labels = c("Invasive", "Native"),
                       guide = guide_legend(reverse=TRUE,
                                            override.aes = list(size = 10)))+
  coord_cartesian(clip = 'off') +
  labs(title="Time series Malaria 1985 - 2022",
       caption="Graphics: FG",color="Status")+ 
  ggthemes::theme_fivethirtyeight()
```

Mosquito have been categorized as `invasive species`, after 2016.

```{r}
#| warning: false
#| message: false
malaria_who_mid_year%>%
  filter(midyear>= 2010)%>%
  ggplot(aes(x=factor(midyear),y=log10(mosquito_number),
             group=midyear,fill=invasive_status))+
  geom_violin()+
  scale_fill_viridis_d(labels = c("Invasive", "Native"),
                       guide = guide_legend(reverse=TRUE,
                                            override.aes = list(size = 10)))+
  ggthemes::theme_fivethirtyeight()+
  theme(legend.position = "top")+
  labs(title="When moqsquito became tag invasive",
       caption="Graphics: FG",fill="Status")
```

The location of cases revealed by consistent `sentinel surveillance` procedure, identify the area in the south est Africa/Asia to be the most affected by the danger of malaria virus to spread across the rest of the World.

Here is a map of the `invasive vector species` in this area:

```{r}
world <- map_data("world")%>%filter(!region=="Antarctica")

malaria_who %>%
  ggplot(mapping=aes(x=longitude,y=latitude))+
  geom_polygon(data=world,
               mapping=aes(x=long,y=lat,group=group),
               linewidth=0.2,
               color="grey80",
               fill="white")+
  geom_point(aes(fill=invasive_status),
             color="grey80",
             shape=21,
             stroke=0.2,
             size=0.7,alpha=0.5)+
  coord_sf(xlim = c(-50, 110), expand = TRUE) +
  scale_fill_viridis_d()+
  labs(title="Invasive vector species from 1985 to 2022",fill="Status")+
  ggthemes::theme_map()+
  theme(plot.background = element_rect(color="steelblue",fill="steelblue"))
```

Let's zoom in to the center of the mass points. Setting the mean range of the latitude and the longitude, to identify the central point, within the mass of points where mosquito were located, and setting a zoom level, a closer focus at the locations is possible, even with a specification of the new range of to be assigned to the map boundaries.[^2]

[^2]: worldbank_data source: <https://databank.worldbank.org/reports.aspx?source=2&series=SH.MLR.INCD.P3&country=#>

```{r}
lon_avg <- mean(range(malaria_who$longitude))
lat_avg <- mean(range(malaria_who$latitude))
lon_avg;lat_avg
```

```{r}
zoom_to <- c(lon_avg,lat_avg)  

zoom_level <- 1.5

lon_span <- 360 / 2^zoom_level
lat_span <- 180 / 2^zoom_level
```

```{r}
lon_bounds <- c(zoom_to[1] - lon_span / 2, zoom_to[1] + lon_span / 2)
lat_bounds <- c(zoom_to[2] - lat_span / 2, zoom_to[2] + lat_span / 2)
```

```{r}
#| message: false
#| warning: false
library(sf)
```

```{r}
#| message: false
#| warning: false
ggplot()+
  geom_polygon(data=world,
               mapping=aes(x=long,y=lat,group=group),
               linewidth=0.2,
               color="grey80",
               fill="white")+
  geom_point(data= malaria_who,
             mapping=aes(x=longitude,y=latitude,
                         fill=invasive_status),
             color="black",
             shape=21,
             stroke=0.2,
             size=3,alpha=0.5)+
  geom_sf_text(data = st_sfc(st_point(zoom_to), crs = 4326),
            label = '.') +
  scale_fill_viridis_d()+
  coord_sf(xlim = lon_bounds, ylim = lat_bounds,expand = TRUE) + 
  labs(title="A closer look at invasive vector species from 1985 to 2022",
       caption="Graphics: FG",
       fill="Status")+
  ggthemes::theme_map()+
  theme(plot.background = element_rect(color="steelblue",fill="steelblue"),
        legend.position = c(0,0.001))
```

## Incidence of malaria

Looking at a different data source, the `Worldbank data` provides a reports with the `incidence of malaria` from 2000 and 2010. [^3]

[^3]: worldbank_data source: <https://databank.worldbank.org/reports.aspx?source=2&series=SH.MLR.INCD.P3&country=#>

```{r}
#| warning: false
#| message: false
malaria_wb <- read_csv(here::here("R_general_resources/Spatials/infectious_container/malaria/data/worldbank_data.csv"))
```

```{r}
#| echo: false
#| include: false
malaria_wb %>% glimpse()
```

```{r}
#| warning: false
#| message: false
malaria_wb_long <- malaria_wb%>%
  janitor::clean_names()%>%
  select(-series_name,-series_code) %>%
  pivot_longer(cols = c(3:14),names_to="year")%>%
  mutate(value=trimws(value),
         value=as.numeric(value),
         value=round(value,3))%>%
  filter(!is.na(value))%>%
  mutate(year=gsub("_yr[0-9]+$","",year),
         year=gsub("^x","",year),
         year=as.integer(year))%>%
  arrange(year)
malaria_wb_long%>%glimpse
```

```{r}
#| echo: false
#| include: false
malaria_wb_long %>%count(year)
```

```{r}
malaria_wb_long%>%
  group_by(year)%>%
  reframe(avg=mean(value))%>%
  mutate(year=as.factor(year))%>%
  ggplot(aes(year,avg,group=1))+
  #geom_col()+ 
  geom_line()+
  labs(title="Incidence of Malaria",
  subtitle="per 1,000 population at risk",
  caption = "Graphics: FG")+
  ggthemes::theme_fivethirtyeight()
```

```{r}
#| echo: false
#| include: false
#| eval: false
p <- malaria_wb_long%>%
  group_by(country_name,year)%>%
  reframe(avg=mean(value))%>%
  mutate(year=as.factor(year)#,color=cut_interval(avg,5)
         )%>%
  ggplot(aes(year,avg,group=country_name))+
  #geom_col()+ 
  geom_line()

plotly::ggplotly(p)
```

```{r}
#| echo: false
#| include: false
#| eval: false
region <- unique(world$region)
country<-unique(malaria_wb$country_name)

setdiff(country,region)


purrr::map2(country,region)
map(match.arg(country[i],region))
    
v<-c() 
for (i in 1:length(country)) {
  v <- country[str_detect(country[i],region)]
  print(v)
}

v

```

## Case Study - antimalarial drug resistance

#### Tracking antimalarial drug resistance using mosquito blood meals: a cross-sectional study

-   Article: <https://www.thelancet.com/journals/lanmic/article/PIIS2666-5247(23)00063-0/fulltext>

-   GitHub repository: <https://github.com/hannaehrlich/maldrugres_SSA>

```{r}
url <- "https://raw.githubusercontent.com/hannaehrlich/maldrugres_SSA/main/Survey_MolecMarker_Data.csv"
```

```{r}
#| message: false
#| warning: false
maldrugres <- read.csv(url)
names(maldrugres)
```

```{r}
#| echo: false
#| include: false
maldrugres%>%glimpse
```

```{r}
#| echo: false
#| include: false
maldrugres%>%
  DataExplorer::profile_missing()
```

```{r}
maldrugres <- maldrugres %>%
  filter(!is.na(Lat),!is.na(Drug),!Drug=="")

maldrugres%>%count(Drug)
```

```{r}
world <- map_data("world")%>%filter(!region=="Antarctica")
africa <- world%>%filter(long >= -50,long< 60)

ggplot(data= maldrugres) +
  geom_polygon(data= africa, 
               mapping=aes(x=long,y=lat,group=group),
               linewidth=0.1,
               color="grey70",fill="grey90")+
  geom_point(mapping=aes(x=Lon,y=Lat,
                         color=Drug,
                         fill=Present),
             shape=21,
             stroke=0.2,
             size=3)+
  scale_color_manual(values = c("steelblue","darkred"))+
  scale_fill_gradient(low=NA,high = "darkred")+
  coord_sf(xlim = c(-20,50),ylim = c(-40,60),clip = "off")+
    labs(title="antiMalarial drug resistance",
       caption="DataSource: GitHub hannaehrlich/maldrugres_SSA | Graphics: FG")+
  ggthemes::theme_map()+
  theme(text = element_text(family="Roboto Condensed"),
        plot.title = element_text(size=30,hjust = 0.5,family="Roboto Condensed"),
        plot.title.position = "plot",
        legend.position = c(-0.6,0.1))
```

```{r}
maldrugres_new <- maldrugres%>%
  mutate(MidYear = round((StartYr+EndYr)/2,0),
         MidYear= as.factor(MidYear))%>%
  select(Country,Site,Lon,Lat,MidYear,Tested,Present,MixedPres,Drug)%>%
  janitor::clean_names()

maldrugres_new%>%head
```

### Tidymodels

#### EDA

```{r}
maldrugres_new%>%
  count(drug)%>%
  ggplot(aes(x=drug,y=n,fill=drug))+
  geom_col()+
  labs(title="Drug class imbalance",
       caption="Graphics: FG")+
  scale_fill_viridis_d()+
  ggthemes::theme_fivethirtyeight()+
  theme(legend.position = "none")
```

```{r}
maldrugres_new%>%
  group_by(country,drug)%>%
  reframe(avg_drug=mean(present))%>%
  ggplot(aes(x=avg_drug,y=fct_reorder(country,avg_drug)))+
  geom_col(aes(fill=drug))+
  scale_fill_viridis_d()+
  labs(title="AntiMalarial Drug Resistance Present",
       caption="Graphics: FG",
       x="Average value by Country",y="")+
   ggthemes::theme_fivethirtyeight()+
  theme(axis.text.x = element_text(angle=0,hjust = 1))
 
```

```{r}
maldrugres_new %>%
  ggplot(aes(present))+
  geom_density()+
  labs(title="Density distribution of antimalarial drug resistance",
       caption = "Graphics: FG")+
  ggthemes::theme_fivethirtyeight()
```

```{r}
maldrugres_new%>%
  group_by(country,drug)%>%
  reframe(avg_drug=mean(present))%>%
  ggplot(aes(x=avg_drug,y=fct_reorder(country,avg_drug)))+
  geom_boxplot()+
  labs(title="AntiMalaria Drug Resistance Present",
       caption="Graphics: FG",
       x="Average value by Country",y="")+
  theme(axis.text.x = element_text(angle=0,hjust = 1))+
  ggthemes::theme_fivethirtyeight()
```

### Spending data

```{r}
#| warning: false
#| message: false
library(tidymodels)
```

```{r}
set.seed(123)
split <- initial_split(maldrugres_new)
training <- training(split)
testing <- testing(split)
cv_folds <- vfold_cv(training,v = 10)
```

### Featuring Engineering

```{r}
rec_pca <- recipe(present ~., training) %>%
  step_dummy(all_nominal_predictors(),keep_original_cols = F)%>%
  step_corr(all_numeric_predictors())%>%
  step_normalize(all_predictors())%>%
  step_pca(all_predictors())
  
rec_pca_df <- rec_pca %>%
prep()%>%
  juice()%>%
  cbind(drug=training$drug,country=training$country)
rec_pca_df%>%head
```

```{r}
#| message: false
#| warning: false
rec_pca_df %>%
ggplot(aes(x=PC1,PC2,group=drug,color=drug))+
  geom_point()+
  geom_smooth(se=F)+
  scale_color_viridis_d()+
  labs(title="Principal Components Analysis",
       caption = "Graphics: FG")+
  ggthemes::theme_fivethirtyeight()+
  theme(axis.title = element_text())
```

```{r}
rec_pca_df %>%
ggplot(aes(x=PC1,y=fct_reorder(country,PC1),group=country))+
  geom_boxplot()+
  labs(title="Principal Components Analysis - boxplot",
       caption = "Graphics: FG",y="")+
  ggthemes::theme_fivethirtyeight()+
    theme(axis.title = element_text(),
          plot.title = element_text(hjust = 1))
```

```{r}
#| message: false
#| warning: false

rec_pca_df %>%
ggplot(aes(x=PC1,y=present))+
  geom_point()+
  scale_y_log10()+
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"))+
  labs(title="Principal Components Analysis - scatterplot",
       caption = "Graphics: FG",y="Present")+
  ggthemes::theme_fivethirtyeight()+
    theme(axis.title = element_text(),
          plot.title = element_text(hjust = 0))
```

```{r}
rec_ica <- recipe(present ~., training) %>%
  step_dummy(all_nominal_predictors(),keep_original_cols = F)%>%
  step_corr(all_numeric_predictors())%>%
  step_normalize(all_predictors())%>%
  step_ica(all_predictors())%>%
  prep()%>%
  juice()%>%
  cbind(drug=training$drug)
```

```{r}
rec_ica %>%
ggplot(aes(x=IC1,IC2,group=drug,color=drug))+
  geom_point()+
  geom_smooth(se=F)+
  scale_colour_viridis_d()+
  labs(title="Independent Components Analysis",
       caption = "Graphics: FG")+
  ggthemes::theme_fivethirtyeight()+
    theme(axis.title = element_text(),
          plot.title = element_text(hjust = 0))
```

```{r}
#| echo: false
#| eval: false
rec_jeo <- recipe(present ~., training) %>%
  step_dummy(all_nominal_predictors(),keep_original_cols = F)%>%
  step_corr(all_numeric_predictors())%>%
  step_normalize(all_predictors())%>%
  step_invlogit(all_predictors())%>%
  step_YeoJohnson()
  

rec_jeo %>%
  prep()%>%
  juice()%>%
  cbind(drug=training$drug)
```

```{r}
#| echo: false
#| include: false
#| eval: false

### Spatial distancing
recipe(present ~., training) %>%
  step_geodist(lat = lat,lon = lon,
                ref_lat = 38.8986312, ref_lon = -77.0062457,
               is_lat_lon = TRUE) %>%
  prep()%>%
  juice()%>%
  head()
  
# step_dummy(all_nominal_predictors(),keep_original_cols = F)%>%
#   step_corr(all_numeric_predictors())%>%
#   step_normalize(all_predictors())%>%
#   step_pca(all_predictors())
```

## Modeling

Following the results of `geom_smooth()` indication on best model type application on `rec_pca` data, here is a model specification for a general additive model (`GAMs`).

```{r}
#| echo: false
#| include: false
#| eval: false
?gen_additive_mod
```

```{r}
mod_pca <- gen_additive_mod() %>%
  set_engine("mgcv") %>%
  set_mode("regression")
```

#### Workflow for GAM modeling

```{r}
#| eval: false

wkf_pca <- workflow() %>%
  add_recipe(rec_pca) %>%
  add_model(mod_pca,formula=present~PC1+PC2)
  
fit_wkf <- wkf_pca%>%
fit(data=training)

fit_wkf %>%
 extract_fit_parsnip()
```

## gstat

```{r}
#| message: false
#| warning: false
library(gstat)
```

```{r}
maldrugres_2<- maldrugres%>%
  janitor::clean_names()%>%
  select(country,site,lon,lat,present,drug)
maldrugres_2%>%head
```

```{r}
library(sf)
```

```{r}
maldrugres_2%>%
  ggplot(aes(lon,lat))+
  geom_point()
```

```{r}
#| message: false
#| warning: false
library(rnaturalearth)
africa <- ne_countries(continent = "Africa",returnclass = 'sf')
crs_afr <- st_crs(africa)
```

```{r}
# View(rgdal::make_EPSG())

rgdal::make_EPSG()%>%
  filter(str_detect(note,"Asia"))
  filter(str_detect(note,"WGS 84"),
         code>=4326)%>%View
```


```{r}
crs <- "EPSG:4326"
maldrugres_sf <- maldrugres_2%>%
  st_as_sf(coords = c(3,4),crs=crs_afr)%>%
  st_transform(crs)
new<- africa %>% st_transform(crs)
afr <- st_make_valid(new)

library(stars) |> suppressPackageStartupMessages()
st_bbox(afr) |>
  st_as_stars() |>
  st_crop(afr) -> grd
grd
```

```{r}
crs<- "ESRI:102023"
maldrugres_sf <- maldrugres_2%>%
  st_as_sf(coords = c(3,4),crs=crs_afr)%>%
  st_transform(crs)
new<- africa %>% st_transform(crs)

```

```{r}
# plot(africa)
ggplot(africa)+
  geom_sf(aes(geometry=geometry))+
  geom_sf(data=maldrugres_sf,aes(geometry=geometry))
```

#### Create a regular 10 km X 10 km grid over Africa

```{r}

afr <- st_make_valid(new)

library(stars) |> suppressPackageStartupMessages()
st_bbox(afr) |>
  st_as_stars(dx = 10000) |>
  st_crop(afr) -> grd
grd

```

#### Inverse distance weigth

```{r}
i <- idw(present~1, maldrugres_sf, grd)

ggplot() + geom_stars(data = i, 
                      aes(fill = var1.pred, x = x, y = y)) + 
    xlab(NULL) + ylab(NULL) +
    geom_sf(data = st_cast(afr, "MULTILINESTRING")) + 
    geom_sf(data = maldrugres_sf)
```

```{r}
v <- variogram(present~1, maldrugres_sf)
```

```{r}
v  %>%
  ggplot(aes(dist,gamma))+
  geom_point(shape=21,stroke=0.5,
             size=3,
             color="steelblue")+
  geom_smooth(se=F,method="lm")+
  geom_text(aes(label=np),hjust=-0.5)+
  labs(x = "distance h [m]",y=expression(gamma(h)))+
  xlim(0, 1.055 * max(v$dist))+
  theme_bw()
```

```{r}
v.m <- fit.variogram(v, vgm(c("Exp", "Mat", "Sph")), 
                     fit.kappa = TRUE)
```

```{r}
plot(variogramLine(v.m, 1.075 * max(v$dist), 10), type = 'l')
```

## kringing

```{r}
doParallel::registerDoParallel()
```

```{r}
k <- krige(present~1, maldrugres_sf, grd, v.m)
```

```{r}
ggplot() + 
  geom_stars(data = k, aes(fill = var1.pred, x = x, y = y)) + 
    xlab(NULL) + ylab(NULL) +
    geom_sf(data = st_cast(afr, "MULTILINESTRING")) + 
    geom_sf(data = maldrugres_sf
            ) +
    coord_sf(lims_method = "geometry_bbox")
```

```{r}
a <- aggregate(maldrugres_sf["present"], by = afr, FUN = mean)
a
```

```{r}
b <- krige(present~1, maldrugres_sf, afr, v.m)
# [using ordinary kriging]
b
```

```{r}
b$sample <- a$present
b$kriging <- b$var1.pred
```

```{r}
b |> select(sample, kriging) |> 
        pivot_longer(1:2, 
                     names_to = "var", values_to = "present") -> b2

b2$var <- factor(b2$var, levels = c("sample", "kriging"))
ggplot() + 
  geom_sf(data = b2, 
          mapping = aes(fill = present)) + 
  facet_wrap(~var) +
     scale_fill_gradientn(colors = sf.colors(20))
```
